---
title: "Linearity Assumption"
execute:
  echo: false
format: html
---

```{r}
#| warning: false

# Load packages
library(scales)

# Set seed
set.seed(1)
```

## Learning Objective

- Explain the linearity assumption of multiple linear regression.
- Explain the consequences of violations of the linearity assumption
of multiple linear regression.
- Evaluate the validity of the linearity assumption of multiple linear regression
in a data set.


## Lesson

### Assumption

I'm going to express the linearity assumption of multiple linear regression
in two ways.
The first is with $\epsilon_i$ denoting the random error.
In multiple linear regression, we assume
$$
Y_i = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \cdots + \beta_p X_{ip} + \epsilon_i
$$
and
$$
E\left(\epsilon_i\middle|X_{i1},\dotsc,X_{ip}\right) = 0 \text{.}
$$
In simple linear regression, we had one explanatory variable.
In multiple linear regression, we have $p$ explanatory variables,
where $p \ge 2$.
We use $X_{ij}$ to denote the value of the $j$th explanatory variable
for the $i$th observation.
We still just have one random error for the $i$th observation,
denoted $\epsilon_i$.
We assume this random error has mean zero,
conditional on the explanatory variables.

The second way to denote the linearity assumption of multiple linear
regression is
$$
E\left(Y_i\middle|X_{i1},\dotsc,X_{ip}\right) = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \cdots + \beta_p X_{ip} \text{.}
$$
I find this representation to be less intuitive, but it will be useful
later, when we encounter logistic regression and Poisson regression.


### Note

Strictly, the linearity assumption is
$$
E(Y_i|X_{11},\dotsc,X_{1p},\dotsc,X_{n1},\dotsc,X_{np}) = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \cdots + \beta_p X_{ip} \text{,}
$$
but that's very cumbersome to write.
It's far simpler using linear algebra:
$$
E\left(\mathbf{Y}\middle|\mathbf{X}\right) = \mathbf{X}\mathbf{\beta} \text{.}
$$
Since linear algebra is not required for this course,
I'll stick to the simpler notation used in the previous section.

### Consequences

If the linearity assumption is met, then the coefficient estimates
(i.e., $\hat{\beta}_0$, $\hat{\beta}_1$, $\hat{\beta}_2$, etc.)
are unbiased.
This means we can interpret them.


### Evaluation

The best and most comprehensive way to check the linearity assumption
of multiple linear regression is to make scatterplots.
In each one, the residuals will be on the vertical axis and a different
explanatory variable will be on the horizontal axis.
We can also make a scatterplot with the residuals on the vertical axis
and the fitted values on the horizontal axis.

Let's look at some examples.

#### Example 1: Not met in either variable

In this example, the outcome doesn't have a linear relationship with
either explanatory variable.
The underlying regression equation is
$$
E\left(Y_i\middle|X_{i1},\dotsc,X_{ip}\right) = 1 + X_{i1}^2 + X_{i2}^2 \text{.}
$$
Note that $\beta_0 = \beta_1 = \beta_2 = 1$.
Let's look at the resulting plots.

```{r}
# Parameters
n <- 1e3
beta_0 <- 1
beta_1 <- 1
beta_2 <- 1

# Generate data
x_1 <- rnorm(n = n,
             sd = 2)
x_2 <- rnorm(n = n,
             sd = 2)
epsilon <- rnorm(n)
y_neither <- beta_0 + beta_1 * x_1^2 + beta_2 * x_2^2 + epsilon

# Fit model
mod_01 <- lm(y_neither ~ x_1 + x_2)

# Plots
plot(x = mod_01$model$x_1,
     y = mod_01$residuals,
     xlab = expression(X[1]),
     ylab = "Residuals",
     pch = 19,
     col = alpha(colour = "black",
                 alpha = 0.2))
abline(h = 0,
       col = "red")

plot(x = mod_01$model$x_2,
     y = mod_01$residuals,
     xlab = expression(X[2]),
     ylab = "Residuals",
     pch = 19,
     col = alpha(colour = "black",
                 alpha = 0.2))
abline(h = 0,
       col = "red")

plot(x = mod_01$fitted.values,
     y = mod_01$residuals,
     xlab = expression(hat(Y)),
     ylab = "Residuals",
     pch = 19,
     col = alpha(colour = "black",
                 alpha = 0.2))
abline(h = 0,
       col = "red")
```
In each plot, there is a non-linear relationship between the mean of the
residuals and the variable on the horizontal axis.

#### Example 2: Met in one variable

In this example, the outcome doesn't have a linear relationship with
one explanatory variable,
but it does with the other.
The underlying regression equation is
$$
E\left(Y_i\middle|X_{i1},\dotsc,X_{ip}\right) = 1 + X_{i1} + X_{i2}^2 \text{.}
$$
Note that $\beta_0 = \beta_1 = \beta_2 = 1$.
Let's look at the resulting plots.

```{r}
# Generate data
y_one <- beta_0 + beta_1 * x_1 + beta_2 * x_2^2 + epsilon

# Fit model
mod_02 <- lm(y_one ~ x_1 + x_2)

# Plots
plot(x = mod_02$model$x_1,
     y = mod_02$residuals,
     xlab = expression(X[1]),
     ylab = "Residuals",
     pch = 19,
     col = alpha(colour = "black",
                 alpha = 0.2))
abline(h = 0,
       col = "red")

plot(x = mod_02$model$x_2,
     y = mod_02$residuals,
     xlab = expression(X[2]),
     ylab = "Residuals",
     pch = 19,
     col = alpha(colour = "black",
                 alpha = 0.2))
abline(h = 0,
       col = "red")

plot(x = mod_02$fitted.values,
     y = mod_02$residuals,
     xlab = expression(hat(Y)),
     ylab = "Residuals",
     pch = 19,
     col = alpha(colour = "black",
                 alpha = 0.2))
abline(h = 0,
       col = "red")
```
For two of the plots (the one with $X_1$ on the horizontal axis
and the one with $\hat{Y}$ on the horizontal axis),
it appears that the residuals have mean 0 regardless of the value
of the variable on the horizontal axis.
For one of the plots (the one with $X_2$ on the horizontal axis),
there is a clear non-linear relationship between the mean of the residuals
and the variable on the horizontal axis.

There are a couple important things to note here.
The first is that in practice, many people only look at a plot of the
residuals as a function of the fitted values.
This saves time, especially if you have many explanatory variables.
However, as shown in this example, if you do this you may not catch
a non-linearity.

Second, it may appear as though the mean of the residuals is not 0
for all values of $X_1$ (or for all fitted values)
since so many of the points lie below the horizontal line at 0.
Recall that the mean of the residuals needs to be 0 throughout the plot,
not the median.
If we look at a histogram of the residuals, we see that they are right-skewed.

```{r}
hist(x = mod_02$residuals,
     xlab = "Residuals",
     main = "")
abline(v = mean(mod_02$residuals),
       col = "blue")
abline(v = median(mod_02$residuals),
       col = "purple")
```
I put a vertical blue line at the mean (which is equal to zero)
and a vertical purple line at the median.
If you can't distinguish blue from purple,
just know that the mean is the right-most vertical line.
Since the residuals are right-skewed,
and the mean is sensitive to outliers,
the mean is greater than the median.

So, in the scatterplots above, the median may lie below the red horizontal line,
but the mean is approximately at the red horizontal line,
no matter the value of $X_1$ or $\hat{Y}$.

#### Example 3: Met in both variables

In this example, the outcome has a linear relationship with
both explanatory variables.
The underlying regression equation is
$$
E\left(Y_i\middle|X_{i1},\dotsc,X_{ip}\right) = 1 + X_{i1} + X_{i2} \text{.}
$$
Note that $\beta_0 = \beta_1 = \beta_2 = 1$.
Let's look at the resulting plots.

```{r}
# Generate data
y_both <- beta_0 + beta_1 * x_1 + beta_2 * x_2 + epsilon

# Fit model
mod_03 <- lm(y_both ~ x_1 + x_2)

# Plots
plot(x = mod_03$model$x_1,
     y = mod_03$residuals,
     xlab = expression(X[1]),
     ylab = "Residuals",
     pch = 19,
     col = alpha(colour = "black",
                 alpha = 0.2))
abline(h = 0,
       col = "red")

plot(x = mod_03$model$x_2,
     y = mod_03$residuals,
     xlab = expression(X[2]),
     ylab = "Residuals",
     pch = 19,
     col = alpha(colour = "black",
                 alpha = 0.2))
abline(h = 0,
       col = "red")

plot(x = mod_03$fitted.values,
     y = mod_03$residuals,
     xlab = expression(hat(Y)),
     ylab = "Residuals",
     pch = 19,
     col = alpha(colour = "black",
                 alpha = 0.2))
abline(h = 0,
       col = "red")
```
In all three plots, the mean of the residuals is approximately zero,
no matter the value of the variable on the horizontal axis.

#### Note

Note that it's not necessarily a problem if
the underlying regression equation is
$$
E\left(Y_i\middle|X_{i1},\dotsc,X_{ip}\right) = 1 + X_{i1}^2 + X_{i2}^2 \text{.}
$$
We can just treat $X_{i1}^2$ and $X_{i2}^2$ as the explanatory variables
instead of $X_{i1}$ and $X_{i2}$.
For the plots below, I've fit this model.

```{r}
# New explanatory variables
x_1_2 <- x_1^2
x_2_2 <- x_2^2

# Fit model
mod_04 <- lm(y_neither ~ x_1_2 + x_2_2)

# Plots
plot(x = mod_04$model$x_1_2,
     y = mod_04$residuals,
     xlab = expression(X[1]^2),
     ylab = "Residuals",
     pch = 19,
     col = alpha(colour = "black",
                 alpha = 0.2))
abline(h = 0,
       col = "red")

plot(x = mod_04$model$x_2_2,
     y = mod_04$residuals,
     xlab = expression(X[2]^2),
     ylab = "Residuals",
     pch = 19,
     col = alpha(colour = "black",
                 alpha = 0.2))
abline(h = 0,
       col = "red")

plot(x = mod_04$fitted.values,
     y = mod_04$residuals,
     xlab = expression(hat(Y)),
     ylab = "Residuals",
     pch = 19,
     col = alpha(colour = "black",
                 alpha = 0.2))
abline(h = 0,
       col = "red")
```
Note that, no matter the value of the variable on the horizontal axis,
the mean of the residuals is approximately zero.

If you're curious about why the residuals are bunched up to the left,
it's because $X_1,X_2\overset{\text{iid}}{\sim}N(0,4)$.
Most of the values of $X_1$ and $X_2$ are thus close to zero.
When they get squared, we double up on the values that are close to zero
but positive.
Recall that $(-x)^2 = x^2 \ge 0$.

## Problems

### Problem 1

According to the following scatterplots,
is the linearity assumption of multiple linear regression met
for these data?

```{r}
# Generate data
y_exp <- beta_0 + beta_1 * x_1 + beta_2 * 1.55^x_2 + epsilon

# Fit model
mod_05 <- lm(y_exp ~ x_1 + x_2)

# Plots
plot(x = mod_05$model$x_1,
     y = mod_05$residuals,
     xlab = expression(X[1]),
     ylab = "Residuals",
     pch = 19,
     col = alpha(colour = "black",
                 alpha = 0.2))
abline(h = 0,
       col = "red")

plot(x = mod_05$model$x_2,
     y = mod_05$residuals,
     xlab = expression(X[2]),
     ylab = "Residuals",
     pch = 19,
     col = alpha(colour = "black",
                 alpha = 0.2))
abline(h = 0,
       col = "red")

plot(x = mod_05$fitted.values,
     y = mod_05$residuals,
     xlab = expression(hat(Y)),
     ylab = "Residuals",
     pch = 19,
     col = alpha(colour = "black",
                 alpha = 0.2))
abline(h = 0,
       col = "red")
```

::: {.callout-tip title="Solution" collapse="true"}

The linearity assumption is not met for these data.
Although the residuals have approximate mean zero for any value of $X_1$
and $\hat{Y}$,
their mean has a non-linear relationship with $X_2$.
The mean of the residuals is above zero when $|X_2| \ge 3$,
and the man of the residuals is below zero when $0 \le X_2 \le 2$.

:::

### Problem 2

Given your answer to Problem 1,
should you interpret the estimated coefficients in that model?

::: {.callout-tip title="Solution" collapse="true"}

No.
Since the linearity assumption was not met,
we should not interpret the coefficient estimates.

Some might argue that even when the linearity assumption is not met
for some of the explanatory variables,
you can interpret the coefficient estimates for the other explanatory
variables.
My suggestion is that you leave this judgment up to a statistician,
at least at this stage in your training.

:::




