---
title: "Homoscedasticity Assumption"
execute:
  echo: false
  warning: false
format: html
---

```{r}
library(ggplot2)
```

## Learning Objective

- Explain the homoscedasticity assumption of multiple linear regression.
- Explain the consequences of violations of the homoscedasticity assumption
of multiple linear regression.
- Evaluate the validity of the homoscedasticity assumption of multiple linear regression
in a data set.


## Lesson

### Assumption

The homoscedasticity assumption of multiple linear regression
can be written in at least two ways.
First, we assume
$$
Y_i = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \cdots + \beta_p X_{ip} + \epsilon_i
$$
and
$$
Var\left(\epsilon_i\middle|X_{i1},\dotsc,X_{ip}\right) = \sigma^2 \text{.}
$$
Note that $\sigma^2$ doesn't depend on $i$ or any of the $X_{ij}$'s;
it's a constant.
As in simple linear regression, "homoscedasticity" means "constant (or same) variance".

The other way of writing this assumption is
$$
Var\left(Y_i\middle|X_{i1},\dotsc,X_{ip}\right) = \sigma^2 \text{.}
$$


### Consequences

We need this assumption for our standard errors to be correct.
If this assumption is not met (but the linearity and iid assumptions are
and either the normality assumption is met or the sample size is large),
we can use robust standard errors to get valid p-values.
The code is the same as for simple linear regression.

### Evaluation

Typically, to evaluate this assumption,
we look at a scatterplot with the residuals on the vertical axis
and the fitted values on the horizontal axis.
What we want to see is constant vertical spread throughout the plot.
This is similar to the evaluation of this assumption for simple linear regression,
except for simple linear regression, we put the explanatory variable
on the horizontal axis.
With multiple linear regression, we have multiple explanatory variables,
so to efficiently check the assumption we put the fitted value on
the horizontal axis.

I don't know of anyone who, in practice, checks for heteroscedasticity
by plotting the residuals against each covariate.
It's good to plot the residuals against each covariate to check
for non-linearity;
if you did this and noticed that some of the plots had non-constant variance
and some didn't,
it would be worth investigating.

One situation that would lead to this happening is if one variable
had no relationship with the outcome.
To make the plots below,
I generated $X$ and $Y$ independently from each other,
then generated $Z$ from the model
$$
Z_i = X_i + X_i \epsilon_i \text{.}
$$
In other symbols,
$$
E(Z_i|X_i) = X_i
$$
but 
$$
Var(Z_i|X_i) = X_i^2 \sigma^2 \text{.}
$$
$Y$ is totally independent of $Z$.

```{r}
# Set seed
set.seed(3)

# Set parameters
n <- 1e2

# Generate data
X <- rnorm(n = n,
           sd = 2)
Y <- rnorm(n = n,
           sd = 2)
epsilon <- rnorm(n)
Z <- X + X * epsilon

# Plots
pairs(data.frame(X,Y,Z))
```

If we make three residual scatterplots, each of which has something different
on the horizontal axis,
we see heteroscedasticity in some of them but not all.

```{r}
mod_01 <- lm(Z ~ X + Y)
plot(x = mod_01$fitted.values,
     y = mod_01$residuals,
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0,
       col = "red")
plot(x = mod_01$model$X,
     y = mod_01$residuals,
     xlab = "X",
     ylab = "Residuals")
abline(h = 0,
       col = "red")
plot(x = mod_01$model$Y,
     y = mod_01$residuals,
     xlab = "Y",
     ylab = "Residuals")
abline(h = 0,
       col = "red")
```

## Problems

### Problem 1

As stated in the
[manual](https://cran.r-project.org/web/packages/Lock5Data/Lock5Data.pdf)
for the `Lock5Data` package,
"The California Department of Developmental Services (DDS) allocates funds to
support developmentally disabled California residents
(such as those with autism, cerebral palsy, or intellectual
disabilities) and their families.
We refer to those supported by DDS as DDS consumers. The
dataset DDS includes data on annual expenditure (in $), ethnicity, age, and
gender for 1000 DDS consumers.
[. . .]
The dataset has been altered slightly for
privacy reasons, but is based on actual DDS consumers."

In order to create the plots below,
I used annual expenditure as the outcome
and ethnicity (Hispanic, White not Hispanic Multi Race,
Asian, Black, Other, American Indian, Native Hawaiian),
age cohort (0-5, 6-12, 13-17, 18-21, 22-50, 51 and above),
and gender (male, female) as explanatory variables.

```{r}
d <- Lock5Data::DDS
mod_02 <- lm(formula = Expenditures ~ AgeCohort + Gender + Ethnicity,
             data = d)
ggplot(mapping = aes(x = mod_02$fitted.values,
                     y = mod_02$residuals)) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 0,
             colour = "red") +
  theme_bw() +
  xlab("Fitted Values") +
  ylab("Residuals")
boxplot(formula = mod_02$residuals ~ mod_02$model$Ethnicity,
        xlab = "Ethnicity",
        ylab = "Residuals")
abline(h = 0,
       col = "red")
boxplot(formula = mod_02$residuals ~ mod_02$model$AgeCohort,
        xlab = "Age Cohort",
        ylab = "Residuals")
abline(h = 0,
       col = "red")
boxplot(formula = mod_02$residuals ~ mod_02$model$Gender,
        xlab = "Gender",
        ylab = "Residuals")
abline(h = 0,
       col = "red")
```



Is the homoscedasticity assumption met for this model?
Justify your answer.

::: {.callout-tip title="Solution" collapse="true"}

The homoscedasticity assumption is not met for this model.
In the plot of the residuals as a function of the fitted values,
we see far lower variability for lower fitted values than
for higher fitted values.

This example is interesting because it's a real-world example where
there's heteroscedasticity for some variables but not others.
When ethnicity and age cohort are on the horizontal axis,
we see non-constant variance;
when gender is on the horizontal axis,
we see equal variance in males and females.
I'm not sure why this is,
but practically, because there's such obvious heteroscedasticity
in the fitted value plot,
we would need to use the robust (sandwich) standard errors.

If we look at the results from the model,
we see that the p-value for Gender is very small,
indicating that there is a statistically significant difference in expenditures
between males and females,
holding all else constant.
So, this example doesn't appear to be the same as the simulated data above.
```{r}
summary(mod_02)
```

Also, sorry that the ethnic groups are not displaying properly
in the boxplot above.
I'll fix that eventually, but it's not a high priority right now.
Even without all the group names displayed, you can tell there's
heteroscedasticity.

:::






