---
title: "Interpretation"
execute:
  echo: false
format: html
---

## Learning Objective

Interpret the coefficients in multiple linear regression.


## Lesson

### General Interpretation

The interpretation of coefficients in multiple linear regression
is very similar to the interpretation of coefficients in simple linear
regression,
with one added detail.
The intercept ($\beta_0$) is still the expected value of the outcome
when all the explanatory variables equal 0.
The coefficient corresponding to the $j$th explanatory variable,
denoted $\beta_j$,
is the expected change in the outcome associated with a one-unit increase
in the $j$th explanatory variable,
*holding all else constant*.
If $\beta_j > 0$, the expected change is an increase;
if $\beta_j < 0$, the expected change is a decrease.
The "else" in "holding all else constant" refers to the other explanatory variables.
Another way of interpreting $\beta_j$ is as the expected difference in the outcome
between two populations that differ by one unit on the $j$th explanatory variable
and have identical values for the other explanatory variables.

### Math

$$
\begin{align*}
& E\left(Y_i\middle|X_{i1},\dotsc,X_{ij} = x + 1,\dotsc,X_{ip}\right) - E\left(Y_i\middle|X_{i1},\dotsc,X_{ij} = x ,\dotsc,X_{ip}\right) \\
&= \beta_0 + \beta_1 X_{i1} + \cdots + \beta_j (x+1) + \cdots + \beta_p X_{ip} - \left(\beta_0 + \beta_1 X_{i1} + \cdots + \beta_j x + \cdots + \beta_p X_{ip}\right) \\
&= \beta_0 + \beta_1 X_{i1} + \cdots + \beta_j x + \beta_j + \cdots + \beta_p X_{ip} - \beta_0 - \beta_1 X_{i1} - \cdots - \beta_j x - \cdots - \beta_p X_{ip} \\
&= \beta_j
\end{align*}
$$


### Example

The following output comes from a regression of heart rate
(in beats per minute)
on age (in years), sex (male or female), admission type (elective or emergency),
and consciousness (conscious, deep stupor, or coma).
The participants were 200 patients admitted to an adult
intensive care unit (ICU).

```{r}
d <- Lock5Data::ICUAdmissions
# sum(complete.cases(d))
d$sex <- "Male"
d$sex[d$Sex == 1] <- "Female"
d$type <- "Elective"
d$type[d$Type == 1] <- "Emergency"
d$consciousness <- "Conscious"
d$consciousness[d$Consciousness == 1] <- "Deep Stupor"
d$consciousness[d$Consciousness == 2] <- "Coma"
mod_01 <- lm(formula = HeartRate ~ Age + sex + type + consciousness,
             data = d)
summary(mod_01)
```

Let's interpret each estimate.

#### Intercept

The estimated mean heart rate among female patients aged zero,
who elected to come to the ICU, and who are in a coma,
is about 72 beats per minute.
Note that this intercept doesn't make sense.
How could a newborn in a coma elect to come to the adult ICU?
Even if a newborn in a coma went to the ICU for a non-urgent reason,
her heart rate wouldn't be 72 beats per minute.
That's way too low.

#### Continuous Explanatory Variable

The estimated mean increase in heart rate associated with a one-year increase
in age, holding all else constant,
is 0.1 beats per minute.

#### Binary Explanatory Variable

The estimated mean difference in heart rate between males and females,
holding all else constant,
is 0.7 beats per minute, with men having the lower estimated mean heart rate.
How did I know that men had the lower estimated mean heart rate?
This coefficient is the expected increase in heart rate associated
with a one-unit increase in the explanatory variable,
holding all else constant.
This particular explanatory variable only has two possible values:
0 (for females) and 1 (for males).
So, a one-unit increase corresponds to a change from female to male.
Since the coefficient estimate is negative, males have a lower estimated
mean heart rate, holding all else constant.

Patients coming to the ICU for an emergency are expected to have
a heart rate 12 beats per minute faster than patients electing
to come to the ICU, holding all else constant.

#### Categorical Explanatory Variable

Conscious patients are expected to have a heart rate that is 1 beat per minute
higher than comatose patients, holding all else equal.
Why did I compare conscious patients to comatose patients?
The dummy variable `consciousnessConscious` has two possible values, 0 and 1.
A 1 can only mean one thing: that patient was conscious.
A 0 can mean one of two things: either that patient was comatose
or that patient was in a deep stupor.
So why is it that the estimated coefficient associated with
`consciousnessConscious` doesn't compare comatose and deep stupor patients
to conscious patients?

Recall that when interpreting a coefficient in multiple linear regression,
we increase the corresponding covariate value by one unit
and hold all other covariate values constant.
So, when interpreting the coefficient estimate for `consciousnessConscious`,
we have to increase `consciousnessConscious` by one unit
and hold `consciousnessDeep Stupor` constant.
As the following table shows,
the only way to increase `consciousnessConscious` by one unit
and hold `consciousnessDeep Stupor` constant
is to increase `consciousnessConscious` by one unit
and hold `consciousnessDeep Stupor` at 0.
This means we compare comatose patients to conscious patients.
 

| Original Level | `consciousnessConscious` | `consciousnessDeep Stupor` |
|:-|-:|-:|
| Coma | 0 | 0 |
| Conscious | 1 | 0 |
| Deep Stupor | 0 | 1 |

We can use the same logic to interpret the last coefficient estimate
in the table.
Patients in a deep stupor are expected to have a heart rate that is
13 beats per minute higher than comatose patients,
all else equal.


## Problems

### Problem 1

In the regression below,
each observation corresponds to a person receiving funds from
the California Department of Developmental Services (DDS).
I used annual expenditure (in dollars) as the outcome
and ethnicity (Hispanic, White not Hispanic, Multi Race,
Asian, Black, Other, American Indian, Native Hawaiian),
age cohort (0-5, 6-12, 13-17, 18-21, 22-50, 51 and above),
age (in years),
and gender (male, female) as explanatory variables.


```{r}
d2 <- Lock5Data::DDS
mod_02 <- lm(formula = Expenditures ~ AgeCohort + Age + Gender + Ethnicity,
             data = d2)
summary(mod_02)
```

Interpret each estimated coefficient.

::: {.callout-tip title="Solution" collapse="true"}

- The estimated mean expenditure for an American Indian newborn female
is -\$2,083.85.
Obviously this can't be right.
- Holding all else constant, 13-17 year-olds are expected to receive
\$12.56 less than 0-5 year-olds.
Note that this is impossible while `Age` is in the model.
You can't have a 13-17 year-old who's the same age as a 0-5 year-old.
- Holding all else constant, 18-21 year-olds are expected to receive
\$5,040.75 more than 0-5 year-olds.
Again, this is impossible while `Age` is in the model.
You can't have an 18-21 year-old who's the same age as a 0-5 year-old.
I won't keep saying this each time.
- Holding all else constant, 22-50 year-olds are expected to receive
\$33,479.46 more than 0-5 year-olds.
- Holding all else constant, people who are 51 or older are expected to receive
\$38,202.87 more than 0-5 year-olds.
- Holding all else constant, 6-12 year-olds are expected to receive
\$497.84 less than 0-5 year-olds.
- Holding all else constant, we would expect an increase of \$212.08
with each additional year of age.
This is possible, even with `AgeCohort` in the model.
You can compare people within the same age cohort who differ by one year in age.
- Holding all else constant,
males are expected to receive \$956.76 less than females.
- Holding all else constant,
Asian recipients are expected to receive \$3,181.00 more than American Indian
recipients.
- Holding all else constant,
Black recipients are expected to receive \$3,424.40 more than American Indian
recipients.
- Holding all else constant,
Hispanic recipients are expected to receive \$3,454.99 more than American Indian
recipients.
- Holding all else constant,
multi race recipients are expected to receive \$3,341.38 more than American Indian
recipients.
- Holding all else constant,
Native Hawaiian recipients are expected to receive \$2,567.08 more than American Indian
recipients.
- Holding all else constant,
recipients who don't fall into the other categories are expected to receive
\$2,498.05 more than American Indian recipients.
- Holding all else constant,
White not Hispanic recipients are expected to receive \$3,116.09 more than American Indian
recipients.


:::





