---
title: "Interpretation"
execute:
  echo: false
format: html
---

## Learning Objective

Interpret the coefficients in multiple linear regression.


## Lesson

### General Interpretation

The interpretation of coefficients in multiple linear regression
is very similar to the interpretation of coefficients in simple linear
regression,
with one added detail.
The intercept ($\beta_0$) is still the expected value of the outcome
when all the explanatory variables equal 0.
The coefficient corresponding to the $j$th explanatory variable,
denoted $\beta_j$,
is the expected change in the outcome associated with a one-unit increase
in the $j$th explanatory variable,
*holding all else constant*.
If $\beta_j > 0$, the expected change is an increase;
if $\beta_j < 0$, the expected change is a decrease.
The "else" in "holding all else constant" refers to the other explanatory variables.
Another way of interpreting $\beta_j$ is as the expected difference in the outcome
between two populations that differ by one unit on the $j$th explanatory variable
and have identical values for the other explanatory variables.

### Math

$$
\begin{align*}
& E\left(Y_i\middle|X_{i1},\dotsc,X_{ij} = x + 1,\dotsc,X_{ip}\right) - E\left(Y_i\middle|X_{i1},\dotsc,X_{ij} = x ,\dotsc,X_{ip}\right) \\
&= \beta_0 + \beta_1 X_{i1} + \cdots + \beta_j (x+1) + \cdots + \beta_p X_{ip} - \left(\beta_0 + \beta_1 X_{i1} + \cdots + \beta_j x + \cdots + \beta_p X_{ip}\right) \\
&= \beta_0 + \beta_1 X_{i1} + \cdots + \beta_j x + \beta_j + \cdots + \beta_p X_{ip} - \beta_0 - \beta_1 X_{i1} - \cdots - \beta_j x - \cdots - \beta_p X_{ip} \\
&= \beta_j
\end{align*}
$$


### Example

The following output comes from a regression of heart rate
(in beats per minute)
on age (in years), sex (male or female), admission type (elective or emergency),
and consciousness (conscious, deep stupor, or coma).
The participants were 200 patients admitted to an adult
intensive care unit (ICU).

```{r}
d <- Lock5Data::ICUAdmissions
# sum(complete.cases(d))
d$sex <- "Male"
d$sex[d$Sex == 1] <- "Female"
d$type <- "Elective"
d$type[d$Type == 1] <- "Emergency"
d$consciousness <- "Conscious"
d$consciousness[d$Consciousness == 1] <- "Deep Stupor"
d$consciousness[d$Consciousness == 2] <- "Coma"
mod_01 <- lm(formula = HeartRate ~ Age + sex + type + consciousness,
             data = d)
summary(mod_01)
```

Let's interpret each estimate.

#### Intercept

The estimated mean heart rate among female patients aged zero,
who elected to come to the ICU, and who are in a coma,
is about 72 beats per minute.
Note that this intercept doesn't make sense.
How could a newborn in a coma elect to come to the adult ICU?
Even if a newborn in a coma went to the ICU for a non-urgent reason,
her heart rate wouldn't be 72 beats per minute.
That's way too low.

#### Continuous Explanatory Variable

The estimated mean increase in heart rate associated with a one-year increase
in age, holding all else constant,
is 0.1 beats per minute.

#### Binary Explanatory Variable

The estimated mean difference in heart rate between males and females,
holding all else constant,
is 0.7 beats per minute, with men having the lower estimated mean heart rate.
How did I know that men had the lower estimated mean heart rate?
This coefficient is the expected increase in heart rate associated
with a one-unit increase in the explanatory variable,
holding all else constant.
This particular explanatory variable only has two possible values:
0 (for females) and 1 (for males).
So, a one-unit increase corresponds to a change from female to male.
Since the coefficient estimate is negative, males have a lower estimated
mean heart rate, holding all else constant.

Patients coming to the ICU for an emergency are expected to have
a heart rate 12 beats per minute faster than patients electing
to come to the ICU, holding all else constant.

#### Categorical Explanatory Variable

Conscious patients are expected to have a heart rate that is 1 beat per minute
higher than comatose patients, holding all else equal.
Why did I compare conscious patients to comatose patients?
The dummy variable `consciousnessConscious` has two possible values, 0 and 1.
A 1 can only mean one thing: that patient was conscious.
A 0 can mean one of two things: either that patient was comatose
or that patient was in a deep stupor.
So why is it that the estimated coefficient associated with
`consciousnessConscious` doesn't compare comatose and deep stupor patients
to conscious patients?

Recall that when interpreting a coefficient in multiple linear regression,
we increase the corresponding covariate value by one unit
and hold all other covariate values constant.
So, when interpreting the coefficient estimate for `consciousnessConscious`,
we have to increase `consciousnessConscious` by one unit
and hold `consciousnessDeep Stupor` constant.
As the following table shows,
the only way to increase `consciousnessConscious` by one unit
and hold `consciousnessDeep Stupor` constant
is to increase `consciousnessConscious` by one unit
and hold `consciousnessDeep Stupor` at 0.
This means we compare comatose patients to conscious patients.
 

| Original Level | `consciousnessConscious` | `consciousnessDeep Stupor` |
|:-|-:|-:|
| Coma | 0 | 0 |
| Conscious | 1 | 0 |
| Deep Stupor | 0 | 1 |

We can use the same logic to interpret the last coefficient estimate
in the table.
Patients in a deep stupor are expected to have a heart rate that is
13 beats per minute higher than comatose patients,
all else equal.


## Problems

### Problem 1

In the regression below,
each observation corresponds to a person receiving funds from
the California Department of Developmental Services (DDS).
I used annual expenditure (in dollars) as the outcome
and ethnicity (Hispanic, White not Hispanic, Multi Race,
Asian, Black, Other, American Indian, Native Hawaiian),
age cohort (0-5, 6-12, 13-17, 18-21, 22-50, 51 and above),
age (in years),
and gender (male, female) as explanatory variables.


```{r}
d2 <- Lock5Data::DDS
mod_02 <- lm(formula = Expenditures ~ AgeCohort + Age + Gender + Ethnicity,
             data = d2)
summary(mod_02)
```

Interpret each estimated coefficient.

::: {.callout-tip title="Solution" collapse="true"}

- The estimated mean expenditure for an American Indian newborn female
is -\$2,083.85.
Obviously this can't be right.
- Holding all else constant, 13-17 year-olds are expected to receive
\$12.56 less than 0-5 year-olds.
Note that this is impossible while `Age` is in the model.
You can't have a 13-17 year-old who's the same age as a 0-5 year-old.
- Holding all else constant, 18-21 year-olds are expected to receive
\$5,040.75 more than 0-5 year-olds.
Again, this is impossible while `Age` is in the model.
You can't have an 18-21 year-old who's the same age as a 0-5 year-old.
I won't keep saying this each time.
- Holding all else constant, 22-50 year-olds are expected to receive
\$33,479.46 more than 0-5 year-olds.
- Holding all else constant, people who are 51 or older are expected to receive
\$38,202.87 more than 0-5 year-olds.
- Holding all else constant, 6-12 year-olds are expected to receive
\$497.84 less than 0-5 year-olds.
- Holding all else constant, we would expect an increase of \$212.08
with each additional year of age.
This is possible, even with `AgeCohort` in the model.
You can compare people within the same age cohort who differ by one year in age.
- Holding all else constant,
males are expected to receive \$956.76 less than females.
- Holding all else constant,
Asian recipients are expected to receive \$3,181.00 more than American Indian
recipients.
- Holding all else constant,
Black recipients are expected to receive \$3,424.40 more than American Indian
recipients.
- Holding all else constant,
Hispanic recipients are expected to receive \$3,454.99 more than American Indian
recipients.
- Holding all else constant,
multi race recipients are expected to receive \$3,341.38 more than American Indian
recipients.
- Holding all else constant,
Native Hawaiian recipients are expected to receive \$2,567.08 more than American Indian
recipients.
- Holding all else constant,
recipients who don't fall into the other categories are expected to receive
\$2,498.05 more than American Indian recipients.
- Holding all else constant,
White not Hispanic recipients are expected to receive \$3,116.09 more than American Indian
recipients.


:::

### Problem 2

Suppose you run the following multiple linear regression
with "country" as the unit of observation
(i.e., each line in the data set is a country).

- Outcome variable
  - Life expectancy (in years)
- Explanatory variables
  - Pharmaceutical spending as a percentage of total health spending (in percentage points)
  - Total pharmaceutical spending (in millions of USD)
  - Pharmaceutical spending per capita (in USD)

"Pharmaceutical spending" refers to spending by consumers on both prescription
and over-the-counter medication.

For the following questions, assume all relevant assumptions have been met.

#### Part (a)

How would you interpret the intercept for this analysis?
Is it meaningful? Why or why not?

::: {.callout-tip title="Solution" collapse="true"}

The intercept is the expected life expectancy  for a country where consumers spend no
money whatsoever on pharmaceuticals.
This would make pharmaceutical spending zero as a percentage of total health spending;
per capita; and in total.
This is not meaningful, because such a country doesn't exist.
Even in very poor countries, people spend money on drugs.

:::

#### Part (b)

How would you interpret the coefficient on
"Pharmaceutical spending as a percentage of total health spending"?
What are the implications of "holding all else equal" for the interpretation
of this coefficient?

::: {.callout-tip title="Solution" collapse="true"}

This is the expected increase in life expectancy associated
with a one percentage point increase in pharmaceutical spending as a percentage
of total health spending, holding all else equal.

If we increase pharmaceutical spending as a percentage of total health spending
while fixing total pharmaceutical spending,
we must be decreasing total health spending.
If we increase pharmaceutical spending as a percentage of total health spending
while fixing total pharmaceutical spending and pharmaceutical spending per capita,
we must be decreasing total health spending and fixing the population of the country.
So, this coefficient is a measure of the effect of decreasing total health spending
(to the extent that we can interpret this causally).

Another way to interpret this coefficient is the expected difference between two countries
that differ in their pharmaceutical spending as a percentage of total health spending
by one percentage point,
but that have identical total pharmaceutical spending,
identical pharmaceutical spending per capita,
and (as a result of the first two) identical populations.
Even in this interpretation, one country must spend more on total health spending than the other
in order to achieve the percentage difference,
and the coefficient is a measure of the effect of decreasing total health spending
(to the extent that we can interpret this causally).
:::

#### Part (c)

How would you interpret the coefficient on
"Total pharmaceutical spending"?
What are the implications of "holding all else equal" for the interpretation
of this coefficient?

::: {.callout-tip title="Solution" collapse="true"}

This is the expected increase in life expectancy associated with a
one million USD increase in total pharmaceutical spending,
holding all else equal.

If we increase total pharmaceutical spending while fixing pharmaceutical spending
per capita, we must be increasing the population as well.
If we increase total pharmaceutical spending while fixing pharmaceutical spending
as a percentage of total health spending,
we must be increasing total health spending as well.
So, this coefficient is measuring the expected change in life expectancy
associated with an increase in total pharmaceutical spending,
population, and total health spending.
That's a lot all at once!
:::

#### Part (d)

How would you interpret the coefficient on
"Pharmaceutical spending per capita"?
What are the implications of "holding all else equal" for the interpretation
of this coefficient?

::: {.callout-tip title="Solution" collapse="true"}

This is the expected increase in life expectancy associated with a
one USD increase in pharmaceutical spending per capita,
holding all else equal.

If we increase pharmaceutical spending per capita while fixing total
pharmaceutical spending, we must be decreasing the population.
If we're fixing total pharmaceutical spending and pharmaceutical spending
as a percentage of total health spending, then we must be fixing total health spending.
So, this coefficient is a measure of the effect of decreasing the population
while holding total pharmaceutical spending and total health spending constant.
        
:::
        
### Problem 3

Consider the paper 
[Association between Traffic Air Pollution and Reduced Forced Vital Capacity:
A Study Using Personal Monitors for Outdoor Workers](https://doi.org/10.1371/journal.pone.0163225).
Suppose we run two analyses, both with forced vital capacity (FVC; in liters) as the outcome.
The first analysis is a simple linear regression with "time in job" (in years)
as the explanatory variable,
and the second is a multiple linear regression with "time in job" (in years)
and age (in years) as the explanatory variables.
Feel free to try this in R!

#### Part (a)

Interpret the intercept for the simple linear regression.

::: {.callout-tip title="Solution" collapse="true"}

This is the expected FVC for an outdoor worker who just started their job.

:::

#### Part (b)

Interpret the intercept for the multiple linear regression.

::: {.callout-tip title="Solution" collapse="true"}

This is the expected FVC for an outdoor worker who just started their job
and is zero years old.

:::

#### Part (c)

Interpret the coefficient on "time in job" for the simple linear regression.

::: {.callout-tip title="Solution" collapse="true"}

This is the expected increase in FVC associated with a one-year increase in time in job.
Alternatively, it's the expected difference in FVC between two workers who
differ by one year in their time in the job.

:::

#### Part (d)

Interpret the coefficient on "time in job" for the multiple linear regression.

::: {.callout-tip title="Solution" collapse="true"}

This is the expected increase in FVC associated with a one-year increase in time in job,
holding age constant.
Alternatively, this is the expected difference in FVC between two workers who have the same age
but differ by one year in their time in the job.

:::

#### Part (e)

Interpret the coefficient on age for the multiple linear regression.

::: {.callout-tip title="Solution" collapse="true"}

This is the expected increase in FVC associated with a one-year increase in age,
holding time in job constant.
Alternatively, this is the expected difference in FVC between two workers who have the same
years of experience but who differ by one year in their age.
        
:::

### Problem 4

Suppose we run two analyses, both with prostate-specific antigen level (PSA; in ng/mL)
as the outcome. The first analysis is a simple linear regression with smallpox vaccination
(yes = 1, no = 0) as the explanatory variable, and the second is a multiple linear regression
with smallpox vaccination and age (in years) as the explanatory variables.

#### Part (a)

Interpret the intercept for the simple linear regression.

::: {.callout-tip title="Solution" collapse="true"}

This is the expected PSA for someone who was not vaccinated against smallpox.

:::

#### Part (b)

Interpret the intercept for the multiple linear regression.

::: {.callout-tip title="Solution" collapse="true"}

This is the expected PSA for someone who was not vaccinated against smallpox
and is zero years old.

:::

#### Part (c)

Interpret the coefficient on smallpox vaccination for the simple linear regression.

::: {.callout-tip title="Solution" collapse="true"}

This is the expected difference in PSA between someone who has not been vaccinated
against smallpox and someone who has.

:::

#### Part (d)

Interpret the coefficient on smallpox vaccination for the multiple linear regression.

::: {.callout-tip title="Solution" collapse="true"}

This is the expected difference in PSA between two people of the same age,
one of whom has not been vaccinated against smallpox and one of whom has.

:::

#### Part (e)

Interpret the coefficient on age in the multiple linear regression.

::: {.callout-tip title="Solution" collapse="true"}

This is the expected increase in PSA associated with a one-year increase in age,
holding vaccination status constant.
Alternatively, this is the expected difference in PSA between
people age $x$ and people age $x + 1$,
where both groups have the same vaccination status.

:::

#### Part (f)

According to the Mayo Clinic,
"[PSA levels normally increase with age](https://www.mayoclinic.org/tests-procedures/psa-test/about/pac-20384731)".
According to the CDC,
"[Routine smallpox vaccination among the American public stopped in 1972 after the disease was eradicated in the United States](https://www.cdc.gov/smallpox/vaccines/index.html)."
What sign (positive, negative, or zero) do you think the slope in the simple linear regression
would have? Justify your answer.

::: {.callout-tip title="Solution" collapse="true"}

People who have been vaccinated against smallpox must be older,
and probably have higher PSA levels.
So, the slope in the simple linear regression should be positive.

:::

### Problem 5

Suppose you wanted to predict weight with height.
Could you include height in feet *and* height in meters as explanatory variables
in the same model?
Why or why not?

::: {.callout-tip title="Solution" collapse="true"}

No. Continuing with the theme of these problems,
we can think about what the interpretation of the coefficients would be.
The coefficient on height in feet would be the expected increase in weight
associated with a one-foot increase in height, holding height in meters constant.
You can't increase height in feet while holding height in meters constant.
Alternatively, the coefficient on height in feet would be the expected
difference in weight between two groups of people that differ in height by one foot
but have the same height in meters.
That's impossible.

That's not really a reason, though; as you'll see in the next question,
it's possible to run a regression even if the coefficients are not interpretable.
The real reason this isn't possible is mathematical.
To summarize, one explanatory variable cannot be a linear combination
of the others.
Problems can even arise if one explanatory variable is close to being a linear combination
of the others.
If you get an error or warning about this while running multiple linear regression,
you may need to remove one or more explanatory variables.
Consult a statistician.

:::

### Problem 6

Suppose you regress salary (in USD) on age (in years) and age$^2$ (in years$^2$).

#### Part (a)

How would you interpret the intercept?

::: {.callout-tip title="Solution" collapse="true"}

This is the expected salary for someone who is zero years old.

:::

#### Part (b)

How would you interpret the coefficient on age?

::: {.callout-tip title="Solution" collapse="true"}

You can't interpret this on its own because it's impossible to increase age by one unit
and not change age$^2$.
Interpreting the coefficients in a polynomial model is tricky;
consult a statistician.
Running this regression is possible because even though age$^2$ is a function of age,
it's not a linear function of age.
So, R will not complain.

:::




