---
title: "Hypothesis Testing"
execute:
  echo: false
format: html
---

## Learning Objective

Interpret the results of a hypothesis test in the context of multiple linear
regression.


## Lesson

### Overview

When conducting simple linear regression in R,
R will automatically compute p-values for each of the coefficients.
For each coefficient,
the null hypothesis is that the coefficient is zero,
and the alternative hypothesis is that the coefficient is not zero.
In symbols, each hypothesis test is
$$
\begin{align*}
H_0 &: \beta_j = 0 \\
H_1 &: \beta_j \ne 0
\end{align*}
$$
where $0 \le j \le p$.

### Example

Let's return to the ICU admissions data.
The outcome is heart rate
(in beats per minute)
and the explanatory variables are
age minus 20 (in years), sex (male or female), admission type (elective or emergency),
and consciousness (conscious, deep stupor, or coma).
The participants were 200 patients admitted to an adult
intensive care unit (ICU).
I subtracted 20 from age because the range of ages in the data set is
16 to 92, with first, second, and third quartiles at
46.75, 63, and 57.55.

```{r}
d <- Lock5Data::ICUAdmissions
# sum(complete.cases(d))
d$sex <- "Male"
d$sex[d$Sex == 1] <- "Female"
d$type <- "Elective"
d$type[d$Type == 1] <- "Emergency"
d$consciousness <- "Conscious"
d$age_20 <- d$Age - 20
d$consciousness[d$Consciousness == 1] <- "Deep Stupor"
d$consciousness[d$Consciousness == 2] <- "Coma"
mod_01 <- lm(formula = HeartRate ~ age_20 + sex + type + consciousness,
             data = d)
summary(mod_01)
```
How do we interpret the hypothesis test results?
Let's use $\alpha = 0.05$ as our threshold for significance.

- The p-value for the intercept is less than 0.05,
so we conclude that the expected heart rate for a twenty-year-old female
who came to the the ICU in a non-urgent coma is not equal to zero.
- The p-value for `age_20` is greater than 0.05,
so we fail to reject the null hypothesis that this coefficient is equal to zero.
There is no linear relationship between age and heart rate,
holding all else constant.
- The p-value for `sexMale` is greater than 0.05,
so there is no difference between the expected heart rate for males and females,
holding all else constant.
- The p-value for `typeEmergency` is less than 0.05,
so the expected heart rate for emergent patients is different
from the expected heart rate for elective patients,
holding all else equal.
- The p-value for `consciousnessConscious` is greater than 0.05,
so conscious patients do not have a different expected heart rate than
comatose patients, holding all else equal.
- The p-value for `consciousnessDeep Stupor` is greater than 0.05,
so patients in a deep stupor do not have a different expected heart rate
than comatose patients, holding all else equal.




## Problems



For the regression below, each data point corresponds to a man who had
his percent body fat estimated by underwater weighing.
The explanatory variables are age (in years),
weight (in pounds),
height (in inches),
neck circumference (in centimeters),
chest circumference (in centimeters),
abdomen circumference (in centimeters),
ankle circumference (in centimeters),
extended biceps circumference (in centimeters),
and wrist circumference (in centimeters).

```{r}
d2 <- Lock5Data::BodyFat
mod_02 <- lm(formula = Bodyfat ~ Age + Weight + Height + Neck + Chest + Abdomen + Ankle + Biceps + Wrist,
             data = d2)
# plot(x = mod_02$fitted.values,
#      y = mod_02$residuals)
# abline(h = 0,
#        col = "red")
# hist(mod_02$residuals)
summary(mod_02)
```




### Problem 1

Interpret the hypothesis test corresponding to each coefficient.
Use $\alpha = 0.05$.

::: {.callout-tip title="Solution" collapse="true"}


- The p-value for the intercept is greater than 0.05,
so we cannot conclude that the expected body fat percentage
for a newborn weighing zero pounds
and having zero for all length measurements is different from zero.
This makes sense.
Someone who effectively doesn't exist should have zero percent body fat.
- The p-value for `Age` is greater than 0.05,
so there is not a linear relationship between age and the expected
body fat percentage.
- The p-value for `Weight` is greater than 0.05,
so there is not a linear relationship between weight and the expected
body fat percentage.
- The p-value for `Height` is greater than 0.05,
so there is not a linear relationship between height and the expected
body fat percentage.
- The p-value for `Neck` is greater than 0.05,
so there is not a linear relationship between neck circumference and the expected
body fat percentage.
- The p-value for `Chest` is greater than 0.05,
so there is not a linear relationship between chest circumference and the expected
body fat percentage.
- The p-value for `Abdomen` is less than 0.05,
so there is a linear relationship beteen abdomen circumference and the expected
body fat percentage.
- The p-value for `Ankle` is greater than 0.05,
so there is not a linear relationship between ankle circumference and the expected
body fat percentage.
- The p-value for `Biceps` is greater than 0.05,
so there is not a linear relationship between biceps circumference and the expected
body fat percentage.
- The p-value for `Wrist` is less than 0.05,
so there is a linear relationship beteen wrist circumference and the expected
body fat percentage.

By the way, all of these conclusions only apply to men,
since all the participants in the study were men.

:::


### Problem 2

Why do you think height was measured in inches but 
the circumferences were measured in centimeters?
I don't have any special knowledge for why this was done,
but it's worth brainstorming ideas.

::: {.callout-tip title="Solution" collapse="true"}

It's possible that this was done because of convention.
Maybe body circumference measurements have traditionally been made
by one profession that uses the metric system,
whereas height is something everyone measures,
and these participants came from a country that uses the imperial system.

There is a computational reason why you might use different units
for different lengths.
Computing the coefficient estimates can get messed up when the variables
are on wildly different units.
(If you're curious why, look up
[condition number](https://en.wikipedia.org/wiki/Condition_number).)
So, if you're running a regression where one of the explanatory variables
has very small values and another has very large values
(like the weight of an elephant and the weight of the medicine it has been administered),
you should consider changing the units of one or both so that the values
are on similar scales
(like measuring the elephant's weight in kilograms and the medicine's weight
in milligrams).
(Apparently,
"[the effective dose of ibuprofen is six milligrams per kilogram of weight for Asian elephants and seven milligrams per kilogram for African elephants, given every 12 hours.](https://news.oregonstate.edu/news/osu-researchers-find-key-pachyderm-pain-relief)"
Who knew?)


:::





