---
title: "Interaction of Two Binary Variables"
execute:
  echo: false
format: html
---

## Learning Objective

- Interpret the coefficients of a multiple linear regression model with an
interaction between two binary explanatory variables.
- Define, use, and interpret the terms "moderation" and
"moderator".
- Interpret the results of a hypothesis test in the context of multiple linear
regression with an interaction term.


## Lesson

### Math

Suppose our linear model is
$$
E(Y_i|X_{i1},X_{i2}) = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \beta_3 X_{i1} X_{i2} \text{,}
$$
where both $X_{i1}$ and $X_{i2}$ can only be 0 or 1.
The product of the two explanatory variables, $X_{i1} X_{i2}$,
is called an interaction.
How do we interpret the coefficients when there's an interaction?

$\beta_0$ is still the expected value of the outcome when all the explanatory
variables are equal to zero:
$$
\begin{align*}
E(Y_i|X_{i1} = 0,X_{i2} = 0) &= \beta_0 + \beta_1 (0) + \beta_2 (0) + \beta_3 (0) (0) \\
&= \beta_0 \text{.}
\end{align*}
$$

$\beta_1$ is the expected difference in the outcome between
observations with $X_{1} = 0$ and $X_{1} = 1$,
holding $X_2$ constant *at zero*:
$$
\begin{align*}
E(Y_i|X_{i1} = 1,X_{i2} = 0) - E(Y_i|X_{i1} = 0,X_{i2} = 0) &= \beta_0 + \beta_1 (1) + \beta_2 (0) + \beta_3 (1) (0) - \left(\beta_0 + \beta_1 (0) + \beta_2 (0) + \beta_3 (0) (0)\right) \\
&= \beta_0 + \beta_1  - \left(\beta_0 \right) \\
&= \beta_1 \text{.}
\end{align*}
$$
Note that with the interaction term, we are not just holding the other variable
constant in general, we are holding it constant at a specific value.

$\beta_2$ is the expected difference in the outcome between
observations with $X_{2} = 0$ and $X_{2} = 1$,
holding $X_1$ constant at zero:
$$
\begin{align*}
E(Y_i|X_{i1} = 0,X_{i2} = 1) - E(Y_i|X_{i1} = 0,X_{i2} = 0) &= \beta_0 + \beta_1 (0) + \beta_2 (1) + \beta_3 (0) (1) - \left(\beta_0 + \beta_1 (0) + \beta_2 (0) + \beta_3 (0) (0)\right) \\
&= \beta_0 + \beta_2 (1)  - \left(\beta_0 \right) \\
&= \beta_2 \text{.}
\end{align*}
$$

$\beta_1 + \beta_3$ is the expected difference in the outcome between
observations with $X_{1} = 0$ and $X_{1} = 1$,
holding $X_2$ constant at 1:
$$
\begin{align*}
E(Y_i|X_{i1} = 1,X_{i2} = 1) - E(Y_i|X_{i1} = 0,X_{i2} = 1) &= \beta_0 + \beta_1 (1) + \beta_2 (1) + \beta_3 (1) (1) - \left(\beta_0 + \beta_1 (0) + \beta_2 (1) + \beta_3 (0) (1)\right) \\
&= \beta_0 + \beta_1 + \beta_2 + \beta_3 - \left(\beta_0  + \beta_2 \right) \\
&= \beta_1 + \beta_3 \text{.}
\end{align*}
$$


$\beta_2 + \beta_3$ is the expected difference in the outcome between
observations with $X_{2} = 0$ and $X_{2} = 1$,
holding $X_1$ constant at 1:
$$
\begin{align*}
E(Y_i|X_{i1} = 1,X_{i2} = 1) - E(Y_i|X_{i1} = 1,X_{i2} = 0) &= \beta_0 + \beta_1 (1) + \beta_2 (1) + \beta_3 (1) (1) - \left(\beta_0 + \beta_1 (1) + \beta_2 (0) + \beta_3 (1) (0)\right) \\
&= \beta_0 + \beta_1 + \beta_2 + \beta_3 - \left(\beta_0 + \beta_1  \right) \\
&= \beta_2 + \beta_3 \text{.}
\end{align*}
$$

### Example

Let's return to the ICU admissions data.
The outcome is heart rate
(in beats per minute).
In the following regression analysis,
I interacted
sex (male or female)
and admission type (elective or emergency).
The participants were 200 patients admitted to an adult
intensive care unit (ICU).


```{r}
d <- Lock5Data::ICUAdmissions
# sum(complete.cases(d))
d$sex <- "Male"
d$sex[d$Sex == 1] <- "Female"
d$type <- "Elective"
d$type[d$Type == 1] <- "Emergency"
mod_01 <- lm(formula = HeartRate ~ sex * type,
             data = d)
summary(mod_01)
```

How do we interpret the coefficient estimates?
To make this a little more clear,
I'm going to write the regression equation with words:
$$
\hat{E}\left(\texttt{HeartRate}\middle|\texttt{sexMale},\texttt{typeEmergency}\right) = 90.4667 + 0.9018 \texttt{sexMale} + 12.2219 \texttt{typeEmergency} - 2.5205 \texttt{sexMale}\times \texttt{typeEmergency} \text{.}
$$

The estimated mean heart rate among females who elected to come to the ICU
is 90.4667 beats per minute:
$$
\begin{align*}
\hat{E}\left(\texttt{HeartRate} \middle|\texttt{sexMale}= 0,\texttt{typeEmergency}= 0\right) &= 90.4667 + 0.9018 (0) + 12.2219 (0) - 2.5205 (0)\times (0) \\
&= 90.4667 \\
\end{align*}
$$

The estimated mean heart rate among males who elected to come to the ICU
is $90.4667 + 0.9018 = 91.3685$ beats per minute:
$$
\begin{align*}
\hat{E}\left(\texttt{HeartRate} \middle|\texttt{sexMale}= 1,\texttt{typeEmergency}= 0\right) &= 90.4667 + 0.9018 (1) + 12.2219 (0) - 2.5205 (1)\times (0) \\
&= 90.4667 + 0.9018 \\
&= 91.3685 \text{.}
\end{align*}
$$

As a result of the previous two interpretations,
the estimated mean difference in heart rate between males and females who
elected to come to the ICU is 0.9018 beats per minute.


The estimated mean heart rate among females who came to the ICU for an emergency
is $90.4667 + 12.2219 = 102.6886$ beats per minute:
$$
\begin{align*}
\hat{E}\left(\texttt{HeartRate} \middle|\texttt{sexMale}= 0,\texttt{typeEmergency}= 1\right) &= 90.4667 + 0.9018 (0) + 12.2219 (1) - 2.5205 (0)\times (1) \\
&= 90.4667 + 12.2219 \\
&= 102.6886 \text{.}
\end{align*}
$$

The estimated mean heart rate among males who came to the ICU for an emergency
is $90.4667 + 0.9018 + 12.2219 - 2.5205 = 101.0699$:
$$
\begin{align*}
\hat{E}\left(\texttt{HeartRate} \middle|\texttt{sexMale}= 1,\texttt{typeEmergency}= 1\right) &= 90.4667 + 0.9018 (1) + 12.2219 (1) - 2.5205 (1)\times (1) \\
&= 90.4667 + 0.9018 + 12.2219 - 2.5205 \\
&= 101.0699 \text{.}
\end{align*}
$$

As a result of the previous two interpretations,
the estimated mean difference in heart rate between males and females who
came to the ICU for an emergency is
$102.6886 - 101.0699 = 2.5205 - 0.9018 = 1.6187$ beats per minute.


### Moderation

An interaction is the statistical way to model moderation.
With moderation, the association of one explanatory variable with the outcome
is moderated by another explanatory variable.
This second explanatory variable,
the one that does the moderating,
is called a moderator.
(Some people say "effect modification" instead of "moderation",
but I'm avoiding the word "effect" because I want to use these terms
even for non-causal data.)

In the example above, the association between coming to the ICU for an emergency
and heart rate is modified by gender.
For men, coming to the ICU for an emergency is associated with a
$101.0699 - 91.3685 = 9.7014$ bpm increase in mean heart rate.
For women, though, coming to the ICU for an emergency is associated with a
$102.6886 - 90.4667 = 12.2219$ bmp increase in mean heart rate.

It would be equivalent to say that the association between gender and heart rate
is modified by visit type.
For elective visits, the estimated mean difference between male and female heart rates
is $91.3685 - 90.4667 = 0.9018$ bpm.
For emergency visits, though, the estimated mean difference between male and female heart rates
is $101.0699 - 102.6886 = -1.6187$ bpm.
(I used a negative here to indicate that the direction is opposite that
for elective visits.
For elective visits, males have the higher estimated mean heart rate,
but for emergency visits, females have the higher estimated mean heat rate.)


### Hypothesis Testing

Now, I've said that gender is a moderator and I've said that visit type is
a moderator,
but I really shouldn't be making that claim without a hypothesis test.
Although the association between gender and heart rate is different
for the two different visit types,
we should check if that difference is statistically significant.
The difference is quantified by the coefficient on the interaction term itself,
so we should check the p-value for that coefficient.
In the output above, we want the p-value associated with
`sexMale:typeEmergency`.
This p-value is 0.786,
which is way higher than any reasonable threshold we would use.
So, we fail to reject the hypothesis that the interaction coefficient is zero.
We can conclude that there is no evidence for moderation of the association
between gender and heart rate by visit type
(or between visit type and heart rate by gender).
For parsimony, we should re-fit the model without the interaction term,
which is what I've done below.

```{r}
mod_02 <- lm(formula = HeartRate ~ sex + type,
             data = d)
summary(mod_02)
```

## Problems

### Problem 1

Suppose a researcher finds that suicidal ideation is a risk factor
for attempted suicide.
For someone who does not own a gun, exhibiting suicidal ideation
is not associated with a large increase in risk of suicide,
but for someone who does own a gun, exhibiting suicidal ideation
is associated with a large increase in risk of suicide.
What is the moderator in this example?

::: {.callout-tip title="Solution" collapse="true"}

Gun ownership.
For those with a gun, there is a strong association between suicidal ideation
and suicide risk,
but for those without a gun,
the association is weaker.

:::

### Problem 2

Jeremy Groves, an anesthetist in the UK,
timed his commute when using two different bikes.
Below are the results of a regression of commute time in minutes
on bike type (steel or carbon) and month of the year (January through July),
with an interaction between the two explanatory variables.
The distance traveled was approximately 27 miles for each commute.
```{r}
d2 <- Lock5Data::BikeCommute
mod_03 <- lm(formula = Minutes ~ Bike * Month,
             data = d2)
summary(mod_03)
```

Interpret each coefficient estimate.

::: {.callout-tip title="Solution" collapse="true"}

**Intercept**

- The estimated mean commute time in January on a carbon bike is 114.755 minutes.

**Comparing Carbon to Steel by Month**

- The estimated mean difference in commute times between carbon and steel
bikes during the month of January is 0.655 minute,
with the steel taking longer.
- The estimated mean difference in commute times between carbon and steel
bikes during the month of February is
$|0.655-5.972| = |-5.317| = 5.317$ minutes,
with the carbon taking longer.
- The estimated mean difference in commute times between carbon and steel
bikes during the month of March is
$|0.655-3.465| = |-2.81| = 2.81$ minutes,
with the carbon taking longer.
- The estimated mean difference in commute times between carbon and steel
bikes during the month of April is
$|0.655+ 5.622| = |6.277| = 6.277$ minutes,
with the steel taking longer.
- The estimated mean difference in commute times between carbon and steel
bikes during the month of May is
$|0.655-2.473| = |-1.818| = 1.818$ minutes,
with the carbon taking longer.
- The estimated mean difference in commute times between carbon and steel
bikes during the month of June is
$|0.655+3.567| = |4.222| = 4.222$ minutes,
with the steel taking longer.
- The estimated mean difference in commute times between carbon and steel
bikes during the month of July is
$|0.655+0.947| = |1.602| = 1.602$ minutes,
with the steel taking longer.

**Comparing Months on a Carbon Bike**

- The estimated mean difference in commute times between January and February
on a carbon bike is 3.588 minutes, with February taking longer.
- The estimated mean difference in commute times between January and March
on a carbon bike is 3.662 minutes, with January taking longer.
- The estimated mean difference in commute times between January and April
on a carbon bike is 11.698 minutes, with January taking longer.
- The estimated mean difference in commute times between January and May
on a carbon bike is 10.215 minutes, with January taking longer.
- The estimated mean difference in commute times between January and June
on a carbon bike is 14.375 minutes, with January taking longer.
- The estimated mean difference in commute times between January and July
on a carbon bike is 9.655 minutes, with January taking longer.

**Comparing Months on a Steel Bike**

- The estimated mean difference in commute times between January and February
on a steel bike is $|3.588-5.972| = |-2.384| = 2.384$ minutes,
with January taking longer.
- The estimated mean difference in commute times between January and March
on a steel bike is $|-3.662-3.465| = |-7.127| = 7.127$ minutes,
with January taking longer.
- The estimated mean difference in commute times between January and April
on a steel bike is $|-11.698 + 5.622| = |-6.076| = 6.076$ minutes,
with January taking longer.
- The estimated mean difference in commute times between January and May
on a steel bike is $|-10.215-2.473| = |-12.688| = 112.688$ minutes,
with January taking longer.
- The estimated mean difference in commute times between January and June
on a steel bike is $|-14.375+3.567| = |-10.808| = 10.808$ minutes,
with January taking longer.
- The estimated mean difference in commute times between January and July
on a steel bike is $|-9.655+0.947| = |-8.708| = 8.708$ minutes,
with January taking longer.

:::




