---
title: "Interaction Between One Binary Variable and One Continuous Variable"
execute:
  echo: false
format: html
---

```{r}
#| warning: false
library(ggplot2)
```

## Learning Objective

Interpret the coefficients of a multiple linear regression model with an
interaction between one binary variable and one continuous variable.


## Lesson

Researchers in Brazil randomly assigned 264 high school students to
activities that involved either synchronized or non-synchronized movements.
The researchers measured pain tolerance with a blood pressure cuff
before and after the activity.
In the model below, I've regressed pain tolerance after the activity
on pain tolerance before, synchronicity (1 for synchronized,
0 for not synchronized),
and the interaction between the two.
```{r}
d <- Lock5Data::SynchronizedMovement
d$Synchronized <- as.numeric(d$Synch == "yes")
mod_01 <- lm(formula = PainTolerance ~ PainToleranceBefore * Synchronized,
             data = d)
summary(mod_01)
```

How do we interpret these coefficients?
We can write the linear model as follows
(please forgive the abbreviations):
$$
\hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef},\texttt{Syn}\right) = 50.06363 + 0.75550 \texttt{TolBef} + 38.94123 \text{Syn} - 0.08173 \texttt{TolBef}\times\texttt{Syn} \text{.}
$$

50.06363 mmHg is the estimated mean pain tolerance among people who were assigned to the
non-synchronized activity and who had zero pain tolerance before
the activity.
$$
\begin{align*}
\hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef}=0,\texttt{Syn}=0\right) &= 50.06363 + 0.75550 (0) + 38.94123 (0) - 0.08173 (0)\times(0) \\
&= 50.06363 \text{.}
\end{align*}
$$

0.75550 mmHg is the estimated mean increase in pain tolerance
associated with a one mmHg increase in prior pain tolerance,
among people assigned to the non-synchronized activity.
In other words, it is the estimated mean difference in pain tolerance
between people with prior pain tolerance $x$ and people with prior
pain tolerance $x + 1$,
assuming all these people were assigned to the non-synchronized activity.
People with higher prior pain tolerance are expected to have the higher
pain tolerance after doing the activity.
$$
\begin{align*}
&\hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef}=x+1,\texttt{Syn}=0\right) - \hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef}=x,\texttt{Syn}=0\right) \\
&= 50.06363 + 0.75550 (x+1) + 38.94123 (0) - 0.08173 (x+1)\times(0) \\
&\qquad - (50.06363 + 0.75550 x + 38.94123 (0) - 0.08173 x\times(0)) \\
&= 50.06363 + 0.75550x + 0.75550 - (50.06363 + 0.75550x) \\
&= 0.75550 \text{.}
\end{align*}
$$

38.94123 mmHg is the estimated mean difference in pain tolerance
between people who were assigned to the synchronized activity and people
who were assigned to the non-synchronized activity,
assuming all these people had prior pain tolerance of zero.
We would expect the people assigned to the synchronized activity to
have higher pain tolerance.
$$
\begin{align*}
&\hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef}=0,\texttt{Syn}=1\right) - \hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef}=0,\texttt{Syn}=0\right) \\
&= 50.06363 + 0.75550 (0) + 38.94123 (1) - 0.08173 (0)\times(1) \\
&\qquad - 50.06363  \\
&= 38.94123 \text{.}
\end{align*}
$$

$0.75550 - 0.08173 = 0.67377$ mmHg is the estimated mean increase in pain tolerance
associated with a one mmHg increase in prior pain tolerance,
among people assigned to the synchronized activity.
In other words, it is the estimated mean difference in pain tolerance
between people with prior pain tolerance $x$ and people with prior
pain tolerance $x + 1$,
assuming all these people were assigned to the synchronized activity.
People with higher prior pain tolerance are expected to have the higher
pain tolerance after doing the activity.
$$
\begin{align*}
&\hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef}=x+1,\texttt{Syn}=1\right) - \hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef}=x,\texttt{Syn}=1\right) \\
&= 50.06363 + 0.75550 (x+1) + 38.94123 (1) - 0.08173 (x+1)\times(1) \\
&\qquad - (50.06363 + 0.75550 x + 38.94123 (1) - 0.08173 x\times(1)) \\
&= 50.06363 + 0.75550 x + 0.75550 + 38.94123  - 0.08173x - 0.08173 \\
&\qquad - (50.06363 + 0.75550 x + 38.94123  - 0.08173 x) \\
&= 0.75550 - 0.08173 \\
&= 0.67377 \text{.}
\end{align*}
$$

$38.94123 - 0.08173x$ mmHg is the estimated mean difference in pain tolerance
between people who were assigned to the synchronized activity and people
who were assigned to the non-synchronized activity,
assuming all these people had prior pain tolerance of $x$.
Since $x$ was only observed between 30 and 300,
$14.42223 \le 38.94123 - 0.08173x \le 36.48933$, and
we would expect the people assigned to the synchronized activity to
have higher pain tolerance.
$$
\begin{align*}
&\hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef}=x,\texttt{Syn}=1\right) - \hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef}=x,\texttt{Syn}=0\right) \\
&= 50.06363 + 0.75550 x + 38.94123 (1) - 0.08173 x\times(1) \\
&\qquad - (50.06363 + 0.75550 x + 38.94123 (0) - 0.08173 x\times(0))  \\
&= 38.94123 - 0.08173x \text{.}
\end{align*}
$$

We can observe this in a plot.
```{r}
#| warning: false
ggplot(data = d,
       mapping = aes(x = PainToleranceBefore,
                     y = PainTolerance,
                     colour = Synch)) +
  geom_point() +
  geom_smooth(method = "lm",
              se = FALSE)
```
If the moderation were stronger, these lines would have different slopes.








