---
title: "Interaction Between One Binary Variable and One Continuous Variable"
execute:
  echo: false
format: html
---

```{r}
#| warning: false
library(ggplot2)
```

## Learning Objective

Interpret the coefficients of a multiple linear regression model with an
interaction between one binary variable and one continuous variable.


## Lesson

Researchers in Brazil randomly assigned 264 high school students to
activities that involved either synchronized or non-synchronized movements.
The researchers measured pain tolerance with a blood pressure cuff
before and after the activity.
In the model below, I've regressed pain tolerance after the activity
on pain tolerance before, synchronicity (1 for synchronized,
0 for not synchronized),
and the interaction between the two.
```{r}
d <- Lock5Data::SynchronizedMovement
d$Synchronized <- as.numeric(d$Synch == "yes")
mod_01 <- lm(formula = PainTolerance ~ PainToleranceBefore * Synchronized,
             data = d)
summary(mod_01)
```

How do we interpret these coefficients?
We can write the linear model as follows
(please forgive the abbreviations):
$$
\hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef},\texttt{Syn}\right) = 50.06363 + 0.75550 \texttt{TolBef} + 38.94123 \text{Syn} - 0.08173 \texttt{TolBef}\times\texttt{Syn} \text{.}
$$

50.06363 mmHg is the estimated mean pain tolerance among people who were assigned to the
non-synchronized activity and who had zero pain tolerance before
the activity.
$$
\begin{align*}
\hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef}=0,\texttt{Syn}=0\right) &= 50.06363 + 0.75550 (0) + 38.94123 (0) - 0.08173 (0)\times(0) \\
&= 50.06363 \text{.}
\end{align*}
$$

0.75550 mmHg is the estimated mean increase in pain tolerance
associated with a one mmHg increase in prior pain tolerance,
among people assigned to the non-synchronized activity.
In other words, it is the estimated mean difference in pain tolerance
between people with prior pain tolerance $x$ and people with prior
pain tolerance $x + 1$,
assuming all these people were assigned to the non-synchronized activity.
People with higher prior pain tolerance are expected to have the higher
pain tolerance after doing the activity.
$$
\begin{align*}
&\hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef}=x+1,\texttt{Syn}=0\right) - \hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef}=x,\texttt{Syn}=0\right) \\
&= 50.06363 + 0.75550 (x+1) + 38.94123 (0) - 0.08173 (x+1)\times(0) \\
&\qquad - (50.06363 + 0.75550 x + 38.94123 (0) - 0.08173 x\times(0)) \\
&= 50.06363 + 0.75550x + 0.75550 - (50.06363 + 0.75550x) \\
&= 0.75550 \text{.}
\end{align*}
$$

38.94123 mmHg is the estimated mean difference in pain tolerance
between people who were assigned to the synchronized activity and people
who were assigned to the non-synchronized activity,
assuming all these people had prior pain tolerance of zero.
We would expect the people assigned to the synchronized activity to
have higher pain tolerance.
$$
\begin{align*}
&\hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef}=0,\texttt{Syn}=1\right) - \hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef}=0,\texttt{Syn}=0\right) \\
&= 50.06363 + 0.75550 (0) + 38.94123 (1) - 0.08173 (0)\times(1) \\
&\qquad - 50.06363  \\
&= 38.94123 \text{.}
\end{align*}
$$

$0.75550 - 0.08173 = 0.67377$ mmHg is the estimated mean increase in pain tolerance
associated with a one mmHg increase in prior pain tolerance,
among people assigned to the synchronized activity.
In other words, it is the estimated mean difference in pain tolerance
between people with prior pain tolerance $x$ and people with prior
pain tolerance $x + 1$,
assuming all these people were assigned to the synchronized activity.
People with higher prior pain tolerance are expected to have the higher
pain tolerance after doing the activity.
$$
\begin{align*}
&\hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef}=x+1,\texttt{Syn}=1\right) - \hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef}=x,\texttt{Syn}=1\right) \\
&= 50.06363 + 0.75550 (x+1) + 38.94123 (1) - 0.08173 (x+1)\times(1) \\
&\qquad - (50.06363 + 0.75550 x + 38.94123 (1) - 0.08173 x\times(1)) \\
&= 50.06363 + 0.75550 x + 0.75550 + 38.94123  - 0.08173x - 0.08173 \\
&\qquad - (50.06363 + 0.75550 x + 38.94123  - 0.08173 x) \\
&= 0.75550 - 0.08173 \\
&= 0.67377 \text{.}
\end{align*}
$$

$38.94123 - 0.08173x$ mmHg is the estimated mean difference in pain tolerance
between people who were assigned to the synchronized activity and people
who were assigned to the non-synchronized activity,
assuming all these people had prior pain tolerance of $x$.
Since $x$ was only observed between 30 and 300,
$14.42223 \le 38.94123 - 0.08173x \le 36.48933$, and
we would expect the people assigned to the synchronized activity to
have higher pain tolerance.
$$
\begin{align*}
&\hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef}=x,\texttt{Syn}=1\right) - \hat{E}\left(\texttt{Tol}\middle|\texttt{TolBef}=x,\texttt{Syn}=0\right) \\
&= 50.06363 + 0.75550 x + 38.94123 (1) - 0.08173 x\times(1) \\
&\qquad - (50.06363 + 0.75550 x + 38.94123 (0) - 0.08173 x\times(0))  \\
&= 38.94123 - 0.08173x \text{.}
\end{align*}
$$

We can observe this in a plot.
```{r}
#| warning: false
ggplot(data = d,
       mapping = aes(x = PainToleranceBefore,
                     y = PainTolerance,
                     colour = Synch)) +
  geom_point() +
  geom_smooth(method = "lm",
              se = FALSE)
```
If the moderation were stronger, these lines would have different slopes.


## Problems

### Problem 1

Check out Figure 2 in
"[Factors in Hand Surgery Access for Rheumatoid Arthritis Before vs After the Patient Protection and Affordable Care Act](https://jamanetwork.com/journals/jamasurgery/fullarticle/2814271)"
(Dwivedi et al., 2024).
This is an example of piecewise linear regression.
For each procedure,
the researchers
fit a simple linear regression before the ACA
and another one after.
This can be achieved through interaction terms.

In my opinion, linear regression was not the appropriate
analysis for these data.
As the authors write in the section "Outcomes of Interest",
"Surgical incidence was defined as the number of hand surgeries performed
each year divided by the population of each cohort."
I think Poisson regression would be more appropriate
for this outcome.
We will learn about Poisson regression later in this course.

For this problem, let's use an outcome that is better-suited
for linear regression.
Suppose our sample consists of $n$ patients who underwent
hand surgery for rheumatoid arthritis (RA).
Define the following:

- $P_i$ is the pain level of patient $i$,
measured on a scale from 1 to 100.
- $Y_i$ is the number of years since 2014.
Assume each patient was only measured once,
so we only have one time value per person.
- $S_i = 1$ if patient $i$ had a synovectomy
and $S_i = 0$ otherwise.
- $A_i = 1$ if $Y_i > 0$ and $A_i = 0$
if $Y_i \le 0$.
You can think of $A_i$ being an indicator
that the observation was *after* 2014.

Suppose our model is
$$
P_i = \beta_0 + \beta_1 Y_i + \beta_2 S_i + \beta_3 A_i + \beta_4 Y_i S_i + \beta_5 Y_i A_i + \beta_6 S_i A_i + \beta_7 Y_i S_i A_i + \epsilon_i \text{.}
$$

#### Part (a)

What is the expected pain level for a patient
who had arthrodesis
$y$ years since 2014, where $y \le 0$?

::: {.callout-tip title="Solution" collapse="true"}

$$
\begin{align*}
    &E(P_i|Y_i = y,S_i = 0,A_i = 0) \\
    &= \beta_0 + \beta_1 y + \beta_2 (0) + \beta_3 (0) + \beta_4 y (0) + \beta_5 y (0) + \beta_6 (0) (0) + \beta_7 y (0) (0) \\
    &= \beta_0 + \beta_1 y 
\end{align*}
$$

:::

#### Part (b)

What is the expected pain level for a patient
who had denervation
$y$ years since 2014, where $y > 0$?

::: {.callout-tip title="Solution" collapse="true"}

$$
\begin{align*}
    &E(P_i|Y_i = y,S_i = 0,A_i = 1) \\
    &= \beta_0 + \beta_1 y + \beta_2 (0) + \beta_3 (1) + \beta_4 y (0) + \beta_5 y (1) + \beta_6 (0) (1) + \beta_7 y (0) (1) \\
    &= \beta_0 + \beta_1 y  + \beta_3   + \beta_5 y   \\
    &= (\beta_0 + \beta_3) + (\beta_1 + \beta_5) y
\end{align*}
$$

:::
        
#### Part (c)

What is the expected pain level for a patient
who had a synovectomy
$y$ years since 2014, where $y \le 0$?

::: {.callout-tip title="Solution" collapse="true"}

$$
\begin{align*}
    &E(P_i|Y_i = y,S_i = 1,A_i = 0) \\
    &= \beta_0 + \beta_1 y + \beta_2 (1) + \beta_3 (0) + \beta_4 y (1) + \beta_5 y (0) + \beta_6 (1) (0) + \beta_7 y (1) (0) \\
    &= \beta_0 + \beta_1 y + \beta_2   + \beta_4 y  \\
    &= (\beta_0 + \beta_2) + (\beta_1 + \beta_4) y
\end{align*}
$$

:::

#### Part (d)

What is the expected pain level for a patient
who had a synovectomy
$y$ years since 2014, where $y > 0$?

::: {.callout-tip title="Solution" collapse="true"}

$$
\begin{align*}
    &E(P_i|Y_i = y,S_i = 1,A_i = 1) \\
    &= \beta_0 + \beta_1 y + \beta_2 (1) + \beta_3 (1) + \beta_4 y (1) + \beta_5 y (1) + \beta_6 (1) (1) + \beta_7 y (1) (1) \\
    &= \beta_0 + \beta_1 y + \beta_2  + \beta_3  + \beta_4 y  + \beta_5 y  + \beta_6  + \beta_7 y  \\
    &= (\beta_0 + \beta_2 + \beta_3 + \beta_6) + (\beta_1 + \beta_4 + \beta_5 + \beta_7) y 
\end{align*}
$$

:::

#### Part (e)

What is the expected difference in pain levels
between a patient who had a synovectomy in 2017
and a patient who had sagittal band reconstruction,
also in 2017?

::: {.callout-tip title="Solution" collapse="true"}

$$
\begin{align*}
    &E(P_i|Y_i = 3,S_i = 1,A_i = 1) - E(P_i|Y_i = 3,S_i = 0,A_i = 1) \\
    &= (\beta_0 + \beta_2 + \beta_3 + \beta_6) + (\beta_1 + \beta_4 + \beta_5 + \beta_7) (3) - ((\beta_0 + \beta_3) + (\beta_1 + \beta_5) (3)) \\
    &= (\beta_2  + \beta_6) + ( \beta_4  + \beta_7) (3)  
\end{align*}
$$
        
:::

### Problem 2

This problem is about the study
"[Association between Traffic Air Pollution and Reduced Forced Vital Capacity: A Study Using Personal Monitors for Outdoor Workers](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0163225)"
(Santos et al., 2016).
Define the following:

- $F_i$ is the forced vital capacity (FVC) for the $i$th
individual, measured in liters.
- $D_i = 1$ if the $i$th individual was a taxi driver
and $D_i = 0$ otherwise.
- $C_i = 1$ if the $i$th individual was a traffic
controller and $C_i = 0$ otherwise.
- $Y_i$ is the number of years
that the $i$th individual has worked in their job.

Suppose our model is
$$
F_i = \beta_0 + \beta_1 D_i + \beta_2 C_i + \beta_3 Y_i + \beta_4 D_i Y_i + \beta_5 C_i Y_i + \epsilon_i \text{.}
$$

#### Part (a)

Why don't we need terms for $D_i C_i$ and $D_i C_i Y_i$?

::: {.callout-tip title="Solution" collapse="true"}

$D_i$ and $C_i$ cannot both be 1,
so any term with their product must be 0.
Each individual has only one occupation
(in this data set).
:::

#### Part (b)

Interpret $\beta_2    + \beta_5  y$ in real-world terms.

::: {.callout-tip title="Solution" collapse="true"}

This is the expected difference in FVC
between a traffic controller and a forest ranger,
both of whom have worked in their job for $y$ years.

$$
\begin{align*}
    &E(F_i|D_i = 0,C_i = 1,Y_i = y) - E(F_i|D_i = 0,C_i = 0,Y_i = y) \\
    &= \beta_0 + \beta_1 (0) + \beta_2 (1) + \beta_3 y + \beta_4 (0) y + \beta_5 (1) y \\
    &\qquad - (\beta_0 + \beta_1 (0) + \beta_2 (0) + \beta_3 y + \beta_4 (0) y + \beta_5 (0) y) \\
    &= \beta_0  + \beta_2  + \beta_3 y  + \beta_5  y  - (\beta_0  + \beta_3 y ) \\
    &=  \beta_2    + \beta_5  y   \\
\end{align*}
$$

:::
        




