---
title: "Overfitting"
execute:
  echo: false
format: html
---

## Learning Objective

Define, use, and interpret the term "overfitting".


## Lesson

### Definition

As we've discussed, we usually want to make claims about a population
but can only observe a sample.
It's important that the conclusions we draw from our sample are generalizable
to the population.
Even if our goal isn't to generalize to some larger population,
but is instead to make predictions,
we want to make sure that we can accurately predict new data,
not just the data in our sample.
"Overfitting" is what happens when you create a model that describes
your sample really well but does a poor job of generalizing to or predicting
new data from the same population.

### Conceptual Example

Suppose your roommate is from Portugal.
You know he speaks Portuguese,
he needs the shower at 7am, he likes to cook dinner,
and on Saturdays he plays Frisbee with friends.
So, you conclude that all Portuguese people speak Portuguese, shower at 7am,
like to cook dinner,
and play Frisbee with their friends on Saturdays.
That's an example of overfitting.
You've done a great job of describing (and predicting) your roommate's behavior,
but there's no reason to believe that the same descriptions and predictions
will work for other people from the same country.
(At least, not all of them;
it is reasonable to predict that other people from Portugal speak Portuguese.)

### Visual Depiction

Overfitting is usually depicted in the following way.
Suppose you have these data:
```{r}
set.seed(1)
n <- 20
x <- rnorm(n)
y <- x + rnorm(n)
plot(x = x,
     y = y)
```

You try fitting a simple linear regression:
```{r}
mod_01 <- lm(y ~ x)
plot(x = x,
     y = y)
abline(mod_01,
       col = "blue")
```

You decide that's not quite good enough.
Although the regression line captures the general trend of the data,
it doesn't go through enough of the points.
So you add a quadratic term.
Now your linear model is
$$
E(Y|X) = \beta_0 + \beta_1 X + \beta_2 X^2 \text{.}
$$
```{r}
mod_02 <- lm(y ~ x + I(x^2))
plot(x = x,
     y = y)
x_seq <- seq(from = min(x),
             to = max(x),
             length.out = 100)
lines(x = x_seq,
      y = predict(object = mod_02,
                  newdata = data.frame(x = x_seq)),
      col = "blue")
```

That's still not quite good enough, so you add a few more polynomial terms.
Now your linear model is
$$
E(Y|X) = \beta_0 + \beta_1 X + \beta_2 X^2 + \beta_3 X^3 + \beta_4 X^4 + \beta_5 X^5 + \beta_6 X^6 + \beta_7 X^7 + \beta_8 X^8 + \beta_9 X^9 + \beta_{10} X^{10} + \beta_{11} X^{11}\text{.}
$$
```{r}
mod_03 <- lm(y ~ x + I(x^2) + I(x^3) + I(x^4) + I(x^5) + I(x^6) + I(x^7) + I(x^8) + I(x^9) + I(x^10) + I(x^11))
plot(x = x,
     y = y)
lines(x = x_seq,
      y = predict(object = mod_03,
                  newdata = data.frame(x = x_seq)),
      col = "blue")
```

This regression line goes through a lot of the points!
But watch what happens when we collect more data:
```{r}
n_new <- n * 100
x_new <- rnorm(n_new)
y_new <- x_new + rnorm(n_new)
```

```{r}
x_seq_new <- seq(from = min(x_new),
                 to = max(x_new),
                 length.out = 100)
plot(x = x_new,
     y = y_new,
     xlab = "x",
     ylab = "y")
lines(x = x_seq_new,
      y = predict(object = mod_03,
                  newdata = data.frame(x = x_seq_new)),
      col = "blue",
      lwd = 2)
```

The eleventh-degree polynomial function, which fit the original data set
of twenty observations really well,
does a terrible job of describing the new data set with 2,000 observations.

### Note

There are lots of things that limit our ability to generalize
from a sample to the population,
including a small sample size or a biased sampling scheme.
"Overfitting" is generally not used to refer to the size of the sample
or how it was collected.
Instead, it refers to the relationship between the model and the sample.
When talking about overfitting,
we're kind of taking the sample for granted.
We can't change the sample now that we have it,
but we can change the model,
and we want to make sure the model isn't overfit to the data.

## Problems

### Problem 1

Propose a method for determining whether a model is overfit to a sample.

::: {.callout-tip title="Solution" collapse="true"}

We could hold out some data from the sample (call this the "test set")
and fit our model to the rest (call this the "training set").
If the model is overfit to the training set,
it will do a poor job of predicting the outcome
in the test set.

:::

### Problem 2

True or false:
A Type II error is the same thing as overfitting.

::: {.callout-tip title="Solution" collapse="true"}

False.
You can make a Type II error even when the model is perfect.
By chance alone, there is always a non-zero chance of making a Type II error.

:::




