---
title: "R-Squared"
execute:
  echo: false
format: html
---

## Learning Objective

- Define, use, and interpret $R^2$.
- Explain the purpose of the adjusted $R^2$.
- Use and interpret the adjusted $R^2$.


## Lesson

### Definition 1

The coefficient of determination, denoted by $R^2$, is defined as follows:
$$
R^2 = 1 - \frac{\sum_{i=1}^n \left(Y_i - \hat{Y}_i\right)^2}{\sum_{i=1}^n \left(Y_i - \bar{Y}\right)^2} \text{.}
$$
As a reminder, $n$ is the sample size, $\hat{Y}_i$ is the fitted value
for the $i$th observation, and $\bar{Y}$ is the sample average.

How do we interpret this?
Imagine that you're pitting two different prediction methods against each other.
In the first, you're just using the sample average to predict every
observation in the data set.
In other words, you're ignoring all the explanatory variables.
The denominator of the fraction,
$$
\sum_{i=1}^n \left(Y_i - \bar{Y}\right)^2 \text{,}
$$
is a measure of how wrong you are using this method.

In the second prediction method, you're using the regression model.
So, for each observation, you're using the explanatory variables
to get a prediction of the outcome.
This prediction is the fitted value.
The numerator of the fraction,
$$
\sum_{i=1}^n \left(Y_i - \hat{Y}_i\right)^2 \text{,}
$$
is a measure of how wrong you are using this method.

If the regression model performs better, i.e., if it's less wrong, then
$$
\sum_{i=1}^n \left(Y_i - \hat{Y}_i\right)^2 < \sum_{i=1}^n \left(Y_i - \bar{Y}\right)^2 \text{.}
$$
This implies
$$
\frac{\sum_{i=1}^n \left(Y_i - \hat{Y}_i\right)^2}{\sum_{i=1}^n \left(Y_i - \bar{Y}\right)^2} < 1
$$
and
$$
R^2 = 1 - \frac{\sum_{i=1}^n \left(Y_i - \hat{Y}_i\right)^2}{\sum_{i=1}^n \left(Y_i - \bar{Y}\right)^2} > 0 \text{.}
$$

If the regression model performs a lot better then
$$
\sum_{i=1}^n \left(Y_i - \hat{Y}_i\right)^2 \ll \sum_{i=1}^n \left(Y_i - \bar{Y}\right)^2 \text{,}
$$
$$
\frac{\sum_{i=1}^n \left(Y_i - \hat{Y}_i\right)^2}{\sum_{i=1}^n \left(Y_i - \bar{Y}\right)^2} \ll 1 \text{,}
$$
and
$$
R^2 = 1 - \frac{\sum_{i=1}^n \left(Y_i - \hat{Y}_i\right)^2}{\sum_{i=1}^n \left(Y_i - \bar{Y}\right)^2} \gg 0 \text{.}
$$
(The symbol $\ll$ means "much less than" and the symbol $\gg$ means
"much greater than".)

All this together implies that the larger $R^2$ is,
the better job it does of predicting the outcome within the sample.

### Definition 2

We can also define $R^2$ like this:
$$
R^2 = \frac{\sum_{i=1}^n \left(\hat{Y}_i - \bar{Y}\right)^2}{\sum_{i=1}^n \left(Y_i - \bar{Y}\right)^2} \text{.}
$$
This definition is mathematically equivalent to the one above,
but it allows for a different interpretation of $R^2$.

In terms of the performance of the regression model,
the worst case scenario is that $\hat{Y}_i = \bar{Y}$,
i.e., incorporating information about the explanatory variables
does no good for predicting the outcome in the data set.
In this case, the numerator of $R^2$ is zero,
so $R^2 = 0$.

The best case scenario is that $\hat{Y}_i = Y_i$,
i.e., the regression model does a perfect job predicting the outcome
in the data set.
In this case, the numerator of $R^2$ is equal to the denominator,
and $R^2 = 1$.

This shows us that $R^2$ ranges from 0 to 1.
If we think of the denominator as a measure of the total variability in the
data set,
we can interpret $R^2$ as the proportion of variance explained by the model.

### Overfitting

Note that $R^2$ is only measuring the performance of the regression model
in the sample,
i.e., the same sample that was used to fit the regression model.
If we try to optimize (i.e., maximize) $R^2$, we may overfit the data.

Mathematically, adding a new explanatory variable to a linear regression model
cannot decrease the $R^2$.
So, if we're trying to increase the $R^2$, we could just keep adding explanatory
variables until we see improvement.
This will probably cause overfitting because some of these explanatory
variables will be related to the outcome in the sample just by chance.
In the population, there may be no relationship,
so the model will perform poorly with new data.

### Adjusted R-Squared

The purpose of the adjusted $R^2$ is to prevent overfitting.
It is defined as follows:
$$
R^2_{\text{adj}} = 1 - (1-R^2) \frac{n - 1}{n - p - 1} \text{,}
$$
where $p$ is the number of explanatory variables.

If we hold $R^2$ and $n$ constant while increasing $p$,
$$
\begin{align*}
p\uparrow &\implies -p \downarrow \\
&\implies n - p - 1 \downarrow \\
&\implies \frac{n - 1}{n - p - 1} \uparrow \\
&\implies (1-R^2) \frac{n - 1}{n - p - 1} \uparrow \\
&\implies -(1-R^2) \frac{n - 1}{n - p - 1} \downarrow \\
&\implies 1 -(1-R^2) \frac{n - 1}{n - p - 1} \downarrow \\
&\implies R^2_{\text{adj}} \downarrow \text{,}
\end{align*}
$$
i.e., the adjusted $R^2$ decreases.
So, the adjusted $R^2$ penalizes adding new explanatory variables to the model
to offset the better fit you get.

### Note

Although I keep saying that the fitted values come from a regression model,
the $R^2$ is agnostic to where the fitted values are coming from.
You could compute the $R^2$ using fitted values from a different prediction
method.
For these other methods, the two definitions above might not be exactly equal,
but it would still be the case that a higher $R^2$ would be better
than a lower $R^2$.

## Problems

### Problem 1

I'm going to randomly generate a data set with twenty independent explanatory
variables.
```{r}
#| echo: true
set.seed(2)
n <- 1e2
p <- 20
d <- matrix(data = rnorm(n = n * p),
            nrow = n,
            ncol = p)
```
Now I'm going to generate an outcome that depends on only the first two
explanatory variables in the data set.
```{r}
#| echo: true
y <- d[ ,1] + d[ ,2] + rnorm(n)
```
Now I'm going to run two regressions,
first with just the first two explanatory variables and second
with all twenty.
```{r}
mod_01 <- lm(y ~ d[ ,1] + d[ ,2])
mod_02 <- lm(y ~ d)
```
Predict the relationships between the $R^2$ and adjusted $R^2$
for both models.

::: {.callout-tip title="Solution" collapse="true"}

```{r}
summary(mod_01)
summary(mod_02)
```

Were your predictions right?

The adjusted $R^2$ for the small model is about the same
as the unadjusted $R^2$ for the small model.
This is desirable;
the small model is the correct model,
so there's no overfitting happening and we wouldn't want to remove
any explanatory variables.

There's a bigger difference between the adjusted and unadjusted $R^2$
for the big model.
Again, this is desirable.
Eighteen of the variables were generated completely independently of the
outcome and should be removed.
There is overfitting happening, so there should be a gap between
the adjusted and unadjusted $R^2$.

The adjusted $R^2$ for the large model is approximately equal
to the adjusted $R^2$ for the small model.
This means that, adjusted for their size, they're performing approximately
equally.
Since we often value parsimony in statistical models
(smaller models are easier to interpret and have higher power to detect effects),
we should probably go with the smaller model.
This is good, since the smaller model is the correct one.


:::

### Problem 2

True or false:
The linearity assumption of multiple linear regression must be true
in order to interpret the $R^2$.

::: {.callout-tip title="Solution" collapse="true"}

False.
The $R^2$ is (essentially) just a measure of how close the fitted values
are to the observed values of the outcome.
Even if none of the assumptions are met, you can still compute
and interpret the $R^2$.

:::




