---
title: "Test Set Error"
execute:
  echo: false
format: html
---

```{r}
#| warning: false
library(caret)
```

## Learning Objective

Define, use, and interpret the following terms:

- Training set
- Test set
- Cross-validation


## Lesson

### Training Set and Test Set

The gold standard method for checking if your model is overfit to your data
is to measure the performance of the model on new data.
If you see a degradation in performance between the original sample and the
new data,
that's evidence for overfitting.
But what if you can't obtain new data?
You can solve this problem by holding out a subset of the data set
for eventual use in testing performance.
This subset is called a "test set".
The remaining data, which you use to fit your model,
is called the "training set".

Let's examine this in a real data set.
The package `Lock5Data` has a data set with 362 observations from
"an in-class survey given to introductory statistics students over several years."
Let's try to predict GPA.
Unfortunately, 17 students are missing GPA,
so we'll have to remove them.

```{r}
d <- Lock5Data::StudentSurvey
d <- d[!is.na(d$GPA), ]
hist(x = d$GPA,
     main = "",
     xlab = "GPA")
```

The first thing I'm going to do is split the data into two subsets,
the training set and the test set.
I'm going to use the package `caret` to do this,
because it automatically performs a stratified random split of the data.
This ensures the training set and test set are not too dissimilar
in terms of the outcome.
For example, you wouldn't want (even by chance) to have the lowest GPAs
in the training set and the highest GPAs in the test set.

```{r}
#| echo: true
set.seed(3)
in_train <- createDataPartition(y = d$GPA,
                                p = 0.8,
                                list = FALSE)
d_train <- d[in_train, ]
d_test <- d[-in_train, ]
```

I put 80% of the observations in the training set.
There is not a hard-and-fast rule about what percentage of the observations
to put in the training set.
As the size of your original data set grows,
you can put a smaller percentage in the test set,
thus allowing more data for training.

Now that we've split our data, I'm going to train two models.
The first is going to contain every possible variable.
I'm not going to think, I'm just going to throw everything in.

```{r}
#| echo: true
mod_01 <- lm(formula = GPA ~ Year + Sex + Smoke + Award + Exercise + TV +
               Height + Weight + Siblings + BirthOrder + VerbalSAT + MathSAT +
               Pulse + Piercings,
             data = d_train)
```

For the second model, I'm only going to include a couple variables that I think
could reasonably be related to GPA.

```{r}
#| echo: true
mod_02 <- lm(formula = GPA ~ Year + VerbalSAT + MathSAT,
             data = d_train)
```

I'm not going to check any assumptions because I'm just pretending
that I only care about prediction.
I don't want to interpret any coefficients or p-values;
I just want to measure how well the two models predict new data.

As a measure of performance, I'm going to use root mean squared error (RMSE):
$$
\sqrt{\frac{1}{m}\sum_{j=1}^m \left(Y_j - \hat{Y}_j\right)^2} \text{.}
$$
Here $m$ is the number of observations in the test set
and $\hat{Y}_j$ is the predicted outcome for the $j$th observation.
I like the RMSE for continuous outcomes because it's in the same units
as the outcome.

```{r}
#| echo: true
y_hat_01 <- predict(object = mod_01,
                    newdata = d_test)
y_hat_02 <- predict(object = mod_02,
                    newdata = d_test)
RMSE_01 <- sqrt(mean(x = (d_test$GPA - y_hat_01)^2,
                     na.rm = TRUE))
RMSE_02 <- sqrt(mean((d_test$GPA - y_hat_02)^2))
RMSE_01
RMSE_02
```

The smaller model, Model 2, has slightly lower error than the larger model,
meaning it does a slightly better job predicting new data.
(Note that I had to remove missing values from the fitted values
from Model 1.
It must've been the case that some observations in the test set
didn't have values for all the covariates used in Model 1,
so it was impossible to obtain fitted values for them.)

How did both models perform in the training set?

```{r}
#| echo: true
sqrt(mean(x = (d_train$GPA - predict(object = mod_01,
                                     newdata = d_train))^2,
          na.rm = TRUE))
sqrt(mean((d_train$GPA - mod_02$fitted.values)^2))
```

The larger model, Model 1, has slightly lower error in the training set
than the smaller model.
This is evidence of overfitting;
the big model did better in the training set but worse in the test set.
It failed to generalize to new data.

### Validation Set

Sometimes, people will split the data into three subsets:
a training set, a validation set, and a test set.
Models are fit on the training set
and their performance is measured in the validation set.
Then, changes are made to the models
and they are re-fit in the training set and re-evaluated in the validation set.
This process proceeds iteratively until a final set of models is obtained.
The performance of these final models is then measured in the test set.

### Cross-Validation

Cross-validation is another method for testing the performance of
a model in "new" data.
In $k$-fold cross-validation,
the original data set is split into $k$ subsets.
Then, the model is fit $k$ times.
In the first iteration,
the model is fit on folds 2 through $k$ and tested on fold 1.
In the second iteration,
the model is fit on fold 1 and folds 3 through $k$ and tested on fold 2.
After $k$ iterations, the test error is summed (or averaged).
The table below depicts this process for $k = 5$.

| Iteration | Fold 1 | Fold 2 | Fold 3 | Fold 4 | Fold 5 |
|:-:|:-:|:-:|:-:|:-:|:-:|
| 1 | [Test]{style="color: red;"} | Train | Train | Train | Train |
| 2 | Train | [Test]{style="color: red;"} | Train | Train | Train |
| 3 | Train | Train | [Test]{style="color: red;"} | Train | Train |
| 4 | Train | Train | Train | [Test]{style="color: red;"} | Train |
| 5 | Train | Train | Train | Train | [Test]{style="color: red;"} |

Just a quick note:
If you're using RMSE as your measure of error
and you're doing cross-validation,
you shouldn't add the $k$ RMSEs directly.
Instead, you should add the sums of squared error
and only take a square root after.

Sometimes, people will split the original data set into 
a training set and a test set
and then do cross-validation on the training set.
This is common for regularized regression (LASSO and ridge),
where you have to choose a penalty parameter,
and ensemble methods, where you have to choose model weights.

## Problems

### Problem 1

When you split your data into a training set and a test set,
which plays the role of "new" data?

::: {.callout-tip title="Solution" collapse="true"}

The test set.
You don't use it to fit the model(s),
only to test the performance of the model(s).

:::

### Problem 2

In $k$-fold cross validation, how many times does each observation
get used for testing?

::: {.callout-tip title="Solution" collapse="true"}

Once.

:::

### Problem 2

In $k$-fold cross validation, how many times does each observation
get used for training?

::: {.callout-tip title="Solution" collapse="true"}

$k - 1$ times.

:::




