---
title: "Likelihood"
execute:
  echo: false
format: html
---

## Learning Objective

- Define "likelihood".
- Explain why you would want to maximize the likelihood.
- Explain when it would be inappropriate to use likelihoods to compare models.


## Lesson

### Normal Example

Suppose you draw a single observation from a normal distribution
with expected value $\mu$ and variance equal to 1.
The value you get is $-1$.
In symbols, $X\sim N(\mu,1)$ and you observe $X = -1$.

The probability density function of $X$ is
$$
f(x) = \frac{1}{\sqrt{2\pi}} \exp\left(- \frac{(x - \mu)^2}{2}\right) \text{.}
$$
We usually think of this as a function of $x$, but it's also a function of $\mu$.
We could write
$$
f(x;\mu) = \frac{1}{\sqrt{2\pi}} \exp\left(- \frac{(x - \mu)^2}{2}\right)
$$
to make it explicit that this function depends on $\mu$.

To make this more concrete, I'm going to plot $f(x;\mu)$ for a variety
of values of $\mu$.
In each one, I'm going to include dotted line segments to show
the value of $f(-1;\mu)$.

```{r}
x <- -1
width <- 2
for(i in (x - width):(x + width)){
  temp <- function(x){
    dnorm(x = x,
          mean = i)
  }
  curve(expr = temp,
        from = x - width - 2,
        to = x + width + 2,
        xlab = "x",
        ylab = "f(x;mu)",
        main = paste0("mu = ",
                      i))
  abline(v = 0)
  segments(x0 = c(-1,-1),
           y0 = c(0,temp(-1)),
           x1 = c(-1,0),
           y1 = c(temp(-1),temp(-1)),
           lty = 2)
}
```

When $\mu = -3$, $f(-1,-3)$ is pretty low;
$f(-1,-2)$ is higher, and $f(-1,-1)$ is highest.
As $\mu$ continues to increase,
the output of the function decreases again;
$f(-1,0)$ is lower and $f(-1,1)$ is lower still.

The probability density function (PDF) is a measure of how likely it is to see
certain values.
The higher the PDF, the more likely that value.
If we don't know the value of $\mu$,
but we know what we observed,
we might want to choose the value of $\mu$ that maximizes the likelihood
of observing what we observed.
This is a way of estimating $\mu$.

In this example, setting $\mu = -1$ maximizes the PDF at $x = -1$.
So, we could argue that $-1$ is a good estimate of $\mu$.

### Definition

"Likelihood" is commonly used in American English, but it has a special
meaning in statistics.
In statistics, the likelihood is the probability density function (PDF) or
probability mass function (PMF), thought of as a function of the parameter(s)
rather than the data.

By fixing the input to the PDF or PMF at the data we observed,
and changing the value(s) of the parameter(s) until we maximize the likelihood,
we find the value(s) of the parameter(s) that maximize(s) the likelihood
of observing what we actually observed.
This is a method for estimating the parameter(s).

### Binomial Example

Suppose a magician has a special coin that lands heads with probability $p$,
where $p \ne 0.5$.
She flips the coin ten times and it lands heads seven times.

In this case, the number of heads follows a binomial distribution;
we can write $X\sim\text{binomial}(10,p)$.
Below, I've plotted the probability mass function (PMF) of $X$
for different values of $p$.
I've included a horizontal dashed line to indicate $P(X = 7;p)$.

```{r}
x <- 7
n <- 10
width <- 2
for(i in (x - width):(x + width)){
  temp <- function(x){
    dbinom(x = x,
           size = n,
           prob = i / n)
  }
  barplot(height = temp(0:10),
          names.arg = 0:10,
          xlab = "x",
          ylab = "P(X = x;p)",
          main = paste0("p = ",
                        i / n),
          ylim = c(0,0.4))
  abline(h = temp(x),
         lty = 2)
}
```

I can also display $P(X = 7;p)$ in a table.

| $p$ | $P(X = 7;p)$ |
|-:|-:|
| 0.5 | `r round(x = dbinom(x = 7,size = 10,prob = 0.5),digits = 3)` |
| 0.6 | `r round(x = dbinom(x = 7,size = 10,prob = 0.6),digits = 3)` |
| 0.7 | `r round(x = dbinom(x = 7,size = 10,prob = 0.7),digits = 3)` |
| 0.8 | `r round(x = dbinom(x = 7,size = 10,prob = 0.8),digits = 3)` |
| 0.9 | `r round(x = dbinom(x = 7,size = 10,prob = 0.9),digits = 3)` |

We can see that $P(X = 7;p)$ is highest for $p = 0.7$.
Although I didn't test all possible values of $p$,
I can tell you that if you use calculus,
you will find that $P(X = 7;p)$ is maximized at $p = 0.7$.

### Different Sample Sizes

In the coin-flipping example above, we were effectively using the likelihood
to compare models.
Each candidate value of $p$ was its own model,
and we choose the one with the highest likelihood.

Suppose now that the magician flipped the coin twenty times
and it landed heads sixteen times.
The value of $p$ that maximizes the likelihood for these twenty flips
is $16 / 20 = 0.8$.
Yet, the value of the likelihood at $p = 0.8$ is
`r round(x = dbinom(x = 16,size = 20,prob = 0.8),digits = 3)`.
This is lower than the maximized likelihood when the number of flips was ten
and $p$ was set to 0.7.
Does that mean we should use $p = 0.7$ instead?

The answer is no.
We can't use the likelihood to compare models that use different samples.
One reason for this is that likelihoods tend to decrease as you add observations.
If we flipped the coin a hundred times and observed eighty heads,
the value of $p$ that would maximize the likelihood is $0.8$,
even though the value of the likelihood for $p = 0.8$ is only 0.099.
With 100 coin flips, we should have more confidence in our estimate,
not less.

By the way, if you want intuition for why the likelihood tends to decrease
with more observations,
it's because you keep multiplying by something less than 1.
Suppose you flip a fair coin one time.
The probability it lands heads is $\frac{1}{2}$.
If you flip the coin again,
the probability of two heads is $\frac{1}{2} \times \frac{1}{2} = \frac{1}{4}$.
If you flip the coin a third time,
the probability of three heads is
$\frac{1}{2} \times \frac{1}{2} \times \frac{1}{2} = \frac{1}{8}$.
This isn't just because three heads in a row is unlikely;
the probability of any length-three sequence of heads and tails is $\frac{1}{8}$.
It doesn't matter if it's heads-heads-heads,
heads-tails-heads, tails-heads-tails, or anything else.

### Model Accuracy

If you're computing a likelihood for a data set, you should believe
the likelihood is accurate.
What I mean is,
it wouldn't make sense to use the binomial PMF for normally distributed data,
or the normal PDF for binomially distributed data.
So, the likelihood should really be reserved for when you're willing to make
the argument that you know the underlying distribution of the data.
(Some statisticians might disagree with me on this point.)

### Maximum-Likelihood Estimator (MLE)

The value of the parameter that maximizes the likelihood for a given data set
is called the maximum-likelihood estimator (MLE).
There is a ton of statistical theory around MLEs.
They have a lot of convenient and desirable properties.

## Problems

### Problem 1

If all four assumptions of simple linear regression are met,
then we can write the likelihood as
$$
f(y_1,\dotsc,y_n;\beta_0,\beta_1,\sigma^2) = \frac{1}{\left(\sqrt{2\pi\sigma^2}\right)^n} \exp\left(-\frac{1}{2\sigma^2} \sum_{i=1}^n \left(y_i - \beta_0 - \beta_1 x_i\right)^2\right) \text{.}
$$
(I'm treating the observed values of the explanatory variable as constants
rather than random variables,
which is why I didn't put them as input into the likelihood function.)

Ignoring $\sigma^2$ for the moment,
how should the maximum-likelihood estimates of $\beta_0$ and $\beta_1$
compare to the ordinary least squares (OLS) estimates?

::: {.callout-tip title="Solution" collapse="true"}

They should be the same.
Watch what happens as we decrease the sum of the squared errors:
$$
\begin{align*}
\sum_{i=1}^n \left(y_i - \beta_0 - \beta_1 x_i\right)^2 \downarrow &\implies - \sum_{i=1}^n \left(y_i - \beta_0 - \beta_1 x_i\right)^2 \uparrow \\
&\implies -\frac{1}{2\sigma^2} \sum_{i=1}^n \left(y_i - \beta_0 - \beta_1 x_i\right)^2 \uparrow \\
&\implies \exp\left(-\frac{1}{2\sigma^2} \sum_{i=1}^n \left(y_i - \beta_0 - \beta_1 x_i\right)^2\right) \uparrow \\
&\implies \frac{1}{\left(\sqrt{2\pi\sigma^2}\right)^n} \exp\left(-\frac{1}{2\sigma^2} \sum_{i=1}^n \left(y_i - \beta_0 - \beta_1 x_i\right)^2\right) \uparrow \\
&\implies f(y_1,\dotsc,y_n;\beta_0,\beta_1,\sigma^2) \uparrow \text{.}
\end{align*}
$$
By minimizing the sum of squared errors, we maximize the likelihood.
The same is true of multiple linear regression.

:::

### Problem 2

Suppose you regress depression, as measured by the Beck Depression Inventory (BDI),
on total sleep over the last seven days.
Call this Model 1.
You then decide that you want to add another explanatory variable,
namely, number of interactions with friends or family over the last seven
days that lasted an hour or more.
Unfortunately, not everyone in your data set has this second explanatory variable
recorded.
You proceed anyone,
regression depression on sleep and social interactions for those participants
who have data for all variables.
Call this Model 2.

Should you compare the likelihoods of Model 1 and Model 2?
Justify your answer.

::: {.callout-tip title="Solution" collapse="true"}

No.
The models were fit on different samples.
Model 1 was fit on the complete data set,
and Model 2 was only fit on those participants who had all values for all
variables.
So, these likelihoods cannot be compared.

:::




