---
title: "AIC"
execute:
  echo: false
format: html
---

## Learning Objective

Use the AIC to evaluate models.


## Lesson

### Formula

The Akaike information criterion (AIC) is a statistic
that is used to compare the performance of different models.
It is computed as follows:
$$
\text{AIC} = 2k - 2 \log \hat{\mathcal{L}} \text{.}
$$
In this formula, $k$ is the number of parameters being estimated.
For multiple linear regression, $k = p + 2$;
this counts the beta coefficients
($p + 1$, for the number of covariates and the intercept)
and $\sigma^2$.

$\hat{\mathcal{L}}$ is the maximized likelihood of the data,
i.e., the value of the likelihood when you set the paramter(s)
equal to the MLE(s).

### Interpretation

A higher likelihood is better, so we want $\hat{\mathcal{L}}$
to be big.
Since $f(x) = \log x$ is an increasing function (see below),
a big $\hat{\mathcal{L}}$ corresponds to a big
$\log \hat{\mathcal{L}}$,
so we want $\log \hat{\mathcal{L}}$ to be big.
This means we want $- 2 \log \hat{\mathcal{L}}$
to be very small.
This means we want the AIC to be small.

```{r}
curve(expr = log,
      from = 0.01,
      to = 2)
abline(h = 0,
       v = 0)
abline(v = 1,
       lty = 2)
```

Why do we add $2k$ to the AIC?
This is how we penalize adding covariates to the model,
similar to with the adjusted $R^2$.
We add $2k$ to prevent overfitting.

### Example

Let's return to the data set collected from introductory statistics
students.

```{r}
#| echo: true
d <- Lock5Data::StudentSurvey
```

I'd like to fit a large model and a small model,
but there's a problem.
I know some observations in this data set are missing values
for some variables.

## Problems

### Problem 1



::: {.callout-tip title="Solution" collapse="true"}



:::

### Problem 2



::: {.callout-tip title="Solution" collapse="true"}



:::




