---
title: "AIC"
execute:
  echo: false
format: html
---

## Learning Objective

Use the AIC to evaluate models.


## Lesson

### Formula

The Akaike information criterion (AIC) is a statistic
that is used to compare the performance of different models.
It is computed as follows:
$$
\text{AIC} = 2k - 2 \log \hat{\mathcal{L}} \text{.}
$$
In this formula, $k$ is the number of parameters being estimated.
For multiple linear regression, $k = p + 2$;
this counts the beta coefficients
($p + 1$, for the number of covariates and the intercept)
and $\sigma^2$.

$\hat{\mathcal{L}}$ is the maximized likelihood of the data,
i.e., the value of the likelihood when you set the paramter(s)
equal to the MLE(s).

### Interpretation

A higher likelihood is better, so we want $\hat{\mathcal{L}}$
to be big.
Since $f(x) = \log x$ is an increasing function (see below),
a big $\hat{\mathcal{L}}$ corresponds to a big
$\log \hat{\mathcal{L}}$,
so we want $\log \hat{\mathcal{L}}$ to be big.
This means we want $- 2 \log \hat{\mathcal{L}}$
to be very small.
This means we want the AIC to be small.

```{r}
curve(expr = log,
      from = 0.01,
      to = 2)
abline(h = 0,
       v = 0)
abline(v = 1,
       lty = 2)
```

Why do we add $2k$ to the AIC?
This is how we penalize adding covariates to the model,
similar to with the adjusted $R^2$.
We add $2k$ to prevent overfitting.

### Example

Let's return to the data set collected from introductory statistics
students.

```{r}
#| echo: true
d <- Lock5Data::StudentSurvey
d <- d[!is.na(d$GPA), ]
```

I'd like to fit a large model and a small model,
but there's a problem.
I know some observations in this data set are missing values
for some variables.

```{r}
#| echo: true
colSums(is.na(d))
```

If I fit a model with `Year`, `VerbalSAT`, and `MathSAT`,
I can use all 345 observations.
If I fit a model with `Height`,
the maximum number of observations I can use is $345 - 7 = 338$.
The two models will be based on two different samples,
meaning I can't compare their likelihoods,
meaning I can't compare their AICs.
So, if I want to use the AIC to compare two models, I have to make sure
they are fit on exactly the same samples first.
In order to achieve this for this example,
I'm going to first create a new data set with complete cases,
i.e., observations with no missing values.

```{r}
#| echo: true
d_complete <- d[complete.cases(d), ]
```

Now I can fit my two models.

```{r}
#| echo: true
mod_01 <- lm(formula = GPA ~ Year + Sex + Smoke + Award + Exercise + TV +
               Height + Weight + Siblings + BirthOrder + VerbalSAT + MathSAT +
               Pulse + Piercings,
             data = d_complete)
mod_02 <- lm(formula = GPA ~ Year + VerbalSAT + MathSAT,
             data = d_complete)
```

Before looking at their AICs, I'm going to look at their
log-likelihooods.
The log-likelihood is just the log of the likelihood;
in this case, it's the log of the maximized likelihood.
Statisticians often prefer the log-likelihood over the likelihood
for reasons that I won't go into here.

```{r}
logLik(mod_01)
logLik(mod_02)
```

Note that the larger model, Model 1, has the higher log likelihood.

Now let's look at the AICs.

```{r}
AIC(mod_01)
AIC(mod_02)
```

Model 1 has the lower AIC!
This is a surprise to me.
It means that, according to the AIC,
the improvement in fit for the larger model is "worth" all those extra variables.
This contrasts with the conclusion from the test set error.

### Warnings

The help file for the R function `AIC`, accessible with the command `?AIC`,
has some helpful warnings:

- "The theory of AIC requires that the log-likelihood
has been maximized:
whereas AIC can be computed for models not fitted by maximum likelihood,
their AIC values should not be compared."
- "Examples of models not 'fitted to the same data' are where the response
is transformed (accelerated-life models are fitted to log-times)
and where contingency tables have been used to summarize data."

I'm not sure whether the first point will come up in this class,
but it's worth keeping in mind when you reach more advanced analyses.
The second point is very important for this class;
when using the AIC to compare models,
make sure those models have the same outcome.
If you transform your outcome, for example, by logging it,
you shouldn't use the AIC to compare the model with the transformed outcome
to the model with the untransformed outcome.

## Problems

### Problem 1

In the example above, I didn't check any assumptions for the two models.
Should I have? Why or why  not?

::: {.callout-tip title="Solution" collapse="true"}

Yes, I should've.
The AIC is based on the likelihood, which is based on all of our
assumptions being met.
So, I should've checked the assumptions for both models.

```{r}
# Model 1
plot(x = mod_01$fitted.values,
     y = mod_01$residuals,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Model 1")
abline(h = 0,
       col = "red")
boxplot(x = mod_01$residuals,
        ylab = "Residuals",
        main = "Model 1")
hist(x = mod_01$residuals,
     xlab = "Residuals",
     main = "Model 1")
qqnorm(y = mod_01$residuals,
       main = "Model 1")
qqline(mod_01$residuals)

# Model 2
plot(x = mod_02$fitted.values,
     y = mod_02$residuals,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Model 2")
abline(h = 0,
       col = "red")
boxplot(x = mod_02$residuals,
        ylab = "Residuals",
        main = "Model 2")
hist(x = mod_02$residuals,
     xlab = "Residuals",
     main = "Model 2")
qqnorm(y = mod_02$residuals,
       main = "Model 2")
qqline(mod_02$residuals)

```

Based on the plots, it seems like the linearity, homoscedasticity,
and normality assumptions are met for both models.
I don't know much about how the data were collected,
but I'm willing to go along with the assumption that the observations
are independent.

:::

### Problem 2

If Model A has an AIC of 400 and Model B has an AIC of 405,
which is better?
Assume all relevant assumptions have been met.

::: {.callout-tip title="Solution" collapse="true"}

Model A is better because it has the lower AIC.

:::




