---
title: "Likelihood Ratio Test"
execute:
  echo: false
format: html
---


## Learning Objective

- Interpret the results of a likelihood ratio test.
- Explain when it would be inappropriate to use a likelihood ratio test.


## Lesson

A likelihood ratio test (LRT) is a hypothesis test that can only be used
with nested models.
Nesting can get somewhat technical,
but this statement will get you pretty far:
If Model A estimates
a subset of the parameters estimated by Model B,
then Model A is nested within Model B.

```{r}
d <- Lock5Data::StudentSurvey
d$MathSATHigher <- as.integer(d$MathSAT > d$VerbalSAT)
d <- d[!is.na(d$GPA), ]
d_complete <- d[complete.cases(d), ]
```


For example, `mod_02` below is nested within `mod_01`.
This is because all the covariates in `mod_02` are also in `mod_01`.

```{r}
#| echo: true
mod_01 <- lm(formula = GPA ~ Sex + Smoke + Award + Exercise + TV +
               Height + Weight + Siblings + BirthOrder + VerbalSAT + MathSAT +
               Pulse + Piercings,
             data = d_complete)
mod_02 <- lm(formula = GPA ~ VerbalSAT + MathSAT,
             data = d_complete)
```

`mod_03` below is not nested within `mod_01`.
`MathSATHigher` is a variable I created that equals 1 if the Math SAT score
is higher than the Verbal SAT score and 0 otherwise.

```{r}
#| echo: true
mod_03 <- lm(formula = GPA ~ MathSATHigher,
             data = d_complete)
```

If you want to know what I mean by "Nesting can get somewhat technical,"
click on the callout below.

::: {.callout-caution collapse="true"}
## Caution: Math

`mod_04`, below, is nested within `mod_01` and `mod_02`,
even though it contains a covariate (`SAT`) that isn't in `mod_01` or `mod_02`.

```{r}
#| echo: true
mod_04 <- lm(formula = GPA ~ SAT,
             data = d_complete)
```

This is because `SAT = VerbalSAT + MathSAT`,
and both `mod_01` and `mod_02` contain `VerbalSAT` and `MathSAT`.

In order to explain this further, I'm going to use `mod_02` instead
of `mod_01` because `mod_02` is simpler.
`mod_02` is this:
$$
E\left(\texttt{GPA}\middle|\texttt{VerbalSAT},\texttt{MathSAT}\right) = \beta_0 + \beta_1 \texttt{VerbalSAT} + \beta_2 \texttt{MathSAT} \text{.}
$$

`mod_04` is this:
$$
\begin{align*}
E\left(\texttt{GPA}\middle|\texttt{SAT}\right) &= \beta_0 + \beta_3 \texttt{SAT} \\
&= \beta_0 + \beta_3 \left(\texttt{VerbalSAT} + \texttt{MathSAT}\right) \\
&= \beta_0 + \beta_3 \texttt{VerbalSAT} + \beta_3 \texttt{MathSAT} \text{.} \\
\end{align*}
$$

So, `mod_04` is `mod_02` but with the additional constraint that $\beta_1 = \beta_2$.
This counts as nesting.

What about `mod_03`?
`mod_03` is
$$
\begin{align*}
E\left(\texttt{GPA}\middle|\texttt{MathSATHigher}\right) &= \beta_0 + \beta_4  \texttt{MathSATHigher.} \\
\end{align*}
$$

There's no way to write `mod_03` as a version of `mod_2`,
but with some additional constraint.
`mod_03` is a discontinuous function of `VerbalSAT` and `MathSAT`,
but `mod_02` is a continuous function of these two variables.

:::

The null hypothesis of the LRT is that the smaller model is the correct one.
The alternative hypothesis is that the larger model is the correct one.

Let's use the LRT to compare `mod_01` to `mod_02`.

```{r}
#| echo: true
anova(mod_01,
      mod_02,
      test = "LRT")
```

The p-value is very small, so we can conclude that the larger model is better.

What happens if we try to get R to conduct a hypothesis test with two
non-nested models?

```{r}
#| echo: true
anova(mod_01,
      mod_03,
      test = "LRT")
```

R was not able to determine that these models are not nested,
which means *you* need to do that before trying to use this test.

## Problems

### Problem 1

If you fail to reject the null hypothesis in an LRT,
what should you do?

::: {.callout-tip title="Solution" collapse="true"}

Generally you would conclude that the larger model is no better than
the smaller one,
and use the smaller one.
We often prefer smaller models because they are easier to interpret
and have greater power.

:::

### Problem 2

If I hadn't removed observations with missing data,
and instead had fit `mod_02` on the data set with 345 observations
instead of 331,
could I still have used the LRT to compare the models?

::: {.callout-tip title="Solution" collapse="true"}

No.
The likelihood ratio test, like its name implies,
uses the likelihood.
So, you can only use it to compare models that have been fit on the same data.

:::




