---
title: "Likelihood Ratio Test"
execute:
  echo: false
format: html
---

## Learning Objective

- Interpret the results of a likelihood ratio test.
- Explain when it would be inappropriate to use a likelihood ratio test.


## Lesson

A likelihood ratio test (LRT) is a hypothesis test that can only be used
with nested models.
Model A is nested within Model B if Model A estimates
a subset of the parameters estimated by Model B.

```{r}
d <- Lock5Data::StudentSurvey
d <- d[!is.na(d$GPA), ]
d_complete <- d[complete.cases(d), ]
```


For example, `mod_02` below is nested within `mod_01`.
This is because all the covariates in `mod_02` are also in `mod_01`.

```{r}
#| echo: true
mod_01 <- lm(formula = GPA ~ Year + Sex + Smoke + Award + Exercise + TV +
               Height + Weight + Siblings + BirthOrder + VerbalSAT + MathSAT +
               Pulse + Piercings,
             data = d_complete)
mod_02 <- lm(formula = GPA ~ Year + VerbalSAT + MathSAT,
             data = d_complete)
```

`mod_03` below is not nested within `mod_01` because it uses a covariate
that isn't in `mod_01`.

```{r}
#| echo: true
mod_03 <- lm(formula = GPA ~ Year + SAT,
             data = d_complete)
```

The null hypothesis of the LRT is that the smaller model is the correct one,
and any difference in fit is due to chance alone.
The alternative hypothesis is that the fit of the larger model is better
than can be explained by chance alone.

Let's use the LRT to compare `mod_01` to `mod_02`.

```{r}
#| echo: true
anova(mod_01,
      mod_02,
      test = "LRT")
```

The p-value is very small, so we can conclude that the larger model is better.

What happens if we try to get R to conduct a hypothesis test with two
non-nested models?

```{r}
#| echo: true
anova(mod_01,
      mod_03,
      test = "LRT")
```

R was not able to determine that these models are not nested,
which means *you* need to do that before trying to use this test.

## Problems

### Problem 1

If you fail to reject the null hypothesis in an LRT,
what should you do?

::: {.callout-tip title="Solution" collapse="true"}

Generally you would conclude that the larger model is no better than
the smaller one,
and use the smaller one.
We often prefer smaller models because they are easier to interpret
and have greater power.

:::

### Problem 2

If I hadn't removed observations with missing data,
and instead had fit `mod_02` on the data set with 345 observations
instead of 331,
could I still have used the LRT to compare the models?

::: {.callout-tip title="Solution" collapse="true"}

No.
The likelihood ratio test, like its name implies,
uses the likelihood.
So, you can only use it to compare models that have been fit on the same data.

:::




