---
title: "Math Refresher"
format: html
---

Doing statistics necessarily requires doing some mathematics,
even if that mathematics is identifying which of two numbers is larger
than the other.
In this section, I've collected those math skills that I think
are most important for understanding the material that follows.

## Equations

Researchers often summarize their regression into a single equation;
it's a compact way to represent the outcome, the covariates,
and how they relate to each other.
Also, interpreting coefficients requires some manipulation
of equations.
So, it's important that we are on the same page about how they work.

An equals sign indicates that the two sides of the equation are exactly
equal to each other:
$$
a = a \text{.}
$$
This implies that if we have an equation, and we do the same exact thing
to both sides, the equality will still be true.
For example,
$$
a + 3 = a + 3 \text{.}
$$
We started from the same place, so if we move the same distance in the same
direction,
we should end up in the same place.

Of course, this applies to other operations besides addition:
$$
\begin{align*}
3a &= 3a \\
a/3 &= a/3 \\
a - 3 &= a - 3 \text{.}
\end{align*}
$$

## Exponents

### Definition

An exponent indicates how many times you should multiply something by itself.
For example,
$$
a^3 = a \times a \times a \text{.}
$$

### Properties

If you multiply two terms with the same base, their exponents add:
$$
\begin{align*}
a^3 a^2 &= (a \times a \times a)(a \times a) \\
&= a \times a \times a \times a \times a \\
&= a^5 \\
&= a^{3+2} \text{.}
\end{align*}
$$
In the example above, $a$ is the base.
For the first term, $a^3$, the exponent is $3$;
for the second term, $a^2$, the exponent is $2$.

If you divide two terms with the same base, the exponents subtract:
$$
\begin{align*}
\frac{a^3}{a^2} &= \frac{a \times a \times a}{a \times a} \\
&= a \\
&= a^1 \\
&= a^{3 - 2} \text{.}
\end{align*}
$$
Look at what is implied by this fact:
$$
\begin{align*}
a^{-3} &= a^{2-5} \\
&= \frac{a^2}{a^5} \\
&= \frac{a \times a}{a\times a \times a \times a \times a} \\
&= \frac{1}{a\times a\times a} \\
&= \frac{1}{a^3} \text{.} \\
\end{align*}
$$


In general,
$$
a^{-b} = \frac{1}{a^b} \text{.}
$$

The next property is one that I see a lot of students get confused about.
It says that
$$
\left(a^b\right)^c = a^{bc} \text{.}
$$
This is easier to see with an example:
$$
\begin{align*}
\left(a^3\right)^2 &= a^3 \times a^3 \\
&= (a\times a\times a) \times (a \times a \times a) \\
&= a\times a\times a \times a \times a \times a \\
&= a^6 \\
&= a^{3\times 2} \text{.}
\end{align*}
$$

### A Special Base

A very common base in statistics is the special number $e$,
which is approximately equal to 2.71828.
It's like $\pi$ in that it is really useful in mathematics
and can't be represented as a ratio of two whole numbers.
We will see $e$ over and over again in this class.
Sometimes we use the notation $\exp(b)$ instead of $e^b$
to make the exponent easier to read.
For example, we will see things like this quite a bit:
$$
e^{a+bx} = \exp(a+bx) \text{.}
$$


## Logarithms

### Relationship to Exponentiation

The logarithm is the opposite of the exponent.
In general,
$$
\log_a c = b
$$
is exactly the same thing as
$$
a^b = c \text{.}
$$
This means that taking a logarithm "undoes" or "reverses" exponentiation.
If we start with $a^b = c$ and take $\log_a$ of both sides, we get
$$
\begin{align*}
a^b &= c \\
\log_a a^b &= \log_a c \\
b &= \log_a c \text{.}
\end{align*}
$$
Why is this true?
The expression $\log_a a^b$ is asking,
"$a$ to what power equals $a^b$?"
The answer to that question is $b$.

It's also true that exponentiation undoes taking a logarithm.
The expression $\log_a c$ is asking the question,
"$a$ to what power equals $c$?"
Whatever that power is, if we raise $a$ to it, we'll get $c$.
So,
$$
a^{\log_a c} = c \text{.}
$$



### A Special Base

You may have learned that $\log$ without a base should be interpreted
as log base 10.
Unfortunately, this is not how statisticians do things.
We use log with base $e$ so often that for us,
$\log$ with no base means $\log_e$.
You may have learned that log with base $e$ is the natural log,
and is written $\ln$.
For whatever reason, statisticians use $\log$ rather than $\ln$.

Just to drive that point home, for a statistican,
$\log e = 1$, since $e^1 = e$.
In this textbook, if I want to use log with base 10,
I'll explicitly note it as $\log_{10}$.

### Properties

Similar to exponents, logarithms (or logs) have special properties.
The properties hold for all bases, but for notational simplicity
I'm just going to use log with base $e$.

The first property is that
$$
\log (ab) = \log a + \log b \text{.}
$$
Why is this true?
Suppose
$$
\begin{align*}
\log(ab) &= x \\
\log a &= y \\
\log b &= z \text{.}
\end{align*}
$$
This is equivalent to
$$
\begin{align*}
e^x &= ab \\
e^y &= a \\
e^z &= b \text{.}
\end{align*}
$$
So, we can say
$$
\begin{align*}
e^x &= e^y e^z \\
e^x &= e^{y+z} \\
\log e^x &= \log e^{y+z} \\
x &= y + z \\
\log(ab) &= \log a + \log b \text{.}
\end{align*}
$$

The second property is that
$$
a \log b = \log b^a \text{.}
$$
Why is this true?
Define
$$
\begin{align*}
a \log b &= x \\
\log b^a &= y \text{.}
\end{align*}
$$
The first implies
$$
\begin{align*}
\log b &= \frac{x}{a} \\
\exp\left(\frac{x}{a}\right) &= b \\
\end{align*}
$$
and the second implies $e^y = b^a$.
We can combine these results to get
$$
\begin{align*}
e^y &= \exp\left(\frac{x}{a}\right)^a \\
e^y &= \exp\left(a\frac{x}{a}\right) \\
e^y &= e^x \\
y &= x \\
\log b^a &= a \log b \text{.}
\end{align*}
$$


## Intercept and Slope

We will talk about intercepts and slopes over and over again in this course.
So, it's important to start with a good foundation.

Consider the line $y = 2x + 4$:
```{r}
x_temp <- seq(from = -3,
              to = 3,
              length.out = 101)
y_temp <- 2 * x_temp + 4
plot(x = x_temp,
     y = y_temp,
     type = "l",
     xlab = "x",
     ylab = "y",
     main = "y = 2x + 4",
     col = "blue")
abline(h = 0)
abline(v = 0)
```

The intercept is 4. This means that when $x = 0$, $y = 4$:
```{r}
plot(x = x_temp,
     y = y_temp,
     type = "l",
     xlab = "x",
     ylab = "y",
     main = "y = 2x + 4",
     col = "blue")
abline(h = 0)
abline(v = 0)
lines(x = c(0,0),
      y = c(0,4),
      col = "red",
      lty = 2)
lines(x = c(-3,0),
      y = c(4,4),
      col = "red",
      lty = 2)
```

The slope is 2. This means that for each one-unit increase in $x$,
$y$ increases by two units.
```{r}
plot(x = x_temp,
     y = y_temp,
     type = "l",
     xlab = "x",
     ylab = "y",
     main = "y = 2x + 4",
     col = "blue")
abline(h = 0)
abline(v = 0)
lines(x = c(0,0),
      y = c(0,4),
      col = "red",
      lty = 2)
lines(x = c(-3,0),
      y = c(4,4),
      col = "red",
      lty = 2)
lines(x = c(1,1),
      y = c(0,6),
      col = "red",
      lty = 2)
lines(x = c(-3,1),
      y = c(6,6),
      col = "red",
      lty = 2)
```

## Problems












