---
title: "Intercept and Slope"
format: html
---

## Learning Objective

Interpret the intercept and slope of a line.


## Lesson

We will talk about intercepts and slopes over and over again in this course.
So, it's important to start with a good foundation.

Consider the line $y = 2x + 4$:
```{r}
#| echo: false

x_temp <- seq(from = -3,
              to = 3,
              length.out = 101)
y_temp <- 2 * x_temp + 4
plot(x = x_temp,
     y = y_temp,
     type = "l",
     xlab = "x",
     ylab = "y",
     main = "y = 2x + 4",
     col = "blue")
abline(h = 0)
abline(v = 0)
```

The intercept is 4. This means that when $x = 0$, $y = 4$:
```{r}
#| echo: false

plot(x = x_temp,
     y = y_temp,
     type = "l",
     xlab = "x",
     ylab = "y",
     main = "y = 2x + 4",
     col = "blue")
abline(h = 0)
abline(v = 0)
lines(x = c(0,0),
      y = c(0,4),
      col = "red",
      lty = 2)
lines(x = c(-3,0),
      y = c(4,4),
      col = "red",
      lty = 2)
```

The slope is 2. This means that for each one-unit increase in $x$,
$y$ increases by two units.
```{r}
#| echo: false

plot(x = x_temp,
     y = y_temp,
     type = "l",
     xlab = "x",
     ylab = "y",
     main = "y = 2x + 4",
     col = "blue")
abline(h = 0)
abline(v = 0)
lines(x = c(0,0),
      y = c(0,4),
      col = "red",
      lty = 2)
lines(x = c(-3,0),
      y = c(4,4),
      col = "red",
      lty = 2)
lines(x = c(1,1),
      y = c(0,6),
      col = "red",
      lty = 2)
lines(x = c(-3,1),
      y = c(6,6),
      col = "red",
      lty = 2)
```

## Problems

### Problem 1

If $x$ is the temperature in degrees Fahrenheit
and $y$ is the temperature in degrees Celsius,
then $y = \frac{5}{9}(x - 32)$.

#### Part (a)

When it's 0 degrees Fahrenheit, what is the temperature in degrees Celsius?

::: {.callout-tip title="Solution" collapse="true"}

$$
\begin{align*}
y &= \frac{5}{9}(0 - 32) \\
&= -\frac{160}{9} \\
&\approx - 17.78
\end{align*}
$$

:::

#### Part (b)

If the temperature increases ten degrees Fahrenheit,
how much does it increase in degrees Celsius?

::: {.callout-tip title="Solution" collapse="true"}

The slope of the line is $\frac{5}{9}$,
so with each one-degree Fahrenheit increase in temperature,
there is a corresponding $\frac{5}{9}$-degree Celsius increase.
This means that with a ten-degree Fahrenheit increase in temperature,
there is a corresponding $\frac{5}{9}\times 10 \approx 5.56$-degree Celsius
increase in temperature.

:::

### Problem 2

There are 2.54 centimeters in every inch.
Write a function connecting $y$,
the length in inches, to $x$,
the length in centimeters.

::: {.callout-tip title="Solution" collapse="true"}

The first time I answered this question, I wrote $y = 2.54x$,
but I knew that couldn't be right because it would imply that we would
always have more inches than centimeters.
But one centimeter is less than one inch,
so the same distance should require more centimeters than inches.
So, it's actually the case that $2.54y = x$, or
$$
\begin{align*}
2.54y &= x \\
y &= \frac{x}{2.54} \\
y &= 0.3937x \text{.}
\end{align*}
$$

:::




