---
title: "Integrals"
execute:
  echo: false
format: html
---

## Learning Objective

Use and interpret the terms “integral” and “integrate”.


## Lesson

First, let me say that you do not need to know calculus
in order to understand this material.
All I want is for you to understand the concept.
Essentially, the integral of a function is the area under it.
In each of the following plots,
the blue area represents the integral of the function
from $a$ to $b$.
```{r}
# Quadratic
f_quad <- function(x){
  -x^2 + 4
}
x <- seq(from = -3,
         to = 3,
         length.out = 500)
y <- f_quad(x)
plot(x = x,
     y = y,
     type = "l",
     main = "a = -2, b = 1")
abline(h = 0)
abline(v = 0)
polygon(x = c(x[x >= -2 & x <= 1],
              1,
              1,
              -2),
        y = c(y[x >= -2 & x <= 1],
              f_quad(1),
              0,
              0),
        col = "blue")

# Sin
f_sin <- function(x){
  x / 2 + sin(x)
}
x_sin <- seq(from = 0,
             to = 6,
             length.out = 250)
y_sin <- f_sin(x_sin)
plot(x = x_sin,
     y = y_sin,
     type = "l",
     xlab = "x",
     ylab = "y",
     main = "a = 2, b = 4")
abline(h = 0)
abline(v = 0)
polygon(x = c(x_sin[x_sin >= 2 & x_sin <= 4],
              4,
              4,
              2),
        y = c(y_sin[x_sin >= 2 & x_sin <= 4],
              f_quad(4),
              0,
              0),
        col = "blue")
```

## Problems

### Problem 1

For this problem, I do want you to compute an integral,
but I promise that you can do it without any knowledge of calculus.
Consider the function $f(x) = x$, or $y = x$,
plotted below.
```{r}
plot(x = x,
     y = x,
     type = "l",
     ylab = "y")
abline(h = 0,
       v = 0)
```
What is the integral of this function from 0 to 1?

::: {.callout-tip title="Solution" collapse="true"}

The integral of the function is the area under the curve.
This curve is a straight line.
It intersects the $y$-axis at $x = 0$.
So, the integral from 0 to 1 will be the area of the triangle
with base 1 and height 1.
That's $1 \times 1 / 2 = \frac{1}{2}$.
```{r}
plot(x = x,
     y = x,
     type = "l",
     ylab = "y")
abline(h = 0,
       v = 0)
polygon(x = c(x[x >= 0 & x <= 1],
              1,
              1,
              0),
        y = c(x[x >= 0 & x <= 1],
              1,
              0,
              0),
        col = "blue")
```

:::

### Problem 2

Consider the function $f(x) = 1$, or $y = 1$,
plotted below.
```{r}
plot(x = x,
     y = rep(x = 1,
             times = length(x)),
     type = "l",
     ylab = "y",
     ylim = c(-3,3))
abline(h = 0,
       v = 0)
```
What is the integral of this function from 0 to 1?

::: {.callout-tip title="Solution" collapse="true"}

This is a square. It has area 1.
```{r}
plot(x = x,
     y = rep(x = 1,
             times = length(x)),
     type = "l",
     ylab = "y",
     ylim = c(-3,3))
abline(h = 0,
       v = 0)
polygon(x = c(x[x >= 0 & x <= 1],
              1,
              1,
              0),
        y = c(rep(x = 1,
                  times = sum(x >= 0 & x <= 1)),
              1,
              0,
              0),
        col = "blue")
```

:::




