---
title: "Derivatives"
execute:
  echo: false
format: html
---

## Learning Objective

Use and interpret the terms “derivative” and “differentiate”.


## Lesson

Again, you do not need to know calculus here;
I just want you to understand the concept.
The derivative of a function at a point is its slope at that point.
To differentiate a function is to find its derivative.

The plot below shows the function $f(x) = -x^2 + 4$.
I've plotted the tangent lines, in red,
at $x = -2$, $x = 0$, and $x = 1$.
```{r}
# Quadratic
f_quad <- function(x){
  -x^2 + 4
}
x <- seq(from = -3,
         to = 3,
         length.out = 500)
y <- f_quad(x)
plot(x = x,
     y = y,
     type = "l")
abline(h = 0)
abline(v = 0)
abline(a = 8,
       b = 4,
       col = "red")
abline(h = 4,
       col = "red")
abline(a = 5,
       b = -2,
       col = "red")
```
The derivative at $x = -2$,
represented by $f'(-2)$, is 4.
So, the tangent line is increasing.
The derivative at $x = 0$ is $f'(0) = 0$,
so the tangent line is flat,
and the derivative at $x = 1$ is $f'(1) = -2$,
so the tangent line is decreasing.


The plot below shows the function $f(x) = \frac{x}{2} + \sin(x)$.
I've plotted the tangent lines, in red,
at $x = 2$ and $x = 4$.
```{r}
# Sin
f_sin <- function(x){
  x / 2 + sin(x)
}
x_sin <- seq(from = 0,
             to = 6,
             length.out = 250)
y_sin <- f_sin(x_sin)
plot(x = x_sin,
     y = y_sin,
     type = "l",
     xlab = "x",
     ylab = "y")
abline(h = 0)
abline(v = 0)
abline(a = sin(2) - 2 * cos(2),
       b = 0.5 + cos(2),
       col = "red")
abline(a = sin(4) - 4 * cos(4),
       b = 0.5 + cos(4),
       col = "red")
```

The derivative at $x = 2$,
represented by $f'(2)$, is around 0.08.
So, the tangent line is increasing.
The derivative at $x = 4$ is $f'(4) \approx -0.15$,
so the tangent line is decreasing.

## Problems

### Problem 1

What is the derivative of the function $f(x) = 2 - 3x$
at $x = 0$?

::: {.callout-tip title="Solution" collapse="true"}

This is a line, so it has the same slope everywhere.
That slope is $-3$.

:::

### Problem 2

What is the sign (positive or negative) of the derivative
of the function below at $x = -2$?
```{r}
plot(x = x,
     y = x^2,
     type = "l")
abline(h = 0)
abline(v = 0)
```

::: {.callout-tip title="Solution" collapse="true"}

When $x = -2$, the function is decreasing, so its slope is negative.

:::




