---
title: "Distribution"
format: html
---

## Learning Objective

Use and interpret the term "distribution".


## Lesson

### Two Meanings

In statistics, there are at least two meanings of the word "distribution".
The first refers to the data.
If someone asks how the data are distributed, you might show them
a histogram or boxplot.
What they want to know is where the points fall on the number line.
Here are a histogram and boxplot of the pH of 53 lakes in Florida:

```{r}
#| echo: false

d <- Lock5Data::FloridaLakes
hist(x = d$pH,
     xlab = "pH",
     main = "Histogram of pH of Florida Lakes")
boxplot(x = d$pH,
        ylab = "pH",
        main = "Boxplot of pH of Florida Lakes")
```
We can see that most of the lakes have a pH between 5 and 8,
none has a pH below 3 or above 10,
and the distribution is symmetric.

The second meaning of the word "distribution" has to do with
where the data come from.
I don't mean who collected them or where they were collected;
I mean the random process that generated the data.
If we know the distribution in this sense,
we know the exact probability of observing any number.
It might help to consider some examples.

### Examples

#### Bernoulli Distribution

The Bernoulli distribution is the simplest possible distribution.
There are only two possible outcomes: 0 and 1.
Suppose $X$ is Bernoulli-distributed with $P(X = 1) = 0.2$.
Now we know the probability that $X$ takes on any value.
If $P(X = 1) = 0.2$ and $X$ can only be 0 or 1,
then $P(X = 0) = 1 - P(X = 1) = 1 - 0.2 = 0.8$.
$$
\begin{align*}
P(X = 0) &= 1 - P(X = 1) \\
&= 1 - 0.2 \\
&= 0.8 \text{.}
\end{align*}
$$
If $x \ne 0$ and $x \ne 1$, then $P(X = x) = 0$.

The Bernoulli distribution describes any random variable that can only
take on two values.
Suppose we follow 100 patients for one year and record whether
each has a heart attack in that year.
If a patient does have a heart attack, we record a value of 1;
if they don't, we record a value of 0.
In this case, we have 100 random variables,
and each follows the Bernoulli distribution.

#### Normal Distribution

The normal distribution is probably the one that gets talked about the most
in introductory statistics courses.
It has a classic bell shape:
```{r}
#| echo: false

curve(expr = dnorm,
      from = -3,
      to = 3,
      ylab = "f(x)",
      ylim = c(0,0.8))
```
The normal distribution always has this shape, although it may
be moved to the left or right or squished together or spread apart.
For example, here is the same normal distribution, but moved one unit
to the right:
```{r}
#| echo: false

temp <- function(x){
  dnorm(x = x,
        mean = 1)
}
curve(expr = temp,
      from = -3,
      to = 3,
      ylab = "f(x)",
      ylim = c(0,0.8))
```
And here is the same normal distribution as in the first plot,
but squished to the center:
```{r}
#| echo: false

temp <- function(x){
  dnorm(x = x,
        sd = 0.5)
}
curve(expr = temp,
      from = -3,
      to = 3,
      ylab = "f(x)")
```

The pH data from above kind of look like they might be normally distributed:
```{r}
#| echo: false

mu_temp <- mean(d$pH)
sd_temp <- sd(d$pH)
temp <- function(x){
  dnorm(x = x,
        mean = mu_temp,
        sd = sd_temp)
}
hist(x = d$pH,
     xlab = "pH",
     main = "Histogram of pH of Florida Lakes",
     freq = FALSE)
curve(expr = temp,
      add = TRUE,
      col = "red")
```
What I mean is that the empirical pH distribution (the first meaning of the word)
appears to match an idealized normal distribution,
so the data may have been generated from a normal distribution
(the second meaning of the word).

### Identically Distributed

If $X$ and $Y$ are identically distributed, it means they have the same
distribution.
This means that if $P(X = 3) = 0.2$, then $P(Y = 3) = 0.2$ as well.
If $P(X \le -4) = 0.8$, then $P(Y \le -4) = 0.8$ as well.
This comes up a lot in statistics,
because we often assume that data are identically distributed.

## Problems

### Problem 1

If I say, "The gene counts are not Poisson-distributed,"
what does that mean?

::: {.callout-tip title="Solution" collapse="true"}

This means that the gene counts differ from the Poisson distribution
in some way.
Perhaps the gene counts are more variable or spread out than the Poisson
distribution.
Perhaps there are more zeroes in the gene counts than you would expect
from the Poisson distribution.
It could mean a lot of things,
but practically, it means we should probably not use the Poisson distribution
to model the gene counts.

:::

### Problem 2

If $X$ and $Y$ are identically distributed and $P(Y \le 10) = 0$,
what is $P(X > 10)$?

::: {.callout-tip title="Solution" collapse="true"}

$$
\begin{align*}
P(X > 10) &= P(Y > 10) \\
&= 1 - P(Y \le 10) \\
&= 1 - 0 \\
&= 1
\end{align*}
$$

:::




