---
title: "Expected Value"
execute:
  echo: false
format: html
---

## Learning Objective

Define, use, and interpret the term "expected value"
in the context of random variables.


## Lesson

### Notation

We use $E(X)$ to represent the expected value of $X$.
You might see "expectation" instead of "expected value".

### Visual Representation

The expected value of a random variable is where you would have to place
a fulcrum to get it to balance.
Let's look at a few examples.

This is a standard normal distribution.
It's expected value is 0.
```{r}
curve(expr = dnorm,
      from = -3,
      to = 3,
      xlab = "x",
      ylab = "f(x)",
      ylim = c(-0.05,0.4))
abline(h = 0)
polygon(x = c(0,0.2,-0.2),
        y = c(0,-0.05,-0.05),
        col = "black")
```

This is a uniform distribution from 0 to 1.
It's expected value is 0.5.
```{r}
plot(x = c(-0.5,0,0,1,1,1.5),
     y = c(0,0,1,1,0,0),
     type = "l",
     xlab = "x",
     ylab = "f(x)",
     ylim = c(-0.05,1))
abline(h = 0)
polygon(x = c(0.5,0.6,0.4),
        y = c(0,-0.05,-0.05),
        col = "black")
```

This is an exponential distribution with expected value equal to 1.
```{r}
curve(expr = dexp,
      from = 0,
      to = 6,
      xlab = "x",
      ylab = "f(x)",
      ylim = c(-0.05,1))
abline(h = 0,
       v = 0)
polygon(x = c(1,1.1,0.9),
        y = c(0,-0.05,-0.05),
        col = "black")
```

This is a Bernoulli distribution with $P(X = 1) = 0.75$.
It's expected value is 0.75.
```{r}
x_B <- c(-0.5,-0.5,0.5,0.5,1.5,1.5)
y_B <- c(0,0.25,0.25,0.75,0.75,0)
plot(x = x_B,
     y = y_B,
     type = "l",
     xlab = "x",
     ylab = "P(X = x)",
     ylim = c(-0.05,1))
segments(x0 = 0.5,
         y0 = 0,
         x1 = 0.5,
         y1 = 0.25)
abline(h = 0)
polygon(x = c(0.75,0.85,0.65),
        y = c(0,-0.05,-0.05),
        col = "black")
```
This example shows that the expected value of a random variable
doesn't need to be in its support.
$E(X)$ can be a value that $X$ can't take.


### Verbal Description

The expected value is a weighted average.
Each possible value of $X$ is weighted by its probability
(or density--a technical term that you don't have to memorize,
but that I can't, in good conscience, not use),
and then these weighted values are summed.


Consider this image.
```{r}
plot(x = c(-5,-5,-5,-3,-1,-1,2,2,2,2,2,3),
     y = c(0.1,0.3,0.5,0.1,0.1,0.3,0.1,0.3,0.5,0.7,0.9,0.1),
     pch = 19,
     ylim = c(-0.05,1),
     cex = 7,
     xlab = "x",
     ylab = "")
abline(h = 0)
```
I'm going to take each value and weight it by it's relative frequency.

| Value | Relative Frequency | Product |
|-:|-:|-:|
| -5 | 3/12 | -15/12 |
| -3 | 1/12 | -3/12 |
| -1 | 2/12 | -2/12 |
| 2 | 5/12 | 10/12 |
| 3 | 1/12 | 3/12 |

The sum of the weighted values is
```{r}
#| echo: true

sum(c(-15 / 12,
      -3 / 12,
      -2 / 12,
      10 / 12,
      3 / 12))
```
This is the expected value of this distribution.

### Math

For a continuous random variable, the expected value is
$$
E(X) = \int_{-\infty}^\infty x f(x)\,dx \text{.}
$$
For a discrete random variable, the expected value is
$$
E(X) = \sum_x x P(X = x) \text{.}
$$

## Problems

### Problem 1

True or false:
The expected value is the most commonly occurring value in a data set.
That's why it's the "expected" value;
it's the one you most expect to see.

::: {.callout-tip title="Solution" collapse="true"}

False.
We reserve "expected value" for random variables,
not for data sets.
The most commonly occurring value in a data set is the mode
of that data set.
The expected value may not even be a value that the random variable can take.

:::

### Problem 2

True or false:
$P(X > E(X))$ is always 0.5.

::: {.callout-tip title="Solution" collapse="true"}

False. This is only true when $E(X)$ is the median of the distribution,
which is not always the case.

:::




