---
title: "Graphical Representations of Probability"
execute:
  echo: false
format: html
---

## Learning Objective

Create, use, and interpret graphical representations of probability.


## Lesson

In statistics, we often want to represent probability visually.
This is especially common when talking about distributions.
In previous statistics courses, you probably used graphical representations
of probability when talking about p-values.

The shortest summary is that we tend to use surface area,
especially area under the curve, to represent
probability.
(This is consistent with the
[principle of proportional ink](https://callingbullshit.org/tools/tools_proportional_ink.html).)
Let's consider the normal distribution to demonstrate this idea,
since many people are familiar with the normal distribution.

Here is a plot of the standard normal distribution,
which has mean 0 and variance 1.
```{r}
x <- seq(from = -3,
         to = 3,
         length.out = 500)
y <- dnorm(x)
plot(x = x,
     y = y,
     ylab = "f(x)",
     type = "l")
```

The area under the entire curve is 1, since the probability
of a standard normal random variable taking on any value is 1.
```{r}
x <- seq(from = -3,
         to = 3,
         length.out = 500)
y <- dnorm(x)
plot(x = x,
     y = y,
     ylab = "f(x)",
     type = "l")
polygon(x = c(x,
              x[length(x)],
              x[1]),
        y = c(y,
              0,
              0),
        col = "blue")
```

The area under the curve for $x \le 0$ is 0.5, since the probability
of a standard normal random variable being less than or equal to 0 is 0.5.
```{r}
x <- seq(from = -3,
         to = 3,
         length.out = 500)
y <- dnorm(x)
plot(x = x,
     y = y,
     ylab = "f(x)",
     type = "l")
polygon(x = c(x[x <= 0],
              0,
              0,
              x[1]),
        y = c(y[x <= 0],
              dnorm(0),
              0,
              0),
        col = "blue")
```

The area under the curve for $x \ge 1.96$ or $x \le -1.96$ is 0.05, since the probability
of a standard normal random variable being greater than or equal to 1.96
or less than or equal to -1.96 is 0.05.
```{r}
x <- seq(from = -3,
         to = 3,
         length.out = 500)
y <- dnorm(x)
plot(x = x,
     y = y,
     ylab = "f(x)",
     type = "l")
polygon(x = c(x[x <= -1.96],
              -1.96,
              -1.96,
              x[1]),
        y = c(y[x <= -1.96],
              dnorm(-1.96),
              0,
              0),
        col = "blue")
polygon(x = c(x[x >= 1.96],
              x[length(x)],
              1.96,
              1.96),
        y = c(y[x >= 1.96],
              0,
              0,
              dnorm(1.96)),
        col = "blue")
```

What this implies is that values near 0 are more likely than values far
from 0,
because integrating under $f(x)$ near 0 produces a larger area
than integrating under $f(x)$ far from 0.



The normal distribution is a continuous distribution, but the principle
is the same for discrete distributions.
Let's consider the Bernoulli distribution.
In the plot below, I've drawn a Bernoulli distribution
where $P(X = 1) = 0.75$.
```{r}
x_B <- c(-0.5,-0.5,0.5,0.5,1.5,1.5)
y_B <- c(0,0.25,0.25,0.75,0.75,0)
plot(x = x_B,
     y = y_B,
     type = "l",
     xlab = "x",
     ylab = "P(X = x)")
segments(x0 = 0.5,
         y0 = 0,
         x1 = 0.5,
         y1 = 0.25)
abline(h = 0)
```

The area of the bar over $x = 1$ is equal to 0.75,
because $P(X = 1) = 0.75$.
```{r}
x_B <- c(-0.5,-0.5,0.5,0.5,1.5,1.5)
y_B <- c(0,0.25,0.25,0.75,0.75,0)
plot(x = x_B,
     y = y_B,
     type = "l",
     xlab = "x",
     ylab = "P(X = x)")
segments(x0 = 0.5,
         y0 = 0,
         x1 = 0.5,
         y1 = 0.25)
abline(h = 0)
polygon(x = c(0.5,0.5,1.5,1.5),
        y = c(0,0.75,0.75,0),
        col = "blue")
```

The area of the bar over $x = 0$ is equal to 0.25,
because $P(X = 0) = 0.25$.
```{r}
x_B <- c(-0.5,-0.5,0.5,0.5,1.5,1.5)
y_B <- c(0,0.25,0.25,0.75,0.75,0)
plot(x = x_B,
     y = y_B,
     type = "l",
     xlab = "x",
     ylab = "P(X = x)")
segments(x0 = 0.5,
         y0 = 0,
         x1 = 0.5,
         y1 = 0.25)
abline(h = 0)
polygon(x = c(-0.5,-0.5,0.5,0.5),
        y = c(0,0.25,0.25,0),
        col = "blue")
```



Why does any of this matter?
We often use probability distributions to represent real-world quantities.
It's important to be able to evaluate whether a distribution is an accurate
representation of a real-world quantity.
For example, suppose I said that I wanted to use this distribution
to represent $T$, the time from birth to death for a randomly-selected
American.
```{r}
temp <- function(x){
  dgamma(x = x,
         shape = 18,
         rate = 1)
}
curve(expr = temp,
      from = 0,
      to = 100,
      ylab = "f(x)")
```
Does that make sense? No!
It would mean that the chance of death is highest around age 18
and lowest around age 0 and after age 40.
That's not an accurate representation of the real world.

At this point, I should point out an exception to the rule stated above.
Although we often use area under the curve to represent probability,
there is one domain in which we commonly use the height of the curve
to represent probability.
That domain is survival analysis.
When the outcome is time to an event (e.g.,
heart attack, death, progression),
we often plot the probability of surviving past a certain time.
It looks something like this:
```{r}
temp <- function(x){
  1 - pexp(q = x)
}
curve(expr = temp,
      from = 0,
      to = 1,
      ylim = c(0,1),
      xlab = "t",
      ylab = "P(T > t)")
```

Sometimes we use $S(t)$ instead of $P(T > t)$, but the meaning is the same.
In this situation, the height of the curve is a probability,
not the area under the curve.
This is why it's always very important to read the axis labels on a plot,
and to include axis labels when creating a plot.

## Problems

### Problem 1

Would it make sense to use this distribution to represent time from birth
until death?
Why or why not?
```{r}
x <- seq(from = -3,
         to = 3,
         length.out = 500)
y <- 0.5 * exp(-abs(x))
plot(x = x,
     y = y,
     ylab = "f(x)",
     type = "l")
```

::: {.callout-tip title="Solution" collapse="true"}

It would not make sense to use this distribution to represent time from birth
until death.
Time from birth until death is positive,
but this distribution places positive probability over negative numbers.

:::

### Problem 2

Would it make sense to use this distribution to represent death
at 1 year post chemotherapy for breast cancer?
Why or why not?
```{r}
x <- seq(from = -3,
         to = 3,
         length.out = 500)
y <- 0.5 * exp(-abs(x))
plot(x = x,
     y = y,
     ylab = "f(x)",
     type = "l")
```

::: {.callout-tip title="Solution" collapse="true"}

It would not make sense.
Death at 1 year post chemotherapy for breast cancer can only take on
two values, death or no death,
whereas with this distribution, any number is possible.

:::




