---
title: "Expected Value"
execute:
  echo: false
format: html
---

## Learning Objective

Define, use, and interpret the term "expected value"
in the context of random variables.


## Lesson

### Notation

We use $E(X)$ to represent the expected value of $X$.
You might see "expectation" instead of "expected value".

### Visual Representation

The expected value of a random variable is where you would have to place
a fulcrum to get it to balance.
Let's look at a few examples.

This is a standard normal distribution.
It's expected value is 0.
```{r}
curve(expr = dnorm,
      from = -3,
      to = 3,
      xlab = "x",
      ylab = "f(x)",
      ylim = c(-0.05,0.4))
abline(h = 0)
polygon(x = c(0,0.2,-0.2),
        y = c(0,-0.05,-0.05),
        col = "black")
```

This is a uniform distribution from 0 to 1.
It's expected value is 0.5.
```{r}
plot(x = c(-0.5,0,0,1,1,1.5),
     y = c(0,0,1,1,0,0),
     type = "l",
     xlab = "x",
     ylab = "f(x)",
     ylim = c(-0.05,1))
abline(h = 0)
polygon(x = c(0.5,0.6,0.4),
        y = c(0,-0.05,-0.05),
        col = "black")
```

This is an exponential distribution with expected value equal to 1.
```{r}
curve(expr = dexp,
      from = 0,
      to = 6,
      xlab = "x",
      ylab = "f(x)",
      ylim = c(-0.05,1))
abline(h = 0,
       v = 0)
polygon(x = c(1,1.1,0.9),
        y = c(0,-0.05,-0.05),
        col = "black")
```

This is a Bernoulli distribution with $P(X = 1) = 0.75$.
It's expected value is 0.75.
```{r}
x_B <- c(-0.5,-0.5,0.5,0.5,1.5,1.5)
y_B <- c(0,0.25,0.25,0.75,0.75,0)
plot(x = x_B,
     y = y_B,
     type = "l",
     xlab = "x",
     ylab = "P(X = x)",
     ylim = c(-0.05,1))
segments(x0 = 0.5,
         y0 = 0,
         x1 = 0.5,
         y1 = 0.25)
abline(h = 0)
polygon(x = c(0.75,0.85,0.65),
        y = c(0,-0.05,-0.05),
        col = "black")
```
This example shows that the expected value of a random variable
doesn't need to be in its support.
$E(X)$ can be a value that $X$ can't take.


### Verbal Description

The expected value is a weighted average.
Each possible value of $X$ is weighted by its probability
(or density--a technical term that you don't have to memorize,
but that I can't, in good conscience, not use),
and then these weighted values are summed.


Consider this image.
```{r}
plot(x = c(-5,-5,-5,-3,-1,-1,2,2,2,2,2,3),
     y = c(0.1,0.3,0.5,0.1,0.1,0.3,0.1,0.3,0.5,0.7,0.9,0.1),
     pch = 19,
     ylim = c(-0.05,1),
     cex = 7,
     xlab = "x",
     ylab = "")
abline(h = 0)
```
I'm going to take each value and weight it by it's relative frequency.

| Value | Relative Frequency | Product |
|-:|-:|-:|
| -5 | 3/12 | -15/12 |
| -3 | 1/12 | -3/12 |
| -1 | 2/12 | -2/12 |
| 2 | 5/12 | 10/12 |
| 3 | 1/12 | 3/12 |

The sum of the weighted values is
```{r}
#| echo: true

sum(c(-15 / 12,
      -3 / 12,
      -2 / 12,
      10 / 12,
      3 / 12))
```
This is the expected value of this distribution.

### Math

For a continuous random variable, the expected value is
$$
E(X) = \int_{-\infty}^\infty x f(x)\,dx \text{.}
$$
For a discrete random variable, the expected value is
$$
E(X) = \sum_x x P(X = x) \text{.}
$$

### Example

Let's consider the simplest (non-degenerate) distribution possible,
the Bernoulli distribution.
Suppose $P(X = 1) = p$ and $P(X = 0) = 1 - p$, with $0 < p < 1$.
What is $E(X)$?
$$
\begin{align*}
E(X) &= \sum_x x P(X = x) \\
&= 0 P(X = 0) + 1 P(X = 1) \\
&= P(X = 1) \\
&= p
\end{align*}
$$
It may be counterintuitive that a probability could be an expected value,
but it's true.
If $X$ can only be 0 or 1,
then $E(X) = P(X = 1)$.

## Problems

### Problem 1

True or false:
The expected value is the most commonly occurring value in a data set.
That's why it's the "expected" value;
it's the one you most expect to see.

::: {.callout-tip title="Solution" collapse="true"}

False.
We reserve "expected value" for random variables,
not for data sets.
The most commonly occurring value in a data set is the mode
of that data set.
The expected value may not even be a value that the random variable can take.

:::

### Problem 2

True or false:
$P(X > E(X))$ is always 0.5.

::: {.callout-tip title="Solution" collapse="true"}

False. This is only true when $E(X)$ is the median of the distribution,
which is not always the case.

:::

### Problem 3

Suppose $Y$ is the (eventual) age at death for a (living) human being selected uniformly at random
from the earth in 2021.

#### Part (a)

A colleague wants to estimate $P(Y = y)$ for various values of $y$.
He finds this graph: [Probability of dying, by age, World, 2021](https://ourworldindata.org/grapher/probability-of-dying-by-age).
Explain to your colleague why that graph is or is not an estimate
of $P(Y = y)$.

::: {.callout-tip title="Solution" collapse="true"}

This graph is not an estimate of $P(Y = y)$,
for at least the following reasons:

- The area under the curve of a distribution needs to integrate or sum to 1.
This is the graphical analogue to the fact
that $P(Y = \text{anything}) = 1$.
In this graph, the bars  sum to a value greater than 1.
If you just add up the top six bars,
moving from the top of the page down,
you get 1.02, which is greater than 1.
- The subtitle of the graph is,
"The annual probability of death among people in a given age group, expressed as a fraction."
So, each bar represents the probability of death
for someone selected from a given age group;
for $Y$, we are sampling someone from the earth at large,
so they could be from any age group.

:::

#### Part (b)

Another colleague wants to estimate $E(Y)$
She finds this graph:
[Life expectancy at birth, total (years)](https://data.worldbank.org/indicator/SP.DYN.LE00.IN).
Explain to your colleague why that graph can or cannot
provide an estimate of $E(Y)$.

::: {.callout-tip title="Solution" collapse="true"}

This graph cannot be used to estimate $E(Y)$.
The title of the graph is "Life expectancy at birth",
so the population is restricted to newborns.
$Y$ is the life expectancy for someone
selected uniformly at random,
so the population includes anyone of any age.
$E(Y)$ should be higher than the life expectancy
for 2021 in this graph;
someone who has already lived to age 10
is going to have a higher life expectancy than
someone who hasn't yet reached age 1.

:::

#### Part (c)

What is the support of $Y$?
Justify your answer.

::: {.callout-tip title="Solution" collapse="true"}

I would say that the support of $Y$
is the interval $(0,\infty)$.
You could select someone of any age,
no matter how close to birth, which is why my interval
goes all the way down to zero.
I didn't include zero because no one can be alive
and dead at the same time.
My interval goes all the way up to infinity because,
even though no one has lived past 122
([Guinness](https://www.guinnessworldrecords.com/world-records/oldest-person)),
we don't know if someone born in 2021 will live longer.

By the way, here is an explanation of the notation I used above.

- $(a,b)$ is the line segment containing all points
between $a$ and $b$, but not $a$ and $b$ themselves.
- $[a,b]$ is the line segment containing all points
between $a$ and $b$, and also $a$ and $b$ themselves.
- $\{a,b\}$ is the set containing only the points
$a$ and $b$.

It's not important that you learn to use this notation for this class.
You can use words instead of symbols to communicate
the support of a random variable.

:::





