---
title: "Violated Assumptions of Linear Regression"
execute:
  echo: false
format: html
---

## Learning Objective

Identify which of the four assumptions of linear regression
are automatically violated when the outcome is Bernoulli.


## Lesson

If our outcome variable is Bernoulli,
some (but not all) of the assumptions of linear regression are automatically
violated.

### Linearity

The linearity assumption of linear regression is not automatically violated
when we have a binary outcome.
Consider this example:
I can simulate $X_1,\dotsc,X_{100}\overset{\text{iid}}{\sim}\text{uniform}(0,1)$
and then simulate
Bernoulli random variables $Y_1,\dotsc,Y_{100}$
such that $E(Y_i|X_i = x) = x$.
In this case, the mean of the outcome is a linear function
of the explanatory variable, which is exactly the linearity assumption.

```{r}
#| echo: true
set.seed(2)
n <- 1e2
x <- runif(n)
y <- rbinom(n = n,
            size = 1,
            prob = x)
mod_01 <- lm(y ~ x)
plot(x = x,
     y = y)
abline(mod_01,
       col = "blue")
summary(mod_01)
```


I admit the residual scatterplot looks weird, but I know from having simulated
the data that the linearity assumption is met.

```{r}
plot(x = x,
     y = mod_01$residuals,
     ylab = "Residuals")
abline(h = 0,
       col = "red")
```

### IID

There's nothing about a binary outcome that makes iid data impossible.
In fact, the type of outcome has nothing to do with whether the data
are iid or not.

### Homoscedasticity

Recall that if $Y\sim\text{Bernoulli}(p)$ then $E(Y) = p$ and $Var(Y) = p(1-p)$.
So, if $Y\in\{0,1\}$ and $E(Y|X)$ is a function of $X$,
then $Var(Y|X) = E(Y|X)\left(1 - E(Y|X)\right)$ will be a function of $X$.
This is a violation of homoscedasticity or equal variance.
If the variance depends on the explanatory variable(s),
then the variance is not constant.
The only way the homoscedasticity assumption would not be violated
is if $X$ was not actually random and could only take on one value,
or if $Y$ and $X$ had no relationship.

### Normality

If $Y$ is Bernoulli, it can only take on the values of 0 or 1.
If $Y = \beta_0 + \beta_1 X_{1} + \cdots + \beta_p X_{p} + \epsilon$,
$\epsilon$ can only take on two values.
Either
$$
\epsilon = 1 - \left(\beta_0 + \beta_1 X_{1} + \cdots + \beta_p X_{p}\right)
$$
or
$$
\epsilon = 0 - \left(\beta_0 + \beta_1 X_{1} + \cdots + \beta_p X_{p}\right) \text{.}
$$
So, obviously $\epsilon$ cannot be normal.



## Problems

### Problem 1

If you perform linear regression with a binary outcome,
are the coefficient estimates unbiased?

::: {.callout-tip title="Solution" collapse="true"}

This depends.
The linearity assumption would have to be met.
As seen in the residual scatterplot above,
this is kind of difficult to check with a residual scatterplot.
You could try checking this with a calibration plot,
which we'll talk about in a later section.

:::

### Problem 2

If you perform linear regression with a binary outcome,
are the standard errors correct?

::: {.callout-tip title="Solution" collapse="true"}

No the default standard errors, because the homoscedasticity assumption is violated.
You could use robust/sandwich standard errors instead.

:::

### Problem 3

If you perform linear regression with a binary outcome,
are the p-values correct?

::: {.callout-tip title="Solution" collapse="true"}

Not the default p-values, because they're computed assuming homoscedasticity,
which is violated.
If you used robust/sandwich standard errors and had a large sample size
(and if the linearity and iid assumptions were met),
you could trust the p-values.

:::




