---
title: "Assumptions of Logistic Regression"
execute:
  echo: false
format: html
---

## Learning Objective

Explain the assumptions of logistic regression.


## Lesson

### Assumptions

Logistic regression has the following assumptions:

1. $Y_i|X_{i1},\dotsc,X_{ip}\sim\text{Bernoulli}(p_i)$.
2. $\log\frac{p_i}{1-p_i} = \beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip}$.
3. The observations are iid.

We've already talked about the Bernoulli distribution
and what it means for observations to be iid,
but the weird function merits some unpacking.

### Logit Function

The function
$$
f(p) = \log\frac{p}{1-p}
$$
is called the logit or log-odds function.
A graph is below.

```{r}
curve(expr = qlogis,
      from = 0.001,
      to = 0.999,
      xlab = "p",
      ylab = "logit(p)")
abline(h = 0,
       v = 0)
```

It's probably more illuminating if we look at the inverse function,
known as the "expit" function.

### Expit Function

The expit function is the inverse of the logit function:
$$
g(q) = \frac{\exp(q)}{1 + \exp(q)} \text{.}
$$
Here's the work if you're curious:
$$
\begin{align*}
q &= \log \frac{p}{1-p} \\
\exp(q) &= \frac{p}{1-p} \\
\exp(q) (1-p) &= p \\
\exp(q) - p \exp(q) &= p \\
\exp(q) &= p\exp(q) + p \\
\exp(q) &= p\left(\exp(q) + 1\right) \\
\frac{\exp(q)}{1 + \exp(q)} &= p \text{.}
\end{align*}
$$

You might also see the expit function written in a different but equivalent
manner:
$$
\begin{align*}
g(q) &= \frac{\exp(q)}{1 + \exp(q)} \\
&= \frac{\exp(q)}{1 + \exp(q)} \frac{\exp(-q)}{\exp(-q)} \\
&= \frac{1}{\exp(-q) + 1} \text{.}
\end{align*}
$$

Again, we can look at a graph:
```{r}
curve(expr = plogis,
      from = -6,
      to = 6,
      xlab = "q",
      ylab = "expit(q)")
```

So, the assumption
$$
\log\frac{p_i}{1-p_i} = \beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip}
$$
can also be written as
$$
p_i = \frac{1}{1 + \exp\left(-\left(\beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip}\right)\right)} \text{.}
$$

Why do we do this?
There are a few reasons.
First, note that the expit function can take in any value
between $-\infty$ and $+\infty$ (exclusive).
This means that
$\beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip}$
can be any number, positive or negative.
Second, the output of the expit function is between 0 and 1.

If we were to use linear regression
and assume $E(Y_i|X_{i1},\dotsc,X_{ip}) = \beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip}$,
we might find that
$\hat{Y}_i = \hat{\beta}_0 + \hat{\beta}_1 X_{i1} + \cdots + \hat{\beta}_p X_{ip}$
is below zero or above one,
which doesn't make sense,
since $\hat{Y}_i = \hat{E}(Y_i|X_{i1},\dotsc,X_{ip})$ should be a probability.

There are other reasons to use this weird function,
which we will get to later.

## Problems

### Problem 1

In the logit function, what is the base of the logarithm?

::: {.callout-tip title="Solution" collapse="true"}

The special number $e$.
The logarithm with base $e$ is sometimes called the natural logarithm.

:::

### Problem 2

What are the possible values the logit function can produce?
You might remember this as the "range" of the function.

::: {.callout-tip title="Solution" collapse="true"}

Any real number, i.e., any positive or negative number,
even those that can't be represented by fractions.

:::




