---
title: "Interpretation"
execute:
  echo: false
format: html
---

## Learning Objective

Interpret the coefficients in logistic regression.


## Lesson

### Intercept

Recall that, in logistic regression, we assume
$$
\log\frac{p_i}{1-p_i} = \beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip} \text{.}
$$

What happens when all the covariates equal zero?
$$
\begin{align*}
\log\frac{p_i}{1-p_i} &= \beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip} \\
\log\frac{p_i}{1-p_i} &= \beta_0 + \beta_1 (0) + \cdots + \beta_p (0) \\
\log\frac{p_i}{1-p_i} &= \beta_0 \\
\frac{p_i}{1-p_i} &= \exp(\beta_0) \\
\end{align*}
$$

So, $\exp(\beta_0)$ is the odds of the outcome being 1 when all the covariates
equal zero.

### Non-Intercept Coefficients

How do we interpret the non-intercept coefficients?
Suppose observation $i$ has $X_{i1} = x$ and observation $j$ has $X_{j1} = x + 1$.
For each $k \ne 1$, $X_{ik} = X_{jk}$.
In words, the two observations are identical except for the first covariate,
where the $j$th observation exceeds the $i$th observation by one unit.
Then
$$
\begin{align*}
\log \frac{\frac{p_j}{1-p_j}}{\frac{p_i}{1-p_i}} &= \log \frac{p_j}{1-p_j} - \log \frac{p_i}{1-p_i} \\
&= \beta_0 + \beta_1 X_{j1} + \cdots + \beta_p X_{jp} - \left(\beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip}\right) \\
&= \beta_0 + \beta_1 (x+1) + \cdots + \beta_p X_{jp} - \left(\beta_0 + \beta_1 x + \cdots + \beta_p X_{ip}\right) \\
&= \beta_0 + \beta_1 x + \beta_1 + \cdots + \beta_p X_{jp} - \left(\beta_0 + \beta_1 x + \cdots + \beta_p X_{ip}\right) \\
&= \beta_1 \text{.}
\end{align*}
$$
By exponentiating both sides of this equation, we see
$$
\begin{align*}
\log \frac{\frac{p_j}{1-p_j}}{\frac{p_i}{1-p_i}} &= \beta_1 \\
\frac{\frac{p_j}{1-p_j}}{\frac{p_i}{1-p_i}} &= \exp(\beta_1) \text{.} \\
\end{align*}
$$

In words, $\exp(\beta_1)$ is the odds ratio comparing two populations
that are identical except that one exceeds the other by one unit
on the first covariate.
If $\exp(\beta_1) > 1$,
the odds that the outcome equals 1 is greater in the population
with the higher value of the first covariate;
if $\exp(\beta_1) < 1$,
the odds that the outcome equals 1 is lesser in the population
with the higher value of the first covariate;
if $\exp(\beta_1) = 1$,
the odds that the outcome equals 1 is the same in both populations.

### Example

```{r}
d <- Lock5Data::NutritionStudy
d$SmokeYes <- ifelse(test = d$Smoke == "Yes",
                     yes = 1,
                     no = 0)
d$YearsSince40 <- d$Age - 40
```

In this data set, "Study subjects were patients who had an elective surgical
procedure during a three-year period to biopsy or
remove a lesion of the lung, colon, breast, skin,
ovary or uterus that was found to be non-cancerous."
([Lock5Data Package](https://cran.r-project.org/web/packages/Lock5Data/Lock5Data.pdf))
Let's regress smoking on years since age 40,
vitamin use (no, occasional, or regular),
and sex (male or female).
I'm using "years since age 40" instead of age directly
so we can get a more meaningful intercept

```{r}
#| echo: true
mod_01 <- glm(formula = SmokeYes ~ YearsSince40 + VitaminUse + Sex,
              family = binomial,
              data = d)
summary(mod_01)
```

How do we interpret these coefficients?
They're actually meaningless in their current form.
We need to exponentiate them first.

```{r}
#| echo: true
exp(mod_01$coefficients)
```

Now we can interpret them.

- The odds of smoking among 40-year-old females who don't use vitamins
is `r exp(mod_01$coefficients[1])`.
This corresponds to a probability of
```{r}
#| echo: true
exp(mod_01$coefficients[1]) / (1 + exp(mod_01$coefficients[1]))
```

- The odds of smoking among people age $x + 1$ is 4% lower
than the odds of smoking among people age $x$,
holding vitamin use and sex constant.
- The odds of smoking among people who occasionally use vitamins
is 36% lower than the odds of smoking among people who don't use vitamins,
holding age and sex constant.
- The odds of smoking among people who regularly use vitamins
is 62% lower than the odds of smoking among people who don't use vitamins,
holding age and sex constant.
- The odds of smoking among males is 95% higher
than the odds of smoking among females,
holding age and vitamin use constant.

### Note

We don't have to explicitly state that we're talking about expected values
or means.
When the outcome is a 1 or 0,
the expected value is a probability,
so just by talking about probabilities and odds,
we know we're talking about expected values.


## Problems

### Problem 1

```{r}
d2 <- Lock5Data::NHANES
d2$OrganicYes <- ifelse(test = d2$Organic == "Yes",
                        yes = 1,
                        no = 0)
d2$Income1000 <- d2$Income / 1e3
```

As described in the [Lock5Data package manual](https://cran.r-project.org/web/packages/Lock5Data/Lock5Data.pdf)
"This dataset is a subset of the 2009-2010 National Health and Nutrition Examination Survey (NHANES).
NHANES is a national survey conducted by the Centers for Disease Control and Prevention (CDC)
on a random sample of Americans. This subset contains data on select variables for the subset of
people with responses to the questions about buying organic food and self-reported health status."

The outcome is a 1 if the respondent bought any food labeled organic
in the past 30 days,
and 0 otherwise.
I've regressed this on self-rated health (excellent, very good,
good, fair, and poor)
and monthly income (in thousands of dollars).

```{r}
mod_02 <- glm(formula = OrganicYes ~ Health + Income1000,
              family = binomial,
              data = d2)
summary(mod_02)
```

Interpret the exponentiated coefficients.
```{r}
#| echo: true
exp(mod_02$coefficients)
```

::: {.callout-tip title="Solution" collapse="true"}

- The odds of having bought organic in the past 30 days
among people who make \$0/month and rate their own health as excellent
is `r exp(mod_02$coefficients[1])`.
- The odds of having bought organic in the past 30 days
among people who rate their health as fair
is 52% lower than the odds of having bought organic in the past 30 days
among people who rate their health as excellent,
holding income constant.
- The odds of having bought organic in the past 30 days
among people who rate their health as good
is 27% lower than the odds of having bought organic in the past 30 days
among people who rate their health as excellent,
holding income constant.
- The odds of having bought organic in the past 30 days
among people who rate their health as poor
is 44% lower than the odds of having bought organic in the past 30 days
among people who rate their health as excellent,
holding income constant.
- The odds of having bought organic in the past 30 days
among people who rate their health as very good
is 6% lower than the odds of having bought organic in the past 30 days
among people who rate their health as excellent,
holding income constant.
- The odds of having bought organic in the past 30 days
among people who make $x + 1000$ dollars per month
is 15% greater than the odds of having bought organice in the past 30 days
among people who make $x$ dollars per month,
holding self-rated health constant.


:::

### Problem 2

```{r}
d3 <- Lock5Data::ICUAdmissions
d3$Died <- d3$Status
d3$YearsSince20 <- d3$Age - 20
```

Let's return to our ICU admissions data set.
Our outcome will be 1 if the patient died and 0 otherwise.
We'll use two covariates:
years since age 20
and whether or not the patient got CPR prior to admission (1 = yes, 0 = no).

```{r}
mod_03 <- glm(formula = Died ~ YearsSince20 + CPR,
              family = binomial,
              data = d3)
summary(mod_03)
```

Intepret the exponentiated coefficients.
```{r}
#| echo: true
exp(mod_03$coefficients)
```

::: {.callout-tip title="Solution" collapse="true"}

- The odds of death for a 20-year-old who did not receive CPR prior to admission
to the ICU is 0.06.
- The odds of death for someone age $x + 1$ is 3% greater than
the odds of death for someone age $x$, holding CPR status constant.
- People who received CPR prior to admission have 5.95 times the odds
of death as people who did not receive CPR prior to admission,
holding age constant.

:::




