---
title: "Calibration"
execute:
  echo: false
format: html
---

## Learning Objective

- Explain calibration in logistic regression.
- Interpret a calibration plot.


## Lesson

### Assumption-Checking

How do we check the assumptions of logistic regression?

1. $Y_i|X_{i1},\dotsc,X_{ip}\sim\text{Bernoulli}(p_i)$.
2. $\log\frac{p_i}{1-p_i} = \beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip}$.
3. The observations are iid.

One of them is easy to check:
We just have to look at the outcome and make sure it's binary.
One of them we've checked before:
We can look at how the data were collected to determine if there
is any dependency between observations.
But what about that weird logistic function?

One way to check the second assumption above is to see if the model
is well-calibrated.
In a well-calibrated model, the fitted values are close to the observed
values.
But how do we know if this is the case?
The fitted values will be probabilities,
but the observed values are 0s and 1s

We can group observations with similar fitted values together,
then compute the proportion of 1s in each group.
The observed proportion of 1s should be close to the
median fitted value in the group.
If we do this, and plot the observed proportion as a function
of the median fitted value, we get a calibration plot.

### Calibration Plot

```{r}
d <- Lock5Data::NutritionStudy
d$SmokeYes <- ifelse(test = d$Smoke == "Yes",
                     yes = 1,
                     no = 0)
d$YearsSince40 <- d$Age - 40
```

Let's walk through an example.
Let's return to the nutrition study,
where we regressed smoking on years since age 40,
vitamin use (no, occasional, or regular),
and sex (male or female).

```{r}
#| echo: true
mod_01 <- glm(formula = SmokeYes ~ YearsSince40 + VitaminUse + Sex,
              family = binomial,
              data = d)
summary(mod_01)
```

The fitted values are obtained like so:
$$
\hat{Y}_i = \frac{1}{1 + \exp\left(-\left(\hat{\beta}_0 + \hat{\beta}_1 X_{i1} + \cdots + \hat{\beta}_p X_{ip}\right)\right)} \text{.}
$$
Of course, we don't have to do that computation by hand.
We can do either of the following:
```{r}
#| echo: true
y_hat_option_1 <- mod_01$fitted.values
y_hat_option_2 <- predict(object = mod_01,
                          newdata = d,
                          type = "response")
all.equal(target = y_hat_option_1,
          current = y_hat_option_2)
```

I'm going to make a new data frame that contains two columns:
the fitted values and the observed outcomes.
```{r}
#| echo: true
cal_data <- data.frame(truth = mod_01$model$SmokeYes,
                       estimate = mod_01$fitted.values)
```

Next, I'm going to sort this data set from lowest to highest fitted value.
```{r}
#| echo: true
cal_data <- cal_data[order(cal_data$estimate), ]
```

Now I'm going to put the rows into ten groups with approximately the same
number of observations.
The number ten is arbitrary;
you can choose a different number for your own data in the future.
```{r}
#| echo: true
cal_bins <- list()
n_bins <- 10
n_obs_per_bin <- floor(nrow(cal_data) / n_bins)
for(i in 1:(n_bins - 1)){
  cal_bins[[i]] <- cal_data[((i - 1) * n_obs_per_bin + 1):(i * n_obs_per_bin), ]
}
cal_bins[[n_bins]] <- cal_data[((n_bins - 1) * n_obs_per_bin + 1):nrow(cal_data), ]
```

Now I'm going to compute the proportion of 1s in each bin
and the median fitted value in each bin.
```{r}
#| echo: true
cal_plot_data <- matrix(data = as.numeric(NA),
                        nrow = n_bins,
                        ncol = 2)
for(i in 1:n_bins){
  cal_plot_data[i,1] <- median(cal_bins[[i]]$estimate)
  cal_plot_data[i,2] <- mean(cal_bins[[i]]$truth)
}
```

Finally, I can plot my results.
```{r}
#| echo: true
plot(x = cal_plot_data[ ,1],
     y = cal_plot_data[ ,2],
     type = "b",
     main = "Calibration Plot",
     xlab = "Median Fitted Value in Bin",
     ylab = "Proportion of 1s in Bin",
     pch = 19)
grid()
abline(a = 0,
       b = 1,
       lty = 2)
```

In a perfectly-calibrated model, the points would fall on the dashed line
because the proportion of 1s in each bin would be exactly equal
to the median fitted value in each bin.
For the most part, the dots on this plot follow the reference line,
with exception of the sixth bin.
I'm not sure what's going on there.
I'm going to try five bins instead of ten.
I've written a function to automate the process.

```{r}
#| echo: true
#' Make a calibration plot.
#' 
#' @param mod The logistic regression model, obtained using the glm function.
#' @param n_bins The number of bins to use.
#' @returns Doesn't return anything; draws a plot.
draw_cal_plot <- function(mod,n_bins){
  cal_data <- data.frame(truth = mod$y,
                         estimate = mod$fitted.values)
  cal_data <- cal_data[order(cal_data$estimate), ]
  cal_bins <- list()
  n_obs_per_bin <- floor(nrow(cal_data) / n_bins)
  for(i in 1:(n_bins - 1)){
    cal_bins[[i]] <- cal_data[((i - 1) * n_obs_per_bin + 1):(i * n_obs_per_bin), ]
  }
  cal_bins[[n_bins]] <- cal_data[((n_bins - 1) * n_obs_per_bin + 1):nrow(cal_data), ]
  cal_plot_data <- matrix(data = as.numeric(NA),
                          nrow = n_bins,
                          ncol = 2)
  for(i in 1:n_bins){
    cal_plot_data[i,1] <- median(cal_bins[[i]]$estimate)
    cal_plot_data[i,2] <- mean(cal_bins[[i]]$truth)
  }
  plot(x = cal_plot_data[ ,1],
       y = cal_plot_data[ ,2],
       type = "b",
       main = "Calibration Plot",
       xlab = "Median Fitted Value in Bin",
       ylab = "Proportion of 1s in Bin",
       pch = 19)
  grid()
  abline(a = 0,
         b = 1,
         lty = 2)
}
```

```{r}
#| echo: true
draw_cal_plot(mod = mod_01,
              n_bins = 5)
```

The bump in the plot doesn't smooth out when using fewer bins.
This tells us that when the model predicts a probability between 0.1 and 0.15,
it's an underestimate.
We should probably find a different model with better calibration.

### Other Plotting Functions

There are packages that have built-in functions for drawing calibration plots.
I'll demonstrate two here.

The first package is `caret`, which has a function `calibration` that you
can use.

```{r}
#| echo: true
library(caret)
cal_obj <- calibration(x = as.factor(d$SmokeYes) ~ mod_01$fitted.values,
                       class = "1")
plot(cal_obj)
```

You can see that this function plots a line all the way above 90%,
even though there aren't any observations with fitted values above 90%.
So, it's a little misleading.

Another function is `cal_plot_breaks` in the package `probably`.


```{r}
#| echo: true
library(probably)
cal_plot_breaks(.data = data.frame(truth = as.factor(mod_01$model$SmokeYes),
                                   estimate = mod_01$fitted.values),
                truth = truth,
                estimate = estimate,
                event_level = "second")
```
Whereas I made bins with equal numbers of observations in each,
this function makes bins that are of equal width on the horizontal axis.
You can see here that the fourth bin has fewer observations
than the other bins,
because the confidence band around its point on the graph is way larger
than the confidence band around other points.
The tick marks on the top of the plot denote observed 1s;
the tick marks on the bottom of the plot denote observed 0s.

### Simulation

I'm going to simulate data
to see a calibration plot for a model where the assumptions
are known to be met,
and another plot for a model where the assumptions are known to not be met.

```{r}
# Set seed
set.seed(6)

# Define parameters
n <- 1e3
beta_0 <- 1
beta_1 <- 2
beta_2 <- 1

# Generate x
x <- rnorm(n = n)
```

I'm going to generate the outcome according to the model
$$
P(Y_i = 1|X_i) = \frac{1}{1 + \exp\left(-\left(\beta_0 + \beta_1 X_i + \beta_2 X_i^2\right)\right)} \text{.}
$$

```{r}
# Generate y
y <- rbinom(n = n,
            size = 1,
            prob = plogis(beta_0 + beta_1 * x + beta_2 * x^2))
```

Next, I'm going to run logistic regression according to the true model
and make a calibration plot.

```{r}
#| echo: true
mod_02 <- glm(formula = y ~ x + I(x^2),
              family = binomial)
summary(mod_02)
draw_cal_plot(mod = mod_02,
              n_bins = 10)
cal_plot_breaks(.data = data.frame(truth = as.factor(mod_02$y),
                                   estimate = mod_02$fitted.values),
                truth = truth,
                estimate = estimate,
                event_level = "second")
```

This model is well-calibrated.
For the most part, the points follow the reference line.

What if our assumption is bad? Let's run a regression without the squared term.

```{r}
#| echo: true
mod_03 <- glm(formula = y ~ x,
              family = binomial)
summary(mod_03)
draw_cal_plot(mod = mod_03,
              n_bins = 10)
cal_plot_breaks(.data = data.frame(truth = as.factor(mod_03$y),
                                   estimate = mod_03$fitted.values),
                truth = truth,
                estimate = estimate,
                event_level = "second")
```

This very clearly does not follow the reference line.
What it's saying is that 100% of observations with fitted values between
0.2 and 0.3 have an outcome of 1. So, our fitted values are very inaccurate!
Somewhere between 80% and 90% of the observations with fitted values
between 0.3 and 0.4 have an outcome of 1.
Our predictions are only accurate for fitted values between about
0.8 and 1.0.

## Problems

### Problem 1

```{r}
d2 <- Lock5Data::NHANES
d2$OrganicYes <- ifelse(test = d2$Organic == "Yes",
                        yes = 1,
                        no = 0)
d2$Income1000 <- d2$Income / 1e3
```

Let's return to the NHANES data set.
The outcome is a 1 if the respondent bought any food labeled organic
in the past 30 days,
and 0 otherwise.
I've regressed this on self-rated health (excellent, very good,
good, fair, and poor)
and monthly income (in thousands of dollars).

```{r}
mod_02 <- glm(formula = OrganicYes ~ Health + Income1000,
              family = binomial,
              data = d2)
summary(mod_02)
```

Interpret the calibration plot.
```{r}
draw_cal_plot(mod = mod_02,
              n_bins = 10)
```

::: {.callout-tip title="Solution" collapse="true"}

This looks great!
The points follow the reference line almost exactly
for fitted values between 0.2 and 0.35 or so.
There are a couple deviations from the fitted line
for fitted values between 0.35 and 0.45,
but these deviations are not big.
This model is well-calibrated.


:::

### Problem 2

```{r}
d3 <- Lock5Data::ICUAdmissions
d3$Died <- d3$Status
d3$YearsSince20 <- d3$Age - 20
```

Let's return to our ICU admissions data set.
Our outcome will be 1 if the patient died and 0 otherwise.
We'll use two covariates:
years since age 20
and whether or not the patient got CPR prior to admission (1 = yes, 0 = no).

```{r}
mod_03 <- glm(formula = Died ~ YearsSince20 + CPR,
              family = binomial,
              data = d3)
summary(mod_03)
```

Interpret the calibration plot.
```{r}
draw_cal_plot(mod = mod_03,
              n_bins = 10)
```

::: {.callout-tip title="Solution" collapse="true"}

This looks pretty good. The points bounce around the line,
but in general, they follow it.

:::




