---
title: "Calibration"
execute:
  echo: false
format: html
---

## Learning Objective

- Explain calibration.
- Interpret a calibration plot.


## Lesson

### Assumption-Checking

How do we check the assumptions of logistic regression?
As a reminder, here they are:

1. $Y_i|X_{i1},\dotsc,X_{ip}\sim\text{Bernoulli}(p_i)$.
2. $\log\frac{p_i}{1-p_i} = \beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip}$.
3. The observations are iid.

The first one is easy to check:
We just have to look at the outcome and make sure it's binary.
The third one we've checked before:
We can look at how the data were collected to determine if there
is any dependency between observations.
But what about that weird logistic function?

One way to check the second assumption above is to see if the model
is well-calibrated.
In a well-calibrated model, the fitted values are close to the observed
values.
But how do we know if this is the case?
The fitted values will be probabilities,
but the observed values are 0s and 1s.
I think the best way to understand the procedure is to walk through the creation
of a calibration plot.

### Calibration Plot

```{r}
d <- Lock5Data::NutritionStudy
d$SmokeYes <- ifelse(test = d$Smoke == "Yes",
                     yes = 1,
                     no = 0)
d$YearsSince40 <- d$Age - 40
```

Let's return to the nutrition study,
where we regressed smoking on years since age 40,
vitamin use (no, occasional, or regular),
and sex (male or female).

```{r}
#| echo: true
mod_01 <- glm(formula = SmokeYes ~ YearsSince40 + VitaminUse + Sex,
              family = binomial,
              data = d)
summary(mod_01)
```

This is a calibration plot for this model.
It was created using the function `cal_plot_breaks` from the package `probably`.
```{r}
#| echo: true
library(probably)
cal_plot_breaks(.data = data.frame(truth = as.factor(mod_01$model$SmokeYes),
                                   estimate = mod_01$fitted.values),
                truth = truth,
                estimate = estimate,
                event_level = "second")
```

How was this plot made? These are the steps:

1. R split the interval from 0 to 1 into ten segments of equal length:
0 to 0.1, 0.1 to 0.2, 0.2 to 0.3, and so forth.
Ten is the default number of segments, but you can change it
by passing a different number to the `num_breaks` parameter in the function.
2. R put observations with similar fitted values into bins.
Any observations with fitted values between 0 and 0.1 were put in the same bin;
any observations with fitted values between 0.1 and 0.2 were put in the same bin;
any observations with fitted values between 0.2 and 0.3 were put in the same bin;
and so forth.
3. R computed the proportion of observations in each bin that had an observed
outcome of 1.
4. R plotted the bin midpoints (0.05, 0.15, 0.25, etc.) on the horizontal
axis and the observed proportions of 1s in the bins on the vertical axis.
This function uses the term "Event Rate" to mean the observed proportion of 1s.
5. R added confidence bands around the points on the plot.
6. R added tick marks to the bottom of the plot to indicate 0s
and tick marks at the top of the plot to indicate 1s.
The locations of the tick marks along the horizontal axis
correspond to the fitted values of the corresponding observations.

You might be wondering where R got the fitted values.
This is the formula
$$
\hat{Y}_i = \frac{1}{1 + \exp\left(-\left(\hat{\beta}_0 + \hat{\beta}_1 X_{i1} + \cdots + \hat{\beta}_p X_{ip}\right)\right)} \text{.}
$$
If you want to see the fitted values,
you don't have to do that computation by hand.
You can do either of the following:
```{r}
#| echo: true
y_hat_option_1 <- mod_01$fitted.values
y_hat_option_2 <- predict(object = mod_01,
                          newdata = d,
                          type = "response")
all.equal(target = y_hat_option_1,
          current = y_hat_option_2)
```

You're probably also wondering how to interpret the plot.
If the fitted values accurately reflect the true underlying probabilities,
then the observed proportion of 1s in each bin should be approximately
equal to the bin midpoint.
In that case, the points on the plot would fall along the dashed reference line.
In the plot above, we see a deviation from the reference line
for the bin midpoint of 0.35.
However, the confidence band around that point contains the reference line.
In fact, the confidence band contains the reference line throughout
the graph.
So, this model is well-calibrated.

Why is the confidence band larger for the bin midpoint of 0.35
than for other bin midpoints?
I can think of two reasons.
First,
there are probably fewer observations with fitted values between 0.3 and 0.4
than in other bins.
We can check that.
```{r}
#| echo: true
sum(y_hat_option_1 < 0.1)
sum(y_hat_option_1 >= 0.1 & y_hat_option_1 < 0.2)
sum(y_hat_option_1 >= 0.2 & y_hat_option_1 < 0.3)
sum(y_hat_option_1 >= 0.3 & y_hat_option_1 < 0.4)
```
We see that there are only 9 observations with fitted values
between 0.3 and 0.4,
but 124 observations with fitted values
between 0 and 0.1.
With fewer observations, the confidence band gets wider.

Second, recall that Bernoulli random variables have their maximum
variance when $p = 0.5$.
So, as the observed proportion gets closer to 0.5,
the variability of the underlying data is higher,
and we have to make the confidence band wider to account for that.

### Simulation

I'm going to simulate data
to see a calibration plot for a model where the assumptions
are known to be met,
and another plot for a model where the assumptions are known to not be met.

```{r}
# Set seed
set.seed(6)

# Define parameters
n <- 1e3
beta_0 <- 1
beta_1 <- 2
beta_2 <- 1

# Generate x
x <- rnorm(n = n)
```

I'm going to generate the outcome according to the model
$$
P(Y_i = 1|X_i) = \frac{1}{1 + \exp\left(-\left(\beta_0 + \beta_1 X_i + \beta_2 X_i^2\right)\right)} \text{.}
$$

```{r}
# Generate y
y <- rbinom(n = n,
            size = 1,
            prob = plogis(beta_0 + beta_1 * x + beta_2 * x^2))
```

Next, I'm going to run logistic regression according to the true model
and make a calibration plot.

```{r}
#| echo: true
mod_02 <- glm(formula = y ~ x + I(x^2),
              family = binomial)
summary(mod_02)
cal_plot_breaks(.data = data.frame(truth = as.factor(mod_02$y),
                                   estimate = mod_02$fitted.values),
                truth = truth,
                estimate = estimate,
                event_level = "second")
```

This model is well-calibrated.
The confidence band contains the reference line for all points.

What if our assumption is bad? Let's run a regression without the squared term.

```{r}
#| echo: true
mod_03 <- glm(formula = y ~ x,
              family = binomial)
summary(mod_03)
cal_plot_breaks(.data = data.frame(truth = as.factor(mod_03$y),
                                   estimate = mod_03$fitted.values),
                truth = truth,
                estimate = estimate,
                event_level = "second")
```

This very clearly does not follow the reference line.
What it's saying is that 100% of observations with fitted values between
0 and 0.3 have an outcome of 1. So, our fitted values are very inaccurate!
Our predictions are only accurate for fitted values between about
0.8 and 1.0.

## Problems

### Problem 1

```{r}
d2 <- Lock5Data::NHANES
d2$OrganicYes <- ifelse(test = d2$Organic == "Yes",
                        yes = 1,
                        no = 0)
d2$Income1000 <- d2$Income / 1e3
```

Let's return to the NHANES data set.
The outcome is a 1 if the respondent bought any food labeled organic
in the past 30 days,
and 0 otherwise.
I've regressed this on self-rated health (excellent, very good,
good, fair, and poor)
and monthly income (in thousands of dollars).

```{r}
mod_04 <- glm(formula = OrganicYes ~ Health + Income1000,
              family = binomial,
              data = d2)
summary(mod_04)
```


```{r}
cal_plot_breaks(.data = data.frame(truth = as.factor(mod_04$y),
                                   estimate = mod_04$fitted.values),
                truth = truth,
                estimate = estimate,
                event_level = "second")
```

Is this model well-calibrated? Justify your answer.

::: {.callout-tip title="Solution" collapse="true"}

This looks great, except for the observations with the lowest fitted values.
What's going on?

If we look at the fitted values less than 0.2,
we see that the smallest fitted value is actually 0.1880,
which is greater than the bin midpoint of 0.15.

```{r}
summary(mod_04$fitted.values[mod_04$fitted.values < 0.2])
```

Below, I've written a function that will do most of what
`cal_plot_breaks` does,
but use the average of the fitted values in each bin
rather than the bin midpoint.

```{r}
#| echo: true
#' Make a calibration plot.
#'   
#' @param mod The logistic regression model, obtained using the glm function.
#' @param n_bins The number of bins to use.
#' @returns Doesn't return anything; draws a plot.
draw_cal_plot <- function(mod,n_bins){
  cal_data <- data.frame(truth = mod$y,
                         estimate = mod$fitted.values)
  cal_bins <- list()
  for(i in 1:(n_bins - 1)){
    cal_bins[[i]] <- cal_data[cal_data$estimate >= (i - 1) / n_bins & cal_data$estimate < i / n_bins, ]
  }
  cal_bins[[n_bins]] <- cal_data[cal_data$estimate >= (n_bins - 1) / n_bins, ]
  cal_plot_data <- matrix(data = as.numeric(NA),
                          nrow = n_bins,
                          ncol = 4)
  for(i in 1:n_bins){
    p_hat <- mean(cal_bins[[i]]$truth)
    n <- nrow(cal_bins[[i]])
    cal_plot_data[i,1] <- mean(cal_bins[[i]]$estimate)
    cal_plot_data[i,2] <- p_hat
    cal_plot_data[i,3] <- p_hat - qnorm(0.95) * sqrt(p_hat * (1 - p_hat) / n)
    cal_plot_data[i,4] <- p_hat + qnorm(0.95) * sqrt(p_hat * (1 - p_hat) / n)
  }
  plot(x = cal_plot_data[ ,1],
       y = cal_plot_data[ ,2],
       type = "b",
       main = "Calibration Plot",
       xlab = "Average Fitted Value in Bin",
       ylab = "Proportion of 1s in Bin",
       pch = 19,
       ylim = c(min(cal_plot_data,
                    na.rm = TRUE),
                max(cal_plot_data,
                    na.rm = TRUE)))
  grid()
  lines(x = cal_plot_data[ ,1],
        y = cal_plot_data[ ,3],
        lty = 3)
  lines(x = cal_plot_data[ ,1],
        y = cal_plot_data[ ,4],
        lty = 3)
  abline(a = 0,
         b = 1,
         lty = 2)
}
```

Now we can apply my function to this model.

```{r}
#| echo: true
draw_cal_plot(mod = mod_04,
              n_bins = 10)
```

My plot is not very pretty,
but even for the smallest fitted values, the reference line is contained
within the confidence band (between the dotted lines).
So, this model is well-calibrated.

:::

### Problem 2

```{r}
d3 <- Lock5Data::ICUAdmissions
d3$Died <- d3$Status
d3$YearsSince20 <- d3$Age - 20
```

Let's return to our ICU admissions data set.
Our outcome will be 1 if the patient died and 0 otherwise.
We'll use two covariates:
years since age 20
and whether or not the patient got CPR prior to admission (1 = yes, 0 = no).

```{r}
mod_05 <- glm(formula = Died ~ YearsSince20 + CPR,
              family = binomial,
              data = d3)
summary(mod_05)
```


```{r}
cal_plot_breaks(.data = data.frame(truth = as.factor(mod_05$y),
                                   estimate = mod_05$fitted.values),
                truth = truth,
                estimate = estimate,
                event_level = "second")
```

Is this model well-calibrated? Justify your answer.

::: {.callout-tip title="Solution" collapse="true"}

This model is well-calibrated.
The confidence band contains the reference line.

:::

### Problem 3

Some courts use algorithms to estimate a criminal defendant's probability
of re-offending. The scores generated from these algorithms--sometimes
on a scale of 1 to 10--are used at various points, including to determine
bond amounts and sentences. Much has been written about these algorithms
(see, for example,
[Machine Bias](https://www.propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing)
and
[Can Racist Algorithms Be Fixed?](https://www.themarshallproject.org/2019/07/01/can-racist-algorithms-be-fixed)).
For this problem, I want you to answer this question:
What would calibration mean in this context?

::: {.callout-tip title="Solution" collapse="true"}

In this context, calibration would mean that defendants
assigned a score of 1 would have a lower probability of re-offending
than defendants assigned a score of 2,
defendants assigned a score of 2 would have a lower probability of re-offending
than defendants assigned a score of 3, and so on.
Furthermore, we would hope that the difference between a score of 1 and 2
would be the same, approximately, as the difference between a score of 2
and 3.

:::


