---
title: "Hypothesis Testing"
execute:
  echo: false
format: html
---

## Learning Objective

Interpret the results of a hypothesis test in the context of logistic
regression.


## Lesson

### Interpretation in General

As in linear regression, R will conduct a hypothesis test
for each coefficient estimated in logistic regression.
The null hypothesis for each coefficient is that that coefficient is zero,
and the alternative hypothesis is that the coefficient is not zero.
It's always a two-sided test.

What does it mean if the coefficient is zero?
The answer to this is different if we're talking about the intercept
versus a non-intercept coefficient.
Let's start with the intercept.
Recall that $\exp(\beta_0)$ is the odds that the outcome equals 1
when all the covariates are equal to zero.
If $\beta_0 = 0$, then $\exp(\beta_0) = 1$.
What does an odds of 1 mean?
$$
\begin{align*}
\frac{p}{1-p} &= 1 \\
p &= 1 - p \\
2p &= 1 \\
p &= \frac{1}{2}
\end{align*}
$$
So, if $\beta_0 = 0$, then the probability that the outcome equals 1
when all the explanatory variables equal 0 is $1/2$.
It's a coin flip.
If the p-value is small, we would reject this hypothesis,
and conclude that the probability that the outcome equals 1
when all the explanatory variables equal 0 is not $1/2$.
Whether or not this is meaningful depends on your goals.

What about for the non-intercept coefficients?
For $j \ne 0$, $\exp(\beta_j)$ is the odds ratio comparing
two populations that are identical except one has $X_{ij} = x + 1$
and the other has $X_{kj} = x$.
Again, if $\beta_j = 0$, then $\exp(\beta_j) = 1$.
If an odds ratio is 1,
then the two odds it's comparing are equal to each other.
So, the null hypothesis corresponds to equal odds for these two populations.
If the p-value is small, we reject the null
and conclude that these two populations have different odds
that the outcome equals 1.

Note that if two odds are equal, the corresponding probabilities
must also be equal.
So, if an odds ratio is 1,
the odds are equivalent,
and the corresponding probabilities are equivalent.

### Example

```{r}
d <- Lock5Data::NutritionStudy
d$SmokeYes <- ifelse(test = d$Smoke == "Yes",
                     yes = 1,
                     no = 0)
d$YearsSince40 <- d$Age - 40
```

Let's return to the nutrition study.
We regressed smoking on years since age 40,
vitamin use (no, occasional, or regular),
and sex (male or female).

```{r}
#| echo: true
mod_01 <- glm(formula = SmokeYes ~ YearsSince40 + VitaminUse + Sex,
              family = binomial,
              data = d)
summary(mod_01)
```

For our hypothesis tests, let's use $\alpha = 0.05$.

- The p-value for the intercept is much less than $\alpha$,
so we can conclude that the odds of smoking
among 40-year-old females who don't use vitamins
is different from 1.
In other words,
the probability of smoking among 40-year-old females who don't use vitamins
is different from 0.5.
- Since the p-value for `YearsSince40` is less than 0.05,
the odds of smoking among people age $x + 1$ is different
than the odds of smoking among people age $x$,
holding vitamin use and sex constant.
- Since the p-value for `VitaminUseOccasional` is greater than 0.05,
we cannot conclude that there is a difference between
the odds of smoking among people who occasionally use vitamins
the odds of smoking among people who don't use vitamins,
holding age and sex constant.
- Since the p-value on `VitaminUseRegular` is less than 0.05,
we can conclude that
the odds of smoking among people who regularly use vitamins
is different than the odds of smoking among people who don't use vitamins,
holding age and sex constant.
- Since the p-value on `SexMale` is greater than 0.05, we cannot conclude that
the odds of smoking among males is different
than the odds of smoking among females,
holding age and vitamin use constant.

You may have noticed that I didn't use the phrase "statistically significant".
I'm not trying to make a statement by doing that.
For the p-values less than 0.05,
we could conclude that the observed differences between the odds
are statistically significant.
For the p-values greater than 0.05,
we could say that although we observed differences in the odds,
those differences are not statistically significant.


## Problems

### Problem 1

```{r}
d2 <- Lock5Data::NHANES
d2$OrganicYes <- ifelse(test = d2$Organic == "Yes",
                        yes = 1,
                        no = 0)
d2$Income1000 <- d2$Income / 1e3
```

Let's return to the NHANES data.
The outcome is a 1 if the respondent bought any food labeled organic
in the past 30 days,
and 0 otherwise.
I've regressed this on self-rated health (excellent, very good,
good, fair, and poor)
and monthly income (in thousands of dollars).

```{r}
mod_02 <- glm(formula = OrganicYes ~ Health + Income1000,
              family = binomial,
              data = d2)
summary(mod_02)
```

Interpret the hypothesis tests associated with the coefficients.
Use $\alpha = 0.05$.


::: {.callout-tip title="Solution" collapse="true"}

- Since the p-value on the intercept is less than 0.05,
the odds of having bought organic in the past 30 days
among people who make \$0/month and rate their own health as excellent
is different from 1 (and the probability is different from 0.5).
- Since the p-value associated with `HealthFair` is less than 0.05,
the odds of having bought organic in the past 30 days
among people who rate their health as fair
is different than the odds of having bought organic in the past 30 days
among people who rate their health as excellent,
holding income constant.
- Since the p-value associated with `HealthGood` is less than 0.05,
the odds of having bought organic in the past 30 days
among people who rate their health as good
is different than the odds of having bought organic in the past 30 days
among people who rate their health as excellent,
holding income constant.
- Since the p-value associated with `HealthPoor` is less than 0.05,
the odds of having bought organic in the past 30 days
among people who rate their health as poor
is different than the odds of having bought organic in the past 30 days
among people who rate their health as excellent,
holding income constant.
- Since the p-value associated with `HealthVery good` is greater than 0.05,
we cannot conclude that
the odds of having bought organic in the past 30 days
among people who rate their health as very good
is different than the odds of having bought organic in the past 30 days
among people who rate their health as excellent,
holding income constant.
- Since the p-value associated with `Income1000` is less than 0.05,
the odds of having bought organic in the past 30 days
among people who make $x + 1000$ dollars per month
is different than the odds of having bought organic in the past 30 days
among people who make $x$ dollars per month,
holding self-rated health constant.


:::

### Problem 2

```{r}
d3 <- Lock5Data::ICUAdmissions
d3$Died <- d3$Status
d3$YearsSince20 <- d3$Age - 20
```

Let's return to our ICU admissions data set.
Our outcome will be 1 if the patient died and 0 otherwise.
We'll use two covariates:
years since age 20
and whether or not the patient got CPR prior to admission (1 = yes, 0 = no).

```{r}
mod_03 <- glm(formula = Died ~ YearsSince20 + CPR,
              family = binomial,
              data = d3)
summary(mod_03)
```

Interpret the hypothesis tests associated with the coefficients.
Use $\alpha = 0.05$.

::: {.callout-tip title="Solution" collapse="true"}

- Since the p-value associated with the intercept is less than 0.05,
the odds of death for a 20-year-old who did not receive CPR prior to admission
to the ICU is different than 1 (and the probability is different than 0.5).
- Since the p-value associated with `YearsSince20` is less than 0.05,
the odds of death for someone age $x + 1$ is different than
the odds of death for someone age $x$, holding CPR status constant.
- Since the p-value associated with `CPR` is less than 0.05,
the odds of death among
people who received CPR prior to admission is different than the odds
of death of people who did not receive CPR prior to admission,
holding age constant.

:::




