---
title: "Performance"
execute:
  echo: false
format: html
---

```{r}
#| warning: false
library(pROC)
```

## Learning Objectives

- Define, compute, and interpret sensitivity and specificity.
- Interpret an ROC curve.
- Define, use, and interpret the area under the curve.


## Lesson

### Discrimination

One measure of the performance of a logistic regression model
is how well it discriminates between observed 0s and observed 1s.
Each observation gets a fitted value;
you can choose some cutoff such that any observation with a fitted value
below the cutoff gets a predicted outcome of 0
and any observation with a fitted value above the cutoff gets
a predicted outcome of 1.
Then you can compare the predicted outcome to the observed outcome.
There are a couple ways to make this comparison.
In order to talk about them,
it might be useful to use the terminology in the following table.
Here, "positive" means a 1 and "negative" means a 0.

| | Predicted Positive | Predicted Negative | Total |
|:-|:-:|:-:|:-:|
| Observed Positive | True Positive (TP) | False Negative (FN) | P |
| Observed Negative | False Positive (FP) | True Negative (TN) | N |


### Sensitivity

Sensitivity is the probability that an observed 1 will have a fitted value above
the cutoff, and thus receive a predicted outcome of 1.
Using the table above, we can compute the sensitivity as
$$
\frac{TP}{P} = \frac{TP}{TP + FN} \text{.}
$$
Sensitivity is also known as the true positive rate or the recall.

Sensitivity is often used in medical testing.
In that context, it's the probability that someone who actually has a disease
will test positive for it.

### Specificity

Specificity is the probability that an observed 0 will have a fitted value
below the cutoff,
and thus receive a predicted outcome of 0.
Using the table above, we can compute the specificity as
$$
\frac{TN}{N} = \frac{TN}{TN + FP} \text{.}
$$
Specificity is also known as the true negative rate.

Specificity is often used in medical testing.
In that context, it's the probability that someone who does not have a disease
will test negative for it.

### ROC Curve

In order to compute the sensitivity and specificity,
we need a cutoff to separate the predicted 1s from the predicted 0s.
Different cutoffs yield different sensitivities and specificities.
If we make a receiver operating characteristic (ROC) curve,
we can look at all these sensitivities and specificities at once.

Let's walk through the creation of an ROC curve for a specific example.
In the future, you will just use an R package to create the ROC curve,
but making one (or at least going through a few of the steps)
helps with understanding how to interpret it.

```{r}
d <- Lock5Data::NutritionStudy
d$SmokeYes <- ifelse(test = d$Smoke == "Yes",
                     yes = 1,
                     no = 0)
d$YearsSince40 <- d$Age - 40
```

Let's return to the nutrition study.
We regressed smoking on years since age 40,
vitamin use (no, occasional, or regular),
and sex (male or female).

```{r}
#| echo: true
mod_01 <- glm(formula = SmokeYes ~ YearsSince40 + VitaminUse + Sex,
              family = binomial,
              data = d)
summary(mod_01)
```

Let's start with an empty plot.
On the horizontal axis, we'll put one minus the specificity
($1 - \text{specificity}$).
I don't know why we use one minus the specificity instead of the specificity
itself, but that's the way it is.
On the vertical axis, we'll put the sensitivity.

```{r}
plot(x = c(0,1),
     y = c(0,1),
     xlab = "1 - Specificity",
     ylab = "Sensitivity",
     main = "ROC Curve",
     type = "l",
     lty = 2)
grid()
```

Next, I'm going to choose the lowest possible cutoff: 0.
Because all the fitted values are above 0,
every observation will get a predicted value of 1.
What are the sensitivity and specificity for this cutoff?
```{r}
#| echo: true
sensitivity <- sum(mod_01$fitted.values > 0 & mod_01$model$SmokeYes == 1) / sum(mod_01$model$SmokeYes == 1)
sensitivity
specificity <- sum(mod_01$fitted.values <= 0 & mod_01$model$SmokeYes == 0) / sum(mod_01$model$SmokeYes == 0)
specificity
```
The sensitivity is 1 because all the observed 1s received predicted a predicted
value of 1,
because *every observation* received a predicted value of 1.
This would be like opening and searching every single bag at the airport;
you wouldn't miss any dangerous objects, but you'd have to search every bag.
The specificity, on the other hand, is 0.
Since every observation received a predicted value of 1,
no observed 0s were correctly categorized as 0s.
In the airport example,
you didn't correctly identify any safe bags as safe.

Anyway, we can add this point to our plot.
```{r}
plot(x = c(0,1),
     y = c(0,1),
     xlab = "1 - Specificity",
     ylab = "Sensitivity",
     main = "ROC Curve",
     type = "l",
     lty = 2)
grid()
points(x = 1 - specificity,
       y = sensitivity,
       pch = 19)
```


Okay, let's do a different cutoff.
I'm going to choose the lowest fitted value.
```{r}
#| echo: true
# Get cutoff
sorted_unique_fitted <- sort(unique(mod_01$fitted.values))
cutoff <- sorted_unique_fitted[1]

# Sensitivity
numerator <- sum(mod_01$fitted.values > cutoff & mod_01$model$SmokeYes == 1)
denominator <- sum(mod_01$model$SmokeYes == 1)
sensitivity <- c(sensitivity,
                 numerator / denominator)
sensitivity[2]

# Specificity
numerator <- sum(mod_01$fitted.values <= cutoff & mod_01$model$SmokeYes == 0)
denominator <- sum(mod_01$model$SmokeYes == 0)
specificity <- c(specificity,
                 numerator / denominator)
specificity[2]
```

By increasing the cutoff, we slightly increased the specificity.
It must be the case that the observation with the lowest fitted value
had an observed outcome of 0.
Let's check:
```{r}
#| echo: true
mod_01$model$SmokeYes[mod_01$fitted.values == cutoff]
```
Yep! So by increasing the cutoff slightly,
we correctly categorized one observed 0 as a predicted 0,
increasing our specificity.
Because all observed 1s remained categorized as predicted 1s,
we didn't decrease our sensitivity.

We can add this new point to our plot.
```{r}
plot(x = c(0,1),
     y = c(0,1),
     xlab = "1 - Specificity",
     ylab = "Sensitivity",
     main = "ROC Curve",
     type = "l",
     lty = 2)
grid()
points(x = 1 - specificity,
       y = sensitivity,
       pch = 19)
```

You can't really tell there's even a new point on the plot,
because it's so close to the first one.

Let's skip ahead and just have R draw the entire ROC curve for us.
There are many packages with functions that do this.
I tend to use the package `pROC`.

```{r}
#| echo: true
mod_01_roc <- roc(response = mod_01$model$SmokeYes,
                  predictor = mod_01$fitted.values)
plot(mod_01_roc)
```

Now that it's plotted, how do we interpret it?
Well, we want specificity and sensitivity to both be high.
So, ideally, we'd like the dark black line to go all the way to the
upper left-hand corner of the plot.
The closer we get to this ideal,
the better the model performs at discriminating 0s and 1s.

What's the worst we could do?
You might think the worst possible ROC curve would be one that
goes all the way to the lower right-hand corner of the plot.
However, if it did,
that would mean that the model does a perfectly bad job of discriminating
0s and 1s.
In other words, it categorizes all observed 0s as predicted 1s
and all observed 1s as predicted 0s.
In that case, you could just flip the predicted labels and do a perfectly
good job.

You may have noticed the following message that the function sent:
```
Setting levels: control = 0, case = 1
Setting direction: controls < cases
```
The `controls < cases` message means that fitted values below
each cutoff were categorized as predicted 0s.
If the function detected that it could perform better
by categorizing fitted values above each cutoff as predicted 0s,
it would do that.
It would tell the user `controls > cases` as well.
In order to prevent this gaming of the system,
you can use the `direction` parameter of the `roc` function.

So, what's the worst you can do?
The worst ROC curve is one that follows the reference line.
it's exactly between perfectly bad and perfectly good.

For the model above, the AUC is:
```{r}
auc(mod_01_roc)
```
This is better than chance.
It's hard to say how good or bad this is without knowing how well
other models perform.

### Area Under the Curve

The area under the curve (AUC) is a single value that summarizes
the ROC curve.
It's literally the area under the ROC curve.
The highest it can be is 1,
and the lowest (assuming you haven't changed the `direction` parameter) is 0.5.
So, a higher AUC is better.

Surprisingly, the AUC has a more intuitive interpretation
than just `high = good` and `low = bad`.
The AUC is the probability that a randomly-selected observed 1
will have a higher fitted value than a randomly-selected observed 0.
This is sometimes called the "probability of superiority"
or the "common language effect size".
This interpretation might make it easier to understand why 0.5
is the worst you can do.
An AUC of 0.5 means that a randomly-selected observed 1
is equally likely to have a fitted value above or below
the fitted value of a randomly-selected
observed 0.


## Problems

### Problem 1

```{r}
# Set seed
set.seed(9)

# Parameters
n <- 1e3
beta_1 <- 2

# Generate x's
x_a <- rnorm(n = n,
             mean = -2.5)
x_b <- rnorm(n)
x_c <- rnorm(n = n,
             mean = 2.5)

# Generate y's
y_a <- rbinom(n = n,
              size = 1,
              prob = plogis(beta_1 * x_a))
y_b <- rbinom(n = n,
              size = 1,
              prob = plogis(beta_1 * x_b))
y_c <- rbinom(n = n,
              size = 1,
              prob = plogis(beta_1 * x_c))

# Fit models
mod_a <- glm(formula = y_a ~ x_a,
             family = binomial)
mod_b <- glm(formula = y_b ~ x_b,
             family = binomial)
mod_c <- glm(formula = y_c ~ x_c,
             family = binomial)
```

Below are three ROC curves.
Each was created from a different simulated data set.
I'll call the data sets A, B, and C,
and I'll describe how they differ.
Your job is to match the data set to the correct plot.

Each data set was simulated according to the model
$$
p_i = \frac{1}{1 + \exp\left(-2 X_i\right)} \text{.}
$$
The three data sets differ only in the distribution of the explanatory variable.

- In data set A,
the explanatory variable is normally distributed with mean -2.5
and variance 1.
This means there are `r sum(y_a == 1)` observations where the outcome is 1
and `r sum(y_a == 0)` observations where the outcome is 0.
- In data set B,
the explanatory variable is normally distributed with mean 0
and variance 1.
This means there are `r sum(y_b == 1)` observations where the outcome is 1
and `r sum(y_b == 0)` observations where the outcome is 0.
- In data set A,
the explanatory variable is normally distributed with mean 2.5
and variance 1.
This means there are `r sum(y_c == 1)` observations where the outcome is 1
and `r sum(y_c == 0)` observations where the outcome is 0.

Simple linear regression was used to estimate the coefficients
and obtain fitted values.

Which plot corresponds to which data set?


```{r}
# Plot ROC curves
plot(roc(response = mod_c$model$y_c,
         predictor = mod_c$fitted.values),
     main = "Plot 1")
plot(roc(response = mod_a$model$y_a,
         predictor = mod_a$fitted.values),
     main = "Plot 2")
plot(roc(response = mod_b$model$y_b,
         predictor = mod_b$fitted.values),
     main = "Plot 3")
```


::: {.callout-tip title="Solution" collapse="true"}


- Plot 1 used data set C.
- Plot 2 used data set A.
- Plot 3 used data set B.

Recall that when we made our ROC curve, we started with a cutoff of 0,
which led to perfect sensitivity and perfectly bad specificity.
When we raised the cutoff slightly,
one observed 0 got correctly categorized as a predicted 0.
This increased the specificity but did not decrease the sensitivity.
If we had a bunch of 0s with low fitted values,
we would continue to see increases in specificity with no corresponding
drops in sensitivity.
Data set A has lots of 0s,
and it's probably the case that a lot of those 0s had low fitted values,
since there was nothing wrong about how we fit the logistic regression.
(For example, we didn't include any unnecessary explanatory variables
or leave out an necessary explanatory variables.)
So, it's reasonable to assume that data set A corresponds to Plot 2,
which has a long, flat line at the top,
indicating increases in specificity without drops in sensitivity.
And this is correct.

If a data set has lots of observed 1s with high fitted values,
then once you reach a certain cutoff,
additional increases in the cutoff will categorize additional observed 1s
as predicted 0s,
decreasing the sensitivity without decreasing the specificity.
This describes data set C and plot 1.

By process of elimination,
the remaining data set matches the remaining plot.

:::






