---
title: "Interactions"
execute:
  echo: false
format: html
---

## Learning Objective

Interpret the coefficients of a logistic regression model with an interaction
term.


## Lesson

### Example

```{r}
d <- Lock5Data::NutritionStudy
d$SmokeYes <- ifelse(test = d$Smoke == "Yes",
                     yes = 1,
                     no = 0)
d$YearsSince40 <- d$Age - 40
```

Let's return to the nutrition data set.
We'll regress smoking on years since age 40,
vitamin use (no, occasional, or regular),
and sex (male or female),
but include an interaction between years since age 40
and vitamin use.

```{r}
#| echo: true
mod_01 <- glm(formula = SmokeYes ~ YearsSince40 * VitaminUse + Sex,
              family = binomial,
              data = d)
summary(mod_01)
```

How do we interpret these coefficients?
In order to save some space, I'm going to use the following notation.
Suppose $i\in\{1,\dotsc,315\}$.
Define:

- $S_i = 1$ if the $i$th participant smokes and $S_i = 0$ otherwise.
- $Y_i$ is years since age 40 for the $i$th participant.
- $O_i = 1$ if the $i$th participant used vitamins occasionally
and $O_i = 0$ otherwise.
- $R_i = 1$ if the $i$th participant used vitamins regularly
and $R_i = 0$ otherwise.
- $M_i = 1$ if the $i$th participant was male and $M_i = 0$ otherwise.

Recall that the $i$th participant's estimated odds of smoking can be
computed like so:
$$
\widehat{\text{odds}}_i = \exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 M_i + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right) \text{.}
$$

**Intercept**

$\exp\left(\hat{\beta}_0\right)$ is the estimated odds of smoking
for a 40-year-old female who doesn't use vitamins.
```{r}
#| echo: true
exp(mod_01$coefficients[1])
```

**Main Effects**


- $\exp\left(\hat{\beta}_1\right)$ is the estimated odds ratio
comparing participants who differ in age by one year,
are the same sex,
and don't use vitamins.
$$
\begin{align*}
\frac{\widehat{\text{odds}}_i}{\widehat{\text{odds}}_j} &= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 M_i + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_j + \hat{\beta}_2 O_j + \hat{\beta}_3 R_j + \hat{\beta}_4 M_j + \hat{\beta}_5 Y_j O_j + \hat{\beta}_6 Y_j R_j\right)} \\
&= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 (y+1) + \hat{\beta}_2 (0) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 (y+1) (0) + \hat{\beta}_6 (y+1) (0)\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 y + \hat{\beta}_2 (0) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 y (0) + \hat{\beta}_6 y (0)\right)} \\
&= \exp\left(\hat{\beta}_1  \right) \\
\end{align*}
$$
```{r}
#| echo: true
exp(mod_01$coefficients[2])
```
- $\exp\left(\hat{\beta}_2\right)$ is the estimated odds ratio
comparing participants who use vitamins occasionally to participants
who don't use vitamins,
where both groups of participants are the same sex
and are 40 years old.
$$
\begin{align*}
\frac{\widehat{\text{odds}}_i}{\widehat{\text{odds}}_j} &= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 M_i + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_j + \hat{\beta}_2 O_j + \hat{\beta}_3 R_j + \hat{\beta}_4 M_j + \hat{\beta}_5 Y_j O_j + \hat{\beta}_6 Y_j R_j\right)} \\
&= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 (0) + \hat{\beta}_2 (1) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 (0) (1) + \hat{\beta}_6 (0) (0)\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 (0) + \hat{\beta}_2 (0) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 (0) (0) + \hat{\beta}_6 (0) (0)\right)} \\
&= \exp\left(\hat{\beta}_2\right)
\end{align*}
$$
```{r}
#| echo: true
exp(mod_01$coefficients[3])
```
- $\exp\left(\hat{\beta}_3\right)$ is the estimated odds ratio
comparing participants who use vitamins regularly to participants
who don't use vitamins,
where both groups of participants are the same sex
and are 40 years old.
$$
\begin{align*}
\frac{\widehat{\text{odds}}_i}{\widehat{\text{odds}}_j} &= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 M_i + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_j + \hat{\beta}_2 O_j + \hat{\beta}_3 R_j + \hat{\beta}_4 M_j + \hat{\beta}_5 Y_j O_j + \hat{\beta}_6 Y_j R_j\right)} \\
&= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 (0) + \hat{\beta}_2 (0) + \hat{\beta}_3 (1) + \hat{\beta}_4 M_i + \hat{\beta}_5 (0) (0) + \hat{\beta}_6 (0) (1)\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 (0) + \hat{\beta}_2 (0) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 (0) (0) + \hat{\beta}_6 (0) (0)\right)} \\
&= \exp\left(\hat{\beta}_3\right)
\end{align*}
$$
```{r}
#| echo: true
exp(mod_01$coefficients[4])
```
- $\exp\left(\hat{\beta}_4\right)$ is the estimated odds ratio
comparing male to non-male participants
who are the same age and have the same vitamin use.
$$
\begin{align*}
\frac{\widehat{\text{odds}}_i}{\widehat{\text{odds}}_j} &= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 M_i + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_j + \hat{\beta}_2 O_j + \hat{\beta}_3 R_j + \hat{\beta}_4 M_j + \hat{\beta}_5 Y_j O_j + \hat{\beta}_6 Y_j R_j\right)} \\
&= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 (1) + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 (0) + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)} \\
&= \exp\left(\hat{\beta}_4\right)
\end{align*}
$$
```{r}
#| echo: true
exp(mod_01$coefficients[5])
```

**Moderation of Age by Vitamin Use**

- $\exp\left(\hat{\beta}_1 + \hat{\beta}_5\right)$ is the estimated odds ratio
comparing participants who differ in age by one year,
are the same sex,
and use vitamins occasionally.
$$
\begin{align*}
\frac{\widehat{\text{odds}}_i}{\widehat{\text{odds}}_j} &= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 M_i + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_j + \hat{\beta}_2 O_j + \hat{\beta}_3 R_j + \hat{\beta}_4 M_j + \hat{\beta}_5 Y_j O_j + \hat{\beta}_6 Y_j R_j\right)} \\
&= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 (y+1) + \hat{\beta}_2 (1) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 (y+1) (1) + \hat{\beta}_6 (y+1) (0)\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 y + \hat{\beta}_2 (1) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 y (1) + \hat{\beta}_6 y (0)\right)} \\
&= \exp\left(\hat{\beta}_1 + \hat{\beta}_5\right) \\
&= \exp\left(\hat{\beta}_1\right) \exp\left(\hat{\beta}_5\right) \\
\end{align*}
$$
```{r}
#| echo: true
exp(mod_01$coefficients[2] + mod_01$coefficients[6])
exp(mod_01$coefficients[2]) * exp(mod_01$coefficients[6])
```
- $\exp\left(\hat{\beta}_1 + \hat{\beta}_6\right)$ is the estimated odds ratio
comparing participants who differ in age by one year,
are the same sex,
and use vitamins regularly.
$$
\begin{align*}
\frac{\widehat{\text{odds}}_i}{\widehat{\text{odds}}_j} &= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 M_i + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_j + \hat{\beta}_2 O_j + \hat{\beta}_3 R_j + \hat{\beta}_4 M_j + \hat{\beta}_5 Y_j O_j + \hat{\beta}_6 Y_j R_j\right)} \\
&= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 (y+1) + \hat{\beta}_2 (0) + \hat{\beta}_3 (1) + \hat{\beta}_4 M_i + \hat{\beta}_5 (y+1) (0) + \hat{\beta}_6 (y+1) (1)\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 y + \hat{\beta}_2 (0) + \hat{\beta}_3 (1) + \hat{\beta}_4 M_i + \hat{\beta}_5 y (0) + \hat{\beta}_6 y (1)\right)} \\
&= \exp\left(\hat{\beta}_1 + \hat{\beta}_6\right) \\
&= \exp\left(\hat{\beta}_1\right) \exp\left(\hat{\beta}_6\right) \\
\end{align*}
$$
```{r}
#| echo: true
exp(mod_01$coefficients[2] + mod_01$coefficients[7])
exp(mod_01$coefficients[2]) * exp(mod_01$coefficients[7])
```

**Moderation of Vitamin Use by Age**

- $\exp\left(\hat{\beta}_2 + \hat{\beta}_5y\right)$ is the estimated odds ratio
comparing participants who use vitamins occasionally
to participants who don't use vitamins,
where both groups are the same sex
and were 40 years old $y$ years ago.
$$
\begin{align*}
\frac{\widehat{\text{odds}}_i}{\widehat{\text{odds}}_j} &= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 M_i + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_j + \hat{\beta}_2 O_j + \hat{\beta}_3 R_j + \hat{\beta}_4 M_j + \hat{\beta}_5 Y_j O_j + \hat{\beta}_6 Y_j R_j\right)} \\
&= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 y + \hat{\beta}_2 (1) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 y (1) + \hat{\beta}_6 y (0)\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 y + \hat{\beta}_2 (0) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 y (0) + \hat{\beta}_6 y (0)\right)} \\
&= \exp\left(\hat{\beta}_2 + \hat{\beta}_5 y\right) \\
&= \exp\left(\hat{\beta}_2\right) \exp\left(\hat{\beta}_5 y\right) \\
\end{align*}
$$
```{r}
temp <- function(x){
  exp(mod_01$coefficients[3] + mod_01$coefficients[6] * (x - 40))
}
curve(expr = temp,
      from = min(d$Age),
      to = max(d$Age),
      xlab = "Age",
      ylab = "Odds Ratio",
      main = "Occasional to No Vitamin Use, Same Sex")
```
- $\exp\left(\hat{\beta}_3 + \hat{\beta}_6y\right)$ is the estimated odds ratio
comparing participants who use vitamins regularly
to participants who don't use vitamins,
where both groups are the same sex
and were 40 years old $y$ years ago.
$$
\begin{align*}
\frac{\widehat{\text{odds}}_i}{\widehat{\text{odds}}_j} &= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 M_i + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_j + \hat{\beta}_2 O_j + \hat{\beta}_3 R_j + \hat{\beta}_4 M_j + \hat{\beta}_5 Y_j O_j + \hat{\beta}_6 Y_j R_j\right)} \\
&= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 y + \hat{\beta}_2 (0) + \hat{\beta}_3 (1) + \hat{\beta}_4 M_i + \hat{\beta}_5 y (0) + \hat{\beta}_6 y (1)\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 y + \hat{\beta}_2 (0) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 y (0) + \hat{\beta}_6 y (0)\right)} \\
&= \exp\left(\hat{\beta}_3 + \hat{\beta}_6 y\right) \\
&= \exp\left(\hat{\beta}_3\right) \exp\left(\hat{\beta}_6 y\right) \\
\end{align*}
$$
```{r}
temp <- function(x){
  exp(mod_01$coefficients[4] + mod_01$coefficients[7] * (x - 40))
}
curve(expr = temp,
      from = min(d$Age),
      to = max(d$Age),
      xlab = "Age",
      ylab = "Odds Ratio",
      main = "Regular to No Vitamin Use, Same Sex")
abline(h = 1,
       lty = 2)
```

### Heuristic

When interpreting terms in a model that has an interaction terms,
these might be some useful questions to ask:

- Is this an intercept or a non-intercept term?
  - If you're interpreting an intercept,
  the interaction term doesn't really change anything.
  Interpreting the intercept always requires setting all the covariates to zero.
  - If you're interpreting a non-intercept term,
  you may need to take into account the interaction term.
- Is this coefficient attached to a single explanatory variable
or a product?
  - It's often really challenging to interpret a coefficient
  that's attached to an interaction term.
  Usually it's easier to interpret coefficients attached to single explanatory
  variables,
  or sums of coefficients,
  where one of the coefficients in the sum is attached to a single explanatory
  variable and the other coefficient in the sum
  is attached to an interaction term.
- What is this thing?
  - For linear regression, the intercept is an expected value
  and each non-intercept term attached to a single explanatory variable
  is an expected difference.
  For logistic regression, the exponentiated intercept is an odds
  and each exponentiated non-intercept coefficient attached to
  a single explanatory variable is an odds ratio.
- Does the explanatory variable attached to this coefficient also appear
in an interaction term?
  - If no, you can interpret the coefficient as though there isn't
  an interaction term in the model.
  - If yes, you have to decide if you want to interpret the coefficient
  by itself or combined with the coefficient attached to the interaction term.
    - If you interpret the coefficient by itself,
    it applies only to a population where the other explanatory variable
    in the interaction term equals zero.
    - If you interpret the coefficient combined with the coefficient
    attached to the interaction term,
    it applies only to a population where the other explanatory variable
    in the interaction term equals one.
    If the other explanatory variable is continuous,
    you can replace it with a lower-case letter
    to represent an algebraic variable
    (rather than a random variable).

## Problems

### Problem 1

```{r}
d2 <- Lock5Data::NHANES
d2$OrganicYes <- ifelse(test = d2$Organic == "Yes",
                        yes = 1,
                        no = 0)
d2$Income1000 <- d2$Income / 1e3
```

Let's return to the NHANES data set.
The outcome is a 1 if the respondent bought any food labeled organic
in the past 30 days,
and 0 otherwise.
I've regressed this on self-rated health (excellent, very good,
good, fair, and poor)
and monthly income (in thousands of dollars),
with an interaction term between the two explanatory variables.

```{r}
mod_02 <- glm(formula = OrganicYes ~ Health * Income1000,
              family = binomial,
              data = d2)
summary(mod_02)
```

Interpret the exponentiated coefficients.
```{r}
#| echo: true
exp(mod_02$coefficients)
```

::: {.callout-tip title="Solution" collapse="true"}

**Intercept**

- The odds of having bought organic food for someone who has no income
and rates their own health as excellent is 0.39.

**Main Effects**

- The odds of someone who rates their own health as fair having bought
organic food are 25% lower than the same odds for someone who rates
their own health as excellent,
provided both these people have no income.
- The odds of someone who rates their own health as good having bought
organic food are 17% lower than the same odds for someone who rates
their own health as excellent,
provided both these people have no income.
- The odds of someone who rates their own health as poor having bought
organic food are 28% lower than the same odds for someone who rates
their own health as excellent,
provided both these people have no income.
- The odds of someone who rates their own health as very good having bought
organic food are 13% higher than the same odds for someone who rates
their own health as excellent,
provided both these people have no income.
- The odds of of having bought organic food for someone who makes $x + 1000$
dollars per month is 20% greater than the odds of having bought organic
food for someone who makes $x$ dollars per month,
provided both these people rate their health as excellent.

**Moderation of Self-Rated Health by Income**

- The odds of someone who rates their own health as fair having bought
organic food are $0.7482764\times 0.8714446^x$ times the same odds for someone who rates
their own health as excellent,
provided both these people make $x$ dollars per month.
The second term gets raised to the $x$th power because
$$
\begin{align*}
\exp\left(\hat{\beta}_1 + \hat{\beta}_6 x\right) &= 
\exp\left(\hat{\beta}_1 \right) \exp\left( \hat{\beta}_6 x\right) \\
&= 
\exp\left(\hat{\beta}_1 \right) \exp\left( \hat{\beta}_6\right)^x \text{.}
\end{align*}
$$
- The odds of someone who rates their own health as good having bought
organic food are $0.8317762\times 0.9699155^x$ times the same odds for someone who rates
their own health as excellent,
provided both these people make $x$ dollars per month.
- The odds of someone who rates their own health as poor having bought
organic food are $0.7233798\times 0.9274793^x$ times the same odds for someone who rates
their own health as excellent,
provided both these people make $x$ dollars per month.
- The odds of someone who rates their own health as very good having bought
organic food are $1.1274843\times 0.9581934^x$ times the same odds for someone who rates
their own health as excellent,
provided both these people make $x$ dollars per month.

**Moderation of Income by Self-Rated Health**

- The odds of of having bought organic food for someone who makes $x + 1000$
dollars per month is $1.1999314\times 0.8714446$ times the odds of having bought organic
food for someone who makes $x$ dollars per month,
provided both these people rate their health as fair.
- The odds of of having bought organic food for someone who makes $x + 1000$
dollars per month is $1.1999314\times 0.9699155$ times the odds of having bought organic
food for someone who makes $x$ dollars per month,
provided both these people rate their health as good.
- The odds of of having bought organic food for someone who makes $x + 1000$
dollars per month is $1.1999314\times 0.7233798$ times the odds of having bought organic
food for someone who makes $x$ dollars per month,
provided both these people rate their health as poor.
- The odds of of having bought organic food for someone who makes $x + 1000$
dollars per month is $1.1999314\times 1.1274843$ times the odds of having bought organic
food for someone who makes $x$ dollars per month,
provided both these people rate their health as very good.



:::

### Problem 2

```{r}
d3 <- Lock5Data::ICUAdmissions
d3$Died <- d3$Status
d3$YearsSince20 <- d3$Age - 20
```

Let's return to our ICU admissions data set.
Our outcome will be 1 if the patient died and 0 otherwise.
We'll use two covariates:
years since age 20
and whether or not the patient got CPR prior to admission (1 = yes, 0 = no).
I've also added an interaction term between the two covariates.

```{r}
mod_03 <- glm(formula = Died ~ YearsSince20 * CPR,
              family = binomial,
              data = d3)
summary(mod_03)
```

Intepret the exponentiated coefficients.
```{r}
#| echo: true
exp(mod_03$coefficients)
```

::: {.callout-tip title="Solution" collapse="true"}

- The odds of death for a twenty-year-old who did not get CPR
is 0.08.
- The odds of death for someone age $x + 21$ is $1.02507578$ times
the odds of death for someone age $x + 20$,
provided both these people did not receive CPR.
- The odds of death for someone who did receive CPR is $0.15894754$ times
the odds of death for someone who didn't receive CPR,
provided both these people are age 20.
- The odds of death for someone age $x + 21$ is $1.02507578\times 1.09876599$ times
the odds of death for someone age $x + 20$,
provided both these people did receive CPR.
- The odds of death for someone who did receive CPR is $0.15894754\times 1.09876599^x$ times
the odds of death for someone who didn't receive CPR,
provided both these people are age $x+20$.

:::




