---
title: "Interactions"
execute:
  echo: false
format: html
---

## Learning Objective

Interpret the coefficients of a logistic regression model with an interaction
term.


## Lesson



```{r}
d <- Lock5Data::NutritionStudy
d$SmokeYes <- ifelse(test = d$Smoke == "Yes",
                     yes = 1,
                     no = 0)
d$YearsSince40 <- d$Age - 40
```

Let's return to the nutrition data set.
We'll regress smoking on years since age 40,
vitamin use (no, occasional, or regular),
and sex (male or female),
but include an interaction between years since age 40
and vitamin use.

```{r}
#| echo: true
mod_01 <- glm(formula = SmokeYes ~ YearsSince40 * VitaminUse + Sex,
              family = binomial,
              data = d)
summary(mod_01)
```

How do we interpret these coefficients?
In order to save some space, I'm going to use the following notation.
Suppose $i\in\{1,\dotsc,315\}$.
Define:

- $S_i = 1$ if the $i$th participant smokes and $S_i = 0$ otherwise.
- $Y_i$ is years since age 40 for the $i$th participant.
- $O_i = 1$ if the $i$th participant used vitamins occasionally
and $O_i = 0$ otherwise.
- $R_i = 1$ if the $i$th participant used vitamins regularly
and $R_i = 0$ otherwise.
- $M_i = 1$ if the $i$th participant was male and $M_i = 0$ otherwise.

Recall that the $i$th participant's estimated odds of smoking can be
computed like so:
$$
\widehat{\text{odds}}_i = \exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 M_i + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right) \text{.}
$$

**Intercept**

$\exp\left(\hat{\beta}_0\right)$ is the estimated odds of smoking
for a 40-year-old female who doesn't use vitamins.
```{r}
#| echo: true
exp(mod_01$coefficients[1])
```

**Main Effects**


- $\exp\left(\hat{\beta}_1\right)$ is the estimated odds ratio
comparing participants who differ in age by one year,
are the same sex,
and don't use vitamins.
$$
\begin{align*}
\frac{\widehat{\text{odds}}_i}{\widehat{\text{odds}}_j} &= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 M_i + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_j + \hat{\beta}_2 O_j + \hat{\beta}_3 R_j + \hat{\beta}_4 M_j + \hat{\beta}_5 Y_j O_j + \hat{\beta}_6 Y_j R_j\right)} \\
&= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 (y+1) + \hat{\beta}_2 (0) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 (y+1) (0) + \hat{\beta}_6 (y+1) (0)\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 y + \hat{\beta}_2 (0) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 y (0) + \hat{\beta}_6 y (0)\right)} \\
&= \exp\left(\hat{\beta}_1  \right) \\
\end{align*}
$$
```{r}
#| echo: true
exp(mod_01$coefficients[2])
```
- $\exp\left(\hat{\beta}_2\right)$ is the estimated odds ratio
comparing participants who use vitamins occasionally to participants
who don't use vitamins,
where both groups of participants are the same sex
and are 40 years old.
$$
\begin{align*}
\frac{\widehat{\text{odds}}_i}{\widehat{\text{odds}}_j} &= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 M_i + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_j + \hat{\beta}_2 O_j + \hat{\beta}_3 R_j + \hat{\beta}_4 M_j + \hat{\beta}_5 Y_j O_j + \hat{\beta}_6 Y_j R_j\right)} \\
&= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 (0) + \hat{\beta}_2 (1) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 (0) (1) + \hat{\beta}_6 (0) (0)\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 (0) + \hat{\beta}_2 (0) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 (0) (0) + \hat{\beta}_6 (0) (0)\right)} \\
&= \exp\left(\hat{\beta}_2\right)
\end{align*}
$$
```{r}
#| echo: true
exp(mod_01$coefficients[3])
```
- $\exp\left(\hat{\beta}_3\right)$ is the estimated odds ratio
comparing participants who use vitamins regularly to participants
who don't use vitamins,
where both groups of participants are the same sex
and are 40 years old.
$$
\begin{align*}
\frac{\widehat{\text{odds}}_i}{\widehat{\text{odds}}_j} &= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 M_i + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_j + \hat{\beta}_2 O_j + \hat{\beta}_3 R_j + \hat{\beta}_4 M_j + \hat{\beta}_5 Y_j O_j + \hat{\beta}_6 Y_j R_j\right)} \\
&= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 (0) + \hat{\beta}_2 (0) + \hat{\beta}_3 (1) + \hat{\beta}_4 M_i + \hat{\beta}_5 (0) (0) + \hat{\beta}_6 (0) (1)\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 (0) + \hat{\beta}_2 (0) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 (0) (0) + \hat{\beta}_6 (0) (0)\right)} \\
&= \exp\left(\hat{\beta}_3\right)
\end{align*}
$$
```{r}
#| echo: true
exp(mod_01$coefficients[4])
```
- $\exp\left(\hat{\beta}_4\right)$ is the estimated odds ratio
comparing male to non-male participants
who are the same age and have the same vitamin use.
$$
\begin{align*}
\frac{\widehat{\text{odds}}_i}{\widehat{\text{odds}}_j} &= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 M_i + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_j + \hat{\beta}_2 O_j + \hat{\beta}_3 R_j + \hat{\beta}_4 M_j + \hat{\beta}_5 Y_j O_j + \hat{\beta}_6 Y_j R_j\right)} \\
&= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 (1) + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 (0) + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)} \\
&= \exp\left(\hat{\beta}_4\right)
\end{align*}
$$
```{r}
#| echo: true
exp(mod_01$coefficients[5])
```

**Moderation of Age by Vitamin Use**

- $\exp\left(\hat{\beta}_1 + \hat{\beta}_5\right)$ is the estimated odds ratio
comparing participants who differ in age by one year,
are the same sex,
and use vitamins occasionally.
$$
\begin{align*}
\frac{\widehat{\text{odds}}_i}{\widehat{\text{odds}}_j} &= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 M_i + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_j + \hat{\beta}_2 O_j + \hat{\beta}_3 R_j + \hat{\beta}_4 M_j + \hat{\beta}_5 Y_j O_j + \hat{\beta}_6 Y_j R_j\right)} \\
&= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 (y+1) + \hat{\beta}_2 (1) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 (y+1) (1) + \hat{\beta}_6 (y+1) (0)\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 y + \hat{\beta}_2 (1) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 y (1) + \hat{\beta}_6 y (0)\right)} \\
&= \exp\left(\hat{\beta}_1 + \hat{\beta}_5\right) \\
&= \exp\left(\hat{\beta}_1\right) \exp\left(\hat{\beta}_5\right) \\
\end{align*}
$$
```{r}
#| echo: true
exp(mod_01$coefficients[2] + mod_01$coefficients[6])
exp(mod_01$coefficients[2]) * exp(mod_01$coefficients[6])
```
- $\exp\left(\hat{\beta}_1 + \hat{\beta}_6\right)$ is the estimated odds ratio
comparing participants who differ in age by one year,
are the same sex,
and use vitamins regularly.
$$
\begin{align*}
\frac{\widehat{\text{odds}}_i}{\widehat{\text{odds}}_j} &= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 M_i + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_j + \hat{\beta}_2 O_j + \hat{\beta}_3 R_j + \hat{\beta}_4 M_j + \hat{\beta}_5 Y_j O_j + \hat{\beta}_6 Y_j R_j\right)} \\
&= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 (y+1) + \hat{\beta}_2 (0) + \hat{\beta}_3 (1) + \hat{\beta}_4 M_i + \hat{\beta}_5 (y+1) (0) + \hat{\beta}_6 (y+1) (1)\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 y + \hat{\beta}_2 (0) + \hat{\beta}_3 (1) + \hat{\beta}_4 M_i + \hat{\beta}_5 y (0) + \hat{\beta}_6 y (1)\right)} \\
&= \exp\left(\hat{\beta}_1 + \hat{\beta}_6\right) \\
&= \exp\left(\hat{\beta}_1\right) \exp\left(\hat{\beta}_6\right) \\
\end{align*}
$$
```{r}
#| echo: true
exp(mod_01$coefficients[2] + mod_01$coefficients[7])
exp(mod_01$coefficients[2]) * exp(mod_01$coefficients[7])
```

**Moderation of Vitamin Use by Age**

- $\exp\left(\hat{\beta}_2 + \hat{\beta}_5y\right)$ is the estimated odds ratio
comparing participants who use vitamins occasionally
to participants who don't use vitamins,
where both groups are the same sex
and were 40 years old $y$ years ago.
$$
\begin{align*}
\frac{\widehat{\text{odds}}_i}{\widehat{\text{odds}}_j} &= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 M_i + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_j + \hat{\beta}_2 O_j + \hat{\beta}_3 R_j + \hat{\beta}_4 M_j + \hat{\beta}_5 Y_j O_j + \hat{\beta}_6 Y_j R_j\right)} \\
&= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 y + \hat{\beta}_2 (1) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 y (1) + \hat{\beta}_6 y (0)\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 y + \hat{\beta}_2 (0) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 y (0) + \hat{\beta}_6 y (0)\right)} \\
&= \exp\left(\hat{\beta}_2 + \hat{\beta}_5 y\right) \\
&= \exp\left(\hat{\beta}_2\right) \exp\left(\hat{\beta}_5 y\right) \\
\end{align*}
$$
```{r}
temp <- function(x){
  exp(mod_01$coefficients[3] + mod_01$coefficients[6] * (x - 40))
}
curve(expr = temp,
      from = min(d$Age),
      to = max(d$Age),
      xlab = "Age",
      ylab = "Odds Ratio",
      main = "Occasional to No Vitamin Use, Same Sex")
```
- $\exp\left(\hat{\beta}_3 + \hat{\beta}_6y\right)$ is the estimated odds ratio
comparing participants who use vitamins regularly
to participants who don't use vitamins,
where both groups are the same sex
and were 40 years old $y$ years ago.
$$
\begin{align*}
\frac{\widehat{\text{odds}}_i}{\widehat{\text{odds}}_j} &= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_i + \hat{\beta}_2 O_i + \hat{\beta}_3 R_i + \hat{\beta}_4 M_i + \hat{\beta}_5 Y_i O_i + \hat{\beta}_6 Y_i R_i\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 Y_j + \hat{\beta}_2 O_j + \hat{\beta}_3 R_j + \hat{\beta}_4 M_j + \hat{\beta}_5 Y_j O_j + \hat{\beta}_6 Y_j R_j\right)} \\
&= \frac{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 y + \hat{\beta}_2 (0) + \hat{\beta}_3 (1) + \hat{\beta}_4 M_i + \hat{\beta}_5 y (0) + \hat{\beta}_6 y (1)\right)}{\exp\left(\hat{\beta}_0 + \hat{\beta}_1 y + \hat{\beta}_2 (0) + \hat{\beta}_3 (0) + \hat{\beta}_4 M_i + \hat{\beta}_5 y (0) + \hat{\beta}_6 y (0)\right)} \\
&= \exp\left(\hat{\beta}_3 + \hat{\beta}_6 y\right) \\
&= \exp\left(\hat{\beta}_3\right) \exp\left(\hat{\beta}_6 y\right) \\
\end{align*}
$$
```{r}
temp <- function(x){
  exp(mod_01$coefficients[4] + mod_01$coefficients[7] * (x - 40))
}
curve(expr = temp,
      from = min(d$Age),
      to = max(d$Age),
      xlab = "Age",
      ylab = "Odds Ratio",
      main = "Regular to No Vitamin Use, Same Sex")
abline(h = 1,
       lty = 2)
```


## Problems

### Problem 1



::: {.callout-tip title="Solution" collapse="true"}




:::

### Problem 2



::: {.callout-tip title="Solution" collapse="true"}



:::




