---
title: "Calibration and ROC"
format: html
---

```{r}
# Load packages
library(probably)
library(ggplot2)
library(pROC)

# Set seed
set.seed(6)
```

## Problem 1



### Part (a)

In the chunk below, I've generated data according to the assumptions
of linear regression.

```{r}
n <- 1e3
x <- rnorm(n)
beta_0 <- 0
beta_1 <- -2
sigma <- 1
epsilon <- rnorm(n = n,
                 sd = sigma)
y <- beta_0 + beta_1 * x + epsilon
```

Use `lm` to regress `y` on `x`. Then, create a scatterplot of the residuals
as a function of `x`. Add a horizontal line at zero.

```{r}

```


### Part (b)

In the chunk below, I've generated new data. I've kept everything the same
except for $Var(Y_i|X_i) = \sigma^2$. I've greatly increased the variance
of the random error.

```{r}
epsilon_new <- rnorm(n = n,
                     sd = 10 * sigma)
y_new <- beta_0 + beta_1 * x + epsilon_new
```

Use `lm` to regress `y_new` on `x`. Then, create a scatterplot of the residuals
as a function of `x`. Add a horizontal line at zero.

```{r}

```




### Part (c)

What is the $R^2$ for the models in Part (b) and Part (c)?

```{r}

```

For the remaining parts of the problem, we will try to see an analog of this
phenomenon with a binary outcome.

### Part (d)

In the chunk below, I've generated a binary outcome according to the assumptions
of logistic regression. I have changed the values of any parameters
or of the explanatory variable.

```{r}
y_binary <- rbinom(n = n,
                   size = 1,
                   prob = plogis(beta_0 + beta_1 * x))
```

Regress `y_binary` on `x` using logistic regression.
Then, create a calibration plot for the model.

```{r}

```

### Part (e)

In the chunk below, I've attenuated $\beta_1$ toward zero.

```{r}
y_binary_new <- rbinom(n = n,
                       size = 1,
                       prob = plogis(beta_0 + beta_1 / 100 * x))
```

Why did I do this? Now, the expected values of the observations are
closer to 0.5.

```{r}
range(plogis(beta_0 + beta_1 * x))
range(plogis(beta_0 + beta_1 / 100 * x))
```

This means that `x` is not very predictive. Even when you know `x`,
whether you observe a 0 or a 1 is basically a coin toss.


Regress `y_binary_new` on `x` using logistic regression.
Then, create a calibration plot for the model.

```{r}

```





### Part (f)

Plot ROC curves and obtain areas under the curves for the models in Part (d)
and Part (e).

```{r}

```


## Problem 2

The purpose of this problem is to get more comfortable working with fitted
values. Make sure to install the package `aplore3` and then run
the chunk below. You don't actually need to load the package `aplore3`.

```{r}
# Load data
d <- aplore3::lowbwt
```

You can use the code `?lowbwt` to read more about this data set.
I find it very annoying to work with factors, so I'm going to create a new
outcome variable consisting of 1's (for those rows that have a low birth weight)
and 0's (for those rows that do not have a low birth weight).

```{r}
# Get outcome of 1's and 0's
d$low_bw <- as.integer(d$bwt < 2500)
```

### Part (a)

Use *linear* regression to regress low birth weight on age,
weight of the mother at last menstrual period, smoking status during
pregnancy, and history of premature labor.

```{r}

```

### Part (b)

Using your coefficient estimates from Part (a), obtain the fitted value
for the 82nd row in the data set. Check what you get against
`mod_2a$fitted.values[82]`.

```{r}

```

### Part (c)

What's troublesome about the answer to Part (b)?




### Part (d)

Repeat Part (a) but use *logistic* regression.

```{r}

```


### Part (e)

Repeat Part (b) but use your coefficient estimates from Part (d).
Check what you get against `mod_2d$fitted.values[82]`.

```{r}

```


### Part (f)

Does the answer to Part (e) have the same problem as the answer to Part (b)?




























