---
title: "Calibration and ROC (Solutions)"
format: html
---

```{r}
# Load packages
library(probably)
library(ggplot2)
library(pROC)

# Set seed
set.seed(6)
```

## Problem 1



### Part (a)

In the chunk below, I've generated data according to the assumptions
of linear regression.

```{r}
n <- 1e3
x <- rnorm(n)
beta_0 <- 0
beta_1 <- -2
sigma <- 1
epsilon <- rnorm(n = n,
                 sd = sigma)
y <- beta_0 + beta_1 * x + epsilon
```

Use `lm` to regress `y` on `x`. Then, create a scatterplot of the residuals
as a function of `x`. Add a horizontal line at zero.

```{r}
# Regression
mod_1a <- lm(y ~ x)

# Plot
plot(x = x,
     y = mod_1a$residuals,
     ylab = "Residuals")
abline(h = 0,
       col = "red")
```


### Part (b)

In the chunk below, I've generated new data. I've kept everything the same
except for $Var(Y_i|X_i) = \sigma^2$. I've greatly increased the variance
of the random error.

```{r}
epsilon_new <- rnorm(n = n,
                     sd = 10 * sigma)
y_new <- beta_0 + beta_1 * x + epsilon_new
```

Use `lm` to regress `y_new` on `x`. Then, create a scatterplot of the residuals
as a function of `x`. Add a horizontal line at zero.

```{r}
# Regression
mod_1b <- lm(y_new ~ x)

# Plot
plot(x = x,
     y = mod_1b$residuals,
     ylab = "Residuals")
abline(h = 0,
       col = "red")
```


Note the difference in the scale of the residuals between the two plots.
Although both data sets meet all of our assumptions, the second one has way
more error.

### Part (c)

What is the $R^2$ for the models in Part (b) and Part (c)?

```{r}
summary(mod_1a)$r.squared
summary(mod_1b)$r.squared
```

For the remaining parts of the problem, we will try to see an analog of this
phenomenon with a binary outcome.

### Part (d)

In the chunk below, I've generated a binary outcome according to the assumptions
of logistic regression. I have changed the values of any parameters
or of the explanatory variable.

```{r}
y_binary <- rbinom(n = n,
                   size = 1,
                   prob = plogis(beta_0 + beta_1 * x))
```

Regress `y_binary` on `x` using logistic regression.
Then, create a calibration plot for the model.

```{r}
# Regression
mod_1d <- glm(formula = y_binary ~ x,
              family = binomial)

# Calibration plot
cal_plot_breaks(.data = data.frame(truth = as.factor(mod_1d$model$y_binary),
                                   estimate = mod_1d$fitted.values),
                truth = truth,
                estimate = estimate,
                event_level = "second")
```

### Part (e)

In the chunk below, I've attenuated $\beta_1$ toward zero.

```{r}
y_binary_new <- rbinom(n = n,
                       size = 1,
                       prob = plogis(beta_0 + beta_1 / 100 * x))
```

Why did I do this? Now, the expected values of the observations are
closer to 0.5.

```{r}
range(plogis(beta_0 + beta_1 * x))
range(plogis(beta_0 + beta_1 / 100 * x))
```

This means that `x` is not very predictive. Even when you know `x`,
whether you observe a 0 or a 1 is basically a coin toss.


Regress `y_binary_new` on `x` using logistic regression.
Then, create a calibration plot for the model.

```{r}
# Regression
mod_1e <- glm(formula = y_binary_new ~ x,
              family = binomial)

# Calibration plot
cal_plot_breaks(.data = data.frame(truth = as.factor(mod_1e$model$y_binary_new),
                                   estimate = mod_1e$fitted.values),
                truth = truth,
                estimate = estimate,
                event_level = "second")

# Distribution of fitted values
hist(mod_1e$fitted.values)

# Calibration plot (more bins, zoom in)
cal_plot_breaks(.data = data.frame(truth = as.factor(mod_1e$model$y_binary_new),
                                   estimate = mod_1e$fitted.values),
                truth = truth,
                estimate = estimate,
                event_level = "second",
                num_breaks = 100) +
  coord_cartesian(xlim = c(0.42,0.57))
```

I don't love the way these plots look.
As great as `cal_plot_breaks` is,
it has some limitations.
In the chunk below I've created some alternative calibration plots.

```{r}
# Part (d)
ggplot(mapping = aes(x = mod_1d$fitted.values,
                     y = y_binary)) +
  geom_point(alpha = 0.3) +
  geom_smooth() +
  geom_abline(slope = 1,
              intercept = 0,
              linetype = "dashed") +
  labs(title = "Part (d)",
       x = "Fitted Values",
       y = "Outcome")

# Part (e)
ggplot(mapping = aes(x = mod_1e$fitted.values,
                     y = y_binary_new)) +
  geom_point(alpha = 0.3) +
  geom_smooth() +
  geom_abline(slope = 1,
              intercept = 0,
              linetype = "dashed") +
  labs(title = "Part (e)",
       x = "Fitted Values",
       y = "Outcome")
```

The blue line is a smoother. You can think of it like a
[moving average](https://www.statisticshowto.com/probability-and-statistics/statistics-definitions/moving-average).
The dashed line has the same meaning as in the original calibration plots.
If the blue line falls on the dashed line, it means that the fitted
values (our estimated probabilities) are approximately equal to the true
probabilities.

Since both data sets meet our assumptions, the blue line in both plots
is approximately on the dashed line.

### Part (f)

Plot ROC curves and obtain areas under the curves for the models in Part (d)
and Part (e).

```{r}
# Part (d)
plot.roc(x = y_binary,
         predictor = mod_1d$fitted.values)
auc(response = y_binary,
    predictor = mod_1d$fitted.values)

# Part (e)
plot.roc(x = y_binary_new,
         predictor = mod_1e$fitted.values)
auc(response = y_binary_new,
    predictor = mod_1e$fitted.values)
```


## Problem 2

The purpose of this problem is to get more comfortable working with fitted
values. Make sure to install the package `aplore3` and then run
the chunk below. You don't actually need to load the package `aplore3`.

```{r}
# Load data
d <- aplore3::lowbwt
```

You can use the code `?lowbwt` to read more about this data set.
I find it very annoying to work with factors, so I'm going to create a new
outcome variable consisting of 1's (for those rows that have a low birth weight)
and 0's (for those rows that do not have a low birth weight).

```{r}
# Get outcome of 1's and 0's
d$low_bw <- as.integer(d$bwt < 2500)
```

### Part (a)

Use *linear* regression to regress low birth weight on age,
weight of the mother at last menstrual period, smoking status during
pregnancy, and history of premature labor.

```{r}
# Linear regression
mod_2a <- lm(formula = low_bw ~ age + lwt + smoke + ptl,
             data = d)
```

### Part (b)

Using your coefficient estimates from Part (a), obtain the fitted value
for the 82nd row in the data set. Check what you get against
`mod_2a$fitted.values[82]`.

```{r}
# Coefficient estimates
mod_2a$coefficients

# 82nd row
d[82, ]

# Fitted value
mod_2a$coefficients[1] + mod_2a$coefficients[2] * d$age[82] +
  mod_2a$coefficients[3] * d$lwt[82] +
  mod_2a$coefficients[4] * (d$smoke[82] == "Yes") +
  mod_2a$coefficients[5] * (d$ptl[82] == "One") +
  mod_2a$coefficients[6] * (d$ptl[82] == "Two, etc.")
mod_2a$fitted.values[82]
```

### Part (c)

What's troublesome about the answer to Part (b)?

The fitted value is negative, but we can't have negative probabilities.


### Part (d)

Repeat Part (a) but use *logistic* regression.

```{r}
# Logistic regression
mod_2d <- glm(formula = low_bw ~ age + lwt + smoke + ptl,
              data = d,
              family = binomial)
```


### Part (e)

Repeat Part (b) but use your coefficient estimates from Part (d).
Check what you get against `mod_2d$fitted.values[82]`.

```{r}
# Coefficient estimates
mod_2d$coefficients

# 82nd row
d[82, ]

# Fitted value
plogis(mod_2d$coefficients[1] + mod_2d$coefficients[2] * d$age[82] +
  mod_2d$coefficients[3] * d$lwt[82] +
  mod_2d$coefficients[4] * (d$smoke[82] == "Yes") +
  mod_2d$coefficients[5] * (d$ptl[82] == "One") +
  mod_2d$coefficients[6] * (d$ptl[82] == "Two, etc."))
mod_2d$fitted.values[82]
```


### Part (f)

Does the answer to Part (e) have the same problem as the answer to Part (b)?

No, the answer to Part (e) is between 0 and 1.


























