---
title: "Interactions in Multiple Linear Regression"
format: html
---

```{r}
# Load packages
library(readxl)
library(emmeans)
```

## Problem 1

This problem uses data from the Heart and Estrogen/progestin Replacement
Study (HERS). You can
[download the data here](https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/excel/hersdata.xls).


```{r}
# Load data
d <- read_xls(path = "hersdata.xls")
```

In the chunk below, I've regressed `LDL` on `statins` and `exercise`,
with an interaction. I'll call this Model 2.

```{r}
mod_02 <- lm(formula = LDL ~ statins * exercise,
             data = d)
summary(mod_02)
```



### Part (a)

Create a scatter plot of the residuals as a function of the fitted values.
Include a horizontal line at zero.

```{r}

```

What do you conclude?




### Part (b)

Plot a boxplot, histogram, and q-q plot of the residuals.

```{r}

```

What do you conclude?






### Part (c)

Interpret the intercept in real-world terms.





### Part (d)

Interpret the estimate for `statinsyes` in real-world terms.
Interpret the results of the hypothesis test associated with this coefficient
as well.





### Part (e)

Interpret the estimate for `exerciseyes` in real-world terms.
Interpret the results of the hypothesis test associated with this coefficient
as well.






### Part (f)

Interpret the estimate for `statinsyes:exerciseyes` in real-world terms.
Interpret the results of the hypothesis test associated with this coefficient
as well.





### Part (g)

Fit a new model (I'll call this Model 3) that's the same as Model 2
but does not have the interaction term.

```{r}

```

### Part (h)

Run this chunk. You'll need to (install and) load the package `emmeans`
for it to work.

```{r}
# emmip(object = mod_02,
#       formula = exercise ~ statins,
#       CIs = TRUE)
# emmip(object = mod_03,
#       formula = exercise ~ statins,
#       CIs = TRUE)
```
Compare and contrast the two plots.






## Problem 2

This problem also uses the HERS data.


### Part (a)

Regress `LDL` on `statins` and `nonwhite`, with an interaction.
This will be Model 4.
Create a scatter plot of the residuals as a function of the fitted values.
Include a horizontal line at zero.

```{r}

```

What do you conclude?



### Part (b)

Plot a boxplot, histogram, and q-q plot of the residuals.

```{r}

```

What do you conclude?






### Part (c)

Interpret the intercept in real-world terms.




### Part (d)

Interpret the estimate for `statinsyes` in real-world terms.
Interpret the results of the hypothesis test associated with this coefficient
as well.





### Part (e)

Interpret the estimate for `nonwhiteyes` in real-world terms.
Interpret the results of the hypothesis test associated with this coefficient
as well.






### Part (f)

Interpret the estimate for `statinsyes:nonwhiteyes` in real-world terms.
Interpret the results of the hypothesis test associated with this coefficient
as well.





### Part (g)

Fit a new model (I'll call this Model 5) that's the same as Model 4
but does not have the interaction term.

```{r}

```

### Part (h)

Run this chunk.

```{r}
# emmip(object = mod_04,
#       formula = nonwhite ~ statins,
#       CIs = TRUE)
# emmip(object = mod_05,
#       formula = nonwhite ~ statins,
#       CIs = TRUE)
```
Compare and contrast the two plots.






























