---
title: "Interactions in Multiple Linear Regression (Solutions)"
format: html
---

```{r}
# Load packages
library(readxl)
library(scales)
library(emmeans)
```

## Problem 1

This problem uses data from the Heart and Estrogen/progestin Replacement
Study (HERS). You can
[download the data here](https://regression.ucsf.edu/sites/g/files/tkssra16191/files/wysiwyg/home/data/excel/hersdata.xls).


```{r}
# Load data
d <- read_xls(path = "hersdata.xls")
```

In the chunk below, I've regressed `LDL` on `statins` and `exercise`,
with an interaction. I'll call this Model 2.

```{r}
mod_02 <- lm(formula = LDL ~ statins * exercise,
             data = d)
summary(mod_02)
```



### Part (a)

Create a scatter plot of the residuals as a function of the fitted values.
Include a horizontal line at zero.

```{r}
plot(x = mod_02$fitted.values,
     y = mod_02$residuals,
     xlab = "Fitted Values",
     ylab = "Residuals",
     col = alpha(colour = "black",
                 alpha = 0.3))
abline(h = 0,
       col = "red")
```

What do you conclude?


It appears that for each fitted value, the residuals average out to zero.
This is actually guaranteed to be the case, because we have a
"saturated" model.

It also appears that there is relatively constant vertical spread
of the residuals throughout the plot, which means the assumption of constant
variance is met.

Note that there are only four possible fitted values.
This is because there are two binary explanatory variables, yielding
four possible combinations of values, i.e., $(0,0)$, $(0,1)$, $(1,0)$,
and $(1,1)$.

Below, I've plotted a boxplot of the residuals for each possible combination
of the explanatory variables. I haven't changed the labeling of the axes
so that you can see what R does by default. This plot also supports
our assumptions of linearity and equal variance.




```{r}
boxplot(formula = mod_02$residuals ~ mod_02$model$statins * mod_02$model$exercise)
```

### Part (b)

Plot a boxplot, histogram, and q-q plot of the residuals.

```{r}
boxplot(x = mod_02$residuals,
        ylab = "Residuals")
hist(x = mod_02$residuals,
     xlab = "Residuals",
     main = "")
qqnorm(mod_02$residuals)
qqline(mod_02$residuals)
```

What do you conclude?


Based on the histogram, the residuals are approximately bell-shaped.
The boxplot reveals that the residuals are right-skewed, and the q-q plot
seems to show that the residuals deviate from normal in their right tail.
Still, the sample size is large, and my understanding is that the observations
are independent (and identically distributed), so I'm not super concerned
about this last assumption being met. Even if it isn't,
we can still interpret the p-values.



### Part (c)

Interpret the intercept in real-world terms.


This is the expected LDL cholesterol for someone who is not on statins
and who does not exercise.


### Part (d)

Interpret the estimate for `statinsyes` in real-world terms.
Interpret the results of the hypothesis test associated with this coefficient
as well.


This is the expected difference in LDL cholesterol between someone who
does use statins and someone who doesn't, where neither of these people
exercise. The fact that it's negative means that we would expect
someone who uses statins to have lower LDL cholesterol than someone who doesn't
(again, assuming neither of these people exercise).

How did I arrive at that interpretation?
Define:

- $L_i$ is the LDL cholesterol for the $i$th individual.
- $S_i = 1$ if the $i$th person uses statins and $S_i = 0$ otherwise.
- $E_i = 1$ if the $i$th person exercises and $E_i = 0$ otherwise.

Our model is
$$
L_i = \beta_0 + \beta_1 S_i + \beta_2 E_i + \beta_3 S_i E_i + \epsilon_i \text{.}
$$
We're trying to interpret $\beta_1$, which is
$$
\begin{align*}
&E\left(L_i\middle|S_i = 1,E_i = 0\right) - E\left(L_i\middle|S_i = 0,E_i = 0\right) \\
&= \beta_0 + \beta_1 (1) + \beta_2 (0) + \beta_3 (1) (0) - \left(\beta_0 + \beta_1 (0) + \beta_2 (0) + \beta_3 (0) (0)\right) \\
&= \beta_0 + \beta_1  - \left(\beta_0 \right) \\
&= \beta_1 \text{.}
\end{align*}
$$

The p-value for this coefficient is less than 0.05, so we reject the null
hypothesis and conclude it's not zero.
(I didn't tell you what alpha level to use, but 0.05 is a reasonable one.)
Thus, we can conclude that there is a statistically significant difference
in LDL cholesterol between people who use statins and people who don't,
assuming all of these people don't exercise.


### Part (e)

Interpret the estimate for `exerciseyes` in real-world terms.
Interpret the results of the hypothesis test associated with this coefficient
as well.


This is the expected difference in LDL cholesterol between someone who
does exercises and someone who doesn't, where neither of these people
use statins. The fact that it's negative means that we would expect
someone who exercises to have lower LDL cholesterol than someone who doesn't
(again, assuming neither of these people use statins).

We're trying to interpret $\beta_2$, which is
$$
\begin{align*}
&E\left(L_i\middle|S_i = 0,E_i = 1\right) - E\left(L_i\middle|S_i = 0,E_i = 0\right) \\
&= \beta_0 + \beta_1 (0) + \beta_2 (1) + \beta_3 (0) (1) - \left(\beta_0 + \beta_1 (0) + \beta_2 (0) + \beta_3 (0) (0)\right) \\
&= \beta_0 + \beta_2  - \left(\beta_0 \right) \\
&= \beta_2 \text{.}
\end{align*}
$$

The p-value for this coefficient is less than 0.05, so we reject the null
hypothesis and conclude it's not zero.
Thus, we can conclude that there is a statistically significant difference
in LDL cholesterol between people who exercise and people who don't,
assuming all of these people don't use statins.



### Part (f)

Interpret the estimate for `statinsyes:exerciseyes` in real-world terms.
Interpret the results of the hypothesis test associated with this coefficient
as well.


This is the expected difference in the expected change in LDL cholesterol
associated with using statins between people who exercise and people who don't.
Because it has the opposite sign of `statinsyes`,
we would expect the effect of statins among people who exercise to be
closer to zero than the effect of statins among people who don't exercise.

We're trying to interpret $\beta_3$, which is
$$
\begin{align*}
&E\left(L_i\middle|S_i = 1,E_i = 1\right) - E\left(L_i\middle|S_i = 0,E_i = 1\right) - \left(E\left(L_i\middle|S_i = 1,E_i = 0\right) - E\left(L_i\middle|S_i = 0,E_i = 0\right)\right) \\
&= \beta_0 + \beta_1 (1) + \beta_2 (1) + \beta_3 (1) (1) - \left(\beta_0 + \beta_1 (0) + \beta_2 (1) + \beta_3 (0) (1)\right) - \beta_1 \\
&= \beta_0 + \beta_1  + \beta_2  + \beta_3  - \left(\beta_0 + \beta_2 \right) - \beta_1 \\
&= \beta_3 \text{.}
\end{align*}
$$

The p-value is high, so we fail to reject the null hypothesis,
i.e., that this coefficient is zero. This tells us that the effect of statins
is the same regardless of exercise status.


### Part (g)

Fit a new model (I'll call this Model 3) that's the same as Model 2
but does not have the interaction term.

```{r}
mod_03 <- lm(formula = LDL ~ statins + exercise,
             data = d)
summary(mod_03)
```

### Part (h)

Run this chunk. You'll need to (install and) load the package `emmeans`
for it to work.

```{r}
emmip(object = mod_02,
      formula = exercise ~ statins,
      CIs = TRUE)
emmip(object = mod_03,
      formula = exercise ~ statins,
      CIs = TRUE)
```
Compare and contrast the two plots.


They're basically the same. In the plot for Model 2, the slopes are different
because of the interaction term. Since the interaction term is close to zero,
the slopes are not that different.

In the plot for Model 3,
the slopes are identical because there's no interaction term.



## Problem 2

This problem also uses the HERS data.


### Part (a)

Regress `LDL` on `statins` and `nonwhite`, with an interaction.
This will be Model 4.
Create a scatter plot of the residuals as a function of the fitted values.
Include a horizontal line at zero.

```{r}
mod_04 <- lm(formula = LDL ~ statins * nonwhite,
             data = d)
summary(mod_04)
plot(x = mod_04$fitted.values,
     y = mod_04$residuals,
     xlab = "Fitted Values",
     ylab = "Residuals",
     col = alpha(colour = "black",
                 alpha = 0.3))
abline(h = 0,
       col = "red")
```

What do you conclude?


It appears that for each fitted value, the residuals average out to zero.
Again, this is guaranteed because the model is saturated.

It also appears that there is relatively constant vertical spread
of the residuals throughout the plot, which means the assumption of constant
variance is met. You could argue that there is lower variance for
the fitted value between 145 and 150, but there are also fewer points
with this fitted value. If points near zero are more likely than points
far from zero, then with fewer observations we would expect to see
fewer points far from zero.

Below, I've plotted a boxplot of the residuals for each possible combination
of the explanatory variables. This plot also supports
our assumptions of linearity and equal variance.




```{r}
boxplot(formula = mod_04$residuals ~ mod_04$model$statins * mod_04$model$nonwhite)
```

### Part (b)

Plot a boxplot, histogram, and q-q plot of the residuals.

```{r}
boxplot(x = mod_04$residuals,
        ylab = "Residuals")
hist(x = mod_04$residuals,
     xlab = "Residuals",
     main = "")
qqnorm(mod_04$residuals)
qqline(mod_04$residuals)
```

What do you conclude?


Based on the histogram, the residuals are approximately bell-shaped.
The boxplot reveals that the residuals are right-skewed, and the q-q plot
seems to show that the residuals deviate from normal in their right tail.
Still, the sample size is large, and the observations
are independent (and identically distributed), so
we can still interpret the p-values.



### Part (c)

Interpret the intercept in real-world terms.


This is the expected LDL cholesterol for someone who is not on statins
and who does is white.


### Part (d)

Interpret the estimate for `statinsyes` in real-world terms.
Interpret the results of the hypothesis test associated with this coefficient
as well.


This is the expected difference in LDL cholesterol between someone who
does use statins and someone who doesn't, where both of these people are white.
The fact that it's negative means that we would expect
someone who uses statins to have lower LDL cholesterol than someone who doesn't
(again, assuming both of these people are white).

Define:

- $N_i = 1$ if the $i$th person is non-white and $N_i = 0$ otherwise.

Our model is
$$
L_i = \beta_0 + \beta_1 S_i + \beta_2 N_i + \beta_3 S_i N_i + \epsilon_i \text{.}
$$
We're trying to interpret $\beta_1$, which is
$$
\begin{align*}
&E\left(L_i\middle|S_i = 1,N_i = 0\right) - E\left(L_i\middle|S_i = 0,N_i = 0\right) \\
&= \beta_0 + \beta_1 (1) + \beta_2 (0) + \beta_3 (1) (0) - \left(\beta_0 + \beta_1 (0) + \beta_2 (0) + \beta_3 (0) (0)\right) \\
&= \beta_0 + \beta_1  - \left(\beta_0 \right) \\
&= \beta_1 \text{.}
\end{align*}
$$

The p-value for this coefficient is less than 0.05, so we reject the null
hypothesis and conclude it's not zero.
Thus, we can conclude that there is a statistically significant difference
in LDL cholesterol between people who use statins and people who don't,
among white people.


### Part (e)

Interpret the estimate for `nonwhiteyes` in real-world terms.
Interpret the results of the hypothesis test associated with this coefficient
as well.


This is the expected difference in LDL cholesterol between someone who
does is not white and someone who is, where neither of these people
use statins. The fact that it's positive means that we would expect
someone who is not white to have higher LDL cholesterol than someone who is white
(again, assuming neither of these people use statins).

We're trying to interpret $\beta_2$, which is
$$
\begin{align*}
&E\left(L_i\middle|S_i = 0,N_i = 1\right) - E\left(L_i\middle|S_i = 0,N_i = 0\right) \\
&= \beta_0 + \beta_1 (0) + \beta_2 (1) + \beta_3 (0) (1) - \left(\beta_0 + \beta_1 (0) + \beta_2 (0) + \beta_3 (0) (0)\right) \\
&= \beta_0 + \beta_2  - \left(\beta_0 \right) \\
&= \beta_2 \text{.}
\end{align*}
$$

The p-value for this coefficient is greater than 0.05, so we cannot reject the null
hypothesis.
Thus, there is not a statistically significant difference
in LDL cholesterol between people who are not white and people who are,
assuming all of these people don't use statins.



### Part (f)

Interpret the estimate for `statinsyes:nonwhiteyes` in real-world terms.
Interpret the results of the hypothesis test associated with this coefficient
as well.


This is the expected difference in the expected change in LDL cholesterol
associated with using statins between people who are not white and people who are.
Because it has the opposite sign of \verb|statinsyes|,
we would expect the effect of statins among nonwhite people to be
closer to zero than the effect of statins among people who are white.

We're trying to interpret $\beta_3$, which is
$$
\begin{align*}
&E\left(L_i\middle|S_i = 1,N_i = 1\right) - E\left(L_i\middle|S_i = 0,N_i = 1\right) - \left(E\left(L_i\middle|S_i = 1,N_i = 0\right) - E\left(L_i\middle|S_i = 0,N_i = 0\right)\right) \\
&= \beta_0 + \beta_1 (1) + \beta_2 (1) + \beta_3 (1) (1) - \left(\beta_0 + \beta_1 (0) + \beta_2 (1) + \beta_3 (0) (1)\right) - \beta_1 \\
&= \beta_0 + \beta_1  + \beta_2  + \beta_3  - \left(\beta_0 + \beta_2 \right) - \beta_1 \\
&= \beta_3 \text{.}
\end{align*}
$$

The p-value is low, so we reject the null hypothesis,
i.e., that this coefficient is zero. This tells us that the effect of statins
depends on race.


### Part (g)

Fit a new model (I'll call this Model 5) that's the same as Model 4
but does not have the interaction term.

```{r}
mod_05 <- lm(formula = LDL ~ statins + nonwhite,
             data = d)
summary(mod_05)
```

### Part (h)

Run this chunk.

```{r}
emmip(object = mod_04,
      formula = nonwhite ~ statins,
      CIs = TRUE)
emmip(object = mod_05,
      formula = nonwhite ~ statins,
      CIs = TRUE)
```
Compare and contrast the two plots.


In the plot for Model 4, the slopes are different
because of the interaction term.

In the plot for Model 5,
the slopes are identical because there's no interaction term.



























