---
title: "Model Selection (Solutions)"
format: html
---

```{r}
# Load packages
library(readxl)
```

This problem uses data from
"[Association between particulate air pollution exposure during pregnancy and postpartum maternal psychological functioning](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0195267)"
(Sheffield et al., 2018).
[This is a direct link to the data](https://doi.org/10.1371/journal.pone.0195267.s001).
Download the data and put the file in the same folder as this Quarto document.

```{r}
# Load data
a <- read_xls("pone.0195267.s001.xls")

# Remove the first column that R added for some unknown reason
a$...1 <- NULL
```

The purpose of this problem is to demonstrate model selection when the goal
is prediction. We will try to predict `pmw1`, which is the "particulate matter
(PM2.5) weekly average from" week 1 of gestation.
(I'm quoting the Data Dictionary in sheet 2 of the Excel file.)
We'll allow ourselves to use any variable in the data set from
`mat_age` through `lookfwd612`. In other words, we're excluding from
consideration `random_id` and all the `pmw` variables.
Just to make that more clear, I'm going to remove anything that's not our outcome
or an eligible explanatory variable.

```{r}
a$randomid <- NULL
exclude <- paste0("pmw",
                  2:40)
a <- a[ ,!(colnames(a) %in% exclude)]
```


### Part (a)

If our goal is prediction, we should evaluate our model based on how well
it predicts new data. So, we should randomly split our data into a training
set and a test set. We can fit our models on the training set and test
them on the test set.

Randomly select 80% of the data set and store it
as a training set. Store the remaining 20% as a test set.

```{r}
n <- nrow(a)
n_train <- round(0.8 * n)
id_train <- sort(sample(x = n,
                        size = n_train))
a_train <- a[id_train, ]
a_test <- a[!(1:n %in% id_train), ]
```


### Part (b)

Use the training set to regress the outcome on only those variables
that you think would be predictive. Note that "predictive" is not the same
as "causal". Many of the explanatory variables were measured after
the outcome, so they could not have caused the outcome. However, they still
might be predictive.

I'll call this model the "judicious" model.

```{r}
# My own personal model
mod_jud <- lm(formula = pmw1 ~ hs_educ + race4 + smoke_preg2 + seasoncb + 
                worry06 + diffslp06,
              data = a_train)
```

### Part (c)

Use the training set to regress the outcome on all candidate variables.
I'll call this the "full" model.

```{r}
mod_full <- lm(formula = pmw1 ~ .,
               data = a_train)
```

### Part (d)

For this part of the problem, I'd like you to use backward selection.
If you try to do this with the full model, R will protest once it removes
a couple variables and realizes that it can include more rows for smaller
models. This is because some of the rows are missing data, so they can't
be included in the full model, but can be included in any model that doesn't
use the variables for which they're missing values.

So, regress the outcome on all possible explanatory variables in the training
set *after removing all rows with missing data*. Then, run backward selection.
I'll call this the "backward" model.

You might want to use `trace = 0` to prevent R from printing out all the
intermediate models in the backward selection process.

```{r}
mod_back <- step(object = lm(formula = pmw1 ~ .,
                             data = a_train[complete.cases(a_train), ]),
                 direction = "backward",
                 trace = 0)
```

### Part (e)

Order the three models (full, judicious, and backward) from best to worst
using the adjusted $R^2$.

```{r}
# Adjusted R^2
summary(mod_jud)$adj.r.squared
summary(mod_full)$adj.r.squared
summary(mod_back)$adj.r.squared
```

The best model is the backward model, the worst model is the
full model, and the judicious model is in the middle.

Do the same using the AIC.

```{r}
# AIC
AIC(mod_jud)
AIC(mod_full)
AIC(mod_back)
```


This was a trick! We can't compare these models using the AIC
because they were fit on different data sets. We didn't use the same rows
in the data set for each because some rows were missing data.

### Part (f)

Use `predict` to get predictions of the outcome for the test set.
Then, find the root mean squared error for each model in the test set.
Rank the models from best to worst using the RMSE.

```{r}
# Get predictions for test data
pred_full <- predict(object = mod_full,
                     newdata = a_test)
pred_jud <- predict(object = mod_jud,
                    newdata = a_test)
pred_back <- predict(object = mod_back,
                     newdata = a_test)

# Get error
sqrt(mean(x = (a_test$pmw1 - pred_jud)^2,
          na.rm = TRUE))
sqrt(mean(x = (a_test$pmw1 - pred_full)^2,
          na.rm = TRUE))
sqrt(mean(x = (a_test$pmw1 - pred_back)^2,
          na.rm = TRUE))
```


The best model is the backward model,
the worst model is the judicious model, and the full model is somewhere in
between. Actually, the full model is pretty close to the backward model.

Perhaps we've identified the utility of backward selection.
When you have lots of candidate explanatory variables to choose from,
and all you care about is prediction, backward selection might be a fast
way to eliminate variables. Still, I'm not sure how it compares to other
variable selection methods like LASSO, which is beyond the scope of this class.







