---
title: "Residuals in Poisson Regression"
format: html
---

## Learning Objective

State that there are different types of residuals for Poisson regression.


## Lesson

Residuals played a big role in linear regression,
but I barely mentioned them for logistic regression.
One of the reasons for that is that we can write the assumptions
of linear regression in terms of random error, i.e., $\epsilon_i$,
but we can't do the same for logistic regression.
We can't write the assumptions of Poisson regression in terms
of random error, either.
Still, it's worth taking a look at the residuals to get a more visceral
sense of why they can be less helpful for regression that isn't linear
regression.

I'm going to start by simulating data according to the assumptions
of Poisson regression.
To simplify things, I'm just going to assume that $t_i = 1$
for every $i$.
```{r}
# Set random seed to make this reproducible
set.seed(3)

# Parameters
n <- 1e3
beta_0 <- 0
beta_1 <- 0.1

# Generate explanatory variable
x <- rnorm(n)

# Generate outcome variable
mu <- exp(beta_0 + beta_1 * x)
y <- rpois(n = n,
           lambda = mu)
```

Now I'm going to perform Poisson regression.
I know all the assumptions are met because I simulated the data that way.
```{r}
# Poisson regression
mod_01 <- glm(formula = y ~ x,
              family = poisson)
summary(mod_01)
```

Now let's talk about residuals.
With Poisson regression, logistic regression, and any form of regression
that uses the `glm` function,
there are many types of residuals.
This is a partial list:

- Response
  - The $i$th residual is $Y_i - \hat{Y}_i$.
  - These are what we know from linear regression.
  - They don't work as well for logistic and Poisson regression
  because logistic and Poisson regression have restrictions on what
  values the outcome can take.
  These restrictions on the outcome create restrictions on the residuals.
  The other types of residuals are attempts at addressing these restrictions
  and making residuals more useful when the outcome is not simply continuous.
- Working
  - The $i$th residual is $\displaystyle \left(\frac{d}{d\mu_i} \left(\beta_0 + \beta_1 X_{i1} +\cdots + \beta_p X_{ip}\right)\right) \left(Y_i - \hat{Y}_i\right)$.
  - These are what you get from `mod_01$residuals` if `mod_01` is the result of a `glm` call.
- Deviance
  - The $i$th residual is $\text{sign}(Y_i - \hat{Y}_i) \sqrt{2 \left(\ell(Y_i,Y_i) - \ell(Y_i,\hat{Y}_i)\right)}$.
  The $\ell$'s are used to denote log-likelihoods for different models.
  - These are what you get from `residuals(mod_01)` if `mod_01` is the result of a `glm` call.
- Pearson
  - The $i$th residual is $\displaystyle \frac{Y_i - \hat{Y}_i}{\sqrt{\widehat{Var}(Y_i)}}$.
  - This is a residual that we have standardized to have variance 1.

Let's look at plots of these different types of residuals for our Poisson
regression model.
Remember: we know the assumptions of Poisson regression are met for these
data.

```{r}
res_res <- residuals(object = mod_01,
                     type = "response")
res_wor <- residuals(object = mod_01,
                     type = "working")
res_dev <- residuals(object = mod_01,
                     type = "deviance")
res_pea <- residuals(object = mod_01,
                     type = "pearson")

# Scatterplot
plot(x = x,
     y = res_res,
     ylab = "Residual",
     main = "Response")
abline(h = 0,
       col = "red")
plot(x = x,
     y = res_wor,
     ylab = "Residual",
     main = "Working")
abline(h = 0,
       col = "red")
plot(x = x,
     y = res_dev,
     ylab = "Residual",
     main = "Deviance")
abline(h = 0,
       col = "red")
plot(x = x,
     y = res_pea,
     ylab = "Residual",
     main = "Pearson")
abline(h = 0,
       col = "red")

# Boxplot
boxplot(x = res_res,
        main = "Response")
boxplot(x = res_wor,
        main = "Working")
boxplot(x = res_dev,
        main = "Deviance")
boxplot(x = res_pea,
        main = "Pearson")

# Histogram
hist(x = res_res,
     xlab = "Residual",
     main = "Response")
hist(x = res_wor,
     xlab = "Residual",
     main = "Working")
hist(x = res_dev,
     xlab = "Residual",
     main = "Deviance")
hist(x = res_pea,
     xlab = "Residual",
     main = "Pearson")

# Q-Q Plot
qqnorm(y = res_res,
       main = "Response")
qqline(y = res_res)
qqnorm(y = res_wor,
       main = "Working")
qqline(y = res_wor)
qqnorm(y = res_dev,
       main = "Deviance")
qqline(y = res_dev)
qqnorm(y = res_pea,
       main = "Pearson")
qqline(y = res_pea)
```

It's hard to know what we should be checking for here.
None of the types of residuals look normal here,
that's for sure.

The whole purpose of this exercise is to demonstrate that there
are multiple types of residuals when it comes to Poisson regression,
but they aren't necessarily useful for checking our assumptions.






