---
title: "Interpretation"
execute:
  echo: false
format: html
---

```{r}
#| warning: false
library(dplyr)
```

## Learning Objective

Interpret the coefficients in Poisson regression.


## Lesson

### Intercept

Recall that, in Poisson regression, we assume
$$
\log\lambda_i = \beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip} \text{.}
$$

What happens when all the covariates equal zero?
$$
\begin{align*}
\log\lambda_i &= \beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip} \\
\log\lambda_i &= \beta_0 + \beta_1 (0) + \cdots + \beta_p (0) \\
\log\lambda_i &= \beta_0 \\
\lambda_i &= \exp(\beta_0) \\
\end{align*}
$$

So, $\exp(\beta_0)$ is the rate when all the covariates
equal zero.

### Non-Intercept Coefficients

How do we interpret the non-intercept coefficients?
Suppose observation $i$ has $X_{i1} = x$ and observation $j$ has $X_{j1} = x + 1$.
For each $k \ne 1$, $X_{ik} = X_{jk}$.
In words, the two observations are identical except for the first covariate,
where the $j$th observation exceeds the $i$th observation by one unit.
Then
$$
\begin{align*}
\log \frac{\lambda_j}{\lambda_i} &= \log \frac{p_j}{1-p_j} - \log \frac{p_i}{1-p_i} \\
&= \beta_0 + \beta_1 X_{j1} + \cdots + \beta_p X_{jp} - \left(\beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip}\right) \\
&= \beta_0 + \beta_1 (x+1) + \cdots + \beta_p X_{jp} - \left(\beta_0 + \beta_1 x + \cdots + \beta_p X_{ip}\right) \\
&= \beta_0 + \beta_1 x + \beta_1 + \cdots + \beta_p X_{jp} - \left(\beta_0 + \beta_1 x + \cdots + \beta_p X_{ip}\right) \\
&= \beta_1 \text{.}
\end{align*}
$$
By exponentiating both sides of this equation, we see
$$
\begin{align*}
\log \frac{\lambda_j}{\lambda_i} &= \beta_1 \\
\frac{\lambda_j}{\lambda_i} &= \exp(\beta_1) \text{.} \\
\end{align*}
$$

In words, $\exp(\beta_1)$ is the rate ratio comparing two populations
that are identical except that one exceeds the other by one unit
on the first covariate.
If $\exp(\beta_1) > 1$,
the rate is greater in the population
with the higher value of the first covariate;
if $\exp(\beta_1) < 1$,
the rate is lesser in the population
with the higher value of the first covariate;
if $\exp(\beta_1) = 1$,
the rate is the same in both populations.

### Example

```{r}
age_group <- rep(x = c("<30",
                       "30-34",
                       "35-39",
                       "40-44",
                       "45-49",
                       "50-54",
                       "55-59",
                       "60-64",
                       "65-69",
                       "70-74",
                       "75-79",
                       "80+"),
                 times = 4)
smoke_group <- rep(x = c("never_smoked",
                         "cigar_and_pipe_only",
                         "cigarette_and_other",
                         "cigarette_only"),
                   each = 12)
population <- c(48,
                739,
                1113,
                656,
                359,
                249,
                632,
                1067,
                897,
                668,
                361,
                274,
                6,
                162,
                187,
                145,
                104,
                98,
                372,
                846,
                949,
                824,
                667,
                537,
                258,
                4596,
                6653,
                4531,
                3030,
                2267,
                4682,
                6052,
                3880,
                2033,
                871,
                345,
                214,
                3185,
                4639,
                3410,
                2239,
                1851,
                3270,
                3791,
                2421,
                1195,
                436,
                113)
deaths <- c(1,
            12,
            14,
            18,
            22,
            19,
            55,
            117,
            170,
            179,
            120,
            128,
            0,
            0,
            1,
            2,
            4,
            3,
            38,
            113,
            173,
            212,
            243,
            253,
            2,
            61,
            121,
            149,
            169,
            193,
            576,
            1001,
            901,
            613,
            337,
            189,
            4,
            56,
            88,
            124,
            140,
            187,
            514,
            778,
            689,
            432,
            214,
            63)
d <- data.frame(age_group,
                smoke_group,
                population,
                deaths)
```

This data set comes from
[A Canadian Study of Smoking and Health--Second Report](https://www.jstor.org/stable/41983444).
These are the variables:

- `age_group`: Age group, with the following levels:
  - 40-44
  - 45-49
  - 50-54
  - 55-59
  - 60-64
  - 65-69
  - 70-74
  - 75-79
  - 80+
- `smoke_group`: Smoking status, which has the following levels:
  - Never smoked
  - Cigar and pipe only
  - Cigarette and other
  - Cigarette only
- `population`: The number of people in that age group who had
that smoking status, at the start of the six-year follow-up period.
- `deaths`: The number of people in that age group who had that
smoking status and who died by the end of the six-year
follow-up period.

Each row corresponds to a unique combination of `age_group`
and `smoke_group`.

These are the modeling decisions I've made for this data set:

- Our outcome will be `deaths`.
Note that this is a count, which is appropriate for the Poisson distribution.
- The offset will be `log(population / 1000)`.
This means we will be modeling the expected number of deaths
per thousand people.
- If I don't make any changes,
the reference age group will be `<30` because it's first alphabetically.
This is not a great idea because it's the smallest age group.
```{r}
#| echo: true
tapply(X = d$population,
       INDEX = d$age_group,
       FUN = sum)
```
I'm going to make `35-39` the reference group because it's
the largest age group.
```{r}
#| echo: true
d$age_group <- relevel(x = as.factor(d$age_group),
                       ref = "35-39")
```
- If I don't make any changes,
the reference smoke group will be `cigar_and_pipe_only`
because it's first alphabetically.
This is not a great idea because it's the smallest smoke group.
```{r}
#| echo: true
tapply(X = d$population,
       INDEX = d$smoke_group,
       FUN = sum)
```
I'm going to make `never_smoked` the reference group because
it's a larger smoke group and because
it's logical to want to compare different forms of smoking
to never having smoked
```{r}
#| echo: true
d$smoke_group <- relevel(x = as.factor(d$smoke_group),
                         ref = "never_smoked")
```
- I'm going to regress the outcome on `age_group` and `smoke_group`
without an interaction term.
I am primarily concerned with the association between smoke group
and deaths, controlling for age.
I'm willing to assume (at least for now)
that the association between smoke group and deaths
is the same, regardless of age group.


```{r}
#| echo: true
mod_01 <- glm(formula = deaths ~ age_group + smoke_group + offset(log(population / 1000)),
              family = poisson,
              data = d)
summary(mod_01)
```

How do we interpret these coefficients?
They're actually meaningless in their current form.
We need to exponentiate them first.

```{r}
#| echo: true
exp(mod_01$coefficients)
```

Now we can interpret them.

- 13.6261893 is the estimated number of deaths per 1,000 people
among 35-39 year-olds who have never smoked.
- Holding smoking group constant,
the estimated number of deaths per 1,000 people under 30
is 0.7422365 times the estimated number of deaths per 1,000 people
age 35-39.
- Holding smoking group constant,
the estimated number of deaths per 1,000 people age 30-34
is 0.8355223 times the estimated number of deaths per 1,000 people
age 35-39.
- Holding smoking group constant,
the estimated number of deaths per 1,000 people age 40-44
is 1.8718787 times the estimated number of deaths per 1,000 people
age 35-39.
- Holding smoking group constant,
the estimated number of deaths per 1,000 people age 45-49
is 3.2573985 times the estimated number of deaths per 1,000 people
age 35-39.
- Holding smoking group constant,
the estimated number of deaths per 1,000 people age 50-54
is 4.9906857 times the estimated number of deaths per 1,000 people
age 35-39.
- Holding smoking group constant,
the estimated number of deaths per 1,000 people age 55-59
is 7.4373868 times the estimated number of deaths per 1,000 people
age 35-39.
- Holding smoking group constant,
the estimated number of deaths per 1,000 people age 60-64
is 9.7883901 times the estimated number of deaths per 1,000 people
age 35-39.
- Holding smoking group constant,
the estimated number of deaths per 1,000 people age 65-69
is 13.8070502 times the estimated number of deaths per 1,000 people
age 35-39.
- Holding smoking group constant,
the estimated number of deaths per 1,000 people age 70-74
is 18.1448406 times the estimated number of deaths per 1,000 people
age 35-39.
- Holding smoking group constant,
the estimated number of deaths per 1,000 people age 75-79
is 24.1906193 times the estimated number of deaths per 1,000 people
age 35-39.
- Holding smoking group constant,
the estimated number of deaths per 1,000 people age 80 and above
is 32.6918510 times the estimated number of deaths per 1,000 people
age 35-39.
- Holding age group constant,
the estimated number of deaths per 1,000 people who
have only smoked cigars or pipes
is 1.0303389 times the estimated number of deaths per 1,000 people
who have never smoked.
- Holding age group constant,
the estimated number of deaths per 1,000 people who
have smoked cigarettes and other things
is 1.2302698 times the estimated number of deaths per 1,000 people
who have never smoked.
- Holding age group constant,
the estimated number of deaths per 1,000 people who
have only smoked cigarettes
is 1.4977956 times the estimated number of deaths per 1,000 people
who have never smoked.




## Problems

### Problem 1

```{r}
d2 <- read.table(file = "https://grodri.github.io/datasets/ships.dat",
                 header = TRUE)
```

This data set comes from page 204 of
McCullagh and Nelder's 1989 book
*Generalized Linear Models*, published by Chapman and Hall.
I downloaded it from
[Germán Rodríguez's website](https://grodri.github.io/glms/datasets/#ship).
These are the variables:

- `type`: Ship type, with five levels: A, B, C, D, and E.
- `construction`: Year of construction, with four levels:
  - `1960-64`
  - `1965-69`
  - `1970-74`
  - `1975-79`
- `operation`: Period of operation, with two levels:
  - `1960-74`
  - `1975-79`
- `months`: Months of service.
- `damage`: Number of damage incidents.

Each row of the data set corresponds to a unique combination
of ship type,
year of construction, and period of operation.

```{r}
mod_02 <- glm(formula = damage ~ type + construction + operation + offset(log(months / 12)),
              family = poisson,
              data = d2)
summary(mod_02)
```

Interpret the exponentiated coefficients.
```{r}
#| echo: true
exp(mod_02$coefficients)
```

::: {.callout-tip title="Solution" collapse="true"}

- 0.01982136 is the estimated number of damage incidents per year
for a type A ship constructed between 1960 and 1964
and operating between 1960 and 1974.
- Holding construction period and operation period constant,
the estimated number of damage incidents per year
for type B ships is 0.58080262 times
the estimated number of damage incidents per year
for type A ships.
- Holding construction period and operation period constant,
the estimated number of damage incidents per year
for type C ships is 0.50288104 times
the estimated number of damage incidents per year
for type A ships.
- Holding construction period and operation period constant,
the estimated number of damage incidents per year
for type D ships is 0.92685196 times
the estimated number of damage incidents per year
for type A ships.
- Holding construction period and operation period constant,
the estimated number of damage incidents per year
for type E ships is 1.38483286 times
the estimated number of damage incidents per year
for type A ships.
- Holding ship type and operation period constant,
the estimated number of damage incidents per year
for ships constructed between 1965 and 1969 is 2.00800246 times
the estimated number of damage incidents per year
for ships constructed between 1960 and 1964.
- Holding ship type and operation period constant,
the estimated number of damage incidents per year
for ships constructed between 1970 and 1974 is 2.26693019 times
the estimated number of damage incidents per year
for ships constructed between 1960 and 1964.
- Holding ship type and operation period constant,
the estimated number of damage incidents per year
for ships constructed between 1975 and 1979 is 1.57369544 times
the estimated number of damage incidents per year
for ships constructed between 1960 and 1964.
- Holding ship type and construction period constant,
the estimated number of damage incidents per year
for ships operating between 1975 and 1979 is 1.46883116 times
the estimated number of damage incidents per year
for ships operating between 1960 and 1974.



:::

### Problem 2

```{r}
d3 <- read.table(file = "https://grodri.github.io/datasets/cancer.dat",
                 header = TRUE)
d3$malignant <- recode_values(x = d3$malignant,
                              "no" ~ 0,
                              "yes" ~ 1)
d3$population <- d3$yes + d3$no
d3$deaths <- d3$no
d3$yes <- NULL
d3$no <- NULL
```

This data set also comes from
[Germán Rodríguez's website](https://grodri.github.io/glms/datasets/#cancer).
It originally appeared in
Bishop, Fienberg, and Holland's 1975 book
*Discrete Multivariate Analysis: Theory and Practice*,
published by MIT Press.
The participants in this study were breast cancer patients.
The variables are:

- `age`: Age group, with three levels:
  - `<50`
  - `50-69`
  - `70+`
- `malignant`: 1 for yes and 0 for no
- `population`: The number of patients at baseline for that
age group and that malignancy status.
- `deaths`: The number of deaths after three years.

#### Part (a)

```{r}
mod_03 <- glm(formula = deaths ~ age + malignant + offset(log(population / 10)),
              family = poisson,
              data = d3)
summary(mod_03)
```

Intepret the exponentiated coefficients.
```{r}
#| echo: true
exp(mod_03$coefficients)
```

::: {.callout-tip title="Solution" collapse="true"}

- 1.149412 is the expected number of deaths per ten
patients under 50 with non-malignant breast cancer.
- The expected number of deaths per ten patients
age 50-69 is 1.639543 times
the expected number of deaths per ten patients
under 50,
holding malignancy constant.
- The expected number of deaths per ten patients
age 70 and above is 2.015077 times
the expected number of deaths per ten patients
under 50,
holding malignancy constant.
- The expected number of deaths per ten patients
with malignant cancer is 1.767224 times
the expected number of deaths per ten patients
with non-malignant cancer,
holding age group constant.

:::

#### Part (b)

If I fit this model, how will the exponentiated coefficients change from part (a)?
```{r}
#| echo: true
mod_04 <- glm(formula = deaths ~ age + malignant + offset(log(population)),
              family = poisson,
              data = d3)
```
::: {.callout-tip title="Solution" collapse="true"}

Only the exponentiated intercept will change.
It will be the estimated number of deaths per person
in the reference group,
rather than per ten people,
so it will be one-tenth of the value for part (a).

Since the non-intercept exponentiated coefficients
represent rate ratios,
using per person or per ten people won't change them.

```{r}
#| echo: true
exp(mod_03$coefficients)
exp(mod_04$coefficients)
```

:::


