---
title: "Hypothesis Testing"
execute:
  echo: false
format: html
---

```{r}
#| warning: false
library(dplyr)
```

## Learning Objective

Interpret the results of a hypothesis test in the context of Poisson
regression.


## Lesson

### Interpretation in General

As in linear and logistic regression, R will conduct a hypothesis test
for each coefficient estimated in Poisson regression.
The null hypothesis for each coefficient is that that coefficient is zero,
and the alternative hypothesis is that the coefficient is not zero.
It's always a two-sided test.

What does it mean if the coefficient is zero?
The answer to this is different if we're talking about the intercept
versus a non-intercept coefficient.
Let's start with the intercept.
Recall that $\exp(\beta_0)$ is the expected rate
when all the covariates are equal to zero.
If $\beta_0 = 0$, then $\exp(\beta_0) = 1$.
So, the null hypothesis for the intercept is that the expected rate is 1.
Whether or not this is meaningful depends on your goals.

What about for the non-intercept coefficients?
For $j \ne 0$, $\exp(\beta_j)$ is the rate ratio comparing
two populations that are identical except one has $X_{ij} = x + 1$
and the other has $X_{kj} = x$.
Again, if $\beta_j = 0$, then $\exp(\beta_j) = 1$.
If a rate ratio is 1,
then the two rates it's comparing are equal to each other.
So, the null hypothesis corresponds to equal rates for these two populations.
If the p-value is small, we reject the null
and conclude that these two populations have different rates.

### Example

```{r}
age_group <- rep(x = c("<30",
                       "30-34",
                       "35-39",
                       "40-44",
                       "45-49",
                       "50-54",
                       "55-59",
                       "60-64",
                       "65-69",
                       "70-74",
                       "75-79",
                       "80+"),
                 times = 4)
smoke_group <- rep(x = c("never_smoked",
                         "cigar_and_pipe_only",
                         "cigarette_and_other",
                         "cigarette_only"),
                   each = 12)
population <- c(48,
                739,
                1113,
                656,
                359,
                249,
                632,
                1067,
                897,
                668,
                361,
                274,
                6,
                162,
                187,
                145,
                104,
                98,
                372,
                846,
                949,
                824,
                667,
                537,
                258,
                4596,
                6653,
                4531,
                3030,
                2267,
                4682,
                6052,
                3880,
                2033,
                871,
                345,
                214,
                3185,
                4639,
                3410,
                2239,
                1851,
                3270,
                3791,
                2421,
                1195,
                436,
                113)
deaths <- c(1,
            12,
            14,
            18,
            22,
            19,
            55,
            117,
            170,
            179,
            120,
            128,
            0,
            0,
            1,
            2,
            4,
            3,
            38,
            113,
            173,
            212,
            243,
            253,
            2,
            61,
            121,
            149,
            169,
            193,
            576,
            1001,
            901,
            613,
            337,
            189,
            4,
            56,
            88,
            124,
            140,
            187,
            514,
            778,
            689,
            432,
            214,
            63)
d <- data.frame(age_group,
                smoke_group,
                population,
                deaths)
d$age_group <- relevel(x = as.factor(d$age_group),
                       ref = "35-39")
d$smoke_group <- relevel(x = as.factor(d$smoke_group),
                         ref = "never_smoked")
```

Let's return to the data from
[A Canadian Study of Smoking and Health--Second Report](https://www.jstor.org/stable/41983444).
These are the variables:

- `age_group`: Age group, with the following levels:
  - 40-44
  - 45-49
  - 50-54
  - 55-59
  - 60-64
  - 65-69
  - 70-74
  - 75-79
  - 80+
- `smoke_group`: Smoking status, which has the following levels:
  - Never smoked
  - Cigar and pipe only
  - Cigarette and other
  - Cigarette only
- `population`: The number of people in that age group who had
that smoking status, at the start of the six-year follow-up period.
- `deaths`: The number of people in that age group who had that
smoking status and who died by the end of the six-year
follow-up period.

These are the results:
```{r}
mod_01 <- glm(formula = deaths ~ age_group + smoke_group + offset(log(population / 1000)),
              family = poisson,
              data = d)
summary(mod_01)
```
How do we interpret the hypothesis tests?
Since we have quite a few coefficients being estimated,
and thus quite a few hypothesis tests being conducted,
I'm going to use a more conservative threshold for the p-values.
Since we're estimating 15 coefficients,
I'm going to use a threshold of $0.05 / 15 = 0.00333$.
(This is called a Bonferroni correction.)

- Since the p-value for the intercept is less than 0.00333,
we can conclude that the expected number of deaths per 1,000 35-39 year-olds
who have never smoked is different from 1.
- Since the p-value for `age_group<30` is greater than 0.00333,
we cannot conclude that
the expected number of deaths per 1,000 people under 30
is different than the expected number of deaths per 1,000 people
age 35-39,
holding smoking group constant.
- Since the p-value for `age_group30-34` is greater than 0.00333,
we cannot conclude that
the expected number of deaths per 1,000 people age 30-34
is different than the expected number of deaths per 1,000 people
age 35-39,
holding smoking group constant.
- Since the p-value for `age_group40-44` is less than 0.00333,
we can conclude that
the expected number of deaths per 1,000 people age 40-44
is different than the expected number of deaths per 1,000 people
age 35-39,
holding smoking group constant.
- Since the p-value for `age_group45-49` is less than 0.00333,
we can conclude that
the expected number of deaths per 1,000 people age 45-49
is different than the expected number of deaths per 1,000 people
age 35-39,
holding smoking group constant.
- Since the p-value for `age_group50-54` is less than 0.00333,
we can conclude that
the expected number of deaths per 1,000 people age 50-54
is different than the expected number of deaths per 1,000 people
age 35-39,
holding smoking group constant.
- Since the p-value for `age_group55-59` is less than 0.00333,
we can conclude that
the expected number of deaths per 1,000 people age 55-59
is different than the expected number of deaths per 1,000 people
age 35-39,
holding smoking group constant.
- Since the p-value for `age_group60-64` is less than 0.00333,
we can conclude that
the expected number of deaths per 1,000 people age 60-64
is different than the expected number of deaths per 1,000 people
age 35-39,
holding smoking group constant.
- Since the p-value for `age_group65-69` is less than 0.00333,
we can conclude that
the expected number of deaths per 1,000 people age 65-69
is different than the expected number of deaths per 1,000 people
age 35-39,
holding smoking group constant.
- Since the p-value for `age_group70-74` is less than 0.00333,
we can conclude that
the expected number of deaths per 1,000 people age 70-74
is different than the expected number of deaths per 1,000 people
age 35-39,
holding smoking group constant.
- Since the p-value for `age_group75-79` is less than 0.00333,
we can conclude that
the expected number of deaths per 1,000 people age 75-79
is different than the expected number of deaths per 1,000 people
age 35-39,
holding smoking group constant.
- Since the p-value for `age_group80+` is less than 0.00333,
we can conclude that
the expected number of deaths per 1,000 people age 80 and above
is different than the expected number of deaths per 1,000 people
age 35-39,
holding smoking group constant.
- Since the p-value for `smoke_groupcigar_and_pipe_only` is greater than 0.00333,
we cannot conclude that
the expected number of deaths per 1,000 people who
have only smoked cigars or pipes
is different than the expected number of deaths per 1,000 people
who have never smoked,
holding age group constant.
- Since the p-value for `smoke_groupcigarette_and_other` is less than 0.00333,
we can conclude that
the expected number of deaths per 1,000 people who
have smoked cigarettes and other things
is different than the expected number of deaths per 1,000 people
who have never smoked,
holding age group constant.
- Since the p-value for `smoke_groupcigarette_only` is less than 0.00333,
we can conclude that
the expected number of deaths per 1,000 people who
have only smoked cigarettes
is different than the expected number of deaths per 1,000 people
who have never smoked,
holding age group constant.

## Problems

### Problem 1

```{r}
d2 <- read.table(file = "https://grodri.github.io/datasets/ships.dat",
                 header = TRUE)
```

Let's return to the ship data.
These are the variables:

- `type`: Ship type, with five levels: A, B, C, D, and E.
- `construction`: Year of construction, with four levels:
  - `1960-64`
  - `1965-69`
  - `1970-74`
  - `1975-79`
- `operation`: Period of operation, with two levels:
  - `1960-74`
  - `1975-79`
- `months`: Months of service.
- `damage`: Number of damage incidents.


```{r}
mod_02 <- glm(formula = damage ~ type + construction + operation + offset(log(months / 12)),
              family = poisson,
              data = d2)
summary(mod_02)
```

Interpret the hypothesis tests for the coefficients in the model.
Use a threshold of $0.05 / 9 = 0.00555$.

::: {.callout-tip title="Solution" collapse="true"}


- Since the p-value for the intercept is less than 0.00555,
we can conclude that
the expected number of damage incidents per year
for a type A ship constructed between 1960 and 1964
and operating between 1960 and 1974 is different than 1.
- Since the p-value for `typeB` is less than 0.00555,
we can conclude that
the expected number of damage incidents per year
for type B ships is different than
the expected number of damage incidents per year
for type A ships,
holding construction period and operation period constant.
- Since the p-value for `typeC` is greater than 0.00555,
we cannot conclude that
the expected number of damage incidents per year
for type C ships is different than
the expected number of damage incidents per year
for type A ships,
holding construction period and operation period constant.
- Since the p-value for `typeD` is greater than 0.00555,
we cannot conclude that
the expected number of damage incidents per year
for type D ships is different than
the expected number of damage incidents per year
for type A ships,
holding construction period and operation period constant.
- Since the p-value for `typeE` is greater than 0.00555,
we cannot conclude that
the expected number of damage incidents per year
for type E ships is different than
the expected number of damage incidents per year
for type A ships,
holding construction period and operation period constant.
- Since the p-value for `construction1965-69` is less than 0.00555,
we can conclude that
the expected number of damage incidents per year
for ships constructed between 1965 and 1969 is different than
the expected number of damage incidents per year
for ships constructed between 1960 and 1964,
holding ship type and operation period constant.
- Since the p-value for `construction1970-74` is less than 0.00555,
we can conclude that
the expected number of damage incidents per year
for ships constructed between 1970 and 1974 is different than
the expected number of damage incidents per year
for ships constructed between 1960 and 1964,
holding ship type and operation period constant.
- Since the p-value for `construction1975-79` is greater than 0.00555,
we cannot conclude that
the expected number of damage incidents per year
for ships constructed between 1975 and 1979 is different than
the expected number of damage incidents per year
for ships constructed between 1960 and 1964,
holding ship type and operation period constant.
- Since the p-value for `operation1975-79` is less than 0.00555,
we can conclude that
the expected number of damage incidents per year
for ships operating between 1975 and 1979 is different than
the expected number of damage incidents per year
for ships operating between 1960 and 1974,
holding ship type and construction period constant.

:::

### Problem 2

```{r}
d3 <- read.table(file = "https://grodri.github.io/datasets/cancer.dat",
                 header = TRUE)
d3$malignant <- recode_values(x = d3$malignant,
                              "no" ~ 0,
                              "yes" ~ 1)
d3$population <- d3$yes + d3$no
d3$deaths <- d3$no
d3$yes <- NULL
d3$no <- NULL
```

Let's return to the breast cancer data.
The variables are:

- `age`: Age group, with three levels:
  - `<50`
  - `50-69`
  - `70+`
- `malignant`: 1 for yes and 0 for no
- `population`: The number of patients at baseline for that
age group and that malignancy status.
- `deaths`: The number of deaths after three years.

```{r}
mod_03 <- glm(formula = deaths ~ age + malignant + offset(log(population / 10)),
              family = poisson,
              data = d3)
summary(mod_03)
```

Interpret the hypothesis tests for the coefficients in the model.
Use a threshold of $0.05 / 4 = 0.0125$.


::: {.callout-tip title="Solution" collapse="true"}

- Since the p-value for the intercept is greater than 0.0125,
we cannot conclude that
the expected number of deaths per ten
patients under 50 with non-malignant breast cancer
is different than 1.
- Since the p-value for `age50-69` is greater than 0.0125,
we cannot conclude that
the expected number of deaths per ten patients
age 50-69 is different than
the expected number of deaths per ten patients
under 50,
holding malignancy constant.
- Since the p-value for `age70+` is greater than 0.0125,
we cannot conclude that,
the expected number of deaths per ten patients
age 70 and above is different than
the expected number of deaths per ten patients
under 50,
holding malignancy constant.
- Since the p-value for `malignant` is greater than 0.0125,
we cannot conclude that
the expected number of deaths per ten patients
with malignant cancer is different than
the expected number of deaths per ten patients
with non-malignant cancer,
holding age group constant.

:::




