---
title: "Quasi-Poisson Regression"
execute:
  echo: false
format: html
---

## Learning Objectives

- Define, use, and interpret the term "overdispersion".
- Interpret a dispersion parameter.
- Explain how and why the output for quasi-Poisson regression differs from
the output for Poisson regression.


## Lesson

### Definitions

Recall that if $Y\sim\text{Poisson}(\mu)$,
then $\mu = E(Y) = Var(Y)$.
It's a very strong assumption to assume that the expectation
is equal to the variance,
and it's often violated in practice.
With real count data,
the variance is often greater than the mean.
This is known as overdispersion.
(The opposite phenomenon, underdispersion,
where the variance is less than the mean,
is much less common.)

Quasi-Poisson regression is a form of Poisson regression
that relaxes the assumption that $Var(Y) = E(Y)$.
In quasi-Poisson regression,
we assume $Var(Y) = \phi E(Y)$,
where $0 < \phi < \infty$.
The parameter $\phi$ is known as a dispersion parameter.
If $\phi > 1$, there is overdispersion.

### Example

```{r}
age_group <- rep(x = c("<30",
                       "30-34",
                       "35-39",
                       "40-44",
                       "45-49",
                       "50-54",
                       "55-59",
                       "60-64",
                       "65-69",
                       "70-74",
                       "75-79",
                       "80+"),
                 times = 4)
smoke_group <- rep(x = c("never_smoked",
                         "cigar_and_pipe_only",
                         "cigarette_and_other",
                         "cigarette_only"),
                   each = 12)
population <- c(48,
                739,
                1113,
                656,
                359,
                249,
                632,
                1067,
                897,
                668,
                361,
                274,
                6,
                162,
                187,
                145,
                104,
                98,
                372,
                846,
                949,
                824,
                667,
                537,
                258,
                4596,
                6653,
                4531,
                3030,
                2267,
                4682,
                6052,
                3880,
                2033,
                871,
                345,
                214,
                3185,
                4639,
                3410,
                2239,
                1851,
                3270,
                3791,
                2421,
                1195,
                436,
                113)
deaths <- c(1,
            12,
            14,
            18,
            22,
            19,
            55,
            117,
            170,
            179,
            120,
            128,
            0,
            0,
            1,
            2,
            4,
            3,
            38,
            113,
            173,
            212,
            243,
            253,
            2,
            61,
            121,
            149,
            169,
            193,
            576,
            1001,
            901,
            613,
            337,
            189,
            4,
            56,
            88,
            124,
            140,
            187,
            514,
            778,
            689,
            432,
            214,
            63)
d <- data.frame(age_group,
                smoke_group,
                population,
                deaths)
d$age_group <- relevel(x = as.factor(d$age_group),
                       ref = "35-39")
d$smoke_group <- relevel(x = as.factor(d$smoke_group),
                         ref = "never_smoked")
```

Let's return to the data from
[A Canadian Study of Smoking and Health--Second Report](https://www.jstor.org/stable/41983444).
These are the variables:

- `age_group`: Age group, with the following levels:
  - 40-44
  - 45-49
  - 50-54
  - 55-59
  - 60-64
  - 65-69
  - 70-74
  - 75-79
  - 80+
- `smoke_group`: Smoking status, which has the following levels:
  - Never smoked
  - Cigar and pipe only
  - Cigarette and other
  - Cigarette only
- `population`: The number of people in that age group who had
that smoking status, at the start of the six-year follow-up period.
- `deaths`: The number of people in that age group who had that
smoking status and who died by the end of the six-year
follow-up period.

These are the results:
```{r}
mod_01 <- glm(formula = deaths ~ age_group + smoke_group + offset(log(population / 1000)),
              family = poisson,
              data = d)
summary(mod_01)
```

Note the line in the output that says,
`(Dispersion parameter for poisson family taken to be 1)`.
This is saying that, since the user asked for Poisson regression,
$\phi = 1$.

Now let's try quasi-Poisson regression and see what changes.

```{r}
mod_02 <- glm(formula = deaths ~ age_group + smoke_group + offset(log(population / 1000)),
              family = quasipoisson,
              data = d)
summary(mod_02)
```

What did and did not change when we used quasi-Poisson regression?

- The coefficient estimates did not change at all.
This is true in general;
using quasi-Poisson will not change
the estimates of the beta coefficients
from what they are in Poisson regression.
- The standard errors, $t$ values, and p-values changed.
For each coefficient,
the standard error in quasi-Poisson regression is smaller than
in Poisson regression.
Since the coefficient estimates did not change,
smaller standard errors mean larger $t$ values and smaller p-values.
This will not necessarily be true each time you run Poisson
and quasi-Poisson regression.
Sometimes the standard errors will increase, and sometimes they will
decrease.
- With quasi-Poisson regression,
the dispersion parameter is estimated to be $0.8579232$.
This is one of those rare scenarios when there appears to be underdispersion.
- With quasi-Poisson regression, there is no AIC.
This is because quasi-Poisson regression is not fit with maximum likelihood
estimation.
In fact, you can't write down a likelihood for the data in quasi-Poisson
regression.
There isn't a real distribution that corresponds to quasi-Poisson regression.
This means that we can't use any likelihood-based methods
to evaluate models fit with quasi-Poisson regression.

### Threshold

Students always ask me how much overdispersion is too much,
i.e., at what point should you use quasi-Poisson regression instead
of Poisson regression.
Unfortunately, I don't think there's a single answer to that question.
It probably depends on your goals and your data set.

I know that's often an unsatisfactory answer,
so I'll use
[An empirical approach to determine a threshold for assessing overdispersion in Poisson and negative binomial models for count data](https://doi.org/10.1080/03610918.2017.1323223)
to offer 1.2 as a threshold.
If $\hat{\phi} \ge 1.2$, consider using quasi-Poisson regression
instead of traditional Poisson regression.



## Problems

### Problem 1

```{r}
d3 <- Lock5Data::PASeniors
```

```{r}
mod_03 <- glm(formula = TextsSent ~ Gender + Age + Communicate + HangHours + HWHours,
              family = quasipoisson,
              data = d3)
summary(mod_03)
```

::: {.callout-tip title="Solution" collapse="true"}



:::

### Problem 2



::: {.callout-tip title="Solution" collapse="true"}



:::




