---
title: "Assumptions of Simple Linear Regression"
format: html
---

## Learning Objective

Explain the four assumptions of simple linear regression.


## Lesson

Suppose we observe $n$ data points, each of which has two measurements.
I'll call these points $\left(X_1,Y_1\right),\dotsc,\left(X_n,Y_n\right)$.
The $Y_i$'s are the outcome variable,
also known as the response or target.
The $X_i$'s are the explanatory variables,
also known as the predictors, covariates, or features.

These are the four assumptions of simple linear regression:

1. Linearity
   a) Words: The relationship between the predictor and the outcome is linear.
   b) Math: $E\left(Y_i\middle|X_i\right) = \beta_0 + \beta_1 X_i$
2. Independence
   a) The observations are independent.
3. Equivariance (or homoscedasticity)
   a) Words: No matter the value of the predictor, the variability
   of the outcome around its mean is the same.
   b) Math: $Var(Y_i|X_i) = \sigma^2$
4. Normality
   a) Words: The distribution of the outcome around its mean is normal.
   b) Math: $Y_i|X_i$ is normally distributed.

I've written them in order from most to least essential,
but sometimes people switch the last two so that they create the acronym
"LINE". This may be helpful as a mnemonic for remembering the assumptions.

### Linearity

The most fundamental assumption of simple linear regression is that
there is a linear relationship between the explanatory variable
and the outcome.
Let's take a look at an example of when that is true and when that's not true.

Here is a plot of simulated data where the linearity assumption is met:
```{r}
#| echo: false
set.seed(1)
n <- 1e3
beta_0 <- 1
beta_1 <- 0.5
x <- rnorm(n = n,
           sd = 2)
epsilon <- rnorm(n = n)
y <- beta_0 + beta_1 * x + epsilon
plot(x = x,
     y = y,
     xlab = "X",
     ylab = "Y",
     main = "E(Y|X) = 1 + 0.5X")
abline(a = beta_0,
       b = beta_1,
       col = "blue")
```
For these data, $E(Y|X) = 1 + 0.5X$.
The blue line is $y = 1 + 0.5x$,
and it goes straight through the cloud of points.

Watch what happens if we focus on a narrow band of points around $X = 2$:
```{r}
#| echo: false

plot(x = x,
     y = y,
     xlab = "X",
     ylab = "Y",
     main = "E(Y|X) = 1 + 0.5X")
abline(a = beta_0,
       b = beta_1,
       col = "blue")
abline(v = 1.9,
       col = "red")
abline(v = 2.1,
       col = "red")
```
I'm going to extract just the points where $1.9 \le X_i \le 2.1$
and make a histogram of their $Y_i$ values:
```{r}
#| echo: false
subset_x_2 <- y[x >= 1.9 & x <= 2.1]
hist(x = subset_x_2,
     xlab = "Y",
     main = "Histogram of Points with 1.9 <= X <= 2.1")
```
We know that $E(Y|X = 2) = 1 + 0.5(2) = 2$.
For the points where $1.9 \le X \le 2.1$,
the sample average is `r mean(subset_x_2)`,
pretty darn close to what it's supposed to be.

Now let's look at an example where this assumption is not met.
```{r}
#| echo: false
y_quadratic <- x^2 + epsilon
plot(x = x,
     y = y_quadratic,
     xlab = "X",
     ylab = "Y",
     main = "E(Y|X) = X^2")
abline(lm(y_quadratic ~ x),
       col = "blue")
```
I've done my best to fit a straight line through these points,
but it's clearly not going through the cloud.
The relationship between the explanatory variable and the response
is U-shaped, not linear.


### Independence

We assume the observations are independent of each other.
This isn't something we can visualize;
it has to do with how the data were collected.

Again, let's consider an example where this assumption is met
and an example where it isn't.
The summer after my first year of college, I worked at a survey organization
called ORC Macro. (I don't think it exists anymore.)
I sat at a computer that dialed ten-digit numbers at random.
If the number belonged to someone and they picked up,
I would ask them questions.
Our biggest survey was commissioned by NOAA, and it was about fishing.
I had to ask people how many fish they had caught in the last year.
The participants in these surveys can probably be considered independent
from each other.
Knowing one person's answers to the survey probably wouldn't tell me
anything about another person's answers.
The respondents had nothing to do with each other.

Now suppose that in addition to asking questions of each person who
answered the phone, I asked questions of everyone in their household.
Now, the respondents are no longer independent.
People in the same household tend to be more similar to each other
than to people in different households.
For the fishing survey, two people in the same household are likely to catch
similar numbers of fish.
They might go fishing together, for example, as a family on vacation.
So, knowing how many fish one person in the household caught
would be be informative about how many fish another person in the household
caught.
This would be a violation of independence.


### Equivariance (or Homoscedasticity)



### Normality

## Problems

### Problem 1



::: {.callout-tip title="Solution" collapse="true"}



:::

### Problem 2



::: {.callout-tip title="Solution" collapse="true"}



:::




