---
title: "Linearity Assumption"
execute:
  echo: false
format: html
---

## Learning Objective

Explain the linearity assumption of simple linear regression.


## Lesson

I'm going to explain the first assumption of simple linear regression in
three ways: first with words, then with a procedural form of math,
and finally with an abstract form of math.

### Words

The first assumption of simple linear regression is that
the relationship between the predictor and the outcome is linear.
(I'm using the term "predictor" now,
but it's also known as the explanatory variable,
covariate, feature, or just $X$.
Also, I'm using the term "outcome",
but it's also known as the response variable,
target, or just $Y$.)

What does this look like in a plot?
Consider the following simulated points:
```{r}
set.seed(1)
n <- 1e3
beta_0 <- 1
beta_1 <- 0.5
x <- rnorm(n = n,
           sd = 2)
epsilon <- rnorm(n = n)
y <- beta_0 + beta_1 * x + epsilon
plot(x = x,
     y = y,
     xlab = "X",
     ylab = "Y")
```
As $X$ increases, we can see $Y$ increasing linearly.
We can draw a straight line through the cloud of points that
describes the relationship well:
```{r}
plot(x = x,
     y = y,
     xlab = "X",
     ylab = "Y")
abline(a = beta_0,
       b = beta_1,
       col = "blue")
```

It's probably helpful to look at some points where the relationship
is not linear, to see the contrast.
Consider these points:
```{r}
y_quadratic <- x^2 + epsilon
plot(x = x,
     y = y_quadratic,
     xlab = "X",
     ylab = "Y")
abline(lm(y_quadratic ~ x),
       col = "blue")
```
I've tried drawing a straight line through them, but it clearly doesn't
describe them well.
Here, the relationship between $X$ and $Y$ is U-shaped,
not linear.

### Procedural Math

Another way of looking at the first assumption of simple linear regression
is as a procedure.
We start with $X_1$.
We might know something about where it came from, or we might not.
In the plots above, I drew $X_1$ from a normal distribution with mean 0
and standard deviation of 2.
In symbols, $X_1\sim N(0,4)$.
(If the standard deviation is 2, the variance is $2^2 = 4$.)
Often, we don't care so much about where $X_1$ came from,
because we're more concerned about its relationship with $Y_1$.
In this particular simulation, $X_1$ turned out to be `r x[1]`.

Our next step is to plug $X_1$ into a formula.
In the first plot above, I used $1 + \frac{1}{2}X_1$.
Note that this is a linear formula.
That brought me to the blue line in the second plot above.
The height of the line at $X_1$ is `r 1 + 0.5 * x[1]`.
```{r}
temp <- function(x){
  1 + 0.5 * x
}
curve(expr = temp,
      xlim = c(-6,8),
      ylim = c(-4,6),
      col = "blue",
      xlab = "X",
      ylab = "Y")
abline(v = x[1],
       col = "red")
abline(h = temp(x[1]),
       col = "red")
```

The next step is to add some random error.
I'm going to use $\epsilon_1$ to represent this random error.
In the simulation above, $\epsilon_1\sim N(0,1)$.
The first assumption doesn't require that the random error be normally
distributed, just that it have mean 0.
On average, we want the points to land on the line.
(Obviously not every point will fall on the line;
we just want that to be the case on average.)
In the simulation above, $\epsilon_1$ turned out to be `r epsilon[1]`.
If we add that value to the line at $X_1$, we get $Y_1$.
```{r}
plot(x = x[1],
     y = y[1],
     xlim = c(-6,8),
     ylim = c(-4,6),
     xlab = "X",
     ylab = "Y")
abline(a = beta_0,
       b = beta_1,
       col = "blue")
segments(x0 = x[1],
         x1 = x[1],
         y0 = y[1] - epsilon[1],
         y1 = y[1],
         col = "red")
```
In symbols, $Y_1 = 1 + \frac{1}{2}X + \epsilon_1$.
Of course, that was only for this example.
In general, we write $Y_1 = \beta_0 + \beta_1 X_1 + \epsilon_1$.

That's the procedure for one observation.
We repeat it for $X_2,X_3,\dotsc,X_n$,
where $n$ is the sample size.
Here are the next few points visualized:
```{r}
for(i in 2:4){
  plot(x = x[1:i],
     y = y[1:i],
     xlim = c(-6,8),
     ylim = c(-4,6),
     xlab = "X",
     ylab = "Y")
abline(a = beta_0,
       b = beta_1,
       col = "blue")
segments(x0 = x[1:i],
         x1 = x[1:i],
         y0 = y[1:i] - epsilon[1:i],
         y1 = y[1:i],
         col = "red")
}
```

To summarize the entire process, I'd write that for $i$ from $1$ to $n$,
$Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$,
where $E(\epsilon_i|X_i) = 0$.

### Abstract Math

The most abstract way to write this first assumption is
$E(Y_i|X_i) = \beta_0 + \beta_1 X_i$.
Given $X_i$, we expect $Y_i$ to fall on the line $\beta_0 + \beta_1 X_i$.
Why use such an abstract representation?
When we move on to logistic regression, Poisson regression,
and really every other form of regression in this class,
we won't have any epsilons.
They're nice for visualizing what's going on and understanding the procedure,
but they cease to be relevant once we depart linear regression.

### Note

Strictly, the linearity assumption is
$$
E(Y_i|X_1,\dotsc,X_n) = \beta_0 + \beta_1 X_i \text{,}
$$
but that's very cumbersome to write,
and it will be more cumbersome once we get to multiple linear regression.
So, I'm simplifying the notation a bit to
$$
E(Y_i|X_i) = \beta_0 + \beta_1 X_i \text{.}
$$
This first assumption is much easier to write with linear algebra,
but linear algebra is not a prerequisite for this course.

## Problems

### Problem 1

Do these data satisfy the first assumption of simple linear regression?
```{r}
y_cos <- cos(2 * x) + epsilon / 2
plot(x = x,
     y = y_cos,
     xlab = "X",
     ylab = "Y")
```

::: {.callout-tip title="Solution" collapse="true"}

No. The relationship between $X$ and $Y$ seems to be cyclical,
not linear.

:::

### Problem 2

Does the first assumption of simple linear regression require
a certain variance for $\epsilon_i$?

::: {.callout-tip title="Solution" collapse="true"}

No. It only requires that the expected value of $\epsilon_i$,
conditional on $X_i$, is zero.

:::




