---
title: "Consequences of the Linearity Assumption"
execute:
  echo: false
format: html
---

## Learning Objective

Explain the consequences of violations of the linearity assumption
of simple linear regression.


## Lesson

### Intuition

If the first assumption of simple linear regression is not met,
the results are basically meaningless.
Consider an example.
The plot below shows mean monthly temperatures in Moscow
for 2023 and 2024.
```{r}
# Data
d <- Lock5Data::CityTemps

# Plot
plot(x = d$Month,
     y = d$Moscow,
     xlab = "Month of the Year",
     ylab = "Temperature in Degrees Celsius",
     main = "Moscow (2023-2024)")

# Simple Linear Regression
mod_01 <- lm(d$Moscow ~ d$Month)
```

If I perform simple linear regression on these data,
I get that each additional month of the year is associated with
a `r round(x = mod_01$coefficients[2],digits = 2)`-degree Celsius
increase in temperature.
```{r}
plot(x = d$Month,
     y = d$Moscow,
     xlab = "Month of the Year",
     ylab = "Temperature in Degrees Celsius",
     main = "Moscow (2023-2024)")
abline(mod_01,
       col = "blue")
```
That's clearly not true.

Simple linear regression is the most basic form of regression,
and it's only appropriate when you have a clear linear relationship
between two variables.

### Bias

In technical terms, we say that if the linearity assumption is not met,
the estimators are biased.
On average, they don't equal the estimand they're estimating.
If the linearity assumption is met,
the estimators are unbiased.

## Problems

### Problem 1

Should you use simple linear regression if a scatterplot of your data
forms the shape of an M?

::: {.callout-tip title="Solution" collapse="true"}

No.

:::






