---
title: "Checking the Linearity Assumption"
execute:
  echo: false
format: html
---

## Learning Objective

Evaluate the validity of the linearity assumption of simple linear regression
in a data set.


## Lesson

It might seem straightforward to check the linearity assumption
of simple linear regression.
You can create a scatterplot of the data and eyeball whether
the relationship between the two variables is linear.
This is, in fact, a fantastic idea.
You should always look at your data before trying to model it.

I will add one thing, though:
It can be easier to evaluate this (and the remaining) assumptions
if we look at the residuals,
particularly when we move on to multiple linear regression.
So, let's start getting practice with that technique now.

Let's take a look at these simulated data again:
```{r}
set.seed(1)
n <- 1e3
beta_0 <- 1
beta_1 <- 0.5
x <- rnorm(n = n,
           sd = 2)
epsilon <- rnorm(n = n)
y <- beta_0 + beta_1 * x + epsilon
plot(x = x,
     y = y,
     xlab = "X",
     ylab = "Y")
abline(a = beta_0,
       b = beta_1,
       col = "blue")
```
Now I'm going to plot the residuals as a function of $X$.
```{r}
mod_01 <- lm(y ~ x)
plot(x = x,
     y = mod_01$residuals,
     xlab = "X",
     ylab = "Residual")
```
Why is this better than just plotting the data?
Well, it's not, at least not yet.
Recall that we're trying to check the assumption that $E(\epsilon_i|X_i) = 0$.
(This is from the "procedural math" way of looking at simple linear regression.)
What this equation means in words is that,
no matter the value of $X_i$, the expected value of the random error should
be zero.
So, no matter the value of $X_i$, the sample average of the residuals
should be zero.

I'm going to add a horizontal line at 0 so we have something to compare to.
```{r}
plot(x = x,
     y = mod_01$residuals,
     xlab = "X",
     ylab = "Residual")
abline(h = 0,
       col = "red")
```
Again, no matter what value we choose on the horizontal axis,
the average of the residuals at that point should be zero.
Let's look at the points around $X = 2$,
specifically, those points were $1.9 \le X_i \le 2.1$.
```{r}
plot(x = x,
     y = mod_01$residuals,
     xlab = "X",
     ylab = "Residual")
abline(h = 0,
       col = "red")
abline(v = 1.9,
       col = "purple")
abline(v = 2.1,
       col = "purple")
```
The residuals in this band should average out to zero.
Do they?
```{r}
#| echo: true
mean(mod_01$residuals[x >= 1.9 & x <= 2.1])
```
Just about! There will always be a little bit of random error,
but this is pretty close to zero.

Let's consider another band.
```{r}
plot(x = x,
     y = mod_01$residuals,
     xlab = "X",
     ylab = "Residual")
abline(h = 0,
       col = "red")
abline(v = -3.1,
       col = "purple")
abline(v = -2.9,
       col = "purple")
```
Do these residuals average out to zero?
```{r}
#| echo: true
mean(mod_01$residuals[x >= -3.1 & x <= -2.9])
```
This average is farther from zero because there are few points in the band,
but it's still close.

In practice, we don't actually cut out vertical slices from the plot
and compute averages.
Instead, we look at the residual scatterplot and eyeball it.
Why is this better than the original scatterplot?
Sometimes zeroing out the slope makes it easier to assess the linearity
assumption.
Also, when we start learning about multiple linear regression,
it will be too difficult to visualize a scatterplot of the outcome
against all the covariates.

## Problems

### Problem 1

Below is a plot of annual expenditures (in dollars) as a function of age
(in years) for 1,000 individuals supported by the California
Department of Developmental Services (DDS).
(The data have been altered slightly to protect privacy.)
Is the first assumption of simple linear regression met for these data?
```{r}
# Scatterplot
d <- Lock5Data::DDS
plot(x = d$Age,
     y = d$Expenditures,
     xlab = "Age (years)",
     ylab = "Annual Expenditures (dollars)")

```

::: {.callout-tip title="Solution" collapse="true"}

No. There is clearly not a linear relationship between age
and expenditures.

:::

### Problem 2

Below are simulated data.
Is the first assumption of simple linear regression met for these data?
```{r}
y_var <- beta_0 + x + x * epsilon
plot(x = x,
     y = y_var,
     xlab = "X",
     ylab = "Y")
mod_02 <- lm(y_var ~ x)
plot(x = x,
     y = mod_02$residuals,
     xlab = "X",
     ylab = "Residuals")
abline(h = 0,
       col = "red")
```

::: {.callout-tip title="Solution" collapse="true"}

Yes. This might be a case where it's a little harder to tell with
the standard scatterplot than the residual scatterplot.

:::




