---
title: "Consequences of the IID Assumption"
execute:
  echo: false
  warning: false
format: html
---

```{r}
library(scales)
```

## Learning Objective

Explain the consequences of violations of the iid assumption
of simple linear regression.


## Lesson

For our purposes, the main consequence of violations of the iid assumption
of simple linear regression is that the standard errors will be wrong.

A standard error is the standard deviation of an estimator;
it's a measure of how variable the estimator is,
or how much it will change from sample to sample.
The higher the standard error,
the less precise our estimator is;
the lower the standard error,
the more precise our estimator is.
Getting a good measure of the variability of the estimator
helps us know how wrong we might be.

In simple linear regression,
we are generally most interested in $\hat{\beta}_1$,
the estimator of the slope.
We often conduct a statistical test where the null hypothesis test
is that $\beta_1 = 0$,
i.e., that there is no (linear) relationship between the explanatory
and outcome variables.
The statistical test is only correct if the standard error is correct.

Let's consider an example.
Below is a plot of commute time (in minutes) as a function of commute distance
(in miles) for a sample of 500 individuals in Atlanta.
I've made the points partially transparent because there's so much overlap,
so darker circles indicate more individuals with the same commute distance
and time.
```{r}
d <- Lock5Data::CommuteAtlanta
plot(x = d$Distance,
     y = d$Time,
     xlab = "Distance (miles)",
     ylab = "Time (minutes)",
     main = "Commuting in Atlanta",
     col = alpha(colour = "black",
                 alpha = 0.2),
     pch = 16)
```

If I run a simple linear regression with time as the outcome
and distance as the explanatory variable,
this is the resulting residual plot.
```{r}
mod_01 <- lm(formula = Time ~ Distance,
             data = d)
plot(x = d$Distance,
     y = mod_01$residuals,
     xlab = "Distance (miles)",
     ylab = "Residuals",
     col = alpha(colour = "black",
                 alpha = 0.2),
     pch = 16)
abline(h = 0,
       col = "red")
```
The linearity assumption appears to be met, although in a real-world setting,
it would behoove us to look into a few of the "weird" points,
for lack of a better word.
Who has a five mile commute that takes two hours?

Since we're focusing on the effects of violating the iid assumption of
simple linear regression, let's take a look at the estimates and their
standard errors.
```{r}
summary(mod_01)
```
The standard errors are in the `Coefficients` table in the 
`Std. Error` column.
If these commuters weren't sampled independently from the same population,
we cannot trust these standard errors.
For example, what if we sampled people who carpool together?
We would have a bunch of people with similar (or even identical) commutes
because they're in the same vehicle
(for at least a portion of the trip).
In that case, the standard errors presented in this table would be wrong.
Since the `t value` column is computed from the `Std. Error` column,
the `t value` column would be wrong,
and since the `Pr(>|t|)` column is computed from the `t value` column,
the `Pr(>|t|)` column would be wrong.

I should say that violations of the iid assumption don't mean you can't
do any regression at all.
There are several subfields within statistics devoted to analyzing
correlated data.
For example, many statisticians focus on longitudinal data,
where participants are observed more than once over time;
in this case, observations within individuals are correlated with each other.
Many educational researchers specialize in hierarchical models,
since students are nested within classrooms, which are nested within schools,
which are nested within districts, etc.

So, what I mean to say is that, when data are correlated,
the standard errors computed from the most straightforward version
of simple linear regression are incorrect.

## Problems

### Problem 1

True or false:
If the commuters in the example above were not independent,
the slope estimate is meaningless and uninterpretable.

::: {.callout-tip title="Solution" collapse="true"}

False. 
This is something that I find counterintuitive,
but a violation of the iid assumption of simple linear regression
doesn't imply that we have to throw out the coefficient estimates.
So long as the linearity assumption is met,
the coefficients are unbiased,
meaning that their expectation is equal to the true value of the population
parameter.
This is a very powerful result in statistics.

:::

### Problem 2

True or false:
If the iid assumption of simple linear regression is not met,
we cannot trust the confidence intervals computed for the coefficients.

::: {.callout-tip title="Solution" collapse="true"}

True.
The confidence interval formula (which we haven't seen)
uses the standard error,
so if the standard error is wrong,
the confidence interval is wrong.



:::




