---
title: "Homoscedasticity Assumption"
execute:
  echo: false
format: html
---

## Learning Objective

Explain the homoscedasticity assumption of simple linear regression.


## Lesson

Similar to with the linearity assumption,
I'm going to try to explain this assumption in three ways.

### Words

The word "homoscedasticity" comes from the Greek words "homos",
[meaning "same"](https://www.merriam-webster.com/dictionary/hom-),
and "skedastikos",
[meaning "able to disperse"](https://www.merriam-webster.com/dictionary/homoscedasticity).
So the word means "same dispersion".
If this assumption is met, then there is equal dispersion,
or equal variability, of the points around the regression line,
no matter the value of the explanatory variable.

It will probably help to look at some pictures.
Consider these simulated data:
```{r}
set.seed(1)
n <- 1e3
beta_0 <- 1
beta_1 <- 0.5
x <- rnorm(n = n,
           sd = 2)
epsilon <- rnorm(n = n)
y <- beta_0 + beta_1 * x + epsilon
plot(x = x,
     y = y,
     xlab = "X",
     ylab = "Y")
abline(a = beta_0,
       b = beta_1,
       col = "blue")
abline(v = 1.9,
       col = "purple")
abline(v = 2.1,
       col = "purple")
abline(v = -0.1,
       col = "purple")
abline(v = 0.1,
       col = "purple")
```
If we look at the vertical spread of the points where $X \approx 0$
and where $X \approx 2$,
it's about the same.
Even though $X$ has changed,
the variability around the regression line has not.
So, these points are homoscedastic.

I have to acknowledge that you might disagree with me on this point.
You might be thinking that the cloud of points looks more like a football,
with lower variability at very high and very low values of $X$.
That's fair.
Having generated the data, I know that the assumption of homoscedasticity is
met.
So why the football shape?
This is because, for any value of $X$, points are more likely to be close
to the line than far away.
If there are a lot of points at a certain value of $X$ (like around $X = 0$),
there will be a few points far from the line,
just by chance.
If there are only a couple points at a certain value of $X$
(like around $X = -6$),
it's unlikely to observe many points far from the line.
So, it appears that the vertical spread of the points tapers off
for high and low values of $X$.

Now look at these points:
```{r}
y_bowtie <- beta_0 + beta_1 * x + epsilon * x
plot(x = x,
     y = y_bowtie,
     xlab = "X",
     ylab = "Y")
abline(a = beta_0,
       b = beta_1,
       col = "blue")
abline(v = 1.9,
       col = "purple")
abline(v = 2.1,
       col = "purple")
abline(v = -0.1,
       col = "purple")
abline(v = 0.1,
       col = "purple")
```
The variability of the points around the regression line when $X \approx 0$
is way different than the variability of the points around the regression
line when $X \approx 2$.
So, these points are heteroscedastic.

### Procedural Math

Recall that we're assuming that $Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$.
This assumption can be written as $Var(\epsilon_i|X_i) = \sigma^2$.
Note that $\sigma^2$ doesn't depend on $X_i$ or even $i$;
it's the same no matter which observation you're talking about.

When we're walking through our procedure
(draw $X_i$ from its distribution;
plug it into $f(x) = \beta_0 + \beta_1 x$;
add random error),
we're not changing the variability of the random error for each observation.
It stays the same.
```{r}
for(i in 1:4){
  plot(x = x[1:i],
     y = y[1:i],
     xlim = c(-6,8),
     ylim = c(-4,6),
     xlab = "X",
     ylab = "Y")
abline(a = beta_0,
       b = beta_1,
       col = "blue")
segments(x0 = x[1:i],
         x1 = x[1:i],
         y0 = y[1:i] - epsilon[1:i],
         y1 = y[1:i],
         col = "red")
}
```
::: {.callout-caution collapse="true"}
## Caution: Math

Some of you might be saying,
"Isn't this assumption included in the second assumption,
when we assume the observations are identically distributed."
Not exactly.
If the data-generating procedure were
$$
Y_i = \beta_0 + \beta_1 X_i + \epsilon_i X_i \text{,}
$$
where $E(\epsilon_i|X_i) = 0$ (assumption 1),
$X_1,\dotsc,X_n$ are iid (assumption 2),
and $\epsilon_1,\dotsc,\epsilon_n$ are iid (assumption 2),
we would satisfy the first two assumptions but not the third.
This is because
$$
\begin{align*}
E\left(\epsilon_i X_i\right) &= E\left(E\left(\epsilon_i X_i\middle|X_i\right)\right) \\
&= E\left(X_i E\left(\epsilon_i \middle|X_i\right)\right) \\
&= E\left(X_i \left(0\right)\right) \\
&= E(0) \\
&= 0
\end{align*}
$$
but
$$
\begin{align*}
Var\left(\epsilon_i X_i\right) &= E\left(Var\left(\epsilon_i X_i\middle|X_i\right)\right) + Var\left(E\left(\epsilon_i X_i\middle|X_i\right)\right) \\
&= E\left(X_i^2 Var\left(\epsilon_i \middle|X_i\right)\right) + Var\left(0\right) \\
&= E\left(X_i^2 \sigma^2\right) \\
&= \sigma^2 E\left(X_i^2\right) \text{.}
\end{align*}
$$

:::


### Abstract Math

In its most abstract form, we can write this assumption as
$Var(Y_i|X_i) = \sigma^2$.
Again, the $\sigma^2$ doesn't depend on $X_i$ or even $i$,
so it's the same across all $X_i$.
Why do we go to this level of abstraction?
It will be useful when we talk about logistic regression
and Poisson regression.

## Problems

### Problem 1

Suppose $Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$
where $Var(\epsilon_i|X_i) = \tau^2$.
Would this satisfy the homoscedasticity assumption of simple linear
regression?
Why or why not?

::: {.callout-tip title="Solution" collapse="true"}

Yes. It's not important which Greek letter you use to represent
the variance of the random error,
just so long as that variance is a constant,
and does not change with $X_i$ or $i$.

:::

### Problem 2

What is the opposite of homoscedasticity?

::: {.callout-tip title="Solution" collapse="true"}

Heteroscedasticity, when the variability of the points around the regression
line varies as a function of $X_i$.

:::




