---
title: "Consequences of the Homoscedasticity Assumption"
execute:
  echo: false
format: html
---

## Learning Objective

Explain the consequences of violations of the homoscedasticity assumption
of simple linear regression.


## Lesson

If the homoscedasticity assumption of simple linear regression is violated,
the standard errors will be wrong.

Consider the following example.
Each observation is a forest fire in Montesinho Park (Portugal)
between January 2000 and December 2003
where at least 0.01 hectare (100 square meters) was burned.
In the figure below,
I've plotted $\log_{10}(\text{area burned})$ as a function of the outside
temperature.
```{r}
d <- Lock5Data::ForestFires
d$Area_No_0 <- d$Area
d$Area_No_0[d$Area_No_0 == 0] <- NA
d$Log10_Area_No_0 <- log10(d$Area_No_0)
plot(x = d$Temp,
     y = d$Log10_Area_No_0,
     xlab = "Outside Temperature (degrees Celsius)",
     ylab = "Log10(Total Burned Area) (log hectares)")
```

If I regress $\log_{10}(\text{area burned})$ on temperature,
this is the resulting residual plot:
```{r}
mod_01 <- lm(formula = Log10_Area_No_0 ~ Temp,
             data = d)
plot(x = mod_01$model$Temp,
     y = mod_01$residuals,
     xlab = "Outside Temperature (degrees Celsius)",
     ylab = "Residuals")
abline(h = 0,
       col = "red")
```

Note that the variance of the residuals is greater when the outside
temperature is 20-25 degrees than when the outside temperature
is 5-10 degrees.

This doesn't appear to be due to the fact that there are more
observations between 20 and 25 degrees than between 5 and 10 degrees.
If we look at vertical slices with approximately equal numbers
of observations, the variance is greater
for higher temperatures.
For example, there are 16 observations with temperature less than 7.5:
```{r}
#| echo: true
sum(mod_01$model$Temp < 7.5)
```
This is a histogram of the residuals for those points:
```{r}
hist(mod_01$residuals[mod_01$model$Temp < 7.5],
     xlab = "Residuals",
     main = "")
```
There are 19 observations with temperature between 25 and 27.5:
```{r}
#| echo: true
sum(mod_01$model$Temp < 27.5 & mod_01$model$Temp >= 25)
```
This is a histogram of the residuals for those points:
```{r}
hist(mod_01$residuals[mod_01$model$Temp < 27.5 & mod_01$model$Temp >= 25],
     xlab = "Residuals",
     main = "")
```

Now let's take a look at the regression results.
```{r}
summary(mod_01)
```
Because of the heteroscedasticity,
we can't trust the values in the `Std. Error` column.
The values in the `t value` column are based on the standard errors,
so those will also be wrong;
and the values in the last column, the p-values,
are based on the $t$ values,
so those will be wrong.


## Problems

### Problem 1

True or false:
In the presence of heteroscedasticity, the estimated coefficients
are biased.

::: {.callout-tip title="Solution" collapse="true"}

False.
Heteroscedasticity does not cause bias.
As long as the first assumption is met,
the regression coefficients will be unbiased.
Heteroscedasticity only affects the standard errors
(and anything calculated from the standard errors,
like the p-values).

:::

### Problem 2

True or false:
If the homoscedasticity assumption is met,
the standard errors are guaranteed to be correct.

::: {.callout-tip title="Solution" collapse="true"}

False.
Remember that the iid assumption is also necessary for the standard
errors to be correct.

:::




