---
title: "Checking the Homoscedasticity Assumption"
execute:
  echo: false
format: html
---

```{r}
#| warning: false
library(scales)
```

## Learning Objective

Evaluate the validity of the homoscedasticity assumption of simple linear regression
in a data set.


## Lesson

In order to evaluate the validity of the homoscedasticity assumption
of simple linear regression,
we plot the residuals as function of the explanatory variable.
If the vertical spread of the residuals varies as a function of $X$,
we have heteroscedasticity.
Let's consider a couple examples.

### Fire Ants

The red imported fire ant has invaded some of the
habitats of fence lizards (e.g., southern Alabama),
but not all (e.g., eastern Arkansas).
The venom from the fire ant can kill a lizard,
whether the ants are attacking the lizard or the lizard
has eaten some ants.
In 2009, Tracy Langkilde published a [study](https://doi.org/10.1890/08-0355.1)
in which she exposed lizards from invaded and uninvaded habitats
to fire ants and counted the number of times the lizards twitched.
Here are side-by-side boxplots of the data:
```{r}
d <- Lock5Data::FireAnts
boxplot(formula = Twitches ~ Habitat,
        data = d)
```

You might be thinking, "How can we do regression with these data?
The outcome is quantitative, but the explanatory variable is
categorical."
It turns out that you can do simple linear regression with a categorical
explanatory variable that only has two levels.
You just have to recode the levels as 0 and 1.
I'll change every "Uninvaded" to a 0,
and every "Invaded" to a 1,
then regress the number of twitches on the new habitat variable.
Here's a plot of the data and the fitted regression line:
```{r}
d$Habitat_01 <- 0
d$Habitat_01[d$Habitat == "Invaded"] <- 1
mod_01 <- lm(formula = Twitches ~ Habitat_01,
             data = d)
plot(x = d$Habitat_01,
     y = d$Twitches,
     xlab = "Habitat",
     ylab = "Twitches",
     col = alpha(colour = "black",
                 alpha = 0.2),
     pch = 19)
abline(mod_01,
       col = "blue")
```

Let's look at boxplots of the residuals:
```{r}
boxplot(formula = mod_01$residuals ~ d$Habitat,
        xlab = "Habitat",
        ylab = "Residuals")
```
It seems like the variability of the residuals is greater for invaded
than uninvaded lizards.
So, I'm not convinced we've satisfied the homoscedasticity assumption
of simple linear regression.

### 

## Problems

### Problem 1



::: {.callout-tip title="Solution" collapse="true"}



:::

### Problem 2



::: {.callout-tip title="Solution" collapse="true"}



:::




