---
title: "Checking the Homoscedasticity Assumption"
execute:
  echo: false
format: html
---

```{r}
#| warning: false
library(scales)
```

## Learning Objective

Evaluate the validity of the homoscedasticity assumption of simple linear regression
in a data set.


## Lesson

In order to evaluate the validity of the homoscedasticity assumption
of simple linear regression,
we plot the residuals as function of the explanatory variable.
If the vertical spread of the residuals varies as a function of $X$,
we have heteroscedasticity.
Let's consider a couple examples.

### Fire Ants

The red imported fire ant has invaded some of the
habitats of fence lizards (e.g., southern Alabama),
but not all (e.g., eastern Arkansas).
The venom from the fire ant can kill a lizard,
whether the ants are attacking the lizard or the lizard
has eaten some ants.
In 2009, Tracy Langkilde published a [study](https://doi.org/10.1890/08-0355.1)
in which she exposed lizards from invaded and uninvaded habitats
to fire ants and counted the number of times the lizards twitched.
Here are side-by-side boxplots of the data:
```{r}
d <- Lock5Data::FireAnts
boxplot(formula = Twitches ~ Habitat,
        data = d)
```

You might be thinking, "How can we do regression with these data?
The outcome is quantitative, but the explanatory variable is
categorical."
It turns out that you can do simple linear regression with a categorical
explanatory variable that only has two levels.
You just have to recode the levels as 0 and 1.
I'll change every "Uninvaded" to a 0,
and every "Invaded" to a 1,
then regress the number of twitches on the new habitat variable.
Here's a plot of the data and the fitted regression line:
```{r}
d$Habitat_01 <- 0
d$Habitat_01[d$Habitat == "Invaded"] <- 1
mod_01 <- lm(formula = Twitches ~ Habitat_01,
             data = d)
plot(x = d$Habitat_01,
     y = d$Twitches,
     xlab = "Habitat",
     ylab = "Twitches",
     col = alpha(colour = "black",
                 alpha = 0.2),
     pch = 19)
abline(mod_01,
       col = "blue")
```

Let's look at boxplots of the residuals:
```{r}
boxplot(formula = mod_01$residuals ~ d$Habitat,
        xlab = "Habitat",
        ylab = "Residuals")
```
It seems like the variability of the residuals is greater for invaded
than uninvaded lizards.
So, I'm not convinced we've satisfied the homoscedasticity assumption
of simple linear regression.

### Weight and Red Blood Cell Density in Mice


Below are data on weight (in grams) and red blood cell density
(I couldn't find the units) for 72 mice.
```{r}
d2 <- Lock5Data::DrugResistance
plot(x = d2$Weight,
     y = d2$RBC,
     xlab = "Weight (g)",
     ylab = "Red Blood Cell Density",
     main = "Mice")
```
Only 5 mice (6.94%) have weights under 5 grams.
```{r}
sum(d2$Weight < 5)
sum(d2$Weight < 5) / nrow(d2) * 100
```

I'm going to remove those mice, because I think they are so different
from the other mice as to represent a different population.

Here is the scatterplot again, without the five tiny mice.
```{r}
d2 <- d2[d2$Weight >= 5, ]
plot(x = d2$Weight,
     y = d2$RBC,
     xlab = "Weight (g)",
     ylab = "Red Blood Cell Density",
     main = "Mice")
```

If I regress red blood cell density on weight,
this is the residual scatter plot.
```{r}
mod_02 <- lm(RBC ~ Weight,
             data = d2)
plot(x = mod_02$model$Weight,
     y = mod_02$residuals,
     xlab = "Weight (g)",
     ylab = "Residuals")
abline(h = 0,
       col = "red")
```

The vertical spread of the residuals is fairly constant across all weights.
So, the homoscedasticity assumption is met.

## Problems

### Problem 1

In the residual plot below, is the homoscedasticity assumption met?
```{r}
set.seed(10)
n <- 1e3
x <- runif(n = n,
           min = -1,
           max = 1)
y <- runif(n = n,
           min = -abs(sin(acos(x))),
           max = abs(sin(acos(x))))
plot(x = x,
     y = y,
     xlab = "x",
     ylab = "Residuals",
     col = alpha(colour = "black",
                 alpha = 0.3),
     pch = 19)
abline(h = 0,
       col = "red")
```

::: {.callout-tip title="Solution" collapse="true"}

It is not met. Variance is highest when $X$ is around 0
and lowest when $|X|$ is around 1.
This is not due to there just being fewer points with $|X|$ around 1.
The points are partially transparent,
so the darker shading at the left and right of the plot means
there are lots of points there.

:::

### Problem 2

In the residual plot below, is the homoscedasticity assumption met?
```{r}
x <- rexp(n)
neg <- rbinom(n = n,
              size = 1,
              prob = 0.5) > 0
x[neg] <- -x[neg]
y <- rexp(n)
neg <- rbinom(n = n,
              size = 1,
              prob = 0.5) > 0
y[neg] <- -y[neg]
plot(x = x,
     y = y,
     xlab = "x",
     ylab = "Residuals",
     col = alpha(colour = "black",
                 alpha = 0.2),
     pch = 19)
abline(h = 0,
       col = "red")
```

::: {.callout-tip title="Solution" collapse="true"}

Yes, the homoscedasticity assumption is met.
It might look like the variance is lower at the far left and far right,
but that's because there are fewer points there,
and points closer to the regression line are more likely than points
far away.

:::




