---
title: "Normality Assumption"
execute:
  echo: false
format: html
---

## Learning Objective

Explain the normality assumption of simple linear regression.


## Lesson

The final assumption of simple linear regression is that the random errors
are normally distributed.
In symbols, we assume $\epsilon_i|X_i$ is normally distributed.
Equivalently, we can say that we assume
$Y_i|X_i$ is normally distributed.

In this plot, all four assumptions are met.
```{r}
set.seed(11)
n <- 1e3
x <- rnorm(n = n,
           sd = 2)
epsilon <- rnorm(n = n,
                 sd = 2)
beta_0 <- 1
beta_1 <- 1
y <- beta_0 + beta_1 * x + epsilon
plot(x = x,
     y = y)
abline(a = beta_0,
       b = beta_1,
       col = "blue")
```

In the next three plots, the first three assumptions are met,
but not the fourth.
I've used three non-normal distributions for the random errors.
```{r}
# Uniform
epsilon_unif <- runif(n = n,
                      min = -2,
                      max = 2)
y_unif <- beta_0 + beta_1 * x + epsilon_unif
plot(x = x,
     y = y_unif,
     ylab = "y")
abline(a = beta_0,
       b = beta_1,
       col = "blue")

# Shifted exponential
epsilon_exp <- rexp(n) - 1
y_exp <- beta_0 + beta_1 * x + epsilon_exp
plot(x = x,
     y = y_exp,
     ylab = "y")
abline(a = beta_0,
       b = beta_1,
       col = "blue")

# Shifted Poisson
epsilon_pois <- rpois(n = n,
                      lambda = 5) - 5
y_pois <- beta_0 + beta_1 * x + epsilon_pois
plot(x = x,
     y = y_pois,
     ylab = "y")
abline(a = beta_0,
       b = beta_1,
       col = "blue")
```

## Problems

### Problem 1

If the errors are Laplace distributed, does that satisfy
the normality assumption of simple linear regression?

::: {.callout-tip title="Solution" collapse="true"}

No. The normality assumption is that the errors are normally distributed.

:::






