---
title: "Consequences of the Normality Assumption"
execute:
  echo: false
format: html
---

## Learning Objective

Explain the consequences of violations of the normality assumption
of simple linear regression.


## Lesson

### Consequence

If the normality assumption of simple linear regression is not satisfied,
the test statistic for each coefficient will not be $t$-distributed,
so the p-value in the output will be wrong.

Let's return to the example where we regressed red blood cell density
on weight in mice.
As before, I'm going to remove the five anomalous mice with very low weights.
```{r}
d2 <- Lock5Data::DrugResistance
d2 <- d2[d2$Weight >= 5, ]
plot(x = d2$Weight,
     y = d2$RBC,
     xlab = "Weight (g)",
     ylab = "Red Blood Cell Density",
     main = "Mice")
```

If I regress red blood cell density on weight,
these are the results.
```{r}
mod_02 <- lm(RBC ~ Weight,
             data = d2)
summary(mod_02)
```
The values in the `Pr(>|t|)` column are p-values.
For them to be correct, all four assumptions must be true.

Let's go a little deeper and look at how the p-values are computed.

Each value in the `t value` column is computed by taking the corresponding
value in the `Estimate` column and dividing it by the corresponding
value in the `Std. Error` column.
So,
$$
\begin{align*}
-\frac{0.57833}{0.66632} &= -0.868 \\
\frac{0.35342}{0.03194} &= 11.066 \text{.}
\end{align*}
$$

These are the test statistics.
If all four assumptions are met, each should follow a $t$ distribution
with $n - 2$ degrees of freedom.
In this example, $n = 67$, so $n - 2 = 65$.

Each value in the `Pr(>|t|)` is computed by integrating under the 
probability density function of a $t_{n-2}$ distribution
from $|\texttt{t value}|$ to $+\infty$ and from
$-|\texttt{t value}|$ to $-\infty$.
I can show this visually for the `(Intercept)` row
of the `Coefficients:` table.
```{r}
x <- seq(from = -3.5,
         to = 3.5,
         length.out = 500)
y <- dt(x = x,
        df = 65)
plot(x = x,
     y = y,
     ylab = "f(x)",
     type = "l")
abline(h = 0)
polygon(x = c(x[x >= 0.868],
              x[length(x)],
              0.868,
              0.868),
        y = c(y[x >= 0.868],
              0,
              0,
              dt(x = 0.868,
                 df = 65)),
        col = "blue")
polygon(x = c(x[x <= -0.868],
              -0.868,
              -0.868,
              x[1]),
        y = c(y[x <= -0.868],
              dt(x = -0.868,
                 df = 65),
              0,
              0),
        col = "blue")
```

I can also directly compute the p-value.
```{r}
#| echo: true

2 * pt(q = -0.868,
       df = 65)
```

If the random errors are not normally distributed,
the test statistics will not be $t$ distributed,
so the p-values won't be correct.

### Qualification

If the first three assumptions are met (linearity, iid, homoscedasticity)
and the sample size is large,
the test statistics will be approximately normally distributed,
with mean 0 and variance 1.
As the sample size gets larger, this approximation gets better.

You may have noticed that the $t_{65}$ distribution looks basically normal.
I'll plot a $t_{65}$ distribution and a standard normal distribution
over each other so you can compare.
```{r}
temp <- function(x){
  dt(x = x,
     df = 65)
}
curve(expr = dnorm,
      from = -3.5,
      to = 3.5,
      ylab = "f(x)")
curve(expr = temp,
      from = -3.5,
      to = 3.5,
      add = TRUE,
      lty = 2)
legend(x = "topright",
       lty = c(1,2),
       legend = c("N(0,1)",
                  "t, df = 65"))
```
Can you tell the difference?
They're nearly identical.

As $n$ increases, $t_{n-2}$ gets closer and closer to $N(0,1)$.
I'll plot a bunch of $t$ distributions with increasing degrees of freedom
so you can see how they change.
```{r}
temp <- function(x){
  dt(x = x,
     df = 5)
}
curve(expr = temp,
      from = -3.5,
      to = 3.5,
      ylab = "f(x)")
for(i in 4:1){
  temp <- function(x){
    dt(x = x,
       df = i)
  }
  curve(expr = temp,
        from = -3.5,
        to = 3.5,
        add = TRUE)
}
```
As the degrees of freedom increases,
the peak rises and the tails flatten.
(Since the tails flatten, and p-values are tail areas,
this means p-values decrease as degrees of freedom increases.)

To summarize:
If the first three assumptions are met and the sample size is large,
the test statistics will be approximately normal.
If the sample size is large, the corresponding $t$ distribution
will be approximately normal.
So, if the first three assumptions are met and the sample size is large,
you can use the p-values in the output,
regardless of whether the random errors are normal or not.

### What's "Large"?

You might be wondering what sample size is large enough for the test
statistics to be approximately normal.
This depends.
Very asymmetric random errors will require a larger sample size
for the test statistic to be approximately normal.
I recommend you consult a statistician.

## Problems

### Problem 1

If the random errors are not normally distributed,
can we trust the p-values?

::: {.callout-tip title="Solution" collapse="true"}

It depends. The more symmetric the random errors are,
and the larger the sample size,
the more trustworthy the p-values become.

:::

### Problem 2

If the random errors are not normally distributed,
can we trust the standard errors?

::: {.callout-tip title="Solution" collapse="true"}

It depends on whether or not the first three assumptions are met.
If yes, then yes.

:::




