---
title: "Checking the Normality Assumption"
execute:
  echo: false
format: html
---

```{r}
#| warning: false
library(scales)
```

## Learning Objective

Evaluate the validity of the normality assumption of simple linear regression
in a data set.


## Lesson

### Three Plots

I like to use three plots to evaluate the validity of the normality assumption.
For each plot, you use the residuals as input.

The first plot is a histogram.
Here is what a histogram looks like when the underlying data are normal.
```{r}
set.seed(13)
n <- 1e3
x <- rnorm(n)
hist(x = x,
     main = "Normal Data")
```

The second plot is a boxplot.
In actuality, I really only use this to check if the data are symmetric or not.
Here is what a boxplot looks like when the underlying data are normal.
```{r}
boxplot(x = x,
        main = "Normal Data")
```

The third plot is a q-q plot.
If the data are normal, the points on the q-q plot will follow the
reference line.
This is what a q-q plot looks like when the underlying data are normal.
```{r}
qqnorm(x)
qqline(x)
```

### Mice Example

Let's return to the example where we regressed red blood cell density
on weight in mice.
As before, I'm going to remove the five anomalous mice with very low weights.
We already know the relationship is linear and the variance is constant.
I'm going to assume the mice are independent.
All we have to check is normality.
Here's the histogram of the residuals.
```{r}
d2 <- Lock5Data::DrugResistance
d2 <- d2[d2$Weight >= 5, ]
mod_02 <- lm(formula = RBC ~ Weight,
             data = d2)
hist(x = mod_02$residuals,
     xlab = "Residuals",
     main = "")
```
This is left skewed, but other than that, it's approximately normal.

Let's check the boxplot.
```{r}
boxplot(x = mod_02$residuals)
```
The asymmetry seems less pronounced than in the histogram,
but it's still present.
The median sits more than halfway up the box, and the bottom whisker is
slightly longer than the top whisker.

Now let's check the q-q plot.
```{r}
qqnorm(mod_02$residuals)
qqline(mod_02$residuals)
```
Most of the points follow the reference line;
it's only when the theoretical quantiles get below about -0.75
that the points diverge from the line.

Ultimately, I would say these residuals are more left-skewed than you
would expect from a normal distribution.
So, I wouldn't say the normality assumption is categorically met.
But, since the sample size is relatively large,
and the residuals are close to normal,
I would trust the p-values.

### Atlanta Commutes

Let's return to the data about commuting in Atlanta.
```{r}
d <- Lock5Data::CommuteAtlanta
plot(x = d$Distance,
     y = d$Time,
     xlab = "Distance (miles)",
     ylab = "Time (minutes)",
     main = "Commuting in Atlanta",
     col = alpha(colour = "black",
                 alpha = 0.2),
     pch = 16)
```

If we regress time on distance, here is the residual scatter plot.
```{r}
mod_01 <- lm(formula = Time ~ Distance,
             data = d)
plot(x = mod_01$model$Distance,
     y = mod_01$residuals,
     xlab = "Distance",
     ylab = "Residuals",
     col = alpha(colour = "black",
                 alpha = 0.2),
     pch = 16)
abline(h = 0,
       col = "red")
```
It's hard for me to assess the variance.
It's possible that there's heteroscedasticity.
Let's check the normality assumption anyway.

Here's a histogram, boxplot, and q-q plot of the residuals.
```{r}
hist(x = mod_01$residuals,
     main = "",
     xlab = "Residuals")
boxplot(mod_01$residuals)
qqnorm(mod_01$residuals)
qqline(mod_01$residuals)
```
These residuals are very right-skewed.
I would say the linearity and iid assumptions are met,
but the homoscedasticity and normality assumptions are not.
So, the estimators of the coefficients are unbiased,
but I wouldn't trust the standard errors and p-values.

This is not directly relevant to assumption-checking,
but I want to know why there seems to be a hard line
below which we don't see any points.
I'll plot the line $y = x$ to show what I mean.
```{r}
plot(x = d$Distance,
     y = d$Time,
     xlab = "Distance (miles)",
     ylab = "Time (minutes)",
     main = "Commuting in Atlanta",
     col = alpha(colour = "black",
                 alpha = 0.2),
     pch = 16)
abline(a = 0,
       b = 1)
```
What's going on here?

## Problems

### Problem 1

Do these residuals satisfy the normality assumption?
Why or why not?
```{r}
z <- rexp(n) - 1
qqnorm(z)
qqline(z)
```

::: {.callout-tip title="Solution" collapse="true"}

Absolutely not. Those points don't follow the reference line at all.

:::

### Problem 2

Do these residuals satisfy the normal assumption?
Why or why not?
```{r}
w <- rexp(n) * (2 * rbinom(n = n,
                           size = 1,
                           prob = 0.5) - 1)
boxplot(w)
```

::: {.callout-tip title="Solution" collapse="true"}

I'd like to see a histogram and q-q plot before making a decision,
but it seems like this distribution is too peaked and heavy-tailed
to be normal.
I say that it's peaked because the box is so small
and I say it's heavy-tailed because there are so many dots beyond
the whiskers.

:::




