---
title: "LINE"
execute:
  echo: false
format: html
---

```{r}
#| warning: false
library(scales)
```

## Learning Objectives

- State and recognize an acronym for the four assumptions.
- Synthesize the four assumptions, their consequences,
and how to check them.


## Lesson

The four assumptions are commonly organized into the acronym "LINE":

- Linearity
- Independence (other texts commonly leave out the assumption of identical
distributions)
- Normality
- Equal variance

When I presented the four assumptions, I put them in descending order
from most to least catastrophic when not met.
Unfortunately, that spells "LIEN", which is harder to remember.

In each of the following plots,
the title contains a letter if it meets the corresponding assumption.

```{r}
set.seed(14)
n <- 1e3
x <- rnorm(n = n,
           sd = 2)
beta_0 <- 1
beta_1 <- 1
group_means <- rep(x = rnorm(n = n / 5),
                   each = 5)
epsilon <- rnorm(n)
epsilon_exp <- rexp(n) - 1
plot_title <- ""
for(i in c(FALSE,TRUE)){
  for(j in c(FALSE,TRUE)){
    for(k in c(FALSE,TRUE)){
      for(l in c(FALSE,TRUE)){
        # Linearity
        if(i){
          y <- beta_0 + beta_1 * x
          plot_title <- paste0(plot_title,"L")
        } else {
          y <- beta_0 + beta_1 / 2 * x^2
        }
        # Independence
        if(j){
          plot_title <- paste0(plot_title,"I")
        } else {
          y <- y + group_means
        }
        # Homoscedasticity and Normality
        if(k & l){
          y <- y + epsilon
          plot_title <- paste0(plot_title,"NE")
        } else if(k){
          y <- y + epsilon_exp
          plot_title <- paste0(plot_title,"E")
        } else if(l){
          y <- y + x * epsilon
          plot_title <- paste0(plot_title,"N")
        } else {
          y <- y + x * epsilon_exp
        }
        plot(x = x,
             y = y,
             main = plot_title,
             col = alpha(colour = "black",
                         alpha = 0.2),
             pch = 16)
        plot_title <- ""
      }
    }
  }
}
```

## Problems

For each problem below,
assume an assumption is met if its corresponding letter
from the acronym is present.
What are the consequences for the analysis?

### Problem 1

IN

::: {.callout-tip title="Solution" collapse="true"}

Since the linearity assumption is not met,
you need to fix your underlying model before doing anything else

:::

### Problem 2

LE

::: {.callout-tip title="Solution" collapse="true"}

If the independence assumption is not met, you need other tools
besides simple linear regression.

:::

### Problem 3

LI

::: {.callout-tip title="Solution" collapse="true"}

The coefficients are unbiased but the standard errors,
test statistics, and p-values are untrustworthy.

:::






