---
title: "Interpretation"
execute:
  echo: false
format: html
---

## Learning Objective

Interpret the coefficients in simple linear regression.


## Lesson

### Overview

Simple linear regression has two coefficients:
the intercept ($\beta_0$) and slope ($\beta_1$).
The intercept is the expected value of the outcome when the explanatory
variable is zero.
The slope is the expected change in the outcome associated with
a one-unit increase in the explanatory variable.
We can also interpret the slope as the expected difference in the outcome
between two populations,
one with the explanatory variable equal to $x$ and the other
with the explanatory variable equal to $x + 1$.

Note that I used the word "expected" in interpreting both coefficients.
We're working with random variables, so we can't guarantee
that the outcome will take on a certain value when the explanatory
variable takes on a certain value.
We can only speak in means and averages.

Also, note that I used "associated with" in the interpretation
of the slope.
In many situations, we do not know if the explanatory variable
causes a change in the outcome variable.
We don't want to make any causal claims unless we're sure they're warranted.

### Atlanta Commutes

Consider the data on commuting in Atlanta.
Although the assumptions of homoscedasticity and normality were not satisfied,
the assumption of linearity was, meaning the coefficient estimates
are unbiased.
So, we can interpret them.
```{r}
d <- Lock5Data::CommuteAtlanta
mod_01 <- lm(formula = Time ~ Distance,
             data = d)
summary(mod_01)
```

Recall that the units for distance were miles and the units
for time were minutes.
So, an intercept estimate of 7.1 means that we would expect
a commute of 7.1 minutes for a distance of zero miles.
Does that sound reasonable? I'm not sure.
Were there any observed commutes of zero miles?
```{r}
#| echo: true
sum(d$Distance == 0)
sort(d$Time[d$Distance == 0])
```
Apparently there were 11 commutes of zero miles,
two of which took ten minutes.
I'm curious how these data were collected . . .

Moving on to the slope, there is an expected increase of 1.2 minutes
associated with each additional mile in the commute.
(Alternatively, we could say that we would expect Atlantans
with a commute of $x + 1$ miles to have a commute that is 1.2 minutes
longer than Atlantans with a commute of $x$ miles.)
Does that sound reasonable?
Sure.


### Fire Ants

Recall the fire ants example.
There were two groups of lizards,
those whose habitats had been invaded by fire ants
and those whose habitats had not.
The outcome was the number of twitches a lizard exhibited when exposed
to fire ants.

In the output below, `Habitat_01` is the recoded habitat variable.
It equals 1 if the habitat is invaded and 0 if the habitat is uninvaded.
```{r}
d2 <- Lock5Data::FireAnts
d2$Habitat_01 <- 0
d2$Habitat_01[d2$Habitat == "Invaded"] <- 1
mod_02 <- lm(formula = Twitches ~ Habitat_01,
             data = d2)
summary(mod_02)
```

The intercept is the expected value of the outcome when
the explanatory variable equals zero.
When the explanatory variable is zero,
that means the habitat is uninvaded (in this example).
So, for lizards whose habitat has not been invaded by fire ants,
we would expect 1.1 twitches when exposed to fire ants.

The slope is the expected change in the outcome variable associated
with a one-unit increase in the explanatory variable.
The only way for the explanatory variable to increase by one unit
is to change from uninvaded to invaded.
So, when the habitat changes from uninvaded to invaded,
we would expect an increase of 1.65 twitches.
In other words, we expect lizards from invaded habitats to exhibit
1.65 more twitches than lizards from uninvaded habitats.

## Problems

### Problem 1

Below are the results of regressing red blood cell density on
weight for 67 mice.
I'm not sure what the units were for red blood cell density,
but weight was measured in grams.
Interpret the intercept and slope.
```{r}
d3 <- Lock5Data::DrugResistance
d3 <- d3[d3$Weight >= 5, ]
mod_03 <- lm(formula = RBC ~ Weight,
             data = d3)
summary(mod_03)
```

::: {.callout-tip title="Solution" collapse="true"}

We would expect a mouse weighing zero grams
to have a red blood cell density of -0.58.
Each additional gram of weight is associated with a 0.35-unit increase
in red blood cell density.
Alternatively,
we would expect mice weighing $x + 1$ grams to have a red blood cell density
that is 0.35 units greater than mice weighting $x$ grams.

:::






