---
title: "Hypothesis Testing"
execute:
  echo: false
format: html
---

## Learning Objective

Interpret the results of a hypothesis test in the context of simple linear
regression.


## Lesson

### Overview

When conducting simple linear regression in R,
R will automatically compute p-values for each of the two coefficients.
For each coefficient,
the null hypothesis is that the coefficient is zero,
and the alternative hypothesis is that the coefficient is not zero.
In symbols, the two hypothesis tests are
$$
\begin{align*}
H_0 &: \beta_0 = 0 \\
H_1 &: \beta_0 \ne 0
\end{align*}
$$
and
$$
\begin{align*}
H_0 &: \beta_1 = 0 \\
H_1 &: \beta_1 \ne 0 \text{.}
\end{align*}
$$

### Red Blood Cell Density and Weight in Mice

Below are the results of regressing red blood cell density on
weight for 67 mice.
Enough of our assumptions were satisfied for these data
that I feel comfortable interpreting the p-values.
```{r}
d3 <- Lock5Data::DrugResistance
d3 <- d3[d3$Weight >= 5, ]
mod_03 <- lm(formula = RBC ~ Weight,
             data = d3)
summary(mod_03)
```

The p-value for the `(Intercept)` line of the `Coefficients` table is 0.389.
This is above any reasonable cutoff we would use in hypothesis testing.
So, we fail to reject the null hypothesis that $\beta_0 = 0$.
This makes sense;
we wouldn't want the expected red blood cell density for a mice weighing
zero grams to be anything other than zero.

The p-value for the `Weight` line of the `Coefficients` table is `<2e-16`,
which means $< 2\times 10^{-16}$.
This is below any reasonable cutoff we would use in hypothesis testing,
so we reject the null hypothesis that $\beta_1 = 0$
and conclude $\beta_1 \ne 0$.
In real-world terms, this means that the expected change in red blood cell
density associated with a one-gram increase in weight
is greater than zero.
We could also say that the increase in red blood cell density
associated with a one-gram increase in weight
is significantly different from zero.

You may have noticed that when I said "significantly different",
I dropped "expected".
This is because I'm describing the statistical significance of an observed
change.
If I'm talking about the population, either the change is zero or it isn't,
so it doesn't make sense to talk about statistical significance.
If I'm talking about the observed change in the sample,
it will almost always be different from zero;
the question, then, is whether that difference is statistically significant.
This is getting deeper into semantics than is practically necessary
(probably).

## Problems

### Problem 1

One hundred men had their neck circumference (in cm)
and biceps circumference (extended, in cm) measured.
Below, I've regressed neck circumference on biceps circumference.
Assuming all necessary assumptions are met,
interpret the results of the hypothesis tests relating to the coefficients.
Use a cutoff of $\alpha = 0.05$.
```{r}
d <- Lock5Data::BodyFat
mod_01 <- lm(formula = Neck ~ Biceps,
             data = d)
summary(mod_01)
```

::: {.callout-tip title="Solution" collapse="true"}

Both p-values are much lower than 0.05,
so we reject the null hypotheses that $\beta_0 = 0$ and $\beta_1 = 0$.
So, we conclude that the expected neck circumference for a man with
a biceps circumference of 0 cm is different from 0,
and that the expected increase in neck circumference associated
with a one-cm increase in biceps circumference is greater than zero.


:::

### Problem 2


Below, I've regressed neck circumference (in cm) on age (in years)
for the same 100 men as above.
Assuming all necessary assumptions are met,
interpret the results of the hypothesis tests relating to the coefficients.
Use a cutoff of $\alpha = 0.05$.
```{r}
mod_02 <- lm(formula = Neck ~ Age,
             data = d)
summary(mod_02)
```

::: {.callout-tip title="Solution" collapse="true"}

The p-value for the intercept is far less than 0.05,
so we reject the null hypothesis that $\beta_0 = 0$
and conclude that the expected neck circumference for a man age 0
is different from zero.

The p-value for age is $0.704 > 0.05$,
so we fail to reject the null hypothesis that $\beta_1 = 0$.
We cannot conclude that neck circumference changes linearly with age.

:::




