---
title: "Ordinary Least Squares"
execute:
  echo: false
format: html
---

## Learning Objective

Explain the term "ordinary least squares (OLS)".


## Lesson

The kind of regression we've been using is called "ordinary least squares",
or "OLS" for short.
This is because we find the line of best fit by minimizing
(hence, "least") the sum of squared errors (hence, "squares") between 
the points and the line.

Let's see what this looks like visually.
Consider these 20 points.
```{r}
set.seed(17)
n <- 20
x <- rnorm(n = n,
           sd = 2)
epsilon <- rnorm(n)
y <- x + epsilon
op <- par()
par(pty = "s")
plot(x = x,
     y = y,
     xlim = c(-4,4),
     ylim = c(-4,4))
```

Now let's look at the points with the regression line.
```{r}
mod_01 <- lm(y ~ x)
par(pty = "s")
plot(x = x,
     y = y,
     xlim = c(-4,4),
     ylim = c(-4,4))
abline(mod_01,
       col = "blue")
```

Now let's visualize the residuals.
```{r}
par(pty = "s")
plot(x = x,
     y = y,
     xlim = c(-4,4),
     ylim = c(-4,4))
abline(mod_01,
       col = "blue")
segments(x0 = x,
         y0 = y,
         x1 = x,
         y1 = mod_01$fitted.values,
         col = "red")
```

Finally, let's visualize the squared residuals.
To find the line, we're minimizing the sum of these squared residuals.
```{r}
par(pty = "s")
plot(x = x,
     y = y,
     xlim = c(-5,5),
     ylim = c(-5,5))
abline(mod_01,
       col = "blue")
segments(x0 = c(x,
                x,
                x - mod_01$residuals,
                x - mod_01$residuals),
         y0 = c(y,
                mod_01$fitted.values,
                mod_01$fitted.values,
                y),
         x1 = c(x,
                x - mod_01$residuals,
                x - mod_01$residuals,
                x),
         y1 = c(mod_01$fitted.values,
                mod_01$fitted.values,
                y,
                y),
         col = "red")
```

That's a little cluttered, but we can focus on just one squared residual
to get the idea.
```{r}
par(pty = "s")
plot(x = x,
     y = y,
     xlim = c(-5,5),
     ylim = c(-5,5))
abline(mod_01,
       col = "blue")
segments(x0 = c(x[4],
                x[4],
                x[4] - mod_01$residuals[4],
                x[4] - mod_01$residuals[4]),
         y0 = c(y[4],
                mod_01$fitted.values[4],
                mod_01$fitted.values[4],
                y[4]),
         x1 = c(x[4],
                x[4] - mod_01$residuals[4],
                x[4] - mod_01$residuals[4],
                x[4]),
         y1 = c(mod_01$fitted.values[4],
                mod_01$fitted.values[4],
                y[4],
                y[4]),
         col = "red")
```

This is where the name "ordinary least squares" comes from.

Why don't we use some other method?
Good question.
Basically, this method is mathematically convenient,
and there's a lot of theory supporting it.

What are the consequences?
Points farther from the line have larger residuals than points
closer to the line,
and much larger squared residuals than points
closer to the line.
Since we want to minimize the squared residuals,
this gives points farther from the line more power.
You may have learned in introductory statistics
that the sample average is sensitive to outliers;
this is the same idea.
And, like the median is robust to outliers,
there are nonparametric forms of regression that are robust to outliers.
We won't cover those.

## Problems

### Problem 1

True or false:
The name "ordinary least squares" indicates that we are choosing the line
that has the least squared error.

::: {.callout-tip title="Solution" collapse="true"}

True.

:::

### Problem 2

True or false:
Typically in regression, we minimize the sum of the absolute values
of the residuals.

::: {.callout-tip title="Solution" collapse="true"}

False.
We minimize the sum of the squares of the residuals.

:::




