---
title: "Log-Transformed Outcomes"
execute:
  echo: false
format: html
---

```{r}
#| warning: false
library(scales)
```

## Learning Objective

Interpret the coefficients when the outcome has been logged.


## Lesson

### Problem

When we log the outcome, the linearity assumption becomes
$$
E\left(\log Y_i \middle| X_i\right) = \beta_0 + \beta_1 X_i \text{.}
$$
This poses problems for interpretation; we'd like to talk about
the untransformed outcome, but
$$
E\left(\log Y_i \middle| X_i\right) \ne \log E\left(Y_i \middle| X_i\right) \text{.}
$$
So, for example, our interpretation of $\beta_0$ is "the expected value
of the log of the outcome when $X_i = 0$."
That's not very useful.
Also,
$$
\begin{align*}
\beta_1 &= \beta_0 + \beta_1(x+1) - \left(\beta_0 + \beta_1 x\right) \\
&= E\left(\log Y_i \middle| X_i = x + 1\right) - E\left(\log Y_i \middle| X_i = x\right) \text{,}
\end{align*}
$$
which isn't very useful, either.

### More Assumptions

Suppose we change our model to
$$
\log Y_i|X_i \sim N\left(\beta_0 + \beta_1 X_i,\sigma^2\right) \text{.}
$$
Then the coefficients are more interpretable.
Specifically,
$$
\begin{align*}
E\left(Y_i\middle|X_i = 0\right) &= \exp\left(\beta_0 + \frac{\sigma^2}{2}\right) \\
\frac{E\left(Y_i\middle|X_i = x+1\right)}{E\left(Y_i\middle|X_i = x\right)} &= \exp\left(\beta_1\right) \text{.}
\end{align*}
$$
In words, when the explanatory variable equals 0, we would expect
the outcome variable to equal
$$
\exp\left(\beta_0 + \frac{\sigma^2}{2}\right) \text{.}
$$
An increase of one unit in the explanatory variable is associated with a
$$
\left(\exp\left(\beta_1\right) - 1\right) \times 100\%
$$
increase in the outcome variable.
(If this is negative, it's a decrease.)
Another way of phrasing this is that a population with the explanatory
variable equal to $x + 1$ is expected to have an outcome $\exp(\beta_1)$
times that of a population with the explanatory variable equal to $x$.

Note that, in order to interpret the coefficients this way,
all four assumptions need to be met in the data.
Also, you need to have used the natural log (base $e$) in order to transform
the outcome.
If you use any other base, the interpretation above is not justified.

### Math

I have a tendency to forget the math behind these interpretations,
so I'm going to store it here.
You don't need to know it;
it's more for me.

::: {.callout-caution collapse="true"}
## Caution: Math

Define $Z_i = \log Y_i$.
Then 
$$
Z_i|X_i \sim N\left(\beta_0 + \beta_1 X_i,\sigma^2\right)
$$
and
$$
f(z|x) = \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(-\frac{\left(z - \beta_0 - \beta_1 x\right)^2}{2\sigma^2}\right) \text{.}
$$
Further,
$$
Y_i = \exp\left(Z_i\right)
$$
and
$$
\begin{align*}
E\left(Y_i\middle|X_i = x\right) &= E\left(\exp\left(Z_i\right)\middle|X_i = x\right) \\
&= \int_{-\infty}^\infty \exp\left(z\right) f(z|x)\,dz \\
&= \int_{-\infty}^\infty \exp\left(z\right) \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(-\frac{\left(z - \beta_0 - \beta_1 x\right)^2}{2\sigma^2}\right) \,dz \\
&= \int_{-\infty}^\infty \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(z-\frac{\left(z - \beta_0 - \beta_1 x\right)^2}{2\sigma^2}\right) \,dz \\
&= \int_{-\infty}^\infty \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(z-\frac{z^2 - 2z\left(\beta_0 + \beta_1 x\right) + \left(\beta_0 + \beta_1 x\right)^2}{2\sigma^2}\right) \,dz \\
&= \int_{-\infty}^\infty \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(-\frac{z^2 - 2z\left(\beta_0 + \beta_1 x\right) - 2z\sigma^2 + \left(\beta_0 + \beta_1 x\right)^2}{2\sigma^2}\right) \,dz \\
&= \int_{-\infty}^\infty \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(-\frac{z^2 - 2z\left(\beta_0 + \beta_1 x+\sigma^2\right)  + \left(\beta_0 + \beta_1 x\right)^2}{2\sigma^2}\right) \,dz \\
&= \int_{-\infty}^\infty \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(-\frac{z^2 - 2z\left(\beta_0 + \beta_1 x+\sigma^2\right) + \left(\beta_0 + \beta_1 x + \sigma^2\right)^2 - \left(\beta_0 + \beta_1 x + \sigma^2\right)^2 + \left(\beta_0 + \beta_1 x\right)^2}{2\sigma^2}\right) \,dz \\
&= \exp\left(\frac{\left(\beta_0 + \beta_1 x + \sigma^2\right)^2 - \left(\beta_0 + \beta_1 x\right)^2}{2\sigma^2}\right) \int_{-\infty}^\infty \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(-\frac{z^2 - 2z\left(\beta_0 + \beta_1 x+\sigma^2\right) + \left(\beta_0 + \beta_1 x + \sigma^2\right)^2 }{2\sigma^2}\right) \,dz \\
&= \exp\left(\frac{2\left(\beta_0 + \beta_1 x\right)\sigma^2 + \sigma^4}{2\sigma^2}\right) \int_{-\infty}^\infty \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(-\frac{\left(z - \beta_0 - \beta_1 x - \sigma^2\right)^2 }{2\sigma^2}\right) \,dz \\
&= \exp\left(\frac{2\left(\beta_0 + \beta_1 x\right)\sigma^2 + \sigma^4}{2\sigma^2}\right)  \\
&= \exp\left(\beta_0 + \beta_1 x + \frac{\sigma^2}{2}\right) \text{.} \\
\end{align*}
$$

Thus,
$$
E\left(Y_i\middle|X_i = 0\right) = \exp\left(\beta_0 + \frac{\sigma^2}{2}\right)
$$
and
$$
\begin{align*}
\frac{E\left(Y_i\middle|X_i = x+1\right)}{E\left(Y_i\middle|X_i = x\right)} &= \frac{\exp\left(\beta_0 + \beta_1 (x+1) + \frac{\sigma^2}{2}\right)}{\exp\left(\beta_0 + \beta_1 x + \frac{\sigma^2}{2}\right)} \\
&= \exp\left(\beta_1\right) \text{.}
\end{align*}
$$

:::

### Forest Fire Example

Let's return to the forest fire example.
Each observation is a forest fire in Montesinho Park (Portugal)
between January 2000 and December 2003.
In the scatterplot below, I've plotted total burned area (in hectares)
as a function of relative humidity (%).
```{r}
d <- Lock5Data::ForestFires
plot(x = d$RH,
     y = d$Area,
     xlab = "Relative Humidity (%)",
     ylab = "Total Burned Area (hectares)",
     pch = 19,
     col = alpha(colour = "black",
                 alpha = 0.2))
```
It's hard to detect a relationship between the two variables.
This might be easier if we log the outcome.
However, we can't take the log of zero,
and 4.7% of the observations have zero as the outcome.
```{r}
#| echo: true
sum(d$Area == 0)
sum(d$Area == 0) / nrow(d)
```
According to the description of the data, a zero indicates that
less than 0.01 hectare (100 square meters) was burned.
So, we can remove the zeros and take the natural log of the outcome.
```{r}
d$Area_No_0 <- d$Area
d$Area_No_0[d$Area_No_0 == 0] <- NA
d$Log_Area_No_0 <- log(d$Area_No_0)
plot(x = d$RH,
     y = d$Log_Area_No_0,
     xlab = "Relative Humidity (%)",
     ylab = "Log(Total Burned Area) (log hectares)",
     col = alpha(colour = "black",
                 alpha = 0.3),
     pch = 19)
```
This scatterplot is easier to interpret.
If we regress the log-transformed total burned area on relative humidity,
we get this residual scatterplot:
```{r}
mod_01 <- lm(formula = Log_Area_No_0 ~ RH,
             data = d)
plot(x = mod_01$model$RH,
     y = mod_01$residuals,
     xlab = "Relative Humidity (%)",
     ylab = "Residuals",
     col = alpha(colour = "black",
                 alpha = 0.3),
     pch = 19)
abline(h = 0,
       col = "red")
```
It appears that the linearity assumption is met,
and I'm willing to assume that the data are iid
(even though it's possible that one forest fire changes the probability
of another).
It appears that there is lower variability in the residuals for higher
relative humidities,
but that may be because there are just fewer observations at
the higher relative humidities.
Let's check the distribution of the residuals.
```{r}
boxplot(x = mod_01$residuals,
        ylab = "Residuals")
hist(x = mod_01$residuals,
     main = "",
     xlab = "Residuals")
qqnorm(mod_01$residuals)
qqline(mod_01$residuals)
```
These look (mostly) normal to me.
So, we can interpret the coefficients.

Here's the output from the regression analysis:
```{r}
summary(mod_01)
```
Note that
$$
\begin{align*}
\hat{\beta}_0 &= 2.064810 \\
\hat{\beta}_1 &= -0.005030 \\
\hat{\sigma} &= 1.528 \text{,}
\end{align*}
$$
so
```{r}
#| include: false

exp(mod_01$coefficients[1] + summary(mod_01)$sigma^2 / 2)
exp(mod_01$coefficients[2])
```
$$
\begin{align*}
\hat{E}\left(Y_i\middle|X_i = 0\right) &= \exp\left(\hat{\beta}_0 + \frac{\sigma^2}{2}\right) \\
&= \exp\left(2.064810 + \frac{1.528^2}{2}\right) \\
&= 25.3 \\
\frac{\hat{E}\left(Y_i\middle|X_i = x+1\right)}{\hat{E}\left(Y_i\middle|X_i = x\right)} &= \exp\left(\hat{\beta}_1\right) \\
&= \exp\left(-0.005030 \right) \\
&= 0.995 \text{.}
\end{align*}
$$
This means that, when the relative humidity is 0%,
we would expect 25.3 hectares to burn.
For each additional percentage point of relative humidity,
we would expect a decrease of 0.5% in total area burned.

## Problems

In the past, when I've forgotten how to interpret regression coefficients
with a log-transformed outcome,
I've found my way to this page from the UCLA Statistical Consulting Group:
[How do I interpret a regression model when some variables are log transformed?](https://stats.oarc.ucla.edu/other/mult-pkg/faq/general/faqhow-do-i-interpret-a-regression-model-when-some-variables-are-log-transformed/).
They recommend using the geometric mean in your interpretations.
I don't think this is a good idea,
since (almost) nobody understands the geometric mean.

There are six variables in this data set:
reading, writing, and math scores;
log math and writing scores;
and gender (or maybe sex).
```{r}
d2 <- read.csv(file = "https://stats.idre.ucla.edu/wp-content/uploads/2016/02/lgtrans.csv")
```


### Problem 1

One of the first things they do is regress the log of the writing score
on gender.
Based on the following two boxplots,
what is the impact of logging the writing score?
```{r}
boxplot(formula = write ~ female,
        data = d2,
        xlab = "Gender",
        ylab = "Writing Score")
boxplot(formula = lgwrite ~ female,
        data = d2,
        xlab = "Gender",
        ylab = "Log Writing Score")
```

::: {.callout-tip title="Solution" collapse="true"}

Logging the writing score makes the distribution of the outcome
more skewed!
Also, whether you log the outcome or not,
the variability of the outcome is different for males and for females,
so we're violating the assumption of homoscedasticity.

:::

### Problem 2

Below are the results of a simple linear regression with log math score
as the outcome and reading score as the explanatory variable.
```{r}
mod_02 <- lm(formula = lgmath ~ read,
             data = d2)
summary(mod_02)
```
How would you interpret the coefficients?
Assume all necessary assumptions have been met.

::: {.callout-tip title="Solution" collapse="true"}

```{r}
#| echo: true
exp(mod_02$coefficients[1] + summary(mod_02)$sigma^2 / 2)
exp(mod_02$coefficients[2])
```
When the reading score is zero, we would expect the math score to be 28.8.
With each one-unit increase in reading score,
we would expect the math score to increase by 1%.

:::

### Problem 3

Below are the results of a simple linear regression with math score
as the outcome and reading score as the explanatory variable.
```{r}
mod_03 <- lm(formula = math ~ read,
             data = d2)
summary(mod_03)
```
How would you interpret the coefficients?
Assume all necessary assumptions have been met.

::: {.callout-tip title="Solution" collapse="true"}

When the reading score is zero, we would expect the math score to be 21.
With each one-unit increase in reading score,
we would expect the math score to increase by 0.6 units.

:::

### Problem 4

Below are the diagnostic plots for the model with reading score
as the explanatory variable and log math score as the outcome variable.
```{r}
plot(x = mod_02$model$read,
     y = mod_02$residuals,
     xlab = "Reading Score",
     ylab = "Residual",
     main = "Outcome is Log-Transformed",
     pch = 19,
     col = alpha(colour = "black",
                 alpha = 0.2))
abline(h = 0,
       col = "red")
boxplot(x = mod_02$residuals,
        ylab = "Residual",
        main = "Outcome is Log-Transformed")
hist(x = mod_02$residuals,
     xlab = "Residual",
     main = "Outcome is Log-Transformed")
qqnorm(mod_02$residuals)
qqline(mod_02$residuals)
```

Below are the diagnostic plots for the model with reading score
as the explanatory variable and math score as the outcome variable.
```{r}
plot(x = mod_03$model$read,
     y = mod_03$residuals,
     xlab = "Reading Score",
     ylab = "Residual",
     main = "Outcome is Not Transformed",
     pch = 19,
     col = alpha(colour = "black",
                 alpha = 0.2))
abline(h = 0,
       col = "red")
boxplot(x = mod_03$residuals,
        ylab = "Residual",
        main = "Outcome is Not Transformed")
hist(x = mod_03$residuals,
     xlab = "Residual",
     main = "Outcome is Not Transformed")
qqnorm(mod_03$residuals)
qqline(mod_03$residuals)
```

Do you think it's better to log-transform this outcome or not?

::: {.callout-tip title="Solution" collapse="true"}

The diagnostic plots are nearly identical whether the outcome is
transformed or not.
I might go with the untransformed outcome just because it's easier to interpret.

:::







