---
title: "Shifting the Data to Make the Intercept Interpretable"
execute:
  echo: false
format: html
---

## Learning Objectives

- Shift the data to make the intercept interpretable.
- Interpret the intercept for the shifted data.


## Lesson

You may have noticed that in some of the examples I've used,
the intercept has not made sense.
For example, when I regressed red blood cell density on weight in mice,
I said,
"We would expect a mouse weighing zero grams
to have a red blood cell density of -0.58."
That was based on these results:
```{r}
d <- Lock5Data::DrugResistance
d <- d[d$Weight >= 5, ]
mod_01 <- lm(formula = RBC ~ Weight,
             data = d)
summary(mod_01)
```
Even though I don't know the units for red blood cell density,
I doubt they can be negative.
Perhaps more important,
among the 67 mice in the data set I used,
none had weight lower than 17.725 grams.
```{r}
#| echo: true
range(d$Weight)
```
Estimating the red blood cell density for a mouse weighing less than
that is extrapolating beyond the data, which is never a good idea.
This may be why the intercept was so weird.

What should we do in this case?
Typically, we subtract the same value
(we can call it $c$)
from the explanatory variable
for every observation.
Then, we refit the model with the new explanatory variable.
The new intercept is the expected value of the outcome
when the new explanatory variable is zero,
which happens when the old explanatory variable is equal to $c$.

Let's look at the math, and then apply it to the mice example.
Suppose $X_i$ is the value of the explanatory variable for the $i$th
observation and $Y_i$ is the value of the explanatory variable for the $i$th
observation.
Define $Z_i = X_i - c$.
Then
$$
\begin{align*}
E\left(Y_i \middle| Z_i = 0\right) &= E\left(Y_i \middle| X_i - c = 0\right) \\
&= E\left(Y_i \middle| X_i = c\right) \text{.}
\end{align*}
$$

Since all the mice in my data set weigh between 17.725 and 23.219 grams,
I'm going to choose $c = 20$.
This is a bit arbitrary,
but it's a round number in the range of my data.
```{r}
#| echo: true
d$Weight_20 <- d$Weight - 20
```

Now I'm going to re-do the regression, but with the new explanatory variable.
```{r}
#| echo: true
mod_02 <- lm(formula = RBC ~ Weight_20,
             data = d)
summary(mod_02)
```
The new intercept is about 6.5.
So, we would expect a red blood cell density of 6.5 for a mouse
weighing 20 grams.
This is far more trustworthy than the original intercept because
we observed mice weighing around 20 grams,
but no mice weighing about 0 grams.

Note that the slope didn't change.
If we shift the explanatory variable to the left or right,
the slope of the line should not change.
Only the intercept should change.

How do you decide what to subtract?
It should be a value in the range of your data,
or not too far from it.
It should be meaningful to your scientific question.
Some people recommend using the sample average of the explanatory variable,
but I don't like to do that because
the sample average changes with every sample.

## Problems

One hundred men had their neck circumference (in cm)
and biceps circumference (extended, in cm) measured.

### Problem 1


Below, I've regressed neck circumference on biceps circumference.
```{r}
d2 <- Lock5Data::BodyFat
mod_03 <- lm(formula = Neck ~ Biceps,
             data = d2)
summary(mod_03)
```
Assuming the necessary assumptions are met,
interpret the intercept and slope.

::: {.callout-tip title="Solution" collapse="true"}

A man with zero biceps circumference would be expected to have
a neck circumference of 20.9 cm.
Each additional cm in biceps circumference is associated with a
0.5-cm increase in neck circumference.

:::

### Problem 2

Below, I've regressed neck circumference on biceps circumference
after subtracting 25 cm from each biceps measurement.
```{r}
range(d2$Biceps)
d2$Biceps_25 <- d2$Biceps - 25
mod_04 <- lm(formula = Neck ~ Biceps_25,
             data = d2)
summary(mod_04)
```
Assuming the necessary assumptions are met,
interpret the intercept and slope.

::: {.callout-tip title="Solution" collapse="true"}

A man with 25 cm biceps circumference would be expected to have
a neck circumference of 34 cm.
Each additional cm in biceps circumference is associated with a
0.5-cm increase in neck circumference.

:::




