---
title: "Robust Standard Errors"
execute:
  echo: false
format: html
---

```{r}
#| warning: false
# Load packages
library(lmtest)
library(sandwich)
```

## Learning Objective

Use robust standard errors in the presence of heteroscedasticity.


## Lesson

### Overview and Example

If the linearity and iid assumptions of simple linear regression are met,
and if you have a large sample size,
but your residuals are heteroscedastic,
you can compute robust standard errors.
"Robust" in statistics typically means "correct even when assumptions
are violated".
You'll need two packages, `lmtest` and `sandwich`.
("Sandwich" comes from the form for the standard errors when
expressed in linear algebra.)

Let's take a look at an example that met the linearity and iid assumptions,
had a large sample size, and had heteroscedastic residuals.
The forest fire example fits these requirements.
Each observation is a forest fire in Montesinho Park (Portugal)
between January 2000 and December 2003
where at least 0.01 hectare (100 square meters) was burned.
In the figure below,
I've plotted $\log_{10}(\text{area burned})$ as a function of the outside
temperature.
```{r}
d <- Lock5Data::ForestFires
d$Area_No_0 <- d$Area
d$Area_No_0[d$Area_No_0 == 0] <- NA
d$Log10_Area_No_0 <- log10(d$Area_No_0)
plot(x = d$Temp,
     y = d$Log10_Area_No_0,
     xlab = "Outside Temperature (degrees Celsius)",
     ylab = "Log10(Total Burned Area) (log hectares)")
```

If I regress $\log_{10}(\text{area burned})$ on temperature,
this is the resulting residual plot:
```{r}
mod_01 <- lm(formula = Log10_Area_No_0 ~ Temp,
             data = d)
plot(x = mod_01$model$Temp,
     y = mod_01$residuals,
     xlab = "Outside Temperature (degrees Celsius)",
     ylab = "Residuals")
abline(h = 0,
       col = "red")
```

Here's the regression results without robust standard errors:
```{r}
summary(mod_01)
```

And here are the results with the robust standard errors.
```{r}
coeftest(x = mod_01,
         vcov = vcovHC(x = mod_01,
                       type = "HC0"))
```

You can see that the `Estimate` column hasn't changed,
but the remaining columns have.
This is because the standard errors have been re-computed,
and the $t$-values and p-values depend on the standard errors.
In this example, the conclusions drawn from the p-values have not changed.
The intercept is still significantly different from zero and
the slope is not.

### Simulation Study

In the example above,
the conclusions didn't change after using robust standard errors.
How heteroscedastic do the errors have to be for the robust standard errors
to make a difference?
Let's conduct a simulation study to find out.

I'm going to generate 500 observations where all the assumptions are met
except homoscedasticity.
Then I'll keep increasing the heteroscedasticity until the robust
standard errors yield a different conclusion than the regular standard
errors.

```{r}
set.seed(20)
n <- 500
x <- rnorm(n = n,
           sd = 2)
beta_0 <- 1
beta_1 <- 1
epsilon <- rnorm(n)
for(i in 1:5){
  y <- beta_0 + beta_1 * x + epsilon * x * i
  plot(x = x,
       y = y)
  mod_03 <- lm(y ~ x)
  print(summary(mod_03))
  print(coeftest(x = mod_03,
                 vcov = vcovHC(x = mod_03,
                               type = "HC0")))
}
```
It took pretty severe heteroscedasticity for the robust standard errors
to yield a different result than the regular standard errors.
Of course, we might get a different pattern with a small sample size
or different parameters.


## Problems

### Problem 1

Let's return to the lizard and ant study,
in which lizards from invaded and uninvaded habitats were exposed
to fire ants.
The outcome variable is the number of times the lizards twitched.
I changed every "Uninvaded" to a 0,
and every "Invaded" to a 1,
then regressed the number of twitches on the new habitat variable.
Here are boxplots of the residuals in each group:
```{r}
d2 <- Lock5Data::FireAnts
d2$Habitat_01 <- 0
d2$Habitat_01[d2$Habitat == "Invaded"] <- 1
mod_02 <- lm(formula = Twitches ~ Habitat_01,
             data = d2)
boxplot(formula = mod_02$residuals ~ d2$Habitat,
        xlab = "Habitat",
        ylab = "Residuals")
```
These are the results without robust standard errors:
```{r}
summary(mod_02)
```

These are the results with robust standard errors:
```{r}
coeftest(x = mod_02,
         vcov = vcovHC(x = mod_02,
                       type = "HC0"))
```

Interpret the hypothesis tests for the coefficients
when robust standard errors were used.
Use $\alpha = 0.05$.

::: {.callout-tip title="Solution" collapse="true"}

Both p-values are far below 0.05,
so in each case we reject the null hypothesis.
We can conclude that the expected number of twitches among lizards
from uninvaded habitats is not equal to zero
and that there is a statistically significant difference
between the average number of twitches by lizards from uninvaded habitats
and lizards from invaded habitats.

Another way of phrasing the conclusion about the slope
is that the expected number of twitches among lizards from uninvaded habitats
is different from the expected number of twitches among lizards
from invaded habitats.

:::

### Problem 2

Let's return to the data about commuting in Atlanta.
Each observation is a single commuter.
If we regress time (in minutes) on distance (in miles), these are the results:
```{r}
d3 <- Lock5Data::CommuteAtlanta
mod_04 <- lm(formula = Time ~ Distance,
             data = d3)
summary(mod_04)
```

These are the results with robust standard errors:
```{r}
coeftest(x = mod_04,
         vcov = vcovHC(x = mod_04,
                       type = "HC0"))
```

Interpret the hypothesis tests for the coefficients
when robust standard errors were used.
Use $\alpha = 0.05$.

::: {.callout-tip title="Solution" collapse="true"}

Both p-values are well below 0.05,
so in each case we reject the null hypothesis.
We can conclude that, for a commute of zero miles,
we would expect the time to be different from zero.
Further, we can conclude that the expected change in time associated
with a one-mile change in distance is different from zero.

:::




