---
title: "Robust Standard Errors"
execute:
  echo: false
format: html
---

```{r}
# Load packages
library(lmtest)
library(sandwich)
```

## Learning Objective

Use robust standard errors in the presence of heteroscedasticity.


## Lesson

If the linearity and iid assumptions of simple linear regression are met,
and if you have a large sample size,
but your residuals are heteroscedastic,
you can compute robust standard errors.
"Robust" in statistics typically means "correct even when assumptions
are violated".
You'll need two packages, `lmtest` and `sandwich`.
("Sandwich" comes from the form for the standard errors when
expressed in linear algebra.)

Let's take a look at an example that met the linearity and iid assumptions,
had a large sample size, and had heteroscedastic residuals.
The forest fire example fits these requirements.
Each observation is a forest fire in Montesinho Park (Portugal)
between January 2000 and December 2003
where at least 0.01 hectare (100 square meters) was burned.
In the figure below,
I've plotted $\log_{10}(\text{area burned})$ as a function of the outside
temperature.
```{r}
d <- Lock5Data::ForestFires
d$Area_No_0 <- d$Area
d$Area_No_0[d$Area_No_0 == 0] <- NA
d$Log10_Area_No_0 <- log10(d$Area_No_0)
plot(x = d$Temp,
     y = d$Log10_Area_No_0,
     xlab = "Outside Temperature (degrees Celsius)",
     ylab = "Log10(Total Burned Area) (log hectares)")
```

If I regress $\log_{10}(\text{area burned})$ on temperature,
this is the resulting residual plot:
```{r}
mod_01 <- lm(formula = Log10_Area_No_0 ~ Temp,
             data = d)
plot(x = mod_01$model$Temp,
     y = mod_01$residuals,
     xlab = "Outside Temperature (degrees Celsius)",
     ylab = "Residuals")
abline(h = 0,
       col = "red")
```

Here's the regression results without robust standard errors:
```{r}
summary(mod_01)
```

And here are the results with the robust standard errors.
```{r}
#| echo: true
coeftest(x = mod_01,
         vcov = vcovHC(x = mod_01,
                       type = "HC0"))
```

You can see that the `Estimate` column hasn't changed,
but the remaining columns have.
This is because the standard errors have been re-computed,
and the $t$-values and p-values depend on the standard errors.
In this example, the conclusions drawn from the p-values have not changed.
The intercept is still significantly different from zero and
the slope is not.



## Problems

### Problem 1



::: {.callout-tip title="Solution" collapse="true"}



:::

### Problem 2



::: {.callout-tip title="Solution" collapse="true"}



:::




